---
title: "6 Statistical Modelling v2 (consolidated): shared LEGs + rank and direction transfer"
output: html_document
---

# Setup

```{r setup, message=FALSE, warning=FALSE}

# ==============================================================================
# Setup
# ==============================================================================
# Consolidated modelling pipeline for mouse-to-human transfer:
#   - two nested feature layers (DGE baseline -> Full + BTM)
#   - leave-one-condition-out (LOCO) cross-validation instead of in-sample metrics
#   - human-DEG universe with ranks recomputed inside the universe
#   - two mouse-LEG scopes: all DEGs (Rank, Direction) and mouse LEGs (Shared)

library(here)
source(here("scripts_notebooks", "required.R"))   # loads packages, colors, theme_vaxgo
library(tidyverse)
library(tidymodels)

# Output folders for the modelling pipeline
dir.create(here("Modelling", "Models"),  showWarnings = FALSE, recursive = TRUE)
dir.create(here("Modelling", "Figures"), showWarnings = FALSE, recursive = TRUE)
dir.create(here("Modelling", "Tables"),  showWarnings = FALSE, recursive = TRUE)
```

# Input parameters

```{r}
# Scope: Infection + Injury only -> 4 conditions -> 4 LOCO folds
treatments_keep <- c("Infection", "Injury")
seed <- 123
```


# Prepare data 

## Load data

```{r}
# Gene-level feature layers (sequence evolution, TFs, cCREs, BTMs) per gene
gene_annotated_layers <- readRDS(here("tables", "human_mouse_statsmodelling_gene_annotated_layers.rds")) %>%
  distinct(hgnc_symbol, .keep_all = TRUE)

# Per-gene expression statistics across conditions and BTM processes
dge_btm_process_genes_diff_bygene_clean_filtered <- readRDS(here("tables", "dge_btm_process_genes_diff_bygene_clean_filtered.rds"))
```

## Feature dictionary

```{r}
# Single source of truth for feature labels: raw column name -> readable label.
# Use `feature_name_original` for selecting columns from the data; use
# `feature_name` (via clean_feat()) only for plot labels.
feature_dictionary <- tibble(
  feature_name_original = c(
    # --- DGE baseline (mouse) ---
    "inverse_se_mouse", "mean_log2fc_mouse", "abs_rank_log2fc_mouse",
    "rank_mouse", "rank_mouse_btm", "dual_rank", "sign_mouse",
    # --- Sequence evolution ---
    "dist_k80", "identity_human2mouse",
    # --- Transcription factors ---
    "n_tf_total", "pct_tf_shared",
    # --- cCRE counts ---
    "n_total_cres_gene", "n_type_PLS", "n_type_pELS", "n_type_dELS",
    # --- cCRE type match ---
    "pct_match_type_PLS", "pct_match_type_pELS", "pct_match_type_dELS",
    # --- cCRE CTCF match ---
    "pct_match_ctcf_PLS_CTCF-bound", "pct_match_ctcf_pELS_CTCF-bound",
    "pct_match_ctcf_dELS_CTCF-bound",
    # --- BTMs ---
    "Not-BTM", "INNATE RESPONSE", "B CELLS", "T CELLS", "PLATELETS",
    "ECM AND MIGRATION", "ENERGY METABOLISM", "MONOCYTES", "CELL CYCLE",
    "IFN", "SIGNAL TRANSDUCTION", "NK CELLS", "NEUTROPHILS"
  ),
  feature_name = c(
    "1/SE", "Log2FC (Mouse)", "|Absolute rank (mouse)|",
    "|Absolute rank (mouse)|", "|Absolute rank (mouse) BTM|", "Dual absolute rank", "Sign (Mouse)",
    "Kimura", "%Protein identity",
    "nTotal TFs", "%Shared TFs",
    "nTotal cCREs", "PLS nCREs", "pELS nCREs", "dELS nCREs",
    "PLS %Match", "pELS %Match", "dELS %Match",
    "PLS CTCF", "pELS CTCF", "dELS CTCF",
    "Not-BTM", "INNATE RESPONSE", "B CELLS", "T CELLS", "PLATELETS",
    "ECM AND MIGRATION", "ENERGY METABOLISM", "MONOCYTES", "CELL CYCLE",
    "IFN", "SIGNAL TRANSDUCTION", "NK CELLS", "NEUTROPHILS"
  ),
  feature_group = c(
    rep("DGE baseline", 7),
    rep("Sequence evolution", 2),
    rep("Transcription factors", 2),
    rep("cCRE counts", 4),
    rep("cCRE type match", 3),
    rep("cCRE CTCF match", 3),
    rep("BTM", 13)
  )
)
```

## Feature sets

```{r}
# --- Building blocks (same categories as the original) ---
feat_base_experiment <- c("treatment", "pathogen", "timepoint_comparison", "hgnc_symbol")
# measurement precision (noise covariate)
feat_noise  <- c("inverse_se_mouse")             
# sequence evolution
feat_seq    <- c("dist_k80", "identity_human2mouse") 
# transcription factor features
feat_tfs    <- c("n_tf_total", "pct_tf_shared")
# cis-regulatory element architecture
# NOTE: use the ORIGINAL column names here (these are selected from the data),
# not the readable labels produced by clean_feat() (used only for plots).
feat_cre    <- c(
  "n_total_cres_gene",
  "n_type_PLS",  "pct_match_type_PLS",  "pct_match_ctcf_PLS_CTCF-bound",
  "n_type_pELS", "pct_match_type_pELS", "pct_match_ctcf_pELS_CTCF-bound",
  "n_type_dELS", "pct_match_type_dELS", "pct_match_ctcf_dELS_CTCF-bound"
)
# blood transcriptional modules
feat_btms <- c(                                      
  "Not-BTM", "INNATE RESPONSE", "B CELLS", "T CELLS", "PLATELETS",
  "ECM AND MIGRATION", "ENERGY METABOLISM", "MONOCYTES", "CELL CYCLE",
  "IFN", "SIGNAL TRANSDUCTION", "NK CELLS", "NEUTROPHILS"
)
# Full + BTM adds all of these
layers <- c(feat_seq, feat_tfs, feat_cre, feat_btms)  

# Mouse-rank scope for the Shared classification (BTM-only rank)
feat_mouse_class <- c("mean_log2fc_mouse", "rank_mouse_btm")

# Framing A = predictive (keeps mouse expression); Framing B = explanatory (no mouse magnitude)
# Shared - classification of shared LEGs (BTM scope, BTM-only rank)
shared_sets_A  <- list("DGE Baseline" = c(feat_base_experiment, feat_noise, feat_mouse_class),
                "Full + BTM"   = c(feat_base_experiment, feat_noise, feat_mouse_class, layers))
shared_sets_B  <- list("DGE Baseline" = c(feat_base_experiment, feat_noise),
                "Full + BTM"   = c(feat_base_experiment, feat_noise, layers))
# Rank - mouse-to-human rank transfer
rank_sets <- list("Mouse only"     = c(feat_base_experiment, "rank_mouse"),
                "Mouse + layers" = c(feat_base_experiment, "rank_mouse", layers))
# Direction - directional concordance
direction_sets <- list("Direction only"     = c(feat_base_experiment, "sign_mouse"),
                "Direction + layers" = c(feat_base_experiment, "sign_mouse", layers))
```

# Build modelling table + targets

# Universe: human DEGs (p_adj_human <= 0.05). Ranks recomputed WITHIN the universe.

```{r}
# ==============================================================================
# Build modelling table + targets
# ==============================================================================
# Two universes are used in this script:
#   1. All DEGs        -> Convergence, Rank and Direction (ranks across all DEGs)
#   2. Mouse LEGs only -> Shared classification (ranks within mouse LEGs)

deg_universe <- "human"   # primary; sensitivity: "any", "all"

# Merge per-gene expression statistics with the biological feature layers
base_tbl <- dge_btm_process_genes_diff_bygene_clean_filtered %>%
  inner_join(gene_annotated_layers, by = "hgnc_symbol") %>%
  filter(treatment %in% treatments_keep)

# ------------------------------------------------------------------------------
# 1. Select the DEG universe
# ------------------------------------------------------------------------------
# human = genes differentially expressed in humans (primary)
# any   = differentially expressed in either species (sensitivity)
# all   = no DEG filter (sensitivity)
deg_filter <- function(tbl, universe) {
  switch(universe,
         human = filter(tbl, p_adj_human <= 0.05),
         any   = filter(tbl, p_adj_mouse <= 0.05 | p_adj_human <= 0.05),
         all   = tbl)
}

# UNIVERSE 1 - all DEGs (used by Convergence, Rank and Direction)
base_deg <- deg_filter(base_tbl, deg_universe)

# ------------------------------------------------------------------------------
# 2. Absolute ranks (all DEGs)
# ------------------------------------------------------------------------------
# Absolute rank = percentile (0-100) of |log2FC| x -log10(adjusted P), per condition.
# High rank = strong AND significant change (direction is ignored).
# One value per gene per condition.
ranks_df <- base_deg %>%
  distinct(treatment, pathogen, timepoint_comparison, hgnc_symbol,
           mean_log2fc_mouse, mean_log2fc_human, p_adj_mouse, p_adj_human) %>%
  group_by(treatment, pathogen, timepoint_comparison) %>%
  mutate(
    rank_mouse = percent_rank(abs(mean_log2fc_mouse) * -log10(pmax(p_adj_mouse, 1e-300))) * 100,
    rank_human = percent_rank(abs(mean_log2fc_human) * -log10(pmax(p_adj_human, 1e-300))) * 100
  ) %>%
  ungroup()

# ------------------------------------------------------------------------------
# 3. Attach ranks and define the targets (all DEGs)
# ------------------------------------------------------------------------------
# relevance       = dual_rank = min(mouse rank, human rank)
#                   high only when the gene is strong in BOTH species
# convergence     = rank_diff = |mouse rank - human rank|
#                   low when the gene sits at a similar rank in both species
# conv_rel        = dual_rank * (1 - rank_diff/100)
#                   combined: high relevance AND high convergence
# sign_concordant = TRUE when mouse and human change in the SAME direction
df_all <- base_deg %>%
  inner_join(ranks_df %>% select(treatment, pathogen, timepoint_comparison, hgnc_symbol, rank_mouse, rank_human),
             by = c("treatment", "pathogen", "timepoint_comparison", "hgnc_symbol")) %>%
  mutate(
    rank_diff = abs(rank_mouse - rank_human),
    dual_rank = pmin(rank_mouse, rank_human),
    conv_rel  = dual_rank * (1 - rank_diff / 100),
    sign_mouse = sign(mean_log2fc_mouse),
    sign_human = sign(mean_log2fc_human),
    sign_concordant = factor(sign_mouse == sign_human, levels = c(FALSE, TRUE)),
    status_original = if_else(is.na(status_original), "Not LEG", as.character(status_original))
  )

# ------------------------------------------------------------------------------
# 4. UNIVERSE 2 - mouse LEGs (used by the Shared classification)
# ------------------------------------------------------------------------------
# Keep only Shared and Mouse-only genes, so both classes are mouse LEGs.
# This controls for mouse LEG status and isolates what makes a mouse LEG shared.
df_mouse_legs <- df_all %>%
  filter(status_original %in% c("Shared", "Mouse only"))

# Genes that belong to at least one blood transcriptional module (BTM)
btm_gene_set <- df_mouse_legs %>%
  filter(!is.na(process), process != "Not-BTM") %>%
  distinct(hgnc_symbol) %>%
  pull(hgnc_symbol)

# Absolute ranks recomputed WITHIN mouse LEGs (classification scope)
btm_ranks <- df_mouse_legs %>%
  filter(hgnc_symbol %in% btm_gene_set) %>%
  distinct(treatment, pathogen, timepoint_comparison, hgnc_symbol,
           mean_log2fc_mouse, mean_log2fc_human, p_adj_mouse, p_adj_human) %>%
  group_by(treatment, pathogen, timepoint_comparison) %>%
  mutate(
    rank_mouse_btm = percent_rank(abs(mean_log2fc_mouse) * -log10(pmax(p_adj_mouse, 1e-300))) * 100,
    rank_human_btm = percent_rank(abs(mean_log2fc_human) * -log10(pmax(p_adj_human, 1e-300))) * 100
  ) %>%
  ungroup() %>%
  select(treatment, pathogen, timepoint_comparison, hgnc_symbol, rank_mouse_btm, rank_human_btm)

df_mouse_legs <- df_mouse_legs %>%
  left_join(btm_ranks, by = c("treatment", "pathogen", "timepoint_comparison", "hgnc_symbol"))

# Attach the BTM-only rank to df_all, so all scores are computed in a single table
df_all <- df_all %>%
  left_join(df_mouse_legs %>%
              distinct(treatment, pathogen, timepoint_comparison, hgnc_symbol, rank_mouse_btm),
            by = c("treatment", "pathogen", "timepoint_comparison", "hgnc_symbol"))

# Quick sanity check
cat("universe:", deg_universe, " | df_all rows:", nrow(df_all),
    " | mouse-LEG rows:", nrow(df_mouse_legs), " | BTM genes:", length(btm_gene_set), "\n")
table(df_all$pathogen)
```

# Helpers

```{r}
# ==============================================================================
# Helpers
# ==============================================================================
# Robust min-max scaling that tolerates zero-variance groups
min_max <- function(x) {
  if (all(is.na(x))) return(rep(0, length(x)))
  rng <- max(x, na.rm = TRUE) - min(x, na.rm = TRUE)
  if (is.na(rng) || rng == 0) return(rep(0, length(x)))
  (x - min(x, na.rm = TRUE)) / rng
}

# ------------------------------------------------------------------------------
# Generic LOCO regression: fit LM + Lasso + RF + NN, extract coefficients/importance,
# evaluate with leave-one-condition-out, and return out-of-fold predictions.
# ------------------------------------------------------------------------------
loco_regression <- function(df, selection, target, set_name, framing) {
  # 1. PREPARE DATASET
  d <- df %>%
    select(all_of(selection), all_of(target)) %>%
    drop_na(all_of(target)) %>% distinct()

  # 2. RECIPE (metadata as id; normalize numeric predictors)
  rec <- recipe(as.formula(paste(target, "~ .")), data = d) %>%
    update_role(any_of(c("hgnc_symbol", "treatment", "pathogen", "timepoint_comparison")), new_role = "id") %>%
    step_dummy(all_nominal_predictors()) %>%
    step_zv(all_predictors()) %>%
    step_normalize(all_numeric_predictors())

  # glmnet requires >= 2 predictors; skip lasso when only one is available
  n_pred <- prep(rec, training = d) %>%
    bake(new_data = NULL) %>%
    select(-all_of(target), -any_of(c("hgnc_symbol", "treatment", "pathogen", "timepoint_comparison"))) %>%
    ncol()
  use_lasso <- n_pred >= 2

  # 3. MODEL FITTING (linear + lasso + random forest + neural network)
  wf_lm <- workflow() %>%
    add_recipe(rec) %>%
    add_model(linear_reg() %>% set_engine("lm"))

  wf_lasso <- workflow() %>%
    add_recipe(rec) %>%
    add_model(linear_reg(penalty = 0.01, mixture = 1) %>%
                set_engine("glmnet") %>%
                set_mode("regression"))

  wf_rf <- workflow() %>%
    add_recipe(rec) %>%
    add_model(rand_forest(trees = 500, mtry = min(8, n_pred), min_n = 5) %>%
                set_engine("ranger", importance = "impurity", num.threads = 1) %>%
                set_mode("regression"))

  wf_nn <- workflow() %>%
    add_recipe(rec) %>%
    add_model(mlp(hidden_units = 10, epochs = 200, penalty = 0.01) %>%
                set_engine("nnet") %>%
                set_mode("regression"))

  fit_lm <- fit(wf_lm, data = d)
  fit_lasso <- if (use_lasso) tryCatch(fit(wf_lasso, data = d), error = function(e) NULL) else NULL
  fit_rf <- fit(wf_rf, data = d)
  fit_nn <- fit(wf_nn, data = d)

  # 4. EXTRACT COEFFICIENTS (standardized effects) & GINI IMPORTANCE
  coefs <- tidy(extract_fit_engine(fit_lm), conf.int = TRUE) %>%
    filter(term != "(Intercept)") %>%
    rename(feature_name_original = term) %>%
    mutate(feature_name_original = str_replace_all(feature_name_original, "`", "")) %>%
    mutate(feature_set = set_name, framing = framing)

  rf_engine <- extract_fit_engine(fit_rf)
  imp <- tibble(feature_name_original = names(rf_engine$variable.importance),
                importance = as.numeric(rf_engine$variable.importance)) %>%
    mutate(feature_set = set_name, framing = framing)

  # 5. LOCO CROSS-VALIDATION (4 folds by condition; save predictions)
  set.seed(seed)
  folds <- group_vfold_cv(d, group = pathogen, v = 4)
  cv_lm    <- fit_resamples(wf_lm,    resamples = folds, control = control_resamples(save_pred = TRUE))
  cv_lasso <- if (use_lasso) {
    tryCatch(fit_resamples(wf_lasso, resamples = folds, control = control_resamples(save_pred = TRUE)),
             error = function(e) NULL)
  } else NULL
  cv_rf    <- fit_resamples(wf_rf,    resamples = folds, control = control_resamples(save_pred = TRUE))
  cv_nn    <- fit_resamples(wf_nn,    resamples = folds, control = control_resamples(save_pred = TRUE))

  cv_metrics <- function(cv, model_name) {
    m <- collect_metrics(cv)
    tibble(model = model_name,
           rmse = m$mean[m$.metric == "rmse"],
           rsq  = m$mean[m$.metric == "rsq"])
  }
  metrics <- bind_rows(cv_metrics(cv_lm, "Linear (LM)"),
                       if (!is.null(cv_lasso)) cv_metrics(cv_lasso, "Lasso"),
                       cv_metrics(cv_rf, "Random Forest"),
                       cv_metrics(cv_nn, "Neural Network")) %>%
    mutate(feature_set = set_name, framing = framing, target = target, task = "regression")

  # 6. OUT-OF-FOLD PREDICTIONS (one per gene and condition)
  d_indexed <- d %>%
    mutate(.row = row_number()) %>%
    select(.row, any_of(c("hgnc_symbol", "treatment", "pathogen", "timepoint_comparison")))
  pred_oof <- d_indexed %>%
    left_join(collect_predictions(cv_lm) %>% select(.row, pred_lm = .pred), by = ".row")
  if (!is.null(cv_lasso)) {
    pred_oof <- pred_oof %>%
      left_join(collect_predictions(cv_lasso) %>% select(.row, pred_lasso = .pred), by = ".row")
  } else {
    pred_oof <- pred_oof %>% mutate(pred_lasso = NA_real_)
  }
  pred_oof <- pred_oof %>%
    left_join(collect_predictions(cv_rf) %>% select(.row, pred_rf = .pred), by = ".row") %>%
    left_join(collect_predictions(cv_nn) %>% select(.row, pred_nn = .pred), by = ".row") %>%
    mutate(feature_set = set_name, framing = framing, target = target)

  list(metrics = metrics, coefficients = coefs, importance = imp,
       fit_lm = fit_lm, fit_lasso = fit_lasso, fit_rf = fit_rf, fit_nn = fit_nn, pred_oof = pred_oof)
}

# ------------------------------------------------------------------------------
# Generic LOCO classification
# fit logistic + Lasso + RF + NN, extract log-odds/importance,
# evaluate (ROC-AUC, PR-AUC) with leave-one-condition-out, and return
# out-of-fold predictions.
# ------------------------------------------------------------------------------
loco_classification <- function(df, selection, target, set_name, framing) {
  # 1. PREPARE DATASET (binary factor, FALSE < TRUE)
  d <- df %>%
    select(all_of(selection), all_of(target)) %>%
    drop_na(all_of(target)) %>%
    distinct() %>%
    mutate(!!target := factor(.data[[target]], levels = c(FALSE, TRUE)))

  # 2. RECIPE (drop near-zero variance and highly correlated predictors)
  rec <- recipe(as.formula(paste(target, "~ .")), data = d) %>%
    update_role(any_of(c("hgnc_symbol", "treatment", "pathogen", "timepoint_comparison")), new_role = "id") %>%
    step_dummy(all_nominal_predictors()) %>%
    step_nzv(all_predictors()) %>%
    step_corr(all_numeric_predictors(), threshold = 0.9) %>%
    step_normalize(all_numeric_predictors())

  # glmnet requires >= 2 predictors; skip lasso when only one is available
  n_pred <- prep(rec, training = d) %>%
    bake(new_data = NULL) %>%
    select(-all_of(target), -any_of(c("hgnc_symbol", "treatment", "pathogen", "timepoint_comparison"))) %>%
    ncol()
  use_lasso <- n_pred >= 2

  # 3. MODEL FITTING (logistic + lasso + random forest + neural network)
  wf_log <- workflow() %>%
    add_recipe(rec) %>%
    add_model(logistic_reg() %>% set_engine("glm"))
  wf_lasso <- workflow() %>%
    add_recipe(rec) %>%
    add_model(logistic_reg(penalty = 0.01, mixture = 1) %>%
                set_engine("glmnet") %>%
                set_mode("classification"))
  wf_rf <- workflow() %>%
    add_recipe(rec) %>%
    add_model(rand_forest(trees = 500, mtry = min(8, n_pred), min_n = 5) %>%
                set_engine("ranger", importance = "impurity", num.threads = 1) %>%
                set_mode("classification"))
  wf_nn <- workflow() %>%
    add_recipe(rec) %>%
    add_model(mlp(hidden_units = 10, epochs = 200, penalty = 0.01) %>%
                set_engine("nnet") %>%
                set_mode("classification"))

  fit_log <- fit(wf_log, data = d)
  fit_lasso <- if (use_lasso) tryCatch(fit(wf_lasso, data = d), error = function(e) NULL) else NULL
  fit_rf <- fit(wf_rf, data = d)
  fit_nn <- fit(wf_nn, data = d)

  # 4. EXTRACT LOG-ODDS & GINI IMPORTANCE
  coefs <- tidy(extract_fit_engine(fit_log), conf.int = TRUE) %>%
    filter(term != "(Intercept)") %>%
    rename(feature_name_original = term) %>%
    mutate(feature_name_original = str_replace_all(feature_name_original, "`", "")) %>%
    mutate(feature_set = set_name, framing = framing)

  rf_engine <- extract_fit_engine(fit_rf)
  imp <- tibble(feature_name_original = names(rf_engine$variable.importance),
                importance = as.numeric(rf_engine$variable.importance)) %>%
    mutate(feature_set = set_name, framing = framing)

  # Lasso (penalized) coefficients at the fitted penalty (0.01); no valid SE/CI
  coefs_lasso <- if (!is.null(fit_lasso)) {
    m <- as.matrix(coef(extract_fit_engine(fit_lasso), s = 0.01))
    tibble(feature_name_original = rownames(m),
           estimate = as.numeric(m[, 1])) %>%
      filter(feature_name_original != "(Intercept)") %>%
      mutate(feature_name_original = str_replace_all(feature_name_original, "`", "")) %>%
      mutate(feature_set = set_name, framing = framing)
  } else NULL

  # 5. LOCO CROSS-VALIDATION (4 folds by condition; save predictions)
  set.seed(seed)
  folds <- group_vfold_cv(d, group = pathogen, v = 4)
  cv_log   <- fit_resamples(wf_log,   resamples = folds, control = control_resamples(save_pred = TRUE))
  cv_lasso <- if (use_lasso) {
    tryCatch(fit_resamples(wf_lasso, resamples = folds, control = control_resamples(save_pred = TRUE)),
             error = function(e) NULL)
  } else NULL
  cv_rf    <- fit_resamples(wf_rf,    resamples = folds, control = control_resamples(save_pred = TRUE))
  cv_nn    <- fit_resamples(wf_nn,    resamples = folds, control = control_resamples(save_pred = TRUE))

  cv_metrics <- function(cv, model_name) {
    p <- collect_predictions(cv)
    tibble(model = model_name,
           roc_auc = roc_auc(p, truth = !!sym(target), .pred_TRUE, event_level = "second")$.estimate,
           pr_auc  = pr_auc(p,  truth = !!sym(target), .pred_TRUE, event_level = "second")$.estimate)
  }
  metrics <- bind_rows(cv_metrics(cv_log, "Logistic Regression"),
                       if (!is.null(cv_lasso)) cv_metrics(cv_lasso, "Lasso"),
                       cv_metrics(cv_rf, "Random Forest"),
                       cv_metrics(cv_nn, "Neural Network")) %>%
    mutate(feature_set = set_name, framing = framing, target = target, task = "classification")

  # 6. OUT-OF-FOLD PREDICTIONS (probability of TRUE, one per gene and condition)
  d_indexed <- d %>%
    mutate(.row = row_number()) %>%
    select(.row, any_of(c("hgnc_symbol", "treatment", "pathogen", "timepoint_comparison")))
  pred_oof <- d_indexed %>%
    left_join(collect_predictions(cv_log) %>% select(.row, pred_log = .pred_TRUE), by = ".row")
  if (!is.null(cv_lasso)) {
    pred_oof <- pred_oof %>%
      left_join(collect_predictions(cv_lasso) %>% select(.row, pred_lasso = .pred_TRUE), by = ".row")
  } else {
    pred_oof <- pred_oof %>% mutate(pred_lasso = NA_real_)
  }
  pred_oof <- pred_oof %>%
    left_join(collect_predictions(cv_rf) %>% select(.row, pred_rf = .pred_TRUE), by = ".row") %>%
    left_join(collect_predictions(cv_nn) %>% select(.row, pred_nn = .pred_TRUE), by = ".row") %>%
    mutate(feature_set = set_name, framing = framing, target = target)

  list(metrics = metrics, coefficients = coefs, importance = imp, coefs_lasso = coefs_lasso,
       fit_log = fit_log, fit_lasso = fit_lasso, fit_rf = fit_rf, fit_nn = fit_nn, pred_oof = pred_oof)
}
```


# Modelling
## Shared vs. Mouse-only

```{r}
# ==============================================================================
# Shared
# ==============================================================================
# status_original is module-specific (a gene can be Shared in one BTM and Mouse
# only in another), so collapse to one row per gene and condition: a gene is
# labelled Shared if it is a shared LEG in at least one module.
df_shared <- df_mouse_legs %>%
  filter(hgnc_symbol %in% btm_gene_set) %>%  # BTM scope only
  group_by(treatment, pathogen, timepoint_comparison, hgnc_symbol) %>%
  mutate(shared = factor(any(status_original == "Shared"), levels = c(FALSE, TRUE))) %>%
  ungroup() %>%
  distinct(treatment, pathogen, timepoint_comparison, hgnc_symbol, .keep_all = TRUE)

shared_res_A <- imap(shared_sets_A, ~ loco_classification(df_shared, .x, "shared", .y, "A_predictive"))
shared_res_B <- imap(shared_sets_B, ~ loco_classification(df_shared, .x, "shared", .y, "B_explanatory"))

shared_metrics   <- bind_rows(map_dfr(shared_res_A, "metrics"), map_dfr(shared_res_B, "metrics"))
shared_coefs     <- bind_rows(map_dfr(shared_res_A, "coefficients"), map_dfr(shared_res_B, "coefficients"))
shared_imp       <- bind_rows(map_dfr(shared_res_A, "importance"),   map_dfr(shared_res_B, "importance"))
shared_coefs_lasso <- bind_rows(map_dfr(shared_res_A, "coefs_lasso"), map_dfr(shared_res_B, "coefs_lasso"))
shared_pred_oof  <- bind_rows(map_dfr(shared_res_A, "pred_oof"), map_dfr(shared_res_B, "pred_oof"))
shared_metrics

# Save tables
saveRDS(shared_metrics,     here("Modelling", "Models", "metrics_shared.rds"))
saveRDS(shared_coefs,       here("Modelling", "Models", "coefs_shared.rds"))
saveRDS(shared_imp,         here("Modelling", "Models", "imp_shared.rds"))
saveRDS(shared_coefs_lasso, here("Modelling", "Models", "coefs_lasso_shared.rds"))
saveRDS(shared_pred_oof,    here("Modelling", "Models", "pred_shared.rds"))
write_csv(shared_metrics, here("Modelling", "Tables", "shared_loco.csv"))

# Save fitted models (framing A, Full + BTM) for reuse (butchered + compressed)
saveRDS(butcher(shared_res_A[["Full + BTM"]]$fit_rf),    here("Modelling", "Models", "rf_model_shared.rds"),    compress = "xz")
saveRDS(butcher(shared_res_A[["Full + BTM"]]$fit_nn),    here("Modelling", "Models", "nn_model_shared.rds"),    compress = "xz")
saveRDS(butcher(shared_res_A[["Full + BTM"]]$fit_lasso), here("Modelling", "Models", "lasso_model_shared.rds"), compress = "xz")
```

## Rank

```{r}
# ==============================================================================
# Rank
# ==============================================================================
#Compute LOCO regression
rank_res <- imap(rank_sets, ~ loco_regression(df_all, .x, "rank_human", .y, "transfer"))
rank_metrics  <- map_dfr(rank_res, "metrics")
rank_coefs    <- map_dfr(rank_res, "coefficients")
rank_imp      <- map_dfr(rank_res, "importance")
rank_pred_oof <- map_dfr(rank_res, "pred_oof")
rank_metrics

# Save tables
saveRDS(rank_metrics,  here("Modelling", "Models", "metrics_rank.rds"))
saveRDS(rank_coefs,    here("Modelling", "Models", "coefs_rank.rds"))
saveRDS(rank_imp,      here("Modelling", "Models", "imp_rank.rds"))
saveRDS(rank_pred_oof, here("Modelling", "Models", "pred_rank.rds"))
write_csv(rank_metrics, here("Modelling", "Tables", "rank_loco.csv"))

# Save fitted models ("Mouse + layers") for reuse (butchered + compressed)
saveRDS(butcher(rank_res[["Mouse + layers"]]$fit_rf),  here("Modelling", "Models", "rf_model_rank.rds"),  compress = "xz")
saveRDS(butcher(rank_res[["Mouse + layers"]]$fit_nn), here("Modelling", "Models", "nn_model_rank.rds"), compress = "xz")
```

## Direction

```{r}
# ==============================================================================
# Direction
# ==============================================================================
direction_res <- imap(direction_sets, ~ loco_classification(df_all, .x, "sign_concordant", .y, "transfer"))
direction_metrics  <- map_dfr(direction_res, "metrics")
direction_coefs    <- map_dfr(direction_res, "coefficients")
direction_imp      <- map_dfr(direction_res, "importance")
direction_pred_oof <- map_dfr(direction_res, "pred_oof")
direction_metrics

# ==============================================================================
# Direction — diagnostics
# ==============================================================================
cat("sign_mouse:\n"); print(table(df_all$sign_mouse, useNA = "ifany"))
cat("sign_human:\n"); print(table(df_all$sign_human, useNA = "ifany"))
cat("sign_concordant:\n"); print(table(df_all$sign_concordant, useNA = "ifany"))

df_all %>%
  group_by(sign_mouse) %>%
  summarise(n = n(), p_concordant = mean(sign_concordant == TRUE, na.rm = TRUE), .groups = "drop") %>%
  print()

# Save tables
saveRDS(direction_metrics,  here("Modelling", "Models", "metrics_direction.rds"))
saveRDS(direction_coefs,    here("Modelling", "Models", "coefs_direction.rds"))
saveRDS(direction_imp,      here("Modelling", "Models", "imp_direction.rds"))
saveRDS(direction_pred_oof, here("Modelling", "Models", "pred_direction.rds"))
write_csv(direction_metrics, here("Modelling", "Tables", "direction_loco.csv"))

# Save fitted models ("Direction + layers") for reuse (butchered + compressed)
saveRDS(butcher(direction_res[["Direction + layers"]]$fit_rf),  here("Modelling", "Models", "rf_model_direction.rds"),  compress = "xz")
saveRDS(butcher(direction_res[["Direction + layers"]]$fit_nn), here("Modelling", "Models", "nn_model_direction.rds"), compress = "xz")
```


# SCORES

## Scores from out-of-fold predictions

```{r}
# ==============================================================================
# SCORES - out-of-fold predictions from the best-performing model per task
# ==============================================================================
# Each score is a model prediction, computed out-of-fold (leave-one-condition-out):
# every gene is predicted by a model trained without its condition.
#   score_shared        -> probability of being Shared (vs Mouse only)
#   score_rank          -> predicted human absolute rank
#   score_direction     -> probability of concordant direction
#   score_translational -> score_rank * score_direction (high rank AND concordant)

# Best model per task (by out-of-fold metric)
best_shared_model <- shared_metrics %>%
  filter(framing == "A_predictive", feature_set == "Full + BTM") %>%
  slice_max(roc_auc, n = 1) %>%
  pull(model)

best_rank_model <- rank_metrics %>%
  filter(feature_set == "Mouse + layers") %>%
  slice_max(rsq, n = 1) %>%
  pull(model)

best_direction_model <- direction_metrics %>%
  filter(feature_set == "Direction + layers") %>%
  slice_max(roc_auc, n = 1) %>%
  pull(model)

cat("Best models -> shared:", best_shared_model,
    "| rank:", best_rank_model,
    "| direction:", best_direction_model, "\n")

# Map model name -> out-of-fold prediction column
model_col <- function(name) {
  case_when(
    name == "Logistic Regression" ~ "pred_log",
    name == "Linear (LM)"         ~ "pred_lm",
    name == "Lasso"               ~ "pred_lasso",
    name == "Random Forest"       ~ "pred_rf",
    name == "Neural Network"      ~ "pred_nn",
    TRUE ~ NA_character_
  )
}

# Extract the out-of-fold prediction of the chosen model and feature set
pick_pred <- function(pred_oof, model_name, framing_keep, set_keep) {
  col <- model_col(model_name)
  pred_oof %>%
    filter(framing == framing_keep, feature_set == set_keep) %>%
    select(treatment, pathogen, timepoint_comparison, hgnc_symbol, pred = all_of(col))
}

score_shared_oof    <- pick_pred(shared_pred_oof,    best_shared_model,    "A_predictive", "Full + BTM")
score_rank_oof      <- pick_pred(rank_pred_oof,      best_rank_model,      "transfer",     "Mouse + layers")
score_direction_oof <- pick_pred(direction_pred_oof, best_direction_model, "transfer",     "Direction + layers")
```

```{r}
# Attach the scores to df_all (one row per gene and condition)
score_tbl <- df_all %>%
  left_join(score_shared_oof %>%
              rename(score_shared = pred),
            by = c("treatment", "pathogen", "timepoint_comparison", "hgnc_symbol"),
            relationship = "many-to-one") %>%
  left_join(score_rank_oof %>%
              rename(score_rank = pred),
            by = c("treatment", "pathogen", "timepoint_comparison", "hgnc_symbol"),
            relationship = "many-to-one") %>%
  left_join(score_direction_oof %>%
              rename(score_direction = pred),
            by = c("treatment", "pathogen", "timepoint_comparison", "hgnc_symbol"),
            relationship = "many-to-one") %>%
  mutate(score_translational = score_rank * score_direction)

score_plot <- score_tbl %>%
  bind_rows(score_tbl %>% mutate(treatment = "Pooled",
                                 pathogen = "",
                                 timepoint_comparison = ""))

# Save
saveRDS(score_plot, here("Modelling", "Models", "score_table_v2.rds"))
```

## Observed vs predicted table

```{r}
# ==============================================================================
# OBSERVED VS PREDICTED - table for exploration (plot, threshold, select genes)
# ==============================================================================
observed_vs_predicted <- df_all %>%
  select(hgnc_symbol, treatment, pathogen, timepoint_comparison,
         status_original, rank_mouse, rank_human, sign_mouse, sign_concordant,
         mean_log2fc_mouse, mean_log2fc_human) %>%
  left_join(score_shared_oof %>%
              rename(pred_shared = pred),
            by = c("treatment", "pathogen", "timepoint_comparison", "hgnc_symbol"),
            relationship = "many-to-one") %>%
  left_join(score_rank_oof %>%
              rename(pred_rank = pred),
            by = c("treatment", "pathogen", "timepoint_comparison", "hgnc_symbol"),
            relationship = "many-to-one") %>%
  left_join(score_direction_oof %>%
              rename(pred_direction = pred),
            by = c("treatment", "pathogen", "timepoint_comparison", "hgnc_symbol"),
            relationship = "many-to-one") %>%
  distinct()

observed_vs_predicted

saveRDS(observed_vs_predicted, here("Modelling", "Tables", "observed_vs_predicted.rds"))
write_csv(observed_vs_predicted, here("Modelling", "Tables", "observed_vs_predicted.csv"))
```

# Priority list (strict) + annotations

```{r}
# ==============================================================================
# Priority list - candidate genes (pooled)
# ==============================================================================
# The table carries all scores, so the ranking can be defined later.
# Example filters (uncomment/adapt as needed):
#   filter(score_shared >= 80)                       # by shared score
#   filter(score_rank >= 80)                         # by predicted human rank
#   filter(score_direction >= 80)                    # by directional concordance
#   filter(score_translational >= 60)                # by combined score
#   filter(score_shared >= 80, score_translational >= 60)   # by several together
#   arrange(desc(score_translational))               # order by any score
pooled <- score_plot %>% filter(treatment == "Pooled")

priority_list <- pooled %>%
  filter(dual_rank >= 75, rank_diff <= 25, p_adj_mouse <= 0.05) %>%
  distinct(hgnc_symbol, .keep_all = TRUE) %>%
  arrange(desc(conv_rel)) %>%
  slice_head(n = 20) %>%
  select(hgnc_symbol, process,
         rank_mouse, rank_human, rank_diff, dual_rank, conv_rel,
         mean_log2fc_mouse, mean_log2fc_human, sign_mouse, sign_human, sign_concordant,
         score_shared, score_rank, score_direction, score_translational,
         status_original, p_adj_mouse)

priority_list
write_csv(priority_list, here("Modelling", "Tables", "priority_relevant_convergent_top20.csv"))

# Quick correlation diagnostics (Spearman rho + BH-adjusted p)
diag_stats <- function(x, y) {
  ok <- complete.cases(x, y)
  x <- x[ok]
  y <- y[ok]
  tibble(rho = cor(x, y, method = "spearman"),
         p_spearman = cor.test(x, y, method = "spearman", exact = FALSE)$p.value)
}
bind_rows(
  diag_stats(pooled$score_shared, pooled$rank_mouse) %>% mutate(comparison = "score_shared vs rank_mouse"),
  diag_stats(pooled$rank_mouse, pooled$rank_human) %>% mutate(comparison = "rank_mouse vs rank_human")
) %>%
  mutate(padj = p.adjust(p_spearman, method = "BH"))
```

# Figures

## Figure helpers

```{r}
library(patchwork)

# Colors for LEG status (Not LEG shown in white)
comp_colors <- c(colors$comparison, "Not LEG" = "white")

# Readable feature labels for plots.
# Derived from feature_dictionary (single source of truth) so labels never drift.
clean_feat <- function(x) {
  map <- setNames(feature_dictionary$feature_name,
                  feature_dictionary$feature_name_original)
  out <- unname(map[as.character(x)])
  ifelse(is.na(out), as.character(x), out)
}

# Forest plot: log-odds or standardized effect with 95% CI
forest_plot <- function(coefs, title, xlab) {
  coefs %>%
    mutate(feature = clean_feat(as.character(feature_name_original))) %>%
    filter(!is.na(feature),
           !is.na(estimate),
           !is.na(conf.low),
           !is.na(conf.high)) %>%
    mutate(feature = make.unique(feature)) %>%
    ggplot() +
    aes(x = estimate, y = fct_reorder(feature, estimate)) +
    geom_vline(xintercept = 0,
               linetype = "dashed",
               color = "gray50") +
    geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                   height = 0.2,
                   alpha = 0.6) +
    geom_point(size = 2,
               pch = 21,
               fill = "#4DBBD5FF") +
    theme_vaxgo() +
    labs(x = xlab,
         y = "",
         title = title)
}

# Gini importance bars (random forest)
gini_plot <- function(imp, title) {
  imp %>%
    mutate(feature = clean_feat(as.character(feature_name_original))) %>%
    filter(!is.na(feature), !is.na(importance)) %>%
    mutate(feature = make.unique(feature)) %>%
    ggplot() +
    aes(x = importance, y = fct_reorder(feature, importance)) +
    geom_col(fill = "#4cc9f0",
             color = "black",
             linewidth = 0.3) +
    theme_vaxgo() +
    labs(x = "Gini importance",
         y = "",
         title = title)
}

# Lasso coefficients (points only; penalized estimates have no valid SE/CI)
lasso_plot <- function(coefs, title, xlab) {
  coefs %>%
    mutate(feature = clean_feat(as.character(feature_name_original))) %>%
    filter(!is.na(feature), !is.na(estimate)) %>%
    mutate(feature = make.unique(feature)) %>%
    ggplot() +
    aes(x = estimate, y = fct_reorder(feature, estimate)) +
    geom_vline(xintercept = 0,
               linetype = "dashed",
               color = "gray50") +
    geom_point(size = 2,
               pch = 21,
               fill = "#4DBBD5FF") +
    theme_vaxgo() +
    labs(x = xlab,
         y = "",
         title = title)
}

# Spearman rho + BH-adjusted p annotation
cor_label <- function(df, x, y) {
  tibble(rho = cor(df[[x]], df[[y]], method = "spearman", use = "complete.obs"),
         p = cor.test(df[[x]], df[[y]], method = "spearman", exact = FALSE)$p.value) %>%
    mutate(padj = p.adjust(p, method = "BH"),
           label = sprintf("rho=%.2f\npadj=%.1e", rho, padj))
}
```

## Multilayer modelling

### Feature layers workflow

### Shared LEG classification

```{r fig.height=4, fig.width=8}
# Classification of Shared vs Mouse only (mouse-LEG universe), LOCO
shared_performance_bars <- shared_metrics %>%
  pivot_longer(c(roc_auc, pr_auc), names_to = "metric", values_to = "value") %>%
  mutate(metric = recode(metric, roc_auc = "ROC-AUC", pr_auc = "PR-AUC"),
         framing = factor(recode(framing, A_predictive = "Predictive (A)", B_explanatory = "Explanatory (B)"),
                          levels = c("Predictive (A)", "Explanatory (B)")))

shared_performance_plot <- shared_performance_bars %>%
  ggplot() +
  aes(x = feature_set, y = value, fill = model) +
  geom_col(position = position_dodge(0.8),
           width = 0.7,
           color = "black",
           linewidth = 0.3) +
  # Annotate the LOCO estimate on each bar
  geom_text(aes(label = sprintf("%.2f", value)),
            position = position_dodge(0.8),
            vjust = 0.5,
            hjust = 1.2,
            angle = 90,
            size = 3,
            color = "black") +
  facet_nested_wrap(metric %>% fct_rev() ~ framing,
                    nrow = 1) +
  scale_y_continuous(limits = c(0, 1), expand = c(0,0)) +
  scale_x_discrete(expand = c(0,0)) +
  scale_fill_manual(values = c("Random Forest" = "#4DBBD5FF",
                               "Logistic Regression" = "gray70",
                               "Lasso" = "#669bbc", "Neural Network" = "#D0D8FB")) +
  theme_vaxgo() +
  theme(legend.position = "top") +
  labs(x = "",
       y = "AUC",
       fill = "Algorithm",
       title = "Shared LEG classification") + 
  guides(fill = guide_legend(nrow = 1))

shared_performance_plot

shared_performance_plot %>% 
ggsave(file = here("Modelling", "Figures", "Shared_Performance_bars.png"),
       width = 8,
       height = 4)

```

### Rank and direction

```{r fig.height=4, fig.width=8}
# LOCO performance for rank transfer (R2) and directional concordance (ROC-AUC)
rank_bars <- rank_metrics %>%
  transmute(metric = "R2", task = "Rank", model, step = feature_set, value = rsq)

direction_bars <- direction_metrics %>%
  pivot_longer(c(roc_auc, pr_auc), names_to = "metric", values_to = "value") %>%
  mutate(metric = recode(metric, roc_auc = "ROC-AUC", pr_auc = "PR-AUC"),
         task = "Direction") %>%
  select(metric, task, model, step = feature_set, value)

rank_direction_data <- bind_rows(rank_bars, direction_bars) %>%
  mutate(task = fct_relevel(task, "Direction", "Rank"),
         metric = fct_relevel(metric,"R2", "ROC-AUC", "PR-AUC"))

rank_direction_plot <- rank_direction_data %>%
  ggplot() +
  aes(x = step %>% fct_rev(), y = value, fill = model) +
  geom_col(position = position_dodge(0.8),
           width = 0.7,
           color = "black",
           linewidth = 0.3) +
  # Annotate the LOCO estimate on each bar
  geom_text(aes(label = sprintf("%.2f", value)),
            position = position_dodge(0.8),
            vjust = -0.4,
            size = 3) +
  facet_nested_wrap(~task+ metric, scales = "free_x") +
    scale_y_continuous(limits = c(0, 1), expand = c(0,0)) +
  scale_x_discrete(expand = c(0,0)) +
  scale_fill_manual(values = c("Random Forest" = "#4DBBD5FF",
                               "Logistic Regression" = "gray70",
                               "Lasso" = "#669bbc", "Neural Network" = "#D0D8FB")) +
  theme_vaxgo() +
  theme(legend.position = "top") +
  labs(x = "",
       y = "Estimate",
       fill = "Algorithm",
       title = "Rank transfer and direction")
#Display
rank_direction_plot

#Save
rank_direction_plot %>% 
ggsave(file = here("Modelling", "Figures", "Rank_Direction_Performance_bars.png"),
       width = 8,
       height = 4)
```

### Shared score vs absolute rank

```{r fig.height=3, fig.width=5}
# score_shared vs absolute rank (mouse), layered points + marginals
score_shared_data <- pooled %>%
  filter(status_original %in% c("Mouse only", "Shared"))
score_shared_label <- cor_label(score_shared_data, "score_shared", "rank_mouse")

score_shared_vs_rank <- score_shared_data %>%
  ggplot() +
  aes(x = score_shared, y = rank_mouse) +
  geom_point(data = . %>% filter(status_original %in% c("Mouse only")),
             aes(fill = status_original),
             shape = 21,
             size = 2,
             alpha = 0.5,
             stroke = 0.3) +
  geom_point(data = . %>% filter(status_original %in% c("Shared")),
             aes(fill = status_original),
             shape = 21,
             size = 2,
             alpha = 1,
             stroke = 0.3) +
  #Right marginal
  geom_violin(aes(y = rank_mouse, x = 1.1, 
                  fill = status_original),
              width = 0.1,
              alpha = 0.5,
              color = NA,
              inherit.aes = FALSE) +
  geom_boxplot(aes(y = rank_mouse, 
                   fill = status_original,
                   x = 1.1),
               width = 0.1,
               alpha = 0,
               color = "black",
               outlier.shape = NA,
               inherit.aes = FALSE) +
  #Top marginal
  geom_violin(aes(x = score_shared, 
                  y = 105, 
                  fill = status_original),
              width = 5,
              alpha = 0.5,
              color = NA,
              inherit.aes = FALSE) +
  geom_boxplot(aes(x = score_shared,
                   fill = status_original,
                   y = 105),
               width = 5,
               alpha = 0,
               color = "black",
               outlier.shape = NA,
               inherit.aes = FALSE) +
  scale_fill_manual(values = comp_colors) +
  geom_text(data = score_shared_label,
            aes(x = 0, y = Inf, label = label),
            hjust = -0.05,
            vjust = 1.2,
            size = 3,
            inherit.aes = FALSE) +
  theme_vaxgo() +
  labs(x = "score_shared",
       y = "Absolute rank (mouse)",
       fill = "LEG status",
       title = "d  Shared LEG score")

score_shared_vs_rank
```

### Dual-species absolute rank

```{r fig.height=6, fig.width=8}
# Absolute rank (mouse) vs absolute rank (human), colored by LEG status
rank_human_label <- cor_label(pooled, "rank_mouse", "rank_human")

rank_mouse_vs_rank_human <- pooled %>%
  ggplot() +
  aes(x = rank_mouse, y = rank_human, fill = status_original) +
  geom_point(shape = 21,
             size = 2,
             alpha = 0.5,
             stroke = 0) +
  geom_abline(slope = 1,
              intercept = 0,
              linetype = "dashed",
              color = "gray50") +
  geom_violin(data = pooled,
              aes(y = rank_human, x = 110, fill = status_original),
              width = 15,
              alpha = 0.5,
              color = NA,
              inherit.aes = FALSE) +
  geom_boxplot(data = pooled,
               aes(y = rank_human, 
                   fill = status_original,
                   x = 110),
               width = 15,
               alpha = 0,
               color = "black",
               outlier.shape = NA,
               inherit.aes = FALSE) +
  geom_violin(data = pooled,
              aes(x = rank_mouse, y = 110, 
                  fill = status_original),
              width = 15,
              alpha = 0.5,
              color = NA,
              inherit.aes = FALSE) +
  geom_boxplot(data = pooled,
               aes(x = rank_mouse, y = 110,
                   fill = status_original),
               width = 15,
               alpha = 0,
               color = "black",
               outlier.shape = NA,
               inherit.aes = FALSE) +
  scale_fill_manual(values = comp_colors) +
  geom_text(data = rank_human_label,
            aes(x = 0, y = Inf, label = label),
            hjust = -0.05,
            vjust = 1.2,
            size = 3,
            inherit.aes = FALSE) +
  theme_vaxgo() +
  labs(x = "Absolute rank (mouse)",
       y = "Absolute rank (human)",
       fill = "LEG status",
       title = "e  Dual-species absolute rank")

rank_mouse_vs_rank_human
```

### Priority genes heatmap

```{r fig.height=6, fig.width=8}
# Clustered heatmap of the prioritized relevant-convergent genes
heat_dat <- pooled %>%
  filter(hgnc_symbol %in% priority_list$hgnc_symbol) %>%
  distinct(hgnc_symbol, .keep_all = TRUE) %>%
  transmute(
    hgnc_symbol,
    `1/SE` = inverse_se_mouse,
    `%Protein identity` = identity_human2mouse,
    `Kimura` = dist_k80,
    `nTotal TFs` = n_tf_total,
    `%Shared TFs` = pct_tf_shared,
    `nTotal cCREs` = n_total_cres_gene,
    `PLS nCREs` = n_type_PLS,
    `pELS nCREs` = n_type_pELS,
    `dELS nCREs` = n_type_dELS,
    `%Match Type (PLS)` = `pct_match_type_PLS`,
    `%Match Type (pELS)` = `pct_match_type_pELS`,
    `%Match Type (dELS)` = `pct_match_type_dELS`,
    `CTCF match` = `pct_match_ctcf_dELS_CTCF-bound`
  )

mat <- heat_dat %>%
  column_to_rownames("hgnc_symbol") %>%
  as.matrix()
mat_z <- scale(mat)
gene_levels <- rownames(mat_z)[hclust(dist(mat_z))$order]
feat_levels <- colnames(mat_z)[hclust(dist(t(mat_z)))$order]

heat_long <- heat_dat %>%
  pivot_longer(-hgnc_symbol, names_to = "feature", values_to = "value") %>%
  group_by(feature) %>%
  mutate(z = as.numeric(scale(value))) %>%
  ungroup() %>%
  mutate(hgnc_symbol = factor(hgnc_symbol, levels = gene_levels),
         feature = factor(feature, levels = feat_levels))

priority_genes_heatmap <- heat_long %>%
  ggplot() +
  aes(x = feature, y = hgnc_symbol, fill = z) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_gradientn(colours = c("#4361ee", "white", "#e5383b"),
                       limits = c(-2.5, 2.5),
                       oob = scales::squish) +
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
        axis.text.y = element_text(size = 7)) +
  labs(x = "",
       y = "",
       fill = "z-score",
       title = "f  Prioritized relevant-convergent genes")

priority_genes_heatmap
```



## Feature contributions

### Shared classification (Lasso coefficients)

```{r fig.height=6, fig.width=8}
# Shared vs Mouse only: Lasso coefficients (best-performing model for this task)
lasso_coefs_plot_shared <- shared_coefs_lasso %>%
  filter(framing == "A_predictive", feature_set == "Full + BTM") %>%
  lasso_plot("Shared: Predictive", "Lasso coefficient") +
  scale_x_continuous(expand = c(0.01,0.01))

lasso_coefs_plot_shared
```

### Rank transfer (standardized effect + Gini)

```{r fig.height=6, fig.width=8}
# Human rank prediction: standardized effects (linear) and Gini importance (random forest)
linear_coefs_plot_rank <- rank_coefs %>%
  filter(feature_set == "Mouse + layers") %>%
  forest_plot("Rank: Mouse + layers", "Std. effect size") +
  scale_x_continuous(expand = c(0.01,0.01))

rf_imp_plot_rank <- rank_imp %>%
  filter(feature_set == "Mouse + layers") %>%
  gini_plot("Rank: Mouse + layers") +
  scale_x_continuous(expand = c(0.01,0.01))

linear_coefs_plot_rank
rf_imp_plot_rank
```

### Direction (log-odds + Gini)

```{r fig.height=6, fig.width=8}
# Directional concordance: log-odds (logistic) and Gini importance (random forest)
log_coefs_plot_direction <- direction_coefs %>%
  filter(feature_set == "Direction + layers") %>%
  forest_plot("Direction + layers", xlab = "Log-odds") +
  scale_x_continuous(expand = c(0.01,0.01))

rf_imp_plot_direction <- direction_imp %>%
  filter(feature_set == "Direction + layers") %>%
  gini_plot("Direction + layers") +
  scale_x_continuous(expand = c(0.01,0.01))

log_coefs_plot_direction
rf_imp_plot_direction
```

### Assembly

```{r fig.height=15, fig.width=13}
feature_contributions_figure = 
  (linear_coefs_plot_rank | rf_imp_plot_rank) /
  (log_coefs_plot_direction | rf_imp_plot_direction) /
  lasso_coefs_plot_shared +
  plot_annotation(title = "Feature contributions")

ggsave(here("Modelling", "Figures", "Models_Coeff_Contributions.png"),
       feature_contributions_figure,
       width = 13,
       height = 15,
       dpi = 300)
feature_contributions_figure
```

