---
title: "Data curation"
output: html_document
---

# Required

Load `here` and source the project-wide library and utility script. Packages are
installed and checked once inside `required.R`, so nothing is installed here.

```{r message=FALSE, warning=FALSE}
# Load the paths helper and source packages, palettes, theme, and helpers --------
library(here)
source(here("scripts_notebooks", "required.R"))
```

# 1. Data curation

## BioProject


Barplot showing the number of immunization datasets per genus, highlighting Homo and Mus in blue.

```{r fig.height=8, fig.width=2, message=FALSE, warning=FALSE}
# Import cleaned data --------
datacuration_clean <- read_csv(here("tables",  "Data curation", "datacuration_step2.csv"))

# Filter for vaccination-related titles, separate genus, and count by genus --------
organisms_datacuration <- datacuration_clean %>%
  filter(!str_detect(str_to_lower(title), "cancer|tumor|autoimmune|drug")) %>% 
  filter(str_detect(str_to_lower(title), "vaccin|immuniz")) %>% 
  separate(organism, into = c("genus", "species"), remove = FALSE) %>% 
  group_by(genus) %>% 
  summarise(n_organism = n()) %>% 
  arrange(-n_organism) %>% 
  mutate(genus = fct_reorder(genus, -n_organism))

# Build the barplot with highlighted Homo and Mus bars --------
organisms_datacuration_barplot <- organisms_datacuration %>% 
  ggplot() +
  aes(x = n_organism,
      y = genus %>% fct_reorder(n_organism)) +
  geom_col(fill = "gray80") + 
  geom_col(data = organisms_datacuration %>% filter(genus %in% c("Homo", "Mus")),
           fill = "#4cc9f0") + 
  theme_light()+
  geom_text(aes(label = n_organism),
            hjust = -0.2,
            size = 3) +
  facet_wrap(~genus, ncol = 1, scales = "free_y") +
  scale_x_continuous(expand = c(0.01,0.01),
                     limits = c(0, 320)) +
  theme(
    legend.position = c(0.95,0.5), 
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.title = element_text(size = 10, face = "bold", margin = margin(0, 0, 1, 2.5)),
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    panel.border = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.line.x = element_blank(),
    axis.line.y = element_blank(),
    axis.ticks.x = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_blank(),
    panel.spacing = unit(0, "cm"),
    strip.text = element_text(color = "black", size = 10,face = "italic", hjust = 0, margin = margin(0, 0, 1, 2)),
    plot.margin = margin(0.2, 0, 0, 0, "cm")) +
    labs(x = "",
       y = "",
       title = "Number of immunization\ndatasets, by genus")

#Display
organisms_datacuration_barplot

# Save the barplot --------
ggsave(organisms_datacuration_barplot, 
       filename = here("Figures", "organisms_datacuration_barplot.png"),
         width = 2, height = 6)
```

### BioProject metadata

Filter for Mus musculus vaccination datasets and retrieve BioProject metadata via Entrez.

```{r}
# Filter for Mus musculus vaccination-related datasets --------
mice_vaccination_mining <- datacuration_clean %>% 
  filter(str_detect(organism,"Mus musculus")) %>% 
  filter(!str_detect(str_to_lower(title), "cancer|tumor|autoimmune|drug")) %>% 
  filter(str_detect(str_to_lower(title), "vaccin|immuniz"))

# Test retrieval for the first BioProject ID --------
bioprojectid <- mice_vaccination_mining$id[[1]]
bioproject_meta <- entrez_summary(db = "bioproject", id = bioprojectid) %>% 
  as.tibble() %>% 
  select(project_name,
         project_description,
         organism_name,
         project_id,
         project_data_type,
         project_target_material,
         project_target_capture,
         organism_name)

# Loop over all BioProject IDs to retrieve metadata --------
bioproject_ids <- mice_vaccination_mining$id
bioproject_metadata_df <- map_dfr(bioproject_ids, ~
  entrez_summary(db = "bioproject", id = .x) %>%
    as.list() %>%
    as_tibble() %>%
    select(
      project_name,
      project_description,
      organism_name,
      project_data_type,
      project_target_material,
      project_target_capture
    ) %>%
    mutate(bioproject_id = .x)
)

# Remove datasets with cancer/autoimmune/drug descriptions --------
bioproject_metadata_df <- bioproject_metadata_df %>% 
  filter(!str_detect(project_description, "cancer|neoantigen|tumor|autoimmune|drug|opioid"))

# Save and reimport the filtered metadata --------
write_excel_csv2(bioproject_metadata_df, here("tables", "Data curation", "datacuration_step2.csv"))

bioproject_metadata_df <- read_delim(here("tables", "Data curation", "datacuration_step2.csv"))

```

## GEO

Retrieve metadata from GEO for individual series and annotate platform information.

```{r}
# Get metadata for a single GEO series (commented out) --------
geo_ids <- "GSE120661"
series <- entrez_search(db = "gds",
                        term = paste(geo_ids, "[ACCN] AND gse[ETYP]", sep = ""))
series_summary <- entrez_summary(db = "gds", id = series$ids)

series_summary_df <- data_frame(bioproject = series_summary$bioproject,
                               uid = series_summary$uid,
                               gse = series_summary$gse,
                               gds = series_summary$gds,
                               title_series = series_summary$title,
                               summary = series_summary$summary,
                               gpl = series_summary$gpl,
                               seriestitle = series_summary$seriestitle,
                               n_samples = series_summary$n_samples,
                               platformtitle = series_summary$platformtitle,
                               platformtaxa = series_summary$platformtaxa,
                               ptechtype = series_summary$ptechtype) %>%
  mutate(gpl = paste0("GPL", gpl))

# Annotate platform by importing and joining GPL annotation files --------
platform_human_1_5000 <- read_csv(here("tables", "raw_data", "platform_human_1-5000.csv"))
platform_human_5000_6821 <- read_csv(here("tables", "raw_data", "platform_human_5000-6821.csv"))
platform_mouse <- read_csv(here("tables", "raw_data", "platform_mouse.csv"))

gpl_annotation_human_mouse <- platform_mouse %>%
  bind_rows(platform_human_1_5000,
            platform_human_5000_6821) %>%
  clean_names() %>%
  rename(gpl = accession,
         title_platform = title)

saveRDS(gpl_annotation_human_mouse, file = here("tables", "raw_data", "gpl_annotation_human_mouse.rds"))


```

Loop over GEO accession IDs to retrieve metadata for multiple datasets.

```{r}
# Import GPL annotation --------
gpl_annotation_human_mouse <- 
  readRDS(file = here("tables", "raw_data", "gpl_annotation_human_mouse.rds"))

# Parse GEO IDs from the downloaded result file --------
geo_ids <- read_delim(here("tables", "Data curation", "gds_result.txt"), delim = "\n", col_names = FALSE) %>%
  mutate(X1 = str_c("GSE", X1, sep = "") %>%
           str_remove(., "200")) %>% 
  pull()

text <- readLines(here("tables","Data curation", "gds_result.txt"))
text_collapsed <- paste(text, collapse = "\n")
gse_ids <- str_extract_all(geo_ids, "GSE\\d+\\b") %>% unlist() %>% unique()

# Loop over GSE IDs to retrieve metadata via Entrez --------
geo_metadata_df <- map_dfr(gse_ids, ~ {
  result <- entrez_search("gds", term = paste(.x, "[ACCN] AND gse[ETYP]", sep = ""))

  if (length(result$ids) == 0) {
    return(tibble(gse = .x, error = "Not found"))
  }

  summary <- entrez_summary(db = "gds", id = result$ids) %>% as.list()

  tibble(field = names(summary), value = summary) %>%
    pivot_wider(names_from = field, values_from = value) %>%
    mutate(gse = .x)
})

# Clean and select relevant metadata columns --------
datacuration_geo_metadata_df <- geo_metadata_df %>% 
  drop_na(uid) %>% 
  rename(title_series = title) %>% 
  mutate(bioproject_id = str_remove(bioproject, "PRJNA"),
           gpl = paste0("GPL", gpl)) %>% 
  select(uid, gse_id = accession, 
         title_series, summary, samples_count,
         gpl, taxon, entrytype, gdstype, pdat, suppfile, pubmedids, bioproject)

# Save GEO metadata --------
saveRDS(geo_metadata_df, file = here("tables", "processed_data", "geo_metadata_df.rds"))

```

## Merge GEO and BioProject metadata

Join the manually curated BioProject annotations with GEO metadata, separate multi-entry fields, and clean GPL identifiers.

```{r}
# Import the manually curated BioProject annotations and the GEO metadata --------
datacuration_step4 <- read_excel(here("tables", "Data curation", "datacuration_step4.xlsx"), 
    sheet = "BioProject_Step4_HumanAtlas_Yes")

geo_metadata_df <-
  readRDS(file = here("tables", "processed_data", "geo_metadata_df.rds")) %>% 
  select(gse_id, gpl, gdstype, taxon, pdat, suppfile, pubmedids) 

# Join annotations with GEO metadata and arrange columns --------
datacuration_annotated <- datacuration_step4 %>% 
   rename(gse_id = gse) %>% 
  inner_join(geo_metadata_df, by = "gse_id") %>% 
  select(
         title,
         project_description,
         gse_id, 
         keywords,
         notes,
         timepoints,
         time_type,
         doses,
         vaccine_type, 
         pathogen_type,
         target_pathogen,
         target_antigen, 
         adjuvanted, 
         viral_vector,
         sample_source, 
         administration_route, 
         sequencing,
         gpl,
         gdstype,
         taxon,
         pdat,
         suppfile,
         pubmedids
         ) %>% 
  separate_rows(sample_source, sep = "//") %>% 
  separate_rows(administration_route, sep = "//") %>%
  separate_rows(target_antigen, target_pathogen, sep = "//") %>% 
  separate_rows(adjuvanted, sep = "//") %>%
  mutate(gpl = str_replace(gpl, ";", "//")) %>%
  separate_rows(gpl, sequencing, sep = "//") %>%
  mutate(gpl = ifelse(grepl("GPL", gpl), gpl, paste0("GPL", gpl))) %>% 
  mutate(sequencing = case_when(gdstype == "Expression profiling by array" ~ "Microarray",
                                gdstype == "Expression profiling by RT-PCR" ~ "RT-PCR",
                                TRUE ~ sequencing)) %>% 
  separate(sample_source, into = c("main_sample_source", "specific_sample_source"),
    sep = " \\(|\\)", remove = TRUE, extra = "drop", fill = "right") %>% 
  separate(target_antigen, into = c("target_antigen", "specific_antigen"),
    sep = " \\(|\\)", remove = TRUE, extra = "drop", fill = "right") %>% 
  separate(target_pathogen, into = c("target_pathogen", "specific_pathogen"),
    sep = " \\(|\\)", remove = TRUE, extra = "drop", fill = "right") %>% 
  distinct() 

# Check for duplicates and view the result --------
datacuration_annotated %>% 
  get_dupes()
  
datacuration_annotated %>% 
  View()

# Save annotated dataset and filter blood samples --------
write_csv(datacuration_annotated, file = here("tables", "Data curation", "datacuration_annotated.csv"))

datacuration_blood <- datacuration_annotated %>% 
  filter(str_detect(main_sample_source, "Blood|PBMCs"))
```

### Selected datasets

Import the annotated dataset and the manually selected datasets, then standardize timepoints and adjacency labels.

```{r}
# Reimport the annotated dataset and standardize adjuvanted column --------
datacuration_annotated <- 
  read_csv(file = here("tables", "Data curation", "datacuration_annotated.csv")) %>% 
  mutate(adjuvant = adjuvanted,
         adjuvanted = if_else(adjuvanted != "No", "Yes", adjuvanted))

# Import the selected datasets and separate multi-entry fields --------
datacuration_selecteddatasets <- read_csv(here("tables", "Data curation", "datacuration_selected.csv")) %>% 
   separate_rows(timepoints, sep = "//") %>% 
  separate_rows(strain, sep = "//")

# Convert timepoints to a common day unit and standardize adjacency --------
datacuration_selecteddatasets_long <- datacuration_selecteddatasets %>% 
  select(gse_id, n_samples, timepoints, time_type, vaccine_type, target_pathogen, target_antigen:sequencing, strain) %>% 
  mutate(timepoints = as.numeric(timepoints),
         n_samples = as.numeric(n_samples),
         timepoints = case_when(time_type == "Hour" ~ timepoints/24,
                                time_type == "Week" ~ timepoints * 7,
                                TRUE ~ timepoints) %>% round(1),
         time_type = "Hour",
         adjuvant = adjuvanted,
         adjuvanted = if_else(adjuvanted != "No", "Yes", adjuvanted))
```


## Visualization

### Sunburst

Prepare a sunburst dataset with hierarchical metadata columns and save for Flourish visualization.

```{r}
# Select hierarchical columns for the sunburst visualization --------
datacuration_annotated_sunburst <- datacuration_annotated %>% 
  select(
         gse_id, 
         title,
         pubmedids,
         pathogen_type, 
         target_pathogen,
         target_antigen, 
         vaccine_type, 
         adjuvanted,
         adjuvant,
         administration_route,
         main_sample_source, 
         specific_sample_source,
         sequencing
         ) %>% 
  distinct()

# Save the sunburst dataset --------
write_csv(datacuration_annotated_sunburst, file = here("tables", "Data curation", "datacuration_annotated_sunburst.csv"))

```

### Sankey diagram

Build each layer of a Sankey diagram from pathogen type through sample source, then bind and save.

#### All

```{r}
# Layer 1: pathogen_type to target_pathogen --------
datacuration_annotated_sankey_level1 <- datacuration_annotated %>% 
  select(source = pathogen_type, target = target_pathogen, gse_id) %>%  
  group_by(source, target) %>% 
  summarise(value = n(),
            step_from = 0, step_to = 1,
            gse_ids = paste0(gse_id, collapse = ";"),
            layer = "pathogen_type to target_pathogen") 

# Layer 2: target_pathogen to target_antigen --------
datacuration_annotated_sankey_level2 <- datacuration_annotated %>% 
  select(source = target_pathogen, target = target_antigen, gse_id) %>%
  group_by(source, target) %>% 
  summarise(value = n(),
            step_from = 1, step_to = 2,
            gse_ids = paste0(gse_id, collapse = ";"),
            layer = "target_pathogen to target_antigen") 

# Layer 3: target_antigen to vaccine_type --------
datacuration_annotated_sankey_level3 <- datacuration_annotated %>% 
  select(source = target_antigen, target = vaccine_type, gse_id) %>%
  group_by(source, target) %>% 
  summarise(value = n(),
            step_from = 2, step_to = 3,
            gse_ids = paste0(gse_id, collapse = ";"),
            layer = "target_antigen to vaccine_type")

# Layer 4: vaccine_type to adjuvanted --------
datacuration_annotated_sankey_level4 <- datacuration_annotated %>% 
  select(source = vaccine_type, target = adjuvanted, gse_id) %>%
  group_by(source, target) %>% 
  summarise(value = n(),
            step_from = 3, step_to = 4,
            gse_ids = paste0(gse_id, collapse = ";"),
            layer = "vaccine_type to adjuvanted") %>% 
  mutate(source = if_else(is.na(source), "No", source),
         target = if_else(is.na(target), "No", target))

# Layer 5: adjuvanted to administration_route --------
datacuration_annotated_sankey_level5 <- datacuration_annotated %>% 
  select(source = adjuvanted, target = administration_route, gse_id) %>%
  group_by(source, target) %>% 
  summarise(value = n(),
            step_from = 4, step_to = 5,
            gse_ids = paste0(gse_id, collapse = ";"),
            layer = "adjuvanted to administration_route") %>% 
  mutate(source = if_else(is.na(source), "No", source),
         target = if_else(is.na(target), "No information", target))

# Layer 6: administration_route to main_sample_source --------
datacuration_annotated_sankey_level6 <- datacuration_annotated %>% 
  select(source = administration_route, target = main_sample_source, gse_id) %>%
  group_by(source, target) %>% 
  summarise(value = n(),
            step_from = 5, step_to = 6,
            gse_ids = paste0(gse_id, collapse = ";"),
            layer = "administration_route to main_sample_source")

# Bind all layers into a single dataframe --------
datacuration_annotated_sankey <- datacuration_annotated_sankey_level1 %>% 
  bind_rows(datacuration_annotated_sankey_level2,
            datacuration_annotated_sankey_level3,
            datacuration_annotated_sankey_level4,
            datacuration_annotated_sankey_level5,
            datacuration_annotated_sankey_level6) %>% 
  distinct()


# Save the Sankey dataset --------
write_csv(datacuration_annotated_sankey, file = here("tables", "Data curation", "datacuration_annotated_sankey_all.csv"))

```

#### Blood only

Filter for blood samples and rebuild the Sankey layers from pathogen type through sample source.

```{r}
# Filter for blood samples only --------
datacuration_annotated_blood <- datacuration_annotated %>% 
  filter(main_sample_source == "Blood")

# Layer 1: pathogen_type to target_pathogen --------
datacuration_annotated_sankey_level1 <- datacuration_annotated_blood %>% 
  select(source = pathogen_type, target = target_pathogen, gse_id) %>%  
  group_by(source, target) %>% 
  summarise(value = n(),
            step_from = 0, step_to = 1,
            gse_ids = paste0(gse_id, collapse = ";"),
            layer = "pathogen_type to target_pathogen") 

# Layer 2: target_pathogen to target_antigen --------
datacuration_annotated_sankey_level2 <- datacuration_annotated_blood %>% 
  select(source = target_pathogen, target = target_antigen, gse_id) %>%
  group_by(source, target) %>% 
  summarise(value = n(),
            step_from = 1, step_to = 2,
            gse_ids = paste0(gse_id, collapse = ";"),
            layer = "target_pathogen to target_antigen") 

# Layer 3: target_antigen to vaccine_type --------
datacuration_annotated_sankey_level3 <- datacuration_annotated_blood %>% 
  select(source = target_antigen, target = vaccine_type, gse_id) %>%
  group_by(source, target) %>% 
  summarise(value = n(),
            step_from = 2, step_to = 3,
            gse_ids = paste0(gse_id, collapse = ";"),
            layer = "target_antigen to vaccine_type")

# Layer 4: vaccine_type to adjuvanted --------
datacuration_annotated_sankey_level4 <- datacuration_annotated_blood %>% 
  select(source = vaccine_type, target = adjuvanted, gse_id) %>%
  group_by(source, target) %>% 
  summarise(value = n(),
            step_from = 3, step_to = 4,
            gse_ids = paste0(gse_id, collapse = ";"),
            layer = "vaccine_type to adjuvanted") %>% 
  mutate(source = if_else(is.na(source), "No", source),
         target = if_else(is.na(target), "No", target))

# Layer 5: adjuvanted to administration_route --------
datacuration_annotated_sankey_level5 <- datacuration_annotated_blood %>% 
  select(source = adjuvanted, target = administration_route, gse_id) %>%
  group_by(source, target) %>% 
  summarise(value = n(),
            step_from = 4, step_to = 5,
            gse_ids = paste0(gse_id, collapse = ";"),
            layer = "adjuvanted to administration_route") %>% 
  mutate(source = if_else(is.na(source), "No", source),
         target = if_else(is.na(target), "No information", target))

# Layer 6: administration_route to main_sample_source --------
datacuration_annotated_sankey_level6 <- datacuration_annotated_blood %>% 
  select(source = administration_route, target = main_sample_source, gse_id) %>%
  group_by(source, target) %>% 
  summarise(value = n(),
            step_from = 5, step_to = 6,
            gse_ids = paste0(gse_id, collapse = ";"),
            layer = "administration_route to main_sample_source")

# Bind all layers into a single dataframe --------
datacuration_annotated_sankey <- datacuration_annotated_sankey_level1 %>% 
  bind_rows(datacuration_annotated_sankey_level2,
            datacuration_annotated_sankey_level3,
            datacuration_annotated_sankey_level4,
            datacuration_annotated_sankey_level5,
            datacuration_annotated_sankey_level6) %>% 
  distinct()
  
# Save the blood-only Sankey dataset --------
write_csv(datacuration_annotated_sankey, file = here("tables", "Data curation", "datacuration_annotated_sankey_blood.csv"))

```

Sankey diagram of blood samples, showing flows from pathogen type through sequencing method per GSE ID.

```{r fig.height=10, fig.width=10}
# Load ggsankey library --------
library(ggsankey)

# Prepare long-format Sankey data --------
sankey_df <- datacuration_annotated_blood %>%
  make_long(pathogen_type, 
         target_pathogen,
         vaccine_type, 
         target_antigen, 
         adjuvanted,
         specific_sample_source,
         sequencing,  
         gse_id) %>% 
  drop_na()

# Build the Sankey ggplot --------
datacuration_blood_sankey_ggplot <- ggplot(sankey_df, aes(x = x, 
               next_x = next_x, 
               node = node, 
               next_node = next_node,
               fill = factor(node),
               label = node
               )) +
  geom_alluvial(flow.alpha = 0.6, 
              space = 1) +
  geom_alluvial_label(size = 3.5, space = 1, 
                     color = "black", 
                     fill = "white", 
                     alpha = 0.5) +
  scale_fill_viridis_d(drop = FALSE) +
  theme_alluvial(base_size = 12) +
  theme(legend.position = "none")

datacuration_blood_sankey_ggplot

```

### Heatmap

#### All datasets

Heatmap of all datasets: columns are metadata variables, rows are clustered by pathogen type with unique color per category.

```{r fig.height=21, fig.width=10}
# Prepare long-format heatmap data with unique GSE IDs --------
heatmap_df <- datacuration_annotated %>%
  select(pathogen_type, target_pathogen, target_antigen, vaccine_type, 
         adjuvanted, administration_route, main_sample_source, 
         specific_sample_source, sequencing, gse_id) %>% 
  group_by(gse_id) %>%
  mutate(gse_id_unique = paste0(gse_id, "_", row_number()))  %>%
  ungroup(gse_id) %>% 
  select(-gse_id) %>% 
  pivot_longer(cols = -gse_id_unique, names_to = "variable", values_to = "value") %>% 
  mutate(variable = fct_relevel(variable, c("pathogen_type", "target_pathogen",
                                            "target_antigen", "vaccine_type",
                                            "adjuvanted", "administration_route",
                                            "main_sample_source", "specific_sample_source",
                                            "sequencing")))

# Generate random colors for each unique value --------
colors_all <- c("#90A4AEFF", "#4DBBD5FF", "#4361ee")
paleta <- colorRampPalette(colors_all)
cores_15 <- paleta(100)
levels_heatmap <- unique(na.omit(heatmap_df$value))
levels_heatmap <- levels_heatmap[!levels_heatmap %in% c("NA", "No")]
set.seed(123)
colors_random <- sample(cores_15, length(levels_heatmap), replace = TRUE)
palette_final <- c(setNames(colors_random, levels_heatmap))

# Cluster rows by pathogen type using hierarchical clustering --------
heatmap_wide <- heatmap_df %>%
  select(gse_id_unique, variable, value) %>%
  pivot_wider(names_from = variable, values_from = value) %>% 
  mutate(gse_id_unique = as.character(gse_id_unique))

heatmap_numeric_virus <- heatmap_wide %>%
  filter(pathogen_type == "Virus") %>% 
  mutate(across(-gse_id_unique, ~as.integer(as.factor(.))))
dist_matrix <- dist(select(heatmap_numeric_virus, -gse_id_unique), method = "euclidean")
hc <- hclust(dist_matrix, method = "ward.D2")
ordered_gse_virus <- heatmap_numeric_virus$gse_id_unique[hc$order]

heatmap_numeric_bacteria <- heatmap_wide %>%
  filter(pathogen_type == "Bacteria") %>% 
  mutate(across(-gse_id_unique, ~as.integer(as.factor(.))))
dist_matrix <- dist(select(heatmap_numeric_bacteria, -gse_id_unique), method = "euclidean")
hc <- hclust(dist_matrix, method = "ward.D2")
ordered_gse_bacteria <- heatmap_numeric_bacteria$gse_id_unique[hc$order]

heatmap_numeric_fungi <- heatmap_wide %>%
  filter(pathogen_type == "Fungi") %>% 
  mutate(across(-gse_id_unique, ~as.integer(as.factor(.))))
dist_matrix <- dist(select(heatmap_numeric_fungi, -gse_id_unique), method = "euclidean")
hc <- hclust(dist_matrix, method = "ward.D2")
ordered_gse_fungi <- heatmap_numeric_fungi$gse_id_unique[hc$order]

heatmap_numeric_protozoa <- heatmap_wide %>%
  filter(pathogen_type == "Protozoa") %>% 
  mutate(across(-gse_id_unique, ~as.integer(as.factor(.))))
dist_matrix <- dist(select(heatmap_numeric_protozoa, -gse_id_unique), method = "euclidean")
hc <- hclust(dist_matrix, method = "ward.D2")
ordered_gse_protozoa <- heatmap_numeric_protozoa$gse_id_unique[hc$order]

heatmap_df <- heatmap_df %>%
  mutate(gse_id_unique = fct_relevel(gse_id_unique, c(ordered_gse_virus, ordered_gse_bacteria, ordered_gse_fungi, ordered_gse_protozoa)) %>% fct_rev())

# Build the heatmap with one geom_tile layer per variable --------
heatmap_plot <- heatmap_df %>% 
  ggplot(aes(x = variable, y = gse_id_unique, label = value)) +
  geom_tile(data = ~filter(.x, variable == "pathogen_type"), aes(fill = value)) +
  scale_fill_manual(values = c(palette_final), na.value = "transparent", name = "pathogen_type") +
  new_scale_fill() +
  geom_tile(data = ~filter(.x, variable == "target_pathogen"), aes(fill = value)) +
  scale_fill_manual(values = c(palette_final), na.value = "transparent", name = "target_pathogen") +
  new_scale_fill() +
  geom_tile(data = ~filter(.x, variable == "target_antigen"), aes(fill = value)) +
  scale_fill_manual(values = c(palette_final), na.value = "transparent", name = "target_antigen") +
  new_scale_fill() +
  geom_tile(data = ~filter(.x, variable == "vaccine_type"), aes(fill = value)) +
  scale_fill_manual(values = c(palette_final), na.value = "transparent", name = "vaccine_type") +
  new_scale_fill() +
  geom_tile(data = ~filter(.x, variable == "adjuvanted"), aes(fill = value), na.rm = TRUE) +
  scale_fill_manual(values = c(palette_final), na.value = "transparent", name = "adjuvanted") +
  new_scale_fill() +
  geom_tile(data = ~filter(.x, variable == "administration_route"), aes(fill = value)) +
  scale_fill_manual(values = c(palette_final), na.value = "transparent", name = "administration_route") +
  new_scale_fill() +
  geom_tile(data = ~filter(.x, variable == "main_sample_source"), aes(fill = value)) +
  scale_fill_manual(values = c(palette_final), na.value = "transparent", name = "main_sample_source") +
  new_scale_fill() +
  geom_tile(data = ~filter(.x, variable == "specific_sample_source"), aes(fill = value)) +
  scale_fill_manual(values = c(palette_final), na.value = "transparent", name = "specific_sample_source") +
  new_scale_fill() +
  geom_tile(data = ~filter(.x, variable == "sequencing"), aes(fill = value)) +
  scale_fill_manual(values = c(palette_final), na.value = "transparent", name = "sequencing") +
  theme_vaxgo() +
  geom_text(size = 2, color = "white", fontface = "bold") +
  theme(strip.text = element_blank(), panel.spacing = unit(0, "lines"),
        legend.position = "none", axis.text.y = element_text(size = 8),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 8))  

heatmap_plot

```


# 2. Demographics

## All

Import the combined human-mouse metadata, standardize timepoint labels and factor levels, and compute sample counts per pathogen and organism.

```{r}
# Import and standardize the combined human and mouse metadata --------
all_human_mouse_metadata <- readRDS(here("tables", "processed_data", "all_human_mouse_metadata.rds")) %>% 
  mutate(pathogen = as.factor(pathogen) %>% fct_relevel("Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         timepoint = if_else(timepoint == "Early hours", "<Day 1", timepoint),
         timepoint = if_else(timepoint == "Day 0", "0h", timepoint),
         timepoint = fct_relevel(timepoint, "0h", "2h", "4h", "6h", "12h", "<Day 1", "24h", "Day 1", "Day 3", "Day 7"),
         sex = str_to_title(sex))
all_human_mouse_metadata

# Count samples by pathogen, organism, and timepoint --------
demography_samples <- all_human_mouse_metadata %>% 
  group_by(pathogen, organism, timepoint, time_unit) %>% 
  summarise(n = n()) %>% 
  ungroup() %>% 
  distinct()
demography_samples
```

### By organism



```{r fig.height=6, fig.width=5}

# Build the multi-pathogen demographic dot plot --------
demography_samples_plot <- demography_samples %>%
  ggplot() +
  aes(x = timepoint, y = organism) +
  facet_wrap(~pathogen, ncol = 1) +
  geom_point(aes(size = n, fill = organism), pch = 21) +
  scale_size_binned(range = c(2, 5)) +
  scale_fill_manual(values = c("Human" = "#90A4AEFF", "Mouse" = "#4DBBD5FF")) +
  theme_vaxgo() +
  theme(legend.position = "right", axis.title.x = element_text(size = 10),
        legend.title = element_text(size = 10), axis.text.x = element_text(angle = 0),
        panel.grid.major.y = element_line(size = 1, linetype = 1, color = "black"),
        axis.line.x = element_line(size = 0.5, colour = "black", linetype = 1),
        axis.line.y = element_line(size = 0, colour = "black", linetype = 1),
        axis.ticks.x = element_line(size = 1, color = "black"),
        axis.ticks.y = element_line(size = 1, color = "black"),
        plot.margin = margin(0, 0, 0, 0, unit = "cm")) +
  labs(x = "Time", y = "", size = "Samples", fill = "Organism") +
  geom_text(aes(label = n), size = 3, hjust = 0.5, vjust = -1,
            data = filter(demography_samples, organism == "Human")) +
  geom_text(aes(label = n), size = 3, hjust = 0.5, vjust = -0.8,
            data = filter(demography_samples, organism == "Mouse")) +
  theme(legend.position = "none", axis.text.y = element_text(vjust = 0.5),
        axis.line.y = element_blank(), panel.spacing.y = unit(0, "cm"),
        panel.background = element_rect(fill = "transparent", color = NA),
        plot.background = element_rect(fill = "transparent", color = NA))

demography_samples_plot

# Save the all-pathogens demography plot --------
ggsave(demography_samples_plot, filename = here("Figures", "all_human_mouse_demography.png"),
         width = 5, height = 6)

```
