---
title: "2_Download_Standardize_Datasets"
output: html_document
---

# Required

Load the `here` package and source the project-wide library and utility script.

```{r}
# Load the `here` package and source the project-wide library and utility script.
library(here)
source(here("scripts_notebooks", "required.R"))
```



# Preprocessing

#===================

# 1. Fluad, Vaccine

## Mouse
Download dataset

Download the Fluad/MF59 mouse dataset (GSE120661) from GEO. Extract metadata from the ExpressionSet, save it as a CSV, then extract and save the raw expression matrix.

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
# Download the GEO ExpressionSet
filename = "GSE120661"
eset <- getGEO(filename = here::here("tables", "raw_data", "GSE120661_series_matrix.txt.gz"))

# Clean and save metadata
influenza_hepatitisb_mouse_metadata = eset %>% 
  pData() %>% 
# Reorder columns
  dplyr::select( 
    geo_sample = "geo_accession", 
    subject_id = "title",
    sex = "gender:ch1", 
    treatment = "treatment:ch1", 
    tissue = "tissue:ch1",
    timepoint = "timepoint:ch1",
    animal_replicate = "animal replicate:ch1",
    strain = "strain/background:ch1") %>% 
  mutate(treatment = fct_relevel(treatment, "saline"),
         timepoint = fct_relevel(timepoint, "0"),
         condition = paste(treatment, timepoint, tissue, sep = "_"))

#Save
saveRDS(influenza_hepatitisb_mouse_metadata, file = here("tables", "processed_data", "influenza_hepatitisb_mouse_metadata.rds"))

# Extract and save expression matrix
eset <- exprs(eset)

#Save
eset %>% 
  as.data.frame() %>% 
  saveRDS(here("tables", "processed_data", "influenza_hepatitisb_mouse_exprs.rds"))

```


Inspect the distribution of expression values for one representative sample (GSM3407136) to confirm the data are on the log2 scale.

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
# Load the expression matrix
influenza_hepatitisb_mouse_exprs = readRDS(here("tables", "processed_data", "influenza_hepatitisb_mouse_exprs.rds"))


# Pivot to long format
GSE120661_exprs_distribution = influenza_hepatitisb_mouse_exprs %>% 
  rownames_to_column("probes") %>% 
  pivot_longer(cols = -probes,
               names_to = "sample_id",
               values_to = "value") 

# Plot expression histogram for the first sample
GSE120661_exprs_distribution %>% 
  filter(sample_id == "GSM3407136") %>% 
  ggplot() +
  aes(x = value) +
  geom_histogram()
```


Import

```{r}
# Load metadata and expression
influenza_hepatitisb_mouse_metadata = readRDS(here("tables", "processed_data", "influenza_mouse_metadata.rds"))
influenza_hepatitisb_mouse_exprs = readRDS(here("tables", "processed_data", "influenza_hepatitisb_mouse_exprs.rds"))

# Check probe count
influenza_hepatitisb_mouse_exprs %>% rownames() %>% length()

# Filter to Fluad PBMC samples, recode timepoint (hours to days)
mouse_fluad_metadata = influenza_hepatitisb_mouse_metadata %>% 
  filter(treatment == "Fluad",
         tissue == "PBMC") %>% 
  mutate(timepoint = as.character(timepoint) %>% as.numeric(),
         timepoint = (timepoint / 24) %>% round(1),
         participant_id = paste0("TriFLU-MF59_BL", animal_replicate)) %>% 
  filter(timepoint %in% c(0, 1, 3, 7))

#Save
mouse_fluad_metadata %>% 
  saveRDS(here("tables", "processed_data", "influenza_mouse_metadata.rds"))

# Subset expression matrix to selected samples
influenza_hepatitisb_mouse_exprs = influenza_hepatitisb_mouse_exprs %>% 
  as.data.frame() %>% 
  select(c(mouse_fluad_metadata$geo_sample)) %>% 
  as.matrix()

#Save
influenza_hepatitisb_mouse_exprs %>% 
  as.data.frame() %>% 
  saveRDS(file = here("tables", "processed_data", "influenza_mouse_exprs.rds"))
```

Annotate genes

```{r}
# Set up Ensembl BioMart for mouse
ensembl_mart <- biomaRt::useEnsembl(
  biomart = "genes",
  dataset = "mmusculus_gene_ensembl",   # humano: "hsapiens_gene_ensembl"
  mirror  = "www"                        # se www cair, tenta useast/asia sozinho
)

# Get probe IDs from the expression matrix
gene_id <- influenza_hepatitisb_mouse_exprs %>% rownames()

# List available attributes
listAttributes(ensembl_mart)

# Retrieve annotations for all probes
annot <- getBM(
  attributes = c("agilent_sureprint_g3_ge_8x60k", 
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "mgi_symbol",
                 "description"),
  filters = "agilent_sureprint_g3_ge_8x60k",
  values = gene_id,
  mart = ensembl_mart)

# Load mouse-to-human orthology lookup table
library(FIT.mouse2man)
HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez %>% 
  clean_names() %>% 
  rename(mouse_entrez = mouse_ortholog)

# Summarise expression per probe, collapse to one row per gene
mouse_fluad_log2summary = influenza_hepatitisb_mouse_exprs %>% #20,845 probes
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
# Compute per-probe summary statistics 
  summarise(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
  # Join BioMart annotation to expression data
inner_join(annot %>% 
              rename(probe = agilent_sureprint_g3_ge_8x60k), 
            by = "probe") %>%  #24.009 rows
  group_by(ensembl_gene_id) %>% 
  # Keep probe with highest median expression per gene
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_gene_id) %>% 
  distinct() %>% 
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup() %>% 
  rename(mouse_entrez = entrezgene_id) %>% 
  # Join human orthologs
  inner_join(HS_MM_Symbol_Entrez, by = c("mouse_entrez", "mgi_symbol")) %>% 
  group_by(hgnc_symbol) %>% 
  # Keep probe with highest median expression per gene
  slice_max(median, n = 1)  %>% #14.396
  ungroup()

# Remove HGNC symbols that map to multiple MGI symbols
hgnc_symbol_dupes = mouse_fluad_log2summary %>% 
  get_dupes(hgnc_symbol)

mouse_fluad_log2summary = mouse_fluad_log2summary %>% 
# Remove genes with ambiguous human symbol mappings
  filter(!hgnc_symbol %in% c(hgnc_symbol_dupes$hgnc_symbol %>% unique())) 

# Build filtered expression matrix with human gene symbols
mouse_fluad_log2_filtered_probes = influenza_hepatitisb_mouse_exprs %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
# Join expression data with annotation by probe ID
  inner_join(mouse_fluad_log2summary, by = "probe") %>%   #214,396
  select(human_symbol = hgnc_symbol,
         human_entrez,
         mouse_entrez,
         c(mouse_fluad_metadata$geo_sample))

mouse_fluad_log2_filtered_probes %>% dim()

mouse_fluad_log2_hgnc_symbol = mouse_fluad_log2_filtered_probes 

# Identify duplicated gene symbols
mouse_fluad_log2_hgnc_symbol %>% get_dupes(human_symbol)

#Save 
mouse_fluad_log2_hgnc_symbol %>% 
  saveRDS(here("tables", "processed_data", "influenza_mouse_exprs_hgnc_symbol.rds"))
```

Create expression set

```{r}
# Load annotation and metadata
mouse_fluad_log2_hgnc_symbol = readRDS(here("tables", "processed_data", "influenza_mouse_exprs_hgnc_symbol.rds"))
mouse_fluad_metadata = readRDS(here("tables", "processed_data", "influenza_mouse_metadata.rds")) 

# Pivot to long, add metadata, compute log2 fold-change vs day 0
mouse_fluad_log2fc_filtered_sample_long = mouse_fluad_log2_hgnc_symbol %>%
  pivot_longer(cols = -human_symbol,
               names_to = "geo_sample",
               values_to = "log2") %>% 
# Attach sample metadata 
  inner_join(mouse_fluad_metadata %>% 
               select(geo_sample, participant_id, timepoint),
             by = "geo_sample") %>% 
  select(participant_id, human_symbol,  timepoint, value = log2) %>%
# Remove duplicate rows
    distinct() %>% 
  # Reshape to wide format with one column per timepoint
  pivot_wider(names_from = timepoint, values_from = value, names_prefix = "day_") %>%
# Reshape columns from wide to long using a common prefix pattern
  pivot_longer(cols = starts_with("day_"),
               names_to = "timepoint",
               names_prefix = "day_",
               values_to = "value") %>%
  group_by(participant_id, human_symbol) %>% 
# Extract baseline (day 0) expression value for each participant × gene pair
  mutate(day0 = value[timepoint == "0"]) %>%
  ungroup() %>%
  # Compute log2 fold-change relative to baseline
mutate(log2fc = value - day0) %>%
  filter(timepoint != "0")

#Save
mouse_fluad_log2fc_filtered_sample_long %>% 
  saveRDS(here("tables", "Differential gene expression", "influenza_mouse_log2fc_sample_clean_long.rds"))

# Pivot to wide format (participants as columns)
mouse_fluad_log2fc_filtered_sample_wide = mouse_fluad_log2fc_filtered_sample_long %>% 
  select(-c(value:day0)) %>% 
  pivot_wider(names_from = "participant_id",
              values_from = "log2fc")

```



## Human

Download dataset
```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
# Download GSE124689 from GEO
GSE124689 <- getGEO("GSE124689", GSEMatrix = TRUE) 

# Extract ExpressionSet
eset <- GSE124689[[1]]

# Expression matrix
influenza_human_exprs <- eset %>% exprs()

# Metadata
influenza_human_metadata = eset %>% pData()

#Save
influenza_human_exprs %>% 
  as.data.frame() %>% 
  saveRDS(here("tables", "processed_data", "influenza_human_exprs.rds"))
```
Import data

Load the raw expression matrix and the curated metadata, then filter to Fluad-treated PBMC samples only.

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
# Load expression matrix
influenza_human_exprs = readRDS(here("tables", "processed_data", "influenza_human_exprs.rds"))

# Load curated pheno data
flu_expFC <- readRDS(here("tables", "FIT", "Flu_Human_expFC.rds"))
flu_metadata = flu_expFC$cur_phenoData
flu_metadata %>% filter(exposure_material_reported == "FLUADD")

# Filter to Fluad PBMC samples
human_samples_pbmc_fluad = flu_metadata %>% 
  filter(exposure_material_reported == "FLUADD")
```

Annotate features

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
# Load platform annotation
probes_annotation <- read.delim(here("tables", "raw_data", "GPL21272-58809.txt"), skip = 19, header = TRUE, sep = "\t", quote = "") %>% 
  select(gene_id = ID, entrez_id = ENTREZ_GENE_ID, ensembl_id = ENSEMBL_ID, gene_symbol = GENE_SYMBOL) %>% 
  mutate(gene_id = as.character(gene_id))

# Join expression with annotation
influenza_human_counts <- influenza_human_exprs %>%
  as.data.frame() %>% 
  rownames_to_column("gene_id") %>% 
  inner_join(probes_annotation, by = "gene_id") %>% 
  select(gene_id, entrez_id, ensembl_id, gene_symbol, c(human_samples_pbmc_fluad$biosample_accession)) %>% 
  mutate(entrez_id = as.character(entrez_id)) 

# Collapse multiple probes per Entrez ID (keep highest median)
influenza_human_counts_genes_selected = influenza_human_counts %>% 
  pivot_longer(cols = -c(gene_id, entrez_id, ensembl_id, gene_symbol),
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(entrez_id, gene_id) %>% 
  drop_na(entrez_id) %>% 
  summarise(median = median(expression),
            median = round(median, 10)) %>% 
  slice_max(median, n = 1) %>% 
  ungroup()

# Merge back with full data
influenza_human_counts_raw = influenza_human_counts %>% 
  inner_join(influenza_human_counts_genes_selected %>% 
               select(gene_id, entrez_id),
             by = join_by("gene_id", "entrez_id"))

# Remove duplicate Entrez IDs by random sampling
dupes_entrez = influenza_human_counts_genes_selected %>% 
  get_dupes(entrez_id) %>% 
  group_by(entrez_id) %>% 
  slice_sample(n = 1)

# Build Entrez-ID-based expression matrix
influenza_human_not_normalized = influenza_human_counts_raw %>%
  select(-gene_symbol) %>% 
  filter(!gene_id %in% c(dupes_entrez$gene_id)) %>% 
  distinct() %>% 
  select(c(entrez_id, human_samples_pbmc_fluad$uid)) %>% 
  column_to_rownames("entrez_id") 

#Save
human_fluadd_not_normalized %>% saveRDS(file = here("tables", "processed_data", "influenza_GSE124689_exprs.rds"))

# Repeat collapse by gene symbol
influenza_human_counts_genes_selected_symbol = influenza_human_counts %>% 
  pivot_longer(cols = -c(gene_id, entrez_id, ensembl_id, gene_symbol),
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(gene_symbol, gene_id) %>% 
  filter(gene_symbol != "") %>% 
  drop_na(gene_symbol) %>% 
  summarise(median = median(expression),
            median = round(median, 10)) %>% 
  slice_max(median, n = 1) %>% 
  ungroup()

influenza_human_counts_raw_symbol = influenza_human_counts %>% 
  inner_join(influenza_human_counts_genes_selected_symbol %>% 
               select(gene_id, gene_symbol),
             by = join_by("gene_id", "gene_symbol"))

# Remove duplicate gene symbols
dupes_symbols = influenza_human_counts_genes_selected_symbol %>% 
  get_dupes(gene_symbol) %>% 
  group_by(gene_symbol) %>% 
  slice_sample(n = 1)

# Build symbol-based expression matrix
human_fluadd_not_normalized_symbol = influenza_human_counts_raw_symbol %>%
  filter(!gene_symbol %in% c(dupes_symbols$gene_symbol)) %>% 
  distinct() %>% 
  select(c(gene_symbol, human_samples_pbmc_fluad$uid)) %>% 
  column_to_rownames("gene_symbol")

# Save symbol-to-Entrez lookup
human_fluadd_symbol_entrez = influenza_human_counts_raw_symbol %>%
  filter(!gene_symbol %in% c(dupes_symbols$gene_symbol)) %>% 
  distinct() %>% 
  select(symbol = gene_symbol, entrez_id)

#Save
human_fluadd_not_normalized_symbol %>% 
  saveRDS(file = here("tables", "processed_data", "influenza_GSE124689_exprs_not_normalized.rds"))

# Inspect results
human_fluadd_not_normalized_symbol %>% 
  rownames_to_column("human_symbol") %>% 
  pivot_longer(columns_to = "geo_sample",
               values_to = "log2fc")

human_fluadd_not_normalized_symbol
```


Clean metadata
```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
# Clean and restructure metadata
flu_metadata_clean = flu_metadata %>% 
  arrange(uid) %>% 
  separate(participant_id, into = c("del", "participant_id")) %>% 
  select(-del) %>% 
  select(sample_id = uid, participant_id, timepoint = study_time_collected, age = age_imputed, sex = gender, race, vaccine = exposure_material_reported, sample_type = cell_type, array = featureSetName2, type = vaccine_type, adjuvant) %>% 
  mutate(age = as.numeric(age),
         sex = as.factor(sex), 
         race = as.factor(race),
         condition = paste(vaccine, "Day", timepoint)) %>%
  filter(sample_id %in% c(influenza_human_counts %>% colnames())) %>% 
  arrange(sample_id) %>% 
# Select all columns: ensures clean column order with key identifier first
  select(condition, everything()) 

#Save
saveRDS(flu_metadata_clean, file = here("tables", "processed_data", "influenza_human_metadata_clean.rds"))
```


Unite and normalize with human

```{r}
# Filter out duplicated MGI symbols
filter_out_mgi = influenza_hepatitisb_mouse_exprs_annotations %>% 
  select(mgi_symbol) %>% 
  get_dupes()

# Clean mouse symbol table
influenza_hepatitisb_mouse_exprs_symbol = influenza_hepatitisb_mouse_exprs_annotations %>% 
  select(-probe, -entrezgene_id) %>% #21,856 rows
  distinct() %>%  #21,666 rows
  filter(mgi_symbol != c(filter_out_mgi$mgi_symbol)) %>% #21,557 rows
  mutate(mgi_symbol = str_to_upper(mgi_symbol)) %>% 
  rename(symbol = mgi_symbol)

# Join human and mouse matrices by symbol, quantile normalize
human_mouse_fluad_matrix = GSE124689_expression_not_normalized_log2_raw_symbol %>% 
  rownames_to_column("symbol") %>% 
  inner_join(influenza_hepatitisb_mouse_exprs_symbol, 
             by = "symbol")  %>% 
  column_to_rownames("symbol") %>% 
  select(human_mouse_fluad_metadata %>% 
           pull(sample_id)) %>% 
  as.matrix() %>% 
  normalizeBetweenArrays(method = "quantile")

human_mouse_fluad_df = human_mouse_fluad_matrix %>% 
  as.data.frame()

# Create joint ExpressionSet
eset_human_mouse <- ExpressionSet(
  assayData = human_mouse_fluad_matrix,
  phenoData = AnnotatedDataFrame(human_mouse_fluad_metadata %>% 
                                   filter(sample_id %in% c(human_mouse_fluad_matrix %>% colnames())) %>% 
                                   column_to_rownames("sample_id"))
)

#Save
eset_human_mouse %>% saveRDS(file = here("tables", "Differential gene expression", "influenza_human_mouse_eset.rds"))

```









#==================
# 2. Engerix B, Vaccine

## Mouse

Prepare data

Filter the GSE120661 dataset to Engerix B-treated PBMC mouse samples, recode timepoints from hours to days, save metadata and expression matrix.

```{r}
# Load expression and metadata
influenza_hepatitisb_mouse_exprs = readRDS(here("tables", "processed_data", "influenza_hepatitisb_mouse_exprs.rds")) %>% 
  as.data.frame()
influenza_hepatitisb_mouse_metadata = readRDS(file = here("tables", "processed_data", "influenza_hepatitisb_mouse_metadata.rds"))

# Filter to Engerix B PBMC samples, recode timepoint (hours to days)
mouse_engerixb_metadata = influenza_hepatitisb_mouse_metadata %>% 
  filter(treatment == "Engerix B",
         tissue == "PBMC") %>% 
  mutate(timepoint = (timepoint / 24) %>% round(1),
         participant_id = paste0("ENG_BL_", animal_replicate)) %>% 
  filter(timepoint %in% c(0, 1, 3, 7))

#Save
mouse_engerixb_metadata %>% 
  saveRDS(file = here("tables", "processed_data", "hepatitisb_mouse_metadata.rds"))

# Subset expression matrix
mouse_engerixb_log2 = influenza_hepatitisb_mouse_exprs %>% 
  select(mouse_engerixb_metadata %>% pull(geo_sample))

```
Annotate genes
Map probe IDs to gene symbols using BioMart and orthology lookup.

Map Agilent probe IDs to human gene symbols using BioMart and FIT.mouse2man orthology lookup, then save the annotated expression matrix.

```{r}
# Set up Ensembl BioMart for mouse
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("mmusculus_gene_ensembl", mart = ensembl_mart)

# Get probe IDs
gene_id <- mouse_engerixb_log2 %>% rownames()

# List available attributes
listAttributes(ensembl_mart)

# Retrieve annotations
annot <- getBM(
  attributes = c("agilent_sureprint_g3_ge_8x60k", #Agilent microarray data
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "mgi_symbol",
                 "description"),
  filters = "agilent_sureprint_g3_ge_8x60k",
  values = gene_id,
  mart = ensembl_mart)

#Save
annot %>% 
  saveRDS(here("tables", "Differential gene expression", "probe_annot_hepatitisb_engerixb_mouse.rds"))

# Load mouse-to-human orthology lookup table
library(FIT.mouse2man)
HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez %>% 
  clean_names() %>% 
  rename(mouse_entrez = mouse_ortholog)

# Summarise expression per probe, collapse to one row per gene
mouse_engerixb_log2summary = mouse_engerixb_log2 %>% #20,845 probes
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarise(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
  inner_join(annot %>% 
              rename(probe = agilent_sureprint_g3_ge_8x60k), 
            by = "probe") %>%  #24.009 rows
  group_by(ensembl_gene_id) %>% 
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_gene_id) %>% 
  distinct() %>% 
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup() %>% 
  rename(mouse_entrez = entrezgene_id) %>% 
  inner_join(HS_MM_Symbol_Entrez, by = c("mouse_entrez", "mgi_symbol")) %>% 
  group_by(hgnc_symbol) %>% 
  slice_max(median, n = 1)  %>% #14.396
  ungroup()

# Remove HGNC symbols that map to multiple MGI symbols
hgnc_symbol_dupes = mouse_engerixb_log2summary %>% 
  get_dupes(hgnc_symbol)

mouse_engerixb_log2summary = mouse_engerixb_log2summary %>% 
  filter(!hgnc_symbol %in% c(hgnc_symbol_dupes$hgnc_symbol %>% unique())) 

# Build filtered expression matrix with human gene symbols
mouse_engerixb_log2_filtered_probes = mouse_engerixb_log2 %>% 
  rownames_to_column("probe") %>% 
  inner_join(mouse_engerixb_log2summary, by = "probe") %>%   #214,396
  select(probe, mouse_entrez, human_symbol = hgnc_symbol, c(mouse_engerixb_metadata %>% pull(geo_sample)))

mouse_engerixb_log2_filtered %>% dim()
mouse_engerixb_log2_hgnc_symbol = mouse_engerixb_log2_filtered_probes %>% 
  select(probe_id = probe, human_symbol, c(mouse_engerixb_metadata$geo_sample)) %>% 
  column_to_rownames("human_symbol")

#Save 
mouse_engerixb_log2_hgnc_symbol %>% 
  saveRDS(here("tables", "processed_data", "hepatitisb_mouse_exprs.rds"))
```


Create expression set

```{r}
# Load data
mouse_engerixb_log2_hgnc_symbol = readRDS(here("tables", "processed_data", "hepatitisb_mouse_exprs.rds"))
mouse_engerixb_metadata = readRDS(file = here("tables", "processed_data", "hepatitisb_mouse_metadata.rds"))

# Create ExpressionSet
eset_mouse_engerixb <- ExpressionSet(
  assayData = mouse_engerixb_log2_hgnc_symbol %>% 
    as.matrix(),
  phenoData = AnnotatedDataFrame(mouse_engerixb_metadata %>% 
                                   column_to_rownames("geo_sample")))

#Save
eset_mouse_engerixb %>% saveRDS(file = here("tables", "Differential gene expression", "hepatitisb_mouse_eset.rds"))
```

Compute log2 fold-change relative to day 0 for each gene and participant, save in long and wide formats.

```{r}
# Compute log2FC
mouse_engerixb_log2fc_filtered_sample_long = mouse_engerixb_log2_hgnc_symbol %>%
  rownames_to_column("human_symbol") %>% 
  pivot_longer(cols = -c(human_symbol, probe_id),
               names_to = "geo_sample",
               values_to = "log2") %>% 
  inner_join(mouse_engerixb_metadata %>% 
               select(geo_sample, participant_id, timepoint),
             by = "geo_sample") %>% 
  select(human_symbol,probe_id, participant_id, timepoint, log2) %>% 
  distinct() %>% 
  select(participant_id,probe_id, human_symbol, timepoint, value = log2) %>%
  pivot_wider(names_from = timepoint, values_from = value, names_prefix = "day_") %>%
  pivot_longer(cols = starts_with("day_"),
               names_to = "timepoint",
               names_prefix = "day_",
               values_to = "value") %>%
  group_by(participant_id, human_symbol) %>% 
  mutate(day0 = value[timepoint == "0"]) %>%
  ungroup() %>%
mutate(log2fc = value - day0) %>%
  filter(timepoint != "0")

#Save
mouse_engerixb_log2fc_filtered_sample_long %>% 
  saveRDS(here("tables", "Differential gene expression", "hepatitisb_mouse_log2fc_sample_clean_long.rds"))

# Pivot to wide
mouse_engerixb_log2fc_filtered_sample_wide = mouse_engerixb_log2fc_filtered_sample_long %>% 
  select(-c(value:day0)) %>% 
  pivot_wider(names_from = "participant_id",
              values_from = "log2fc")

```


## Human

Download directly from GEO. The data is too large.
Prepare data

```{r}
#EngerixB: GSE124533 -----
filename = "GSE124533"
GSE124533 <- getGEO(filename = here("tables", "raw_data", "GSE124533_series_matrix.txt.gz"), GSEMatrix = F)

#Metadata ------
hepatitisb_human_metadata = GSE124533@phenoData %>% 
  pData() %>% 
  dplyr::select( 
    geo_sample = "geo_accession", 
    participant_id = "participant:ch1",
    sex = "gender:ch1",
    treatment = "treatment:ch1",
    ethnicity = "ethnicity:ch1",
    race = "race:ch1",
    age = "age:ch1",
    timepoint = "day:ch1") %>% 
  mutate(condition = treatment,
         organism = "Human",
         time_unit = "Day")
#Save
hepatitisb_human_metadata %>% saveRDS(file = here("tables", "processed_data", "hepatitisb_human_metadata.rds"))

# Load data
hepatitisb_human_metadata = readRDS(file = here("tables", "processed_data", "hepatitisb_human_metadata.rds"))


#Filter metadata
hepatitisb_human_metadata_filtered = hepatitisb_human_metadata %>%
  filter(treatment %in% c("ENGERIXB1"),
         timepoint %in% c("0", "1" , "3", "7"))

#Save
hepatitisb_human_metadata_filtered %>% 
  saveRDS(file = here("tables", "processed_data", "hepatitisb_human_metadata.rds"))


#Counts matrix
hepatitisb_human_exprs = GSE124533@assayData[["exprs"]] %>% as.data.frame()

hepatitisb_human_exprs_engerixb = GSE124533_exprs %>% 
  select(c(hepatitisb_human_metadata_filtered$geo_sample))
#Save
hepatitisb_human_exprs_engerixb %>% saveRDS(file = here("tables", "processed_data", "hepatitisb_human_exprs.rds"))
```



Import
Load the required data objects from disk.
```{r}
# Load data
human_exprs_engerixb = readRDS(file = here("tables", "processed_data", "hepatitisb_human_exprs.rds"))
human_engerixb_metadata = readRDS(file = here("tables", "processed_data", "hepatitisb_human_metadata.rds"))

########### Filter and annotate probes -----
probes_annotation <- read.delim(here("tables", "raw_data", "GPL21272-58809.txt"), skip = 19, header = TRUE, sep = "\t", quote = "") %>% 
  select(gene_id = ID, entrez_id = ENTREZ_GENE_ID, ensembl_id = ENSEMBL_ID, gene_symbol = GENE_SYMBOL) %>% 
  mutate(gene_id = as.character(gene_id))


#Get annotations for each probe 
human_engerix_log2summary = human_exprs_engerixb %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarise(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
  inner_join(probes_annotation %>% 
              rename(probe = gene_id,
                     entrezgene_id = entrez_id,
                     human_symbol = gene_symbol), 
            by = "probe") %>% 
  group_by(ensembl_id) %>% 
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_id) %>% 
  distinct() %>% 
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup() 

#Filter out genes with multiple hgnc_symbols
hgnc_symbol_dupes = human_engerix_log2summary %>% 
  get_dupes(human_symbol)

human_engerix_log2summary = human_engerix_log2summary %>% 
  filter(!human_symbol %in% c(hgnc_symbol_dupes$human_symbol %>% unique())) 

# Build filtered expression matrix with human symbols
human_engerix_log2_filtered_probes = human_exprs_engerixb %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  inner_join(human_engerix_log2summary, by = "probe") %>%   #214,396
  select(human_symbol,
         c(human_engerixb_metadata$geo_sample))
#14395
human_engerix_log2_filtered_probes %>% dim()

human_engerix_log2_hgnc_symbol = human_engerix_log2_filtered_probes 

human_engerix_log2_hgnc_symbol %>% 
  get_dupes(human_symbol)

#Save 
human_engerix_log2_hgnc_symbol %>% 
  saveRDS(here("tables", "processed_data", "hepatitisb_human_exprs_hgnc_symbol.rds"))
```




Load the Engerix B human dataset from saved RDS files.

```{r}

# Load data
human_engerix_log2_hgnc_symbol = readRDS(here("tables", "processed_data", "hepatitisb_human_exprs_hgnc_symbol.rds"))
human_metadata_filtered = readRDS(file = here("tables", "processed_data", "hepatitisb_human_metadata.rds"))

#Get value from time 0 and calculate Log2FC
human_engerixb_log2fc_filtered_sample_long = human_engerix_log2_hgnc_symbol %>%
  ungroup() %>% 
# Reshape expression data from wide (columns = samples) to long (rows = probe × sample) format for tidyverse workflows
  pivot_longer(cols = -(human_symbol),
               names_to = "geo_sample",
               values_to = "log2") %>% 
# Attach sample metadata (participant ID, timepoint, condition) to expression values by GEO sample ID
  inner_join(human_metadata_filtered %>% 
               select(geo_sample, participant_id, timepoint),
             by = "geo_sample") %>% 
  select(human_symbol, participant_id, timepoint, log2) %>% 
# Remove duplicate rows: eliminates redundant probe-to-gene mappings (one probe may match multiple transcript clusters)
  distinct() %>% 
  select(participant_id, human_symbol, timepoint, value = log2) 

# #Get over duplicated symbols
# overduplicated = human_engerixb_log2fc_filtered_sample_long %>% 
#   select(human_symbol) %>% 
#   group_by(human_symbol) %>% 
#   dplyr::count() %>% 
#   arrange(-n) 

#Calculate log2fc
human_engerixb_log2fc_filtered_sample_long = human_engerixb_log2fc_filtered_sample_long %>%
  # filter(!human_symbol %in% c(overduplicated %>% 
  #                               ungroup() %>% 
  #                               filter(n > median(n)) %>% 
  #                               pull(human_symbol))) %>% 
  # Reshape to wide format with one column per timepoint
  pivot_wider(names_from = timepoint, values_from = value, names_prefix = "day_") %>% 
# Reshape columns from wide to long using a common prefix pattern: separates metric type from timepoint via names_pattern
  pivot_longer(cols = starts_with("day_"),
               names_to = "timepoint",
               names_prefix = "day_",
               values_to = "value") %>%
  group_by(participant_id, human_symbol) %>% 
# Extract baseline (day 0) expression value for each participant × gene pair: reference for fold-change calculation
  mutate(day0 = value[timepoint == "0"]) %>%
  ungroup() %>%
  # Compute log2 fold-change relative to baseline
mutate(log2fc = value - day0) %>%
# Remove baseline rows: only post-vaccination/infection timepoints are relevant for differential expression analysis
  filter(timepoint != "0")


#Annotate sample and pivot longer
human_engerixb_log2fc_filtered_sample_long_annotated = human_engerixb_log2fc_filtered_sample_long %>% 
# Attach sample metadata (participant ID, timepoint, condition) to expression values by GEO sample ID
  inner_join(human_engerixb_metadata %>% 
               select(sample_id = geo_sample,
                      participant_id, timepoint), by = join_by("participant_id", "timepoint"))

#Save
human_engerixb_log2fc_filtered_sample_long_annotated %>% 
# Save R object to disk in native RDS format: preserves data types and structure for fast loading
  saveRDS(file = here("tables", "Differential gene expression", "hepatitisb_human_log2fc_sample_clean_long.rds"))

# Play notification sound: alerts that a long-running chunk has finished execution
beep()

#Pivot wider
human_engerixb_log2fc_filtered_sample_wide = human_engerixb_log2fc_filtered_sample_long %>% 
  select(-c(value:day0)) %>% 
# Widen data: reshape so that each timepoint becomes a separate column (required format for t-test input)
  pivot_wider(names_from = "participant_id",
              values_from = "log2fc")
  
```


Create expression set
Build an ExpressionSet with the expression matrix and phenotype data.
Build an ExpressionSet with the expression matrix and phenotype data.
```{r}
# Load data
human_engerix_log2_hgnc_symbol = readRDS(here("tables", "processed_data", "hepatitisb_human_exprs_hgnc_symbol.rds"))
human_metadata_filtered = readRDS(file = here("tables", "processed_data", "hepatitisb_human_metadata.rds"))

human_metadata_filtered = human_metadata_filtered %>% 
  mutate(age = str_remove(age, "y") %>% as.numeric() %>% scale(),
         timepoint = as.factor(timepoint),
         sex = as.factor(sex),
         race = as.factor(race))

#Create Expression Set
eset_human <- ExpressionSet(
  assayData = human_engerix_log2_hgnc_symbol %>% 
# Move identifier column to row names: required format for ExpressionSet and Limma (matrix must have rownames)
    column_to_rownames("human_symbol") %>% 
    as.matrix(),
# Convert metadata to AnnotatedDataFrame: formal phenotype data structure required by ExpressionSet
  phenoData = AnnotatedDataFrame(human_metadata_filtered))

#Save
eset_human %>% saveRDS(file = here("tables", "Differential gene expression", "hepatitisb_human_eset.rds"))
```








#==================
# 3. S. aureus, Infection
## Mouse

Retrieve Exprs and Metadata from GEO
```{r}
#S aureus: GSE19668 -----
filename = "saureus_GSE19668"
# Download GEO ExpressionSet using GEOquery package: retrieves both expression data and phenotype metadata
GSE19668 <- getGEO("GSE19668", GSEMatrix = TRUE) 
# Unlist the GEO download result: extract the ExpressionSet object from the list wrapper
eset <- GSE19668[[1]]

#Metadata ------
GSE19668_metadata = eset %>% 
  pData() %>% 
# Reorder columns: place comparison/condition column first for readability
  dplyr::select( 
    geo_sample = "geo_accession", 
    subject_id = "title",
    # sex = "gender:ch1", 
    treatment = "infection:ch1", 
    # tissue = "tissue:ch1",
    timepoint = "time point:ch1",
    # animal_replicate = "animal replicate:ch1",
    strain = "strain:ch1") %>% 
# Relevel treatment factor: set saline/control as the reference level for proper contrast encoding
  mutate(treatment = fct_relevel(treatment, "none"),
         timepoint = fct_relevel(timepoint, "0"),
         condition = paste(treatment, strain, timepoint, sep = "_"))

#Save
GSE19668_metadata %>% saveRDS(file = here("tables", paste0(filename, "_metadata.rds")))


#Get counts matrix ------
GSE_exprs <- eset %>% exprs()

#Save as RDS
GSE_exprs %>% as.data.frame() %>% saveRDS(here("tables", paste0(filename, "_exprs.rds")))
```

Clean metadata
Clean and restructure the metadata table.
Clean and restructure the metadata table.
```{r}
# Load data
GSE19668_metadata = readRDS(file = here("tables", "processed_data", "saureus_GSE19668_metadata.rds"))

GSE19668_metadata %>% 
  select(strain)

#Filter metadata
GSE19668_metadata_filtered = GSE19668_metadata %>%
  filter(strain == "C57BL/6J") %>% 
  separate(subject_id, into = c("tissue", "strain1", "timepoint1", "participant_id"), sep = " ") %>% 
  select(!c(tissue:timepoint1)) %>% 
  arrange(participant_id, timepoint) %>% 
  separate(timepoint, into = c("timepoint_number", "time_unit")) %>% 
  mutate(time_unit = "hours",
         timepoint = timepoint_number %>% as.numeric(),
         organism = "Mouse") %>% 
  arrange(timepoint) %>% 
  select(-timepoint_number) %>% 
# Select all columns: ensures clean column order with key identifier first
  select(geo_sample, participant_id, condition, timepoint, time_unit, everything())

#Save
GSE19668_metadata_filtered %>% saveRDS(file = here("tables", "processed_data", "saureus_mouse_metadata_filtered.rds"))

```


Annotate dataset
Load the S. aureus mouse dataset from saved RDS files.

```{r}
# Load data
saureus_mouse_exprs = readRDS(here("tables", "processed_data", "saureus_GSE19668_exprs.rds"))
saureus_mouse_metadata = readRDS(file = here("tables", "processed_data", "saureus_mouse_metadata_filtered.rds"))

#Inputs
filename = "saureus_mouse"
mouse_exprs = saureus_mouse_exprs 
mouse_metadata = saureus_mouse_metadata
```


Annotate gene symbols and map probes to human orthologs via BioMart.

```{r}
#Annotate genes ------
#Mouse -----
#Annotate data and calculate log2fc
mouse_exprs_metadata  = mouse_exprs %>% 
  select(c(mouse_metadata$geo_sample)) %>%
  rownames_to_column("probe_id") %>% 
  pivot_longer(cols = c(mouse_metadata$geo_sample),
               names_to = "geo_sample",
               values_to = "value_raw") %>% 
# Attach sample metadata (participant ID, timepoint, condition) to expression values by GEO sample ID
  inner_join(mouse_metadata, 
             by = "geo_sample") 


#Get value from day 0 and calculate Log2FC
mouse_exprs_log2fc_sample = mouse_exprs_metadata %>% 
  filter(timepoint == 0) %>% 
  select(probe_id, participant_id, day0_value_raw = value_raw) %>% 
  #Filter batch
  inner_join(mouse_exprs_metadata %>% 
# Remove baseline rows: only post-vaccination/infection timepoints are relevant for differential expression analysis
               filter(timepoint != "0"),
             by = join_by("probe_id", "participant_id")) %>% 
  # Compute log2 fold-change relative to baseline
mutate(log2fc = value_raw - day0_value_raw) %>% 
# Remove duplicate rows: eliminates redundant probe-to-gene mappings (one probe may match multiple transcript clusters)
  distinct() 

#Check duplicates
mouse_exprs_log2fc_sample %>% 
  filter(participant_id == "rep1",
         probe_id == "1415670_at")


mouse_exprs_log2fc_sample %>% 
# Save R object to disk in native RDS format: preserves data types and structure for fast loading
  saveRDS(here("tables", "Differential gene expression", "saureus_mouse_log2fc_sample_clean_long.rds"))
```


Annotate probes with gene symbols by querying the Ensembl BioMart database.

```{r}
########### Filter and annotate probes -----
# Configure Ensembl BioMart
ensembl_mart <- useMart("ensembl")
# Select the mouse genome (mmusculus_gene_ensembl) as the annotation dataset in BioMart
ensembl_mart <- useDataset("mmusculus_gene_ensembl", mart = ensembl_mart)

# Get probe IDs from expression matrix
gene_id <- mouse_exprs %>% 
  rownames_to_column("probe_id") %>% 
  pull(probe_id) %>% unique()

# listAttributes(ensembl_mart)
# Annotate data
annot <- getBM(
  attributes = c("affy_mouse430_2",
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "mgi_symbol",
                 "description"),
  filters = "affy_mouse430_2", #Probes
  values = gene_id,
  mart = ensembl_mart)

annot %>% 
# Save R object to disk in native RDS format: preserves data types and structure for fast loading
  saveRDS(here("tables", "Differential gene expression", "probe_annot_saureus_infection_mouse.rds"))

# Load mouse-to-human orthology lookup
# (Install orthology package if needed)
# library(devtools)
# devtools::install_github('shenorrLabTRDF/FIT.mouse2man')
library(FIT.mouse2man)
HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez %>% 
# Standardise column names to snake_case using janitor::clean_names for consistent tidyverse syntax
  clean_names() %>% 
  rename(mouse_entrez = mouse_ortholog)

#Get annotations for each probe 
### Filter probes with highest median expression
mouse_log2summary = mouse_exprs %>% #20,845 probes
  rownames_to_column("probe_id") %>% 
  rename(probe = probe_id) %>% 
# Reshape expression data from wide (columns = samples) to long (rows = probe × sample) format for tidyverse workflows
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
# Group by probe ID to collapse multiple probe-to-gene mappings later
  group_by(probe) %>% 
# Compute per-probe summary statistics (mean, median) across all samples to select the most representative probe
  summarise(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
# Join BioMart gene annotation to expression data by probe ID: adds gene symbol, Entrez ID, and description
  inner_join(annot %>% 
              rename(probe = affy_mouse430_2), 
            by = "probe") %>%  #24.009 rows
  # Collapse to one row per Ensembl ID
  group_by(ensembl_gene_id) %>% 
  # Keep probe with highest median expression per gene
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_gene_id) %>% 
# Remove duplicate rows: eliminates redundant probe-to-gene mappings (one probe may match multiple transcript clusters)
  distinct() %>% 
  
  # Collapse to one row per Entrez ID
  group_by(entrezgene_id) %>% 
  # Keep probe with highest median expression per gene
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup() %>% 
  rename(mouse_entrez = entrezgene_id) %>% 
  
  # Join human orthologs, collapse to one row per HGNC symbol
  inner_join(HS_MM_Symbol_Entrez, by = c("mouse_entrez", "mgi_symbol")) %>% 
# Group by human gene symbol: after orthology mapping, collapse mouse genes that map to the same human gene
  group_by(hgnc_symbol) %>% 
  # Keep probe with highest median expression per gene
  slice_max(median, n = 1)  %>% #14.396
  ungroup()

# Remove HGNC symbols with multiple MGI mappings
hgnc_symbol_dupes = mouse_log2summary %>% 
  #filter ensembl id with multiple mgi_symbols
  get_dupes(hgnc_symbol)

#Filter out probes with multiple mgi and hgnc symbols
probe_dupes = mouse_log2summary %>% get_dupes(probe)

mouse_log2summary = mouse_log2summary %>% 
  select(probe, hgnc_symbol) %>% 
# Remove genes with ambiguous human symbol mappings: retains only one-to-one mouse-to-human orthologs
  filter(!hgnc_symbol %in% c(hgnc_symbol_dupes$hgnc_symbol %>% unique()),
         !probe %in% c(probe_dupes$probe %>% unique())) 

# Build filtered expression matrix with human symbols

# Join expression with annotation, keep only mapped probes
mouse_log2_filtered_probes = mouse_exprs %>% 
  rownames_to_column("probe_id") %>% 
  rename(probe = probe_id) %>% 
# Join expression data with annotation by probe ID: adds gene symbols to raw probe-level measurements
  inner_join(mouse_log2summary, by = "probe") %>%   #214,396
  select(human_symbol = hgnc_symbol,
         probe,
         c(mouse_metadata %>% 
             pull(geo_sample)))
#Check
mouse_log2_filtered_probes %>% dim()
mouse_log2_hgnc_symbol = mouse_log2_filtered_probes 
# Identify duplicated gene symbols: marks genes that map to multiple Entrez IDs or probes for removal
mouse_log2_hgnc_symbol %>% get_dupes(probe)

#Save
mouse_log2_hgnc_symbol %>% saveRDS(file = here("tables", "processed_data", "saureus_mouse_exprs_hgnc_symbol.rds"))
```

Pivot data and create expression set
```{r}
# Load data
mouse_log2_hgnc_symbol = readRDS(file = here("tables", "processed_data", "saureus_mouse_exprs_hgnc_symbol.rds"))
mouse_metadata = readRDS(file = here("tables", "processed_data", "saureus_mouse_metadata_filtered.rds"))

#Pivot to long and wide format, and annotate
mouse_log2_samples_log2fc_long = mouse_log2_hgnc_symbol %>% 
  # Reshape expression data from wide to long format
  pivot_longer(cols = -c(human_symbol, probe),
               names_to = "geo_sample",
               values_to = "log2fc") %>% 
# Attach sample metadata (participant ID, timepoint, condition) to expression values by GEO sample ID
  inner_join(mouse_metadata,
             by = "geo_sample")


mouse_log2_samples_log2fc_wide = mouse_log2_samples_log2fc_long %>% 
  select(human_symbol, probe, participant_id, timepoint, log2fc) %>%
# Widen data: reshape so that each timepoint becomes a separate column (required format for t-test input)
  pivot_wider(names_from = "participant_id",
              values_from = "log2fc") %>% 
  mutate(timepoint = as.character(timepoint) %>% fct_reorder(timepoint %>% as.numeric()))


#Create Expression Set
# Create ExpressionSet from expression matrix and phenotype data
eset_mouse <- ExpressionSet(
  assayData = mouse_log2_hgnc_symbol %>% 
    select(-probe) %>% 
# Move identifier column to row names: required format for ExpressionSet and Limma (matrix must have rownames)
    column_to_rownames("human_symbol") %>% 
    as.matrix(),
# Convert metadata to AnnotatedDataFrame: formal phenotype data structure required by ExpressionSet
  phenoData = AnnotatedDataFrame(mouse_metadata))

#Save
eset_mouse %>% saveRDS(file = here("tables", "Differential gene expression", "saureus_mouse_eset.rds"))

```











## Human
Prepare the S. aureus human dataset: download, import, and annotate.

```{r}
#S. aureus: GSE33341 -----
filename = "GSE33341"
GSE <- getGEO("GSE33341", GSEMatrix = TRUE) 
eset <- GSE[[2]] #The matrix #1 the mouse annotated

#Metadata ------
GSE_metadata = eset %>% 
  pData() %>% 
# Reorder columns: place comparison/condition column first for readability
  dplyr::select( 
    geo_sample = "geo_accession", 
    participant_id = "title",
    sex = "host gender:ch1",
    condition = "source_name_ch1",
    pathogen = "pathogen:ch1",
    description = "source_name_ch1", 
    ethnicity = "ethnicity:ch1",
    tissue_infection = "anatomic site of infection:ch1",
    batch_experimental = "experimental batch:ch1",
    age = "host age (yr):ch1") %>% 
  mutate(condition = pathogen,
         organism = "Human",
         timepoint = 1,
         time_unit = "Day",
         tissue = "Blood",
         participant_id = str_extract(participant_id, "\\[.*?\\]") %>% 
           str_remove("\\[") %>% 
           str_remove("\\]"),
         condition = case_when(description == "blood patient with E. coli bacteremia prior antibiotic treatment (Emergency Department)" ~ "E. coli prior antibiotic",
                               description == "blood patient with S. aureus bacteremia prior antibiotic treatment (Emergency Department)" ~ "S aureus prior antibiotic",
                               description == "blood human NOT infected" ~ "No infection",
                               T ~ description),
         pathogen = if_else(pathogen == "-", "Healthy", pathogen))
#Save
GSE_metadata %>% saveRDS(file = here("tables", "processed_data", "saureus_human_metadata.rds"))

#Filter metadata
GSE_metadata_filtered = GSE_metadata %>%
  filter(pathogen %in% c("Staphylococcus aureus", "Healthy")) %>% 
# Select all columns: ensures clean column order with key identifier first
  select(geo_sample, participant_id, condition, timepoint, everything()) %>% 
  mutate(ethnicity = str_to_title(ethnicity)) %>% 
  filter(batch_experimental != "2739",
         ethnicity != "Unknown",
         age != "unknown")

GSE_metadata_filtered %>% 
# Save R object to disk in native RDS format: preserves data types and structure for fast loading
  saveRDS(file = here("tables", "processed_data", "saureus_human_metadata.rds"))



#Expression data----
# Extract raw expression matrix from the ExpressionSet: probes in rows, samples in columns
GSE_exprs <- exprs(eset) %>% as.data.frame()

#Save
GSE_exprs %>% saveRDS(file = here("tables", "processed_data", "saureus_human_exprs.rds"))
```

Prepare data
Load and filter data for the current condition.
Filter and prepare expression data for the current condition.
```{r}
# Load data
saureus_human_exprs = readRDS(file = here("tables", "processed_data", "saureus_human_exprs.rds"))
saureus_human_metadata = readRDS(file = here("tables", "processed_data", "saureus_human_metadata.rds"))
#Inputs
filename = "saureus_human"
human_exprs = saureus_human_exprs 
human_metadata = saureus_human_metadata
```


Annotate gene symbols and map probes to human orthologs via BioMart.

```{r}
#Annotate genes ------
#Human -----
#Annotate data and calculate log2fc
human_exprs_metadata  = human_exprs %>% 
  rownames_to_column("probe_id") %>% 
# Reshape expression data from wide (columns = samples) to long (rows = probe × sample) format for tidyverse workflows
  pivot_longer(cols = -probe_id,
               names_to = "geo_sample",
               values_to = "value_raw") %>% 
# Attach sample metadata (participant ID, timepoint, condition) to expression values by GEO sample ID
  inner_join(human_metadata, 
             by = "geo_sample") 

#Get value from healthy and calculate Log2FC
human_exprs_log2fc_sample = human_exprs_metadata %>% 
  filter(pathogen == "Healthy") %>% 
  select(probe_id, value_raw_ref = value_raw) %>% 
# Group by probe ID to collapse multiple probe-to-gene mappings later
  group_by(probe_id) %>% 
  summarise(value_raw_ref = mean(value_raw_ref)) %>% 
# Attach sample metadata (participant ID, timepoint, condition) to expression values by GEO sample ID
  inner_join(human_exprs_metadata %>% 
               filter(pathogen != "Healthy"),
             by = join_by("probe_id")) %>% 
  filter(condition != "No infection") %>% 
  # Compute log2 fold-change relative to baseline
mutate(log2fc = value_raw - value_raw_ref) %>% 
# Remove duplicate rows: eliminates redundant probe-to-gene mappings (one probe may match multiple transcript clusters)
  distinct() 

#Check duplicates
human_exprs_log2fc_sample %>% 
  filter(participant_id == "CSD4620.CEL",
         probe_id == "1007_s_at")

human_exprs_log2fc_sample %>% 
# Save R object to disk in native RDS format: preserves data types and structure for fast loading
  saveRDS(here("tables", "Differential gene expression", "saureus_human_log2fc_sample_clean_long.rds"))
```


Annotate probes with gene symbols by querying the Ensembl BioMart database.

```{r}
########### Filter and annotate probes -----
# Configure Ensembl BioMart
ensembl_mart <- useMart("ensembl")
# listDatasets(ensembl_mart)
ensembl_mart <- useDataset("hsapiens_gene_ensembl", mart = ensembl_mart)

# Get probe IDs from expression matrix
gene_id <- human_exprs_log2fc_sample$probe_id %>% unique()

# Annotate data
annot <- getBM(
  attributes = c("affy_hg_u133a_2",
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "hgnc_symbol",
                 "description"),
  filters = "affy_hg_u133a_2",
  values = gene_id,
  mart = ensembl_mart) %>% 
  #Standardize
  rename(probe = affy_hg_u133a_2,
                     entrezgene_id = entrezgene_id,
                     human_symbol = hgnc_symbol,
                     ensembl_id = ensembl_gene_id)

#Filter genes with highest median expression
human_log2summary = human_exprs %>% 
  rownames_to_column("probe") %>% 
# Reshape expression data from wide (columns = samples) to long (rows = probe × sample) format for tidyverse workflows
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
# Group by probe ID to collapse multiple probe-to-gene mappings later
  group_by(probe) %>% 
# Compute per-probe summary statistics (mean, median) across all samples to select the most representative probe
  summarise(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
# Join BioMart gene annotation to expression data by probe ID: adds gene symbol, Entrez ID, and description
  inner_join(annot, 
            by = "probe") %>% 
  # Collapse to one row per Ensembl ID
  group_by(ensembl_id) %>% 
  # Keep probe with highest median expression per gene
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_id) %>% 
# Remove duplicate rows: eliminates redundant probe-to-gene mappings (one probe may match multiple transcript clusters)
  distinct() %>% 
  
  # Collapse to one row per Entrez ID
  group_by(entrezgene_id) %>% 
  # Keep probe with highest median expression per gene
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup()  


#Filter out genes with multiple hgnc_symbols
hgnc_symbol_dupes = human_log2summary %>% 
  #filter ensembl id with multiple hgnc_symbols
  get_dupes(human_symbol)

human_log2summary = human_log2summary %>% 
  filter(!human_symbol %in% c(hgnc_symbol_dupes$human_symbol %>% unique())) 

# Build filtered expression matrix with human symbols

# Join expression with annotation, keep only mapped probes
human_log2_filtered_probes = human_exprs %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
# Join expression data with annotation by probe ID: adds gene symbols to raw probe-level measurements
  inner_join(human_log2summary, by = "probe") %>%   #214,396
  select(human_symbol, probe,
         c(human_metadata$geo_sample))
#14395
human_log2_filtered_probes %>% dim()

human_log2_filtered_probes = human_log2_filtered_probes 

human_log2_filtered_probes %>% 
# Identify duplicated gene symbols: marks genes that map to multiple Entrez IDs or probes for removal
  get_dupes(human_symbol)

#Save 
human_log2_filtered_probes %>% 
# Save R object to disk in native RDS format: preserves data types and structure for fast loading
  saveRDS(here("tables", "processed_data", "saureus_human_exprs_hgnc_symbol.rds"))
```


Load the S. aureus human dataset from saved RDS files.

```{r}
# Load data
human_log2_hgnc_symbol = readRDS(file = here("tables", "processed_data", "saureus_human_exprs_hgnc_symbol.rds"))
human_metadata = readRDS(file = here("tables", "processed_data", "saureus_human_metadata.rds"))

#Pivot to long and wide format, and annotate
human_log2_samples_log2fc_long = human_log2_hgnc_symbol %>% 
  # Reshape expression data from wide to long format
  pivot_longer(cols = -c(human_symbol, probe),
               names_to = "geo_sample",
               values_to = "log2fc") %>% 
# Attach sample metadata (participant ID, timepoint, condition) to expression values by GEO sample ID
  inner_join(human_metadata,
             by = "geo_sample")


human_log2_samples_log2fc_wide = human_log2_samples_log2fc_long %>% 
  select(human_symbol, probe, participant_id, timepoint, log2fc) %>%
# Widen data: reshape so that each timepoint becomes a separate column (required format for t-test input)
  pivot_wider(names_from = "participant_id",
              values_from = "log2fc") %>% 
  mutate(timepoint = as.character(timepoint) %>% fct_reorder(timepoint %>% as.numeric()))


#Create Expression Set
human_metadata_eset = human_metadata %>% 
  mutate(age = as.numeric(age) %>% scale(),
         timepoint = as.factor(timepoint),
         sex = as.factor(sex),
         pathogen = fct_relevel(pathogen, "Healthy"),
         batch_experimental = as.factor(batch_experimental),
         ethnicity = as.factor(ethnicity)) 

# Create ExpressionSet from expression matrix and phenotype data
eset_human <- ExpressionSet(
  assayData = human_log2_hgnc_symbol %>% 
    select(-probe) %>% 
# Move identifier column to row names: required format for ExpressionSet and Limma (matrix must have rownames)
    column_to_rownames("human_symbol") %>% 
    select(c(human_metadata_eset$geo_sample)) %>% 
    as.matrix(),
# Convert metadata to AnnotatedDataFrame: formal phenotype data structure required by ExpressionSet
  phenoData = AnnotatedDataFrame(human_metadata_eset))

#Save
eset_human %>% saveRDS(file = here("tables", "Differential gene expression", "saureus_human_eset.rds"))

```





#==================
# 4. E. coli, infection

### Human and Mouse, same dataset

E. coli: GSE33341 -----.

```{r}
#E. coli: GSE33341 -----
filename = "GSE33341"
# Download GEO ExpressionSet using GEOquery package: retrieves both expression data and phenotype metadata
GSE33341 <- getGEO("GSE33341", GSEMatrix = TRUE) 
eset_mouse <- GSE33341[[1]]
eset_human <- GSE33341[[2]]

#Metadata ------
#Mouse
ecoli_mouse_metadata = eset_mouse %>% 
  pData() %>% 
# Reorder columns: place comparison/condition column first for readability
  dplyr::select( 
    geo_sample = "geo_accession", 
    #Host
    subject_title = "title",
    subject_num = "replicate number:ch1",
    batch = "experimental batch:ch1",
    host_strain = "host strain:ch1",
    organism = "organism_ch1",
    age = "host age (yr):ch1",
    sex = "host gender:ch1", 
    #Pathogen
    pathogen = "pathogen:ch1",
    status = "infection status:ch1",
    site_infection = "anatomic site of infection:ch1",
    route_infection = "inoculum route:ch1",
    dosage = "pathogen   dosage cfu/g:ch1",
    timepoint = "time after infection (h):ch1",
    bac_strain = "bacterial strain:ch1"
    ) %>% 
  filter(pathogen %in% c("Escherichia  coli", "-")) %>%
  mutate(pathogen = if_else(pathogen == "Escherichia  coli", "escherichia coli", "none"),
         subject_id = paste0("Mouse_", host_strain, "_Animal", subject_num, "_Batch", batch),
         status = str_replace(status, "-", "Healthy"),
         condition = paste("Mouse", host_strain, timepoint, sep = "_"),
         comparison = paste0("Mouse_", host_strain, "_Batch", batch)) %>% 
# Select all columns: ensures clean column order with key identifier first
  select(geo_sample, comparison, condition, subject_id, everything())


#Human
ecoli_human_metadata = eset_human %>% 
  pData() %>% 
  filter(`pathogen:ch1` %in% c("Escherichia coli", "-")) %>%
# Reorder columns: place comparison/condition column first for readability
  dplyr::select( 
    geo_sample = "geo_accession", 
    #Host
    subject_title = "title",
    # subject_num = "replicate number:ch1",
    batch = "experimental batch:ch1",
    # host_strain = "host strain:ch1",
    organism = "organism_ch1",
    age = "host age (yr):ch1",
    sex = "host gender:ch1", 
    ethnicity = "ethnicity:ch1",
    #Pathogen
    pathogen = "pathogen:ch1",
    # status = "infection status:ch1",
    site_infection = "anatomic site of infection:ch1"
    # route_infection = "inoculum route:ch1",
    # dosage = "pathogen   dosage cfu/g:ch1",
    # timepoint = "time after infection (h):ch1",
    # bac_strain = "bacterial strain:ch1"
    ) %>% 
  mutate(site_infection = str_to_lower(site_infection),
         organism = "Human",
         pathogen = if_else(pathogen == "-", "Healthy", pathogen),
         status = if_else(pathogen == "Healthy", pathogen, "infected"),
         timepoint = "0",
         subject_id = if_else(pathogen == "Escherichia coli", 
                              paste0("Ecoli_human_", site_infection,"_",geo_sample), 
                              paste0("Healthy_human_", geo_sample)),
         site_infection = if_else(site_infection == "-", "none", site_infection),
         condition = subject_id) %>% 
# Select all columns: ensures clean column order with key identifier first
  select(geo_sample, condition, subject_id, everything()) %>% 
  mutate(timepoint = if_else(pathogen == "Escherichia coli", "24", timepoint))

ecoli_human_metadata

#Unite metadata
ecoli_mouse_human_metadata = ecoli_human_metadata %>% 
  bind_rows(ecoli_mouse_metadata)

ecoli_mouse_human_metadata %>% 
# Save R object to disk in native RDS format: preserves data types and structure for fast loading
  saveRDS(file = here("tables", "processed_data", "ecoli_human_mouse_metadata.rds"))
```

Load the E. coli mouse dataset from saved RDS files.

```{r}
# Load data
ecoli_mouse_human_metadata =
# Load dataset from disk or GEO: reads pre-saved expression matrix / metadata into the R environment
  readRDS(file = here("tables", "processed_data", "ecoli_human_mouse_metadata.rds"))


ecoli_mouse_human_metadata %>% 
  select(strain, treatment, tissue) %>% 
# Remove duplicate rows: eliminates redundant probe-to-gene mappings (one probe may match multiple transcript clusters)
  distinct()

#Get expression data -----
#Mouse
# Extract raw expression matrix from the ExpressionSet: probes in rows, samples in columns
mouse_exprs <- exprs(eset_mouse) %>% 
  as.data.frame() %>% 
  select(GSE33341_ecoli_mouse_human_metadata %>% 
           filter(organism == "Mus musculus") %>% 
           pull(geo_sample)) %>% 
  rownames_to_column("probe_id") 

mouse_exprs %>% 
  as.data.frame() %>% 
# Save R object to disk in native RDS format: preserves data types and structure for fast loading
  saveRDS(here("tables", "processed_data", "ecoli_mouse_exprs.rds"))

#Human
# Extract raw expression matrix from the ExpressionSet: probes in rows, samples in columns
human_exprs <- exprs(eset_human) %>% 
  as.data.frame() %>% 
  select(GSE33341_ecoli_mouse_human_metadata %>% 
           filter(organism == "Human") %>% 
           pull(geo_sample)) %>% 
  rownames_to_column("probe_id")

human_exprs %>% 
  as.data.frame() %>% 
# Save R object to disk in native RDS format: preserves data types and structure for fast loading
  saveRDS(here("tables", "processed_data", "ecoli_human_exprs.rds"))

```

#### Mouse

Annotate genes

This procedure was done using only one batch (8)

```{r}
# Load data
ecoli_human_exprs = readRDS(here("tables", "processed_data", "ecoli_human_exprs.rds"))
ecoli_mouse_exprs = readRDS(here("tables", "processed_data", "ecoli_mouse_exprs.rds"))
ecoli_mouse_human_metadata = readRDS(file = here("tables", "processed_data", "ecoli_human_mouse_metadata.rds"))

#Annotate genes ------
#Mouse -----
#Annotate data and calculate log2fc
mouse_expression_metadata  = ecoli_mouse_exprs %>% 
  pivot_longer(cols = 2:125,
               names_to = "geo_sample",
               values_to = "value_raw") %>% 
# Attach sample metadata (participant ID, timepoint, condition) to expression values by GEO sample ID
  inner_join(ecoli_mouse_human_metadata, 
             by = "geo_sample") 

mouse_metadata =  ecoli_mouse_human_metadata %>% 
               filter(host_strain == "C57BL/6J",
                      batch == "8")

mouse_metadata %>% 
# Save R object to disk in native RDS format: preserves data types and structure for fast loading
  saveRDS(file = here("tables", "processed_data", "ecoli_mouse_metadata.rds"))

#Get value from day 0 and calculate Log2FC
log2fc_sample = mouse_expression_metadata %>% 
  #Filter batch
  filter(organism == "Mus musculus",
         status == "Healthy",
         host_strain == "C57BL/6J",
         batch == "8") %>% 
  select(probe_id, batch, subject_num, day0_value_raw = value_raw) %>% 
  #Get the mean expression value for healthy
  group_by(probe_id, batch) %>% 
  summarise(n = n(),
            #Mean of healthy
            day0_value_raw = sum(day0_value_raw)/n,
            batch) %>% 
  select(probe_id, day0_value_raw) %>% 
  #Filter batch
  inner_join(mouse_expression_metadata %>% 
               filter(timepoint != "0",
                      host_strain == "C57BL/6J",
                      batch == "8"),
             by = join_by("probe_id", "batch")) %>% 
  # Compute log2 fold-change relative to baseline
mutate(log2fc = value_raw - day0_value_raw) %>% 
# Remove duplicate rows: eliminates redundant probe-to-gene mappings (one probe may match multiple transcript clusters)
  distinct() 

#Check duplicates
log2fc_sample %>% 
  filter(subject_id == "Mouse_C57BL/6J_Animal1_Batch8",
         probe_id == "1415670_at")


#Save
log2fc_sample %>% 
# Save R object to disk in native RDS format: preserves data types and structure for fast loading
  saveRDS(here("tables", "Differential gene expression", "ecoli_mouse_log2fc_sample_clean_long.rds"))
```

Annotate probes with gene symbols by querying the Ensembl BioMart database.

```{r}
########### Filter and annotate probes -----
# Configure Ensembl BioMart
ensembl_mart <- useMart("ensembl")
# Select the mouse genome (mmusculus_gene_ensembl) as the annotation dataset in BioMart
ensembl_mart <- useDataset("mmusculus_gene_ensembl", mart = ensembl_mart)

# Get probe IDs from expression matrix
gene_id <- log2fc_sample$probe_id %>% unique()
# listAttributes(ensembl_mart)
# Annotate data
annot <- getBM(
  attributes = c("affy_mouse430_2",
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "mgi_symbol",
                 "description"),
  filters = "affy_mouse430_2",
  values = gene_id,
  mart = ensembl_mart)

annot %>% 
# Save R object to disk in native RDS format: preserves data types and structure for fast loading
  saveRDS(here("tables", "Differential gene expression", "probe_annot_ecoli_infection_mouse.rds"))

# Load mouse-to-human orthology lookup
# (Install orthology package if needed)
# library(devtools)
# devtools::install_github('shenorrLabTRDF/FIT.mouse2man')
library(FIT.mouse2man)
HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez %>% 
# Standardise column names to snake_case using janitor::clean_names for consistent tidyverse syntax
  clean_names() %>% 
  rename(mouse_entrez = mouse_ortholog)

#Get annotations for each probe 
### Filter probes with highest median expression
mouse_log2summary = ecoli_mouse_exprs %>% #20,845 probes
  as.data.frame() %>% 
  rename(probe = probe_id) %>% 
# Reshape expression data from wide (columns = samples) to long (rows = probe × sample) format for tidyverse workflows
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
# Group by probe ID to collapse multiple probe-to-gene mappings later
  group_by(probe) %>% 
# Compute per-probe summary statistics (mean, median) across all samples to select the most representative probe
  summarise(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
# Join BioMart gene annotation to expression data by probe ID: adds gene symbol, Entrez ID, and description
  inner_join(annot %>% 
              rename(probe = affy_mouse430_2), 
            by = "probe") %>%  #24.009 rows
  # Collapse to one row per Ensembl ID
  group_by(ensembl_gene_id) %>% 
  # Keep probe with highest median expression per gene
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_gene_id) %>% 
# Remove duplicate rows: eliminates redundant probe-to-gene mappings (one probe may match multiple transcript clusters)
  distinct() %>% 
  
  # Collapse to one row per Entrez ID
  group_by(entrezgene_id) %>% 
  # Keep probe with highest median expression per gene
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup() %>% 
  rename(mouse_entrez = entrezgene_id) %>% 
  
  # Join human orthologs, collapse to one row per HGNC symbol
  inner_join(HS_MM_Symbol_Entrez, by = c("mouse_entrez", "mgi_symbol")) %>% 
# Group by human gene symbol: after orthology mapping, collapse mouse genes that map to the same human gene
  group_by(hgnc_symbol) %>% 
  # Keep probe with highest median expression per gene
  slice_max(median, n = 1)  %>% #14.396
  ungroup()

# Remove HGNC symbols with multiple MGI mappings
hgnc_symbol_dupes = mouse_log2summary %>% 
  #filter ensembl id with multiple mgi_symbols
  get_dupes(hgnc_symbol)

mouse_log2summary = mouse_log2summary %>% 
  select(probe, hgnc_symbol) %>% 
# Remove genes with ambiguous human symbol mappings: retains only one-to-one mouse-to-human orthologs
  filter(!hgnc_symbol %in% c(hgnc_symbol_dupes$hgnc_symbol %>% unique())) 

# Build filtered expression matrix with human symbols

# Join expression with annotation, keep only mapped probes
mouse_log2_filtered_probes = ecoli_mouse_exprs %>% 
  as.data.frame() %>% 
  rename(probe = probe_id) %>% 
# Join expression data with annotation by probe ID: adds gene symbols to raw probe-level measurements
  inner_join(mouse_log2summary, by = "probe") %>%   #214,396
  select(human_symbol = hgnc_symbol, 
         probe,
         c(mouse_metadata %>% 
             pull(geo_sample)))
#14395
mouse_log2_filtered_probes %>% dim()

mouse_log2_hgnc_symbol = mouse_log2_filtered_probes 

mouse_log2_hgnc_symbol %>% 
# Identify duplicated gene symbols: marks genes that map to multiple Entrez IDs or probes for removal
  get_dupes(human_symbol)

#Save
mouse_log2_hgnc_symbol %>% 
# Save R object to disk in native RDS format: preserves data types and structure for fast loading
  saveRDS(file = here("tables", "processed_data", "ecoli_mouse_exprs_hgnc_symbol_long.rds"))
```

Import
Load the required data objects from disk.
Load the required data objects from disk.

```{r}
# Load data
mouse_counts_log2_summary_annotated_genes = 
# Load dataset from disk or GEO: reads pre-saved expression matrix / metadata into the R environment
  readRDS(file = here("tables", "processed_data", "ecoli_mouse_exprs_hgnc_symbol_long.rds"))

#Pivot wider
log2fc_sample_wide = mouse_counts_log2_summary_annotated_probes %>% 
  select(probe = probe_id, entrezgene_id, batch:human_symbol) %>% 
# Remove duplicate rows: eliminates redundant probe-to-gene mappings (one probe may match multiple transcript clusters)
  distinct() %>% 
  select(human_symbol, subject_id, timepoint, log2fc) %>%
  group_by(human_symbol, timepoint) %>% 
# Widen data: reshape so that each timepoint becomes a separate column (required format for t-test input)
  pivot_wider(names_from = subject_id,
              values_from = log2fc) %>% 
  ungroup()  %>% 
  mutate(timepoint = fct_reorder(timepoint, timepoint %>% as.numeric()))
  
#Print
log2fc_sample_wide
```





#### Human
There are batches

```{r}

ecoli_human_exprs = readRDS(here("tables", "processed_data", "ecoli_human_exprs.rds"))
ecoli_mouse_exprs = readRDS(here("tables", "processed_data", "ecoli_mouse_exprs.rds"))
ecoli_mouse_human_metadata = readRDS(file = here("tables", "processed_data", "ecoli_human_mouse_metadata.rds"))


#Annotate genes ------
#Mouse -----
#Annotate data and calculate log2fc
human_expression_metadata  = ecoli_human_exprs %>% 
  pivot_longer(cols = 2:63,
               names_to = "geo_sample",
               values_to = "value_raw") %>% 
# Attach sample metadata (participant ID, timepoint, condition) to expression values by GEO sample ID
  inner_join(ecoli_mouse_human_metadata, 
             by = "geo_sample") 

human_expression_metadata %>% 
# Save R object to disk in native RDS format: preserves data types and structure for fast loading
  saveRDS(file = here("tables", "processed_data", "ecoli_human_exprs_metadata.rds"))


human_expression_metadata =
# Load dataset from disk or GEO: reads pre-saved expression matrix / metadata into the R environment
  readRDS(file = here("tables", "processed_data", "ecoli_human_exprs_metadata.rds"))

human_metadata =  human_expression_metadata %>% 
  filter(organism == "Human") %>% 
# Remove duplicate rows: eliminates redundant probe-to-gene mappings (one probe may match multiple transcript clusters)
  select(-value_raw, - probe_id) %>% distinct()

# Save R object to disk in native RDS format: preserves data types and structure for fast loading
human_metadata %>% saveRDS(file = here("tables", "processed_data", "ecoli_human_metadata.rds"))

#Get value from day 0 and calculate Log2FC
human_log2fc_sample = human_expression_metadata %>% 
  #Filter batch
  filter(organism == "Human",
         status == "Healthy") %>% 
  select(probe_id, participant_id = subject_id, status, day0_value_raw = value_raw) %>% 
# Group by probe ID to collapse multiple probe-to-gene mappings later
  group_by(probe_id, status) %>% 
  summarise(n = n(),
            #Mean of healthy
            day0_value_raw = sum(day0_value_raw)/n) %>% 
  select(probe_id, day0_value_raw) %>% 
  #Join by probe id to infected participants
  inner_join(human_expression_metadata %>% 
               filter(site_infection == "urinary tract"), #1 lung, 19 urinary tract),
             by = join_by("probe_id")) %>% 
  # Compute log2 fold-change relative to baseline
mutate(log2fc = value_raw - day0_value_raw) %>% 
  select(probe_id:subject_id, batch:status, log2fc)

#Check duplicates
human_log2fc_sample %>% 
  filter(subject_id == "Ecoli_human_urinary tract_GSM824707",
         probe_id == "1007_s_at") #OK

human_log2fc_sample %>% 
# Save R object to disk in native RDS format: preserves data types and structure for fast loading
  saveRDS(here("tables", "Differential gene expression", "ecoli_human_log2fc_sample_clean_long.rds"))

```

Annotate probes with gene symbols by querying the Ensembl BioMart database.

```{r}
########### Filter and annotate probes -----
# Configure Ensembl BioMart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("hsapiens_gene_ensembl", mart = ensembl_mart)

# Get probe IDs from expression matrix
gene_id <- human_log2fc_sample$probe_id %>% unique()
# listAttributes(ensembl_mart)%>% View()
# Annotate data
#[HG-U133A_2] Affymetrix Human Genome U133A 2.0 Array
annot <- getBM(
  attributes = c("affy_hg_u133a_2",
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "hgnc_symbol",
                 "description"),
  filters = "affy_hg_u133a_2",
  values = gene_id,
  mart = ensembl_mart)

#Get annotations for each probe 
### Filter probes with highest median expression
#Get annotations for each probe 
### Filter probes with highest median expression
human_ecoli_log2summary = human_exprs %>% 
  rename(probe = probe_id) %>% 
# Reshape expression data from wide (columns = samples) to long (rows = probe × sample) format for tidyverse workflows
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
# Group by probe ID to collapse multiple probe-to-gene mappings later
  group_by(probe) %>% 
# Compute per-probe summary statistics (mean, median) across all samples to select the most representative probe
  summarise(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
# Join BioMart gene annotation to expression data by probe ID: adds gene symbol, Entrez ID, and description
  inner_join(annot %>% 
              rename(probe = affy_hg_u133a_2), 
            by = "probe") %>% 
  # Collapse to one row per Ensembl ID
  group_by(ensembl_gene_id) %>% 
  # Keep probe with highest median expression per gene
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_gene_id) %>% 
# Remove duplicate rows: eliminates redundant probe-to-gene mappings (one probe may match multiple transcript clusters)
  distinct() %>% 
  
  # Collapse to one row per Entrez ID
  group_by(entrezgene_id) %>% 
  # Keep probe with highest median expression per gene
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup() %>% 
  rename(human_symbol = hgnc_symbol) %>% 
  filter(human_symbol != "") %>%
# Group by Entrez gene ID: collapse multiple Ensembl IDs per Entrez ID after orthology mapping
  group_by(entrezgene_id) %>%
# Randomly select one representative entry from each group of duplicates (last resort for ambiguous mappings)
  slice_sample(n = 1) %>%  # one line per entrezgene_id
  ungroup() %>%
  distinct(probe, .keep_all = TRUE)  # remove dup probes

human_ecoli_log2summary %>% 
# Identify duplicated gene symbols: marks genes that map to multiple Entrez IDs or probes for removal
  get_dupes(probe)



# Join expression with annotation, keep only mapped probes
human_ecoli_log2_filtered_probes = human_exprs %>%
  rename(probe = probe_id) %>% 
# Join expression data with annotation by probe ID: adds gene symbols to raw probe-level measurements
  inner_join(human_ecoli_log2summary, by = "probe") %>%   #214,396
  select(human_symbol, probe,
         c(human_metadata$geo_sample))
#14395
human_ecoli_log2_filtered_probes %>% dim()

human_ecoli_log2_filtered_symbol = human_ecoli_log2_filtered_probes 

human_ecoli_log2_filtered_symbol %>% 
# Identify duplicated gene symbols: marks genes that map to multiple Entrez IDs or probes for removal
  get_dupes(human_symbol)

#Save 
human_ecoli_log2_filtered_symbol %>% 
# Save R object to disk in native RDS format: preserves data types and structure for fast loading
  saveRDS(here("tables", "processed_data", "ecoli_human_exprs_hgnc_symbol.rds"))
```












#====================

#5. Burn and Trauma, Injuries


##Download datasets

### Human

#### Trauma, Human (GSE36809)
Clinical metadata were curated and standardized to ensure consistency across the dataset. The time elapsed since injury, originally recorded in hours, was converted to days and rounded to one decimal place. To account for the irregular temporal distribution of samples, timepoints were discretized into seven categorical bins: <1 day (Early Hours), 1–2 days, 2–5 days, 5–10 days, 10–18 days, 18–25 days, and >25 days. Samples were stratified into 'Trauma' and 'Control' groups based on subject identifiers. Rigorous quality control was applied by excluding samples explicitly flagged in the metadata for low RNA quality or incomplete temporal data. Finally, temporal bins were treated as ordered factors to facilitate downstream differential expression modeling.

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
#Download datasets
filename = "trauma_human"
GSE_matrix <- getGEO(filename = here("tables", "raw_data", "GSE36809_series_matrix.txt.gz"), 
                     GSEMatrix = T)

#Get Expression Set
eset <- GSE_matrix

#Get metadata -------
GSE_metadata = eset %>% Biobase::pData()

#Select, clean, and standardize attributes
GSE_metadata_annotated = GSE_metadata %>% 
  select(geo_sample = "geo_accession",
         sample_id = "source_name_ch1",
         description = title,
         tissue = `tissue:ch1`,
         sex = `Sex:ch1`,
         age = "age:ch1",
         timepoint = `hours_since_injury:ch1`,
         platform_id
         ) %>% 
  separate(sample_id, sep = ", ", into = c("subject_id", "delete")) %>% 
  select(-delete) %>% 
    mutate(timepoint = if_else(timepoint == "N/A", "0", timepoint),
         timepoint = (as.numeric(timepoint)/24) %>% round(x = ., digits = 1),
         subject_id = str_remove(subject_id, "Subject "),
         geo = "GSE36809",
         organism = "Human",
         treatment = if_else(subject_id != "Control", "Trauma", "Control")) %>% 
  filter(!description %in% c("Please notice that this sample with low RNA quality.", 
                            "Sample from patients with incomplete time points.")) %>% 
  mutate(
    
    # Binned timepoints
    timepoints_new = case_when(
      timepoint < 1            ~ "Early_Hours",
      timepoint >= 1 & timepoint < 2  ~ "Day_1",
      timepoint >= 2 & timepoint < 5  ~ "Day_3",
      timepoint >= 5 & timepoint < 10 ~ "Day_7",
      timepoint >= 10 & timepoint < 18 ~ "Day_14",
      timepoint >= 18 & timepoint < 25 ~ "Day_21",
      timepoint >= 25          ~ "Day_28",
      TRUE                         ~ "Other"
    ),
    
    #Explicit bins
    timepoints_bin = case_when(
      timepoint < 1            ~ "<Day_1",
      timepoint >= 1 & timepoint < 2  ~ "Day_1_>=1 to <2",
      timepoint >= 2 & timepoint < 5  ~ "Day_3_>=2 to <5",
      timepoint >= 5 & timepoint < 10 ~ "Day_7_>=5 to <10",
      timepoint >= 10 & timepoint < 18 ~ "Day_14_>=10 to <18",
      timepoint >= 18 & timepoint < 25 ~ "Day_21_>=18 to <25",
      timepoint >= 25          ~ "Day_28_>25",
      TRUE                         ~ "Other"
    ),
    
    timepoints_new = factor(timepoints_new, levels = c("Early_Hours", "Day_1", "Day_3", "Day_7", "Day_14", "Day_21", "Day_28"))
  )



GSE_metadata_annotated %>% 
  group_by(subject_id)%>%
  arrange(subject_id, timepoint) %>%
  View()

GSE_metadata_annotated %>% 
  filter(subject_id != "Control") %>% 
  group_by(subject_id)%>%
  summarise(n_timepoints = n()) %>% 
  arrange(subject_id, n_timepoints) %>% 
  filter(n_timepoints > 1) # -10 participants


GSE_metadata_annotated %>% 
  group_by(timepoints_new, treatment)%>%
  summarise(n_sample = n()) %>% 
  arrange(timepoints_new) %>% 
# Create ggplot object: maps data to aesthetic (x = expression value)
  ggplot() +
  aes(x = timepoints_new,
      size = n_sample,
      y = treatment,
      color = treatment) +
# Add scatter plot points: each point represents a sample in PCA/MDS space
  geom_point()

GSE_metadata_annotated_filtered = GSE_metadata_annotated %>% 
  filter(timepoints_new %in% c("Early_Hours", "Day_1", "Day_3", "Day_7"))


#Save metadata
saveRDS(GSE_metadata_annotated_filtered, file = here("tables", "processed_data", paste0(filename, "_metadata.rds")))

#Get counts matrix ------
GSE_exprs <- eset %>% exprs()

#Save as RDS
GSE_exprs %>% 
  as.data.frame() %>% 
  select(GSE_metadata_annotated_filtered$geo_sample) %>% 
  saveRDS(here("tables", "processed_data", paste0(filename, "_exprs.rds")))

```












#### Burn, Human (GSE37069)

Download the raw dataset from GEO using GEOquery.

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
#Download datasets
filename = "burn_human"
GSE_matrix <- getGEO(filename = here("tables", "raw_data", "GSE37069_series_matrix.txt.gz"), 
                     GSEMatrix = T)

#Get Expression Set
eset <- GSE_matrix

#Get metadata -------
GSE_metadata = eset %>% Biobase::pData()

#Select, clean, and standardize attributes
GSE_metadata_annotated = GSE_metadata %>% 
  select(geo_sample = "geo_accession",
         treatment = "source_name_ch1",
         organism = organism_ch1,
         age = `age:ch1`,
         tissue = `tissue:ch1`,
         sex = `Sex:ch1`,
         timepoint = `hours_since_injury:ch1`,
         platform_id,
         ) %>% 
  separate(treatment, sep = ", ", into = c("subject_id", "delete")) %>% 
  select(-delete) %>% 
    mutate(subject_id = str_remove(subject_id, "Subject "),
           timepoint = if_else(timepoint == "N/A", "0", timepoint),
         timepoint = (as.numeric(timepoint)/24) %>% round(x = ., digits = 1),
         geo = "GSE37069",
         treatment = if_else(subject_id != "Control", "Burn", "Control")) %>% 
  mutate(
    
    # Binned timepoints
    timepoints_new = case_when(
      timepoint < 1            ~ "Early_Hours",
      timepoint >= 1 & timepoint < 2  ~ "Day_1",
      timepoint >= 2 & timepoint < 5  ~ "Day_3",
      timepoint >= 5 & timepoint < 10 ~ "Day_7",
      timepoint >= 10 & timepoint < 18 ~ "Day_14",
      timepoint >= 18 & timepoint < 25 ~ "Day_21",
      timepoint >= 25          ~ "Day_28",
      TRUE                         ~ "Other"
    ),
    
    #Explicit bins
    timepoints_bin = case_when(
      timepoint < 1            ~ "<Day_1",
      timepoint >= 1 & timepoint < 2  ~ "Day_1_>=1 to <2",
      timepoint >= 2 & timepoint < 5  ~ "Day_3_>=2 to <5",
      timepoint >= 5 & timepoint < 10 ~ "Day_7_>=5 to <10",
      timepoint >= 10 & timepoint < 18 ~ "Day_14_>=10 to <18",
      timepoint >= 18 & timepoint < 25 ~ "Day_21_>=18 to <25",
      timepoint >= 25          ~ "Day_28_>25",
      TRUE                         ~ "Other"
    ),
    
    timepoints_new = factor(timepoints_new, levels = c("Early_Hours", "Day_1", "Day_3", "Day_7", "Day_14", "Day_21", "Day_28"))
  )


GSE_metadata_annotated %>% 
  group_by(timepoints_new, treatment)%>%
  summarise(n_sample = n()) %>% 
  arrange(timepoints_new) %>% 
  ggplot() +
  aes(x = timepoints_new,
      size = n_sample,
      y = treatment,
      color = treatment) +
  geom_point()

GSE_metadata_annotated %>%
  group_by(timepoint, treatment)%>%
  summarise(n_sample = n()) %>%
  arrange(timepoint) %>%
  ggplot() +
  aes(x = timepoint,
      color = treatment) +
  geom_histogram()

GSE_metadata_annotated_filtered = GSE_metadata_annotated %>% 
  filter(timepoints_new %in% c("Early_Hours", "Day_1", "Day_3", "Day_7"))
  

#Save metadata
write_csv(GSE_metadata_annotated_filtered, file = here("tables",  "processed_data", paste0(filename, "_metadata.csv")))

#Get counts matrix ------
GSE_exprs <- eset %>% exprs()

#Save as RDS
GSE_exprs %>% 
  as.data.frame() %>% 
  select(GSE_metadata_annotated_filtered$geo_sample) %>% 
  saveRDS(here("tables", "processed_data", paste0(filename, "_exprs.rds")))

```

### Mouse
#### Trauma and Burn, Mouse (GSE7404)

Download the raw dataset from GEO using GEOquery.

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
#Download datasets
filename = "trauma_burn_mouse"
GSE_matrix <- getGEO(filename = here("tables", "raw_data", "GSE7404_series_matrix.txt.gz"), 
                     GSEMatrix = T)

#Get Expression Set
eset <- GSE_matrix

#Get metadata
GSE_metadata = eset %>% Biobase::pData()

#Select, clean, and standardize attributes
GSE_metadata_annotated = GSE_metadata %>% 
  select(geo_sample = "geo_accession",
         sample_id = "source_name_ch1",
         participant_description = title,
         characteristics = characteristics_ch1,
         timepoint = `time:ch1`,
         genetic = `genetic background:ch1`,
         tissue = `source_name_ch1`,
         organism = organism_ch1,
         description = description,
         # sex = `Sex:ch1`,
         platform_id
         ) %>% 
# Filter to PBMC samples only
  filter(tissue == "mouse leukocytes") %>% 
  fill(genetic, .direction = "up") %>% 
  mutate(treatment = case_when(str_detect(description, "burn") ~ "Burn", 
                               str_detect(description, "trauma") ~ "Trauma", 
                               str_detect(description, "LPS") ~ "LPS",
                               T~description),
         treatment = if_else(str_detect(participant_description, "Sham"), paste0(treatment, "_", "Control"), treatment),
         tissue = str_remove(tissue, "mouse "), 
         sex = "male",
         geo = "GSE7404") %>% 
  extract(
    col = participant_description, 
    into = c("timepoint", "participant_id"), 
    regex = "(\\d+\\s?(?:day|Day|hr))\\s+(rep\\s?\\d+)", 
    remove = T 
  ) %>% 
  mutate(timepoint = str_replace(timepoint, "hr", "Hour") %>% str_replace("day", "Day"),
         timepoint = fct_reorder(timepoint, parse_number(timepoint))) %>% 
  select(geo_sample, participant_id, timepoint, sex, 
         strain = genetic, tissue, organism, treatment, platform_id) 


GSE_metadata_annotated %>% 
  group_by(timepoint, treatment)%>%
  summarise(n_sample = n()) %>% 
  arrange(timepoint) %>% 
  ggplot() +
  aes(x = timepoint,
      size = n_sample,
      y = treatment,
      color = treatment) +
  geom_point()
  

#Save metadata
write_csv(GSE_metadata_annotated, file = here("tables", "processed_data", paste0(filename, "_metadata.csv")))

#Get counts matrix
GSE_exprs <- eset %>% exprs()

#Save as RDS
GSE_exprs %>% as.data.frame() %>% 
  saveRDS(here("tables", "processed_data", paste0(filename, "_exprs.rds")))

```


## Annotate genes

#### Mouse 

##### Trauma and Burn

Import previously saved analysis results for combined processing.

```{r}
#Import datasets
mouse_exprs = readRDS(here("tables", "processed_data", "trauma_burn_mouse_exprs.rds")) %>% 
  as.matrix() %>%
  { log2(. + 1) } %>% 
  as.data.frame()
mouse_metadata = read_csv(file = here("tables", "processed_data", "trauma_burn_mouse_metadata.csv")) %>% 
  mutate(status = if_else(str_detect(treatment, "_Control"), "Control", "Treated"),
         treatment = str_remove(treatment, "_Control"))

#Inputs
filename = "trauma_burn_mouse"

#Annotate genes ------
mouse_exprs_metadata  = mouse_exprs %>% 
  rownames_to_column("probe_id") %>% 
  pivot_longer(cols = 2:145,
               names_to = "geo_sample",
               values_to = "value_raw") %>% 
# Attach sample metadata 
  inner_join(mouse_metadata, 
             by = "geo_sample") 

#Get value from day 0 and calculate Log2FC (non-paired)
mouse_exprs_log2fc_sample = mouse_exprs_metadata %>% 
  filter(status == "Control") %>% 
  select(probe_id, treatment, timepoint, subject_num = participant_id, value_raw_ref = value_raw) %>% 
  distinct() %>% 
  group_by(timepoint, treatment, probe_id) %>% 
  summarise(value_raw_ref = mean(value_raw_ref)) %>% 
  ungroup() %>% 
  inner_join(mouse_exprs_metadata %>% 
               filter(status != "Control"),
             by = join_by("probe_id", "treatment", "timepoint")) %>% 
mutate(log2fc = value_raw - value_raw_ref) %>% 
  distinct() 

#Check duplicates
mouse_exprs_log2fc_sample %>% 
  filter(participant_id == "rep 1",
         treatment == "Burn",
         probe_id == "1415670_at")
```


Annotate probes with gene symbols by querying the Ensembl BioMart database.

```{r}
########### Filter and annotate probes -----
# Configure Ensembl BioMart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("mmusculus_gene_ensembl", mart = ensembl_mart)

# Get probe IDs from expression matrix
gene_id <- rownames(mouse_exprs) %>% unique()

# listAttributes(ensembl_mart) %>% View()
# Annotate data
annot <- getBM(
  attributes = c("affy_mouse430_2",
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "mgi_symbol",
                 "description"),
  filters = "affy_mouse430_2", #Probes
  values = gene_id,
  mart = ensembl_mart)

# Load mouse-to-human orthology lookup
library(FIT.mouse2man)
HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez %>% 
  clean_names() %>% 
  rename(mouse_entrez = mouse_ortholog)

#Get annotations for each probe 
### Filter probes with highest median expression
mouse_log2summary = mouse_exprs %>% #20,845 probes
  as.data.frame() %>% 
  rownames_to_column("probe_id") %>% 
  rename(probe = probe_id) %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarise(median = median(expression, na.rm = T)) %>%  
  inner_join(annot %>% 
              rename(probe = affy_mouse430_2), 
            by = "probe") %>%  #24.009 rows
  group_by(ensembl_gene_id) %>% 
  # Keep probe with highest median expression per gene
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_gene_id) %>% 
  distinct() %>% 
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup() %>% 
  rename(mouse_entrez = entrezgene_id) %>% 

  # Join human orthologs, collapse to one row per HGNC symbol
  inner_join(HS_MM_Symbol_Entrez, by = c("mouse_entrez", "mgi_symbol")) %>% 
  group_by(hgnc_symbol) %>% 
  # Keep Entrez with highest median expression per gene
  slice_max(median, n = 1)  %>% #14.396
  ungroup()

# Remove HGNC symbols with multiple MGI mappings
hgnc_symbol_dupes = mouse_log2summary %>% 
  get_dupes(hgnc_symbol)

mouse_log2summary = mouse_log2summary %>% 
  select(probe, hgnc_symbol) %>% 
  filter(!hgnc_symbol %in% c(hgnc_symbol_dupes$hgnc_symbol %>% unique())) 

# Build filtered expression matrix with human symbols

# Join expression with annotation, keep only mapped probes
mouse_log2_filtered_probes = mouse_exprs %>% 
  rownames_to_column("probe_id") %>% 
  as.data.frame() %>% 
  rename(probe = probe_id) %>% 
  inner_join(mouse_log2summary, by = "probe") %>%   #214,396
  select(human_symbol = hgnc_symbol,
         probe,
         c(mouse_metadata %>% 
             pull(geo_sample)))
#Check
mouse_log2_filtered_probes %>% dim()
mouse_log2_hgnc_symbol = mouse_log2_filtered_probes 
mouse_log2_hgnc_symbol %>% get_dupes(human_symbol)

#Save
mouse_log2_hgnc_symbol %>% saveRDS(file = here("tables", paste0(filename, "_exprs_hgnc_symbol.rds")))

mouse_log2_hgnc_symbol = readRDS(file = here("tables", paste0(filename, "_exprs_hgnc_symbol.rds")))

mouse_exprs_log2fc_sample = mouse_exprs_log2fc_sample %>% 
  rename(probe = probe_id) %>% 
  inner_join(mouse_log2_hgnc_symbol %>% 
               select(human_symbol, probe),
             by = join_by(probe))
#Save
mouse_exprs_log2fc_sample %>% 
  saveRDS(here("tables", "Differential gene expression", "trauma_burn_mouse_log2fc_sample_clean_long.rds"))

```




#### Human
##### Trauma

Import previously saved analysis results for combined processing.

```{r}
#Import datasets
human_exprs = readRDS(here("tables", "processed_data", "trauma_human_exprs.rds"))
human_metadata = read_csv(file = here("tables", "processed_data", "trauma_human_metadata.csv"))

#Inputs
filename = "trauma_human"

#Annotate genes ------
#Annotate data and calculate log2fc
human_exprs_metadata  = human_exprs %>% 
  rownames_to_column("probe_id") %>% 
  pivot_longer(cols = 2:681,
               names_to = "geo_sample",
               values_to = "value_raw") %>% 
  inner_join(human_metadata, 
             by = "geo_sample")

#Get value from day 0 and calculate Log2FC (non-paired)
human_exprs_log2fc_sample = human_exprs_metadata %>% 
  filter(treatment == "Control") %>% 
  select(probe_id, treatment, value_raw_ref = value_raw) %>% 
  distinct() %>% 
  group_by(probe_id) %>% 
  summarise(value_raw_ref = mean(value_raw_ref)) %>% 
  ungroup() %>% 
  inner_join(human_exprs_metadata %>% 
               filter(treatment != "Control"),
             by = join_by("probe_id")) %>% 
  # Compute log2 fold-change relative to baseline
mutate(log2fc = value_raw - value_raw_ref) %>% 
  distinct() 

human_trauma_log2fc_sample = human_exprs_log2fc_sample

#Check for duplicates
human_exprs_log2fc_sample %>% 
  filter(description == "Blood, subject 15429976, hours since injury 5.8",
         probe_id == "1007_s_at")
```


Annotate probes with gene symbols by querying the Ensembl BioMart database.

```{r}
########### Filter and annotate probes -----
# Configure Ensembl BioMart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("hsapiens_gene_ensembl", mart = ensembl_mart)

# Get probe IDs from expression matrix
gene_id <- rownames(human_exprs) %>% unique()

# listAttributes(ensembl_mart) %>% View()
# Annotate data
annot <- getBM(
  attributes = c("affy_hg_u133_plus_2",
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "hgnc_symbol",
                 "description"),
  filters = "affy_hg_u133_plus_2", #Probes
  values = gene_id,
  mart = ensembl_mart) %>% #Standardize
  rename(probe = affy_hg_u133_plus_2,
                     entrezgene_id = entrezgene_id,
                     human_symbol = hgnc_symbol,
                     ensembl_id = ensembl_gene_id)

#Filter genes with highest median expression
human_log2summary = human_exprs %>% 
  rownames_to_column("probe") %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarise(median = median(expression, na.rm = T)) %>%  
  inner_join(annot, 
            by = "probe") %>% 
  
  group_by(ensembl_id) %>% 
  # Keep probe with highest median expression per gene
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_id) %>% 
  distinct() %>% 
  # Collapse to one row per Entrez ID
  group_by(entrezgene_id) %>% 
  # Keep probe with highest median expression per gene
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup()  


#Filter out genes with multiple hgnc_symbols
hgnc_symbol_dupes = human_log2summary %>% 
  get_dupes(human_symbol)

human_log2summary = human_log2summary %>% 
  filter(!human_symbol %in% c(hgnc_symbol_dupes$human_symbol %>% unique())) 

# Join expression with annotation, keep only mapped probes
human_log2_filtered_probes = human_exprs %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  inner_join(human_log2summary, by = "probe") %>%   #214,396
  select(human_symbol, probe,
         c(human_metadata$geo_sample)) %>% 
  drop_na()
#18124
human_log2_filtered_probes %>% dim()

human_log2_filtered_probes = human_log2_filtered_probes 

human_log2_filtered_probes %>% 
  get_dupes(human_symbol)

#Save 
human_log2_filtered_probes %>% 
# Save
  saveRDS(here("tables", "processed_data", "trauma_human_exprs_hgnc_symbol.rds"))

#Annotate probes from log2fc table
human_exprs_log2fc_sample = human_trauma_log2fc_sample %>% 
  rename(probe = probe_id) %>% 
  inner_join(human_log2_filtered_probes %>% 
               select(probe,
                      human_symbol),
             by = "probe")

human_exprs_log2fc_sample %>% head()

human_exprs_log2fc_sample %>% 
# Save R object to disk in native RDS format: preserves data types and structure for fast loading
  saveRDS(here("tables", "Differential gene expression", "trauma_human_log2fc_sample_clean_long.rds"))


```



##### Burn

Import previously saved analysis results for combined processing.

```{r}
#Import datasets
human_exprs = readRDS(here("tables", "processed_data", "burn_human_exprs.rds")) #Log2 values
human_metadata = read_csv(file = here("tables", "processed_data", "burn_human_metadata.csv"))

#Inputs
filename = "burn_human"

#Annotate genes ------
#Mouse -----
#Annotate data and calculate log2fc
human_exprs_metadata  = human_exprs %>% 
  rownames_to_column("probe_id") %>% 
  pivot_longer(cols = 2:329,
               names_to = "geo_sample",
               values_to = "value_raw") %>% 
  inner_join(human_metadata, 
             by = "geo_sample")

#Get value from day 0 and calculate Log2FC (non-paired)
human_exprs_log2fc_sample = human_exprs_metadata %>% 
  filter(treatment == "Control") %>% 
  select(probe_id, treatment, value_raw_ref = value_raw) %>% 
  distinct() %>% 
  group_by(probe_id) %>% 
  summarise(value_raw_ref = mean(value_raw_ref)) %>% 
  ungroup() %>% 
  inner_join(human_exprs_metadata %>% 
               filter(treatment != "Control"),
             by = join_by("probe_id")) %>% 
  # Compute log2 fold-change relative to baseline
mutate(log2fc = value_raw - value_raw_ref) %>% 
  distinct() 

#Save
human_burn_log2fc_sample = human_exprs_log2fc_sample

#Check for duplicates
human_exprs_log2fc_sample %>% 
  filter(subject_id == "11710452",
         probe_id == "1007_s_at")

########### Filter and annotate probes -----
# Configure Ensembl BioMart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("hsapiens_gene_ensembl", mart = ensembl_mart)

# Get probe IDs from expression matrix
gene_id <- rownames(human_exprs) %>% unique()

# listAttributes(ensembl_mart) %>% View()
# Annotate data
annot <- getBM(
  attributes = c("affy_hg_u133_plus_2",
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "hgnc_symbol",
                 "description"),
  filters = "affy_hg_u133_plus_2", #Probes
  values = gene_id,
  mart = ensembl_mart) %>% #Standardize
  rename(probe = affy_hg_u133_plus_2,
                     entrezgene_id = entrezgene_id,
                     human_symbol = hgnc_symbol,
                     ensembl_id = ensembl_gene_id)

#Filter genes with highest median expression
human_log2summary = human_exprs %>% 
  rownames_to_column("probe") %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarise(median = median(expression, na.rm = T)) %>%  
  inner_join(annot, 
            by = "probe") %>% 
  group_by(ensembl_id) %>% 
  # Keep probe with highest median expression per gene
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_id) %>% 
  distinct() %>% 
  # Collapse to one row per Entrez ID
  group_by(entrezgene_id) %>% 
  # Keep probe with highest median expression per gene
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup()  


#Filter out genes with multiple hgnc_symbols
hgnc_symbol_dupes = human_log2summary %>% 
  get_dupes(human_symbol)

human_log2summary = human_log2summary %>% 
  filter(!human_symbol %in% c(hgnc_symbol_dupes$human_symbol %>% unique())) 

# Join expression with annotation, keep only mapped probes
human_log2_filtered_probes = human_exprs %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  inner_join(human_log2summary, by = "probe") %>%   #214,396
  select(human_symbol, probe,
         c(human_metadata$geo_sample))
#18196
human_log2_filtered_probes %>% dim()

human_log2_filtered_probes = human_log2_filtered_probes 

human_log2_filtered_probes %>% 
  get_dupes(human_symbol)

#Save 
human_log2_filtered_probes %>% 
  saveRDS(here("tables", "processed_data", "burn_human_exprs_hgnc_symbol.rds"))


#Annotate probes with log2fc
human_exprs_log2fc_sample = human_exprs_log2fc_sample %>% 
  rename(probe = probe_id) %>% 
  inner_join(human_log2_filtered_probes %>% 
               select(probe,
                      human_symbol),
             by = "probe")

human_exprs_log2fc_sample %>% 
saveRDS(here("tables", "Differential gene expression", "burn_human_log2fc_sample_clean_long.rds"))
```




















# ===================
# 6. Duchenne
## Mouse

Download dataset

Download the Duchenne mouse dataset (GSE1025) from GEO. Extract metadata from the ExpressionSet, save it as a CSV, then extract and save the raw expression matrix.

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
# Download the GEO ExpressionSet
filename = "GSE1025"
GSE1025 <- getGEO("GSE1025", GSEMatrix = TRUE) 
eset <- GSE1025[[1]]

# Clean and save metadata
duchenne_metadata = eset %>% 
  pData() %>% 
  dplyr::select( 
    geo_sample = "geo_accession", 
    subject_id = "title",
    tissue = "source_name_ch1",
    platform_id,
    participant_id = subject_id
    ) %>% 
        mutate(
          tissue = "Muscle",
          temp_col = str_to_lower(subject_id),
          treatment = case_when(
            str_detect(subject_id, "control|Control") ~ "Control",
            str_detect(subject_id, "mdx") ~ "DMD",
            T ~ NA
          )
        ) %>%
          extract(
            col = temp_col,
            into = c("timepoint", "group", "replicate"),
            regex = "leg (\\d+) day ([a-z]+) rep (\\d+)",
            convert = TRUE
          ) %>%
          mutate(
            treatment = fct_relevel(treatment, "Control"),
            condition = paste(treatment, timepoint, sep = "_")
          ) %>% 
  select(!(group))

#Save
saveRDS(duchenne_metadata, file = here("tables", "processed_data", "duchenne_mouse_metadata.rds"))

# Extract and save expression matrix
eset <- exprs(eset)

#Save
eset %>% 
  as.data.frame() %>% 
  saveRDS(here("tables", "processed_data", "duchenne_mouse_exprs.rds"))

```


Inspect the distribution of expression values for one representative sample (GSM3407136) to confirm the data are on the log2 scale.

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
# Load the expression matrix
duchenne_exprs = readRDS(here("tables", "processed_data", "duchenne_mouse_exprs.rds"))
duchenne_counts_not_normalized_log2 = readRDS(here("tables", "processed_data", "duchenne_mouse_exprs.rds"))

# Pivot to long format
duchenne_exprs_distribution = duchenne_exprs %>% 
  rownames_to_column("probes") %>% 
  pivot_longer(cols = -probes,
               names_to = "sample_id",
               values_to = "value") 

# Plot expression histogram for the first sample
duchenne_exprs_distribution %>% 
  mutate(value_log = log2(value)) %>% 
  filter(sample_id == "GSM15862") %>% 
  ggplot() +
  aes(x = value_log) +
  geom_histogram()

```


Import
Load the required data objects from disk.
```{r}
# Load metadata and expression
duchenne_metadata = readRDS(here("tables", "processed_data", "duchenne_mouse_metadata.rds"))
duchenne_exprs = readRDS(here("tables", "processed_data", "duchenne_mouse_exprs.rds")) %>% 
  as.matrix() %>%
  { log2(. + 1) } %>% 
  as.data.frame()

# Check probe count
duchenne_exprs %>% rownames() %>% length()
duchenne_exprs %>% dim()

# Subset expression matrix to selected samples
duchenne_exprs = duchenne_exprs %>% 
  as.data.frame() %>% 
  select(c(duchenne_metadata$geo_sample)) %>% 
  as.matrix()

#Inputs
filename = "duchenne_mouse"

#Annotate genes ------
mouse_exprs_metadata  = duchenne_exprs %>% 
  as.data.frame() %>% 
  rownames_to_column("probe_id") %>% 
  pivot_longer(cols = 2:37,
               names_to = "geo_sample",
               values_to = "value_raw") %>% 
  inner_join(duchenne_metadata, 
             by = "geo_sample") 

#Get value from day 0 and calculate Log2FC (non-paired)
mouse_exprs_log2fc_sample = mouse_exprs_metadata %>% 
  filter(treatment == "Control") %>% 
  select(probe_id, treatment, timepoint, subject_num = participant_id, value_raw_ref = value_raw) %>% 
  distinct() %>% 
  group_by(timepoint, treatment, probe_id) %>% 
  summarise(value_raw_ref = mean(value_raw_ref)) %>% 
  ungroup() %>% 
  inner_join(mouse_exprs_metadata %>% 
               filter(treatment != "Control") %>% 
               select(-treatment),
             by = join_by("probe_id", "timepoint")) %>% 
  # Compute log2 fold-change relative to baseline
 mutate(log2fc = value_raw - value_raw_ref) %>% 
  distinct() 

#Check duplicates
mouse_exprs_log2fc_sample %>% 
  filter(participant_id == "Leg 7 Day mdx rep 1",
         probe_id == "100001_at")


mouse_exprs_log2fc_sample

#Save
duchenne_exprs %>% 
  as.data.frame() %>% 
# Save R object to disk in native RDS format: preserves data types and structure for fast loading
  saveRDS(file = here("tables", "processed_data", "duchenne_mouse_exprs.rds"))
```

Annotate genes

Map Agilent probe IDs to human gene symbols using Ensembl BioMart and the FIT.mouse2man orthology lookup table.

```{r}
# Set up Ensembl BioMart for mouse
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("mmusculus_gene_ensembl", mart = ensembl_mart)

# Get probe IDs from the expression matrix
gene_id <- duchenne_exprs %>% rownames()

# List available attributes
listAttributes(ensembl_mart)

# Affymetrix Murine Genome U74A Version 2 Array

# Retrieve annotations for all probes
annot <- getBM(
  attributes = c("affy_mg_u74av2", #Agilent microarray data (design ID 028005, retrieved from GEO)
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "mgi_symbol",
                 "description"),
  filters = "affy_mg_u74av2",
  values = gene_id,
  mart = ensembl_mart)

# Load mouse-to-human orthology lookup table
library(FIT.mouse2man)

HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez %>% 
  clean_names() %>% 
  rename(mouse_entrez = mouse_ortholog)

# Summarise expression per probe, collapse to one row per gene
mouse_log2summary = duchenne_exprs %>% #20,845 probes
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
# Compute per-probe summary statistics
  summarise(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
  inner_join(annot %>% 
              rename(probe = affy_mg_u74av2), 
            by = "probe") %>%  #24.009 rows
  group_by(ensembl_gene_id) %>% 
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_gene_id) %>% 
# Remove duplicate rows
  distinct() %>% 
  
  # Collapse multiple Ensembl IDs per Entrez ID
  group_by(entrezgene_id) %>% 
  # Keep probe with highest median expression per gene
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup() %>% 
  rename(mouse_entrez = entrezgene_id) %>% 
  
  # Join human orthologs, collapse to one row per HGNC symbol
  inner_join(HS_MM_Symbol_Entrez, by = c("mouse_entrez", "mgi_symbol")) %>% 
  group_by(hgnc_symbol) %>% 
  # Keep probe with highest median expression per gene
  slice_max(median, n = 1)  %>% #14.396
  ungroup()

# Remove HGNC symbols that map to multiple MGI symbols
hgnc_symbol_dupes = mouse_log2summary %>% 
  get_dupes(hgnc_symbol)

mouse_log2summary = mouse_log2summary %>% 
  filter(!hgnc_symbol %in% c(hgnc_symbol_dupes$hgnc_symbol %>% unique())) 

# Build filtered expression matrix with human gene symbols
mouse_log2_filtered_probes = duchenne_exprs %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
# Join expression data with annotation by probe ID
  inner_join(mouse_log2summary, by = "probe") %>%   #214,396
  select(human_symbol = hgnc_symbol,
         probe_id = probe,
         human_entrez,
         mouse_entrez,
         c(duchenne_metadata$geo_sample))

#14395mouse_log2_filtered_probes
mouse_log2_filtered_probes %>% dim()

mouse_log2_hgnc_symbol = mouse_log2_filtered_probes 

# Identify duplicated gene symbols
mouse_log2_hgnc_symbol %>% get_dupes(human_symbol)

mouse_exprs_log2fc_sample = mouse_exprs_log2fc_sample %>% 
  inner_join(mouse_log2_hgnc_symbol %>% 
               select(human_symbol, probe_id),
             by = join_by(probe_id))

mouse_exprs_log2fc_sample %>% 
  saveRDS(here("tables", "Differential gene expression", "duchenne_mouse_log2fc_sample_clean_long.rds"))

#Save 
mouse_log2_hgnc_symbol %>% 
  saveRDS(here("tables", "processed_data", "duchenne_mouse_exprs_hgnc_symbol.rds"))
```







