---
title: "2_Comparing_Human_Mouse"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    embed-resources: true
    fig-cap-location: bottom
    theme: [cosmo, mouse2human-html.scss]
---

# Required

```{r}
#| message: false
#| warning: false
# install.packages("here") if needed
# install.packages("here")
library(here)
source(here("scripts_notebooks", "required.R"))
```

# Unified analyses

## 1. Differential gene expression analysis

### Pivot DEGs table

```{r}
#| fig-height: 20
#| fig-width: 20
#Import datasets ==============
#DEGs (matched)
all_degs_human_mouse =
  readRDS(here("tables", "Differential gene expression", "all_human_mouse_dge_limma_degs_matched.rds"))
#DEGs (unmatched)
all_degs_human_mouse_unmatched =
  readRDS(here("tables", "Differential gene expression", "all_human_mouse_dge_limma_degs.rds"))
#DEGs (matched with controls)
all_degs_human_mouse_matched_bionull =
  readRDS(file = here("tables", "Differential gene expression", "all_human_mouse_dge_limma_degs_matched_control.rds"))

#Check
all_degs_human_mouse %>% 
  dplyr::count(timepoint, organism, treatment, pathogen) 

#Functional modules
btm_annotation_genes = read_csv(file = here("tables", "Gene sets", "btm_annotation_genes.csv")) %>% 
  clean_names() %>% 
  filter(module_category == "immune")

# Different timepoints, same species

human_mouse_log2fc_avg_wide_all =

  #Condition 1
  all_degs_human_mouse_unmatched %>% 
  select(organism, pathogen:human_symbol, mean_l2fc:sd, sd_sampling_stability) %>% 
  rename(mean_log2fc = mean_l2fc, p_adj = adj_p_val, 
    tstatistic = t, bstatistic = b) %>% 
  rename_with(~ paste0(.x, "_1"), all_of(c("organism", "condition", "timepoint", "mean_log2fc", "ci_lower", "ci_upper", "ave_expr", "tstatistic", "p_value", "p_adj", "bstatistic", "se", "sd", "sd_sampling_stability"))) %>% 
  inner_join(
  #Condition 2
   all_degs_human_mouse_unmatched %>% 
  select(organism, pathogen:human_symbol, mean_l2fc:sd, sd_sampling_stability) %>% 
  rename(mean_log2fc = mean_l2fc, p_adj = adj_p_val, 
    tstatistic = t, bstatistic = b) %>% 
  rename_with(~ paste0(.x, "_2"), all_of(c("organism", "condition", "timepoint", "mean_log2fc", "ci_lower", "ci_upper", "ave_expr", "tstatistic", "p_value", "p_adj", "bstatistic", "se", "sd", "sd_sampling_stability"))),
  
  by = join_by("pathogen", "treatment", "human_symbol")
  ) %>% 
  
  
  # Filter for strict 1:1 ortholog pairs with complete expression profiles
  drop_na(mean_log2fc_1, mean_log2fc_2) %>% 
  
  # Compute expression divergence and volatility metrics
  mutate(
    # Cross-species effect size delta
    mean_log2fc_diff = mean_log2fc_1 - mean_log2fc_2,
    
    # Coefficient of Variation - CV
    cv_1 = sd_1 / abs(mean_log2fc_1),
    cv_2 = sd_2 / abs(mean_log2fc_2),
    cv_mean  = (cv_1 + cv_2) / 2, # Combined cross-species relative noise
    
    # Statistical Weights for Downstream Weighted Correlation
    weight_sd_inverse = 1 / (sd_1^2 + sd_2^2),
    weight_se_inverse = 1 / (se_1^2 + se_2^2),
    
    se_combined = sqrt(se_1^2 + se_2^2),
    weight_se_inverse_combined = 1/(se_combined^2)
  ) %>% 
  
  
  #Factor reordering
    mutate(pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "Staphylococcus aureus", "Escherichia coli", "Burn", "Trauma"),
       timepoint_comparison = paste0(timepoint_1,  sep = "-", timepoint_2)) %>% 
  # filter(!c(pathogen == "Staphylococcus aureus" & timepoint_mouse == "24h")) %>%
  mutate(timepoint_2 = as.character(timepoint_2),
         timepoint_2 = fct_relevel(timepoint_2, c("Early", "Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         timepoint_1 = as.character(timepoint_1),
         timepoint_1 = fct_relevel(timepoint_1, c("Early", "Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         treatment = case_when(pathogen %in% c("Hepatitis B", "Influenza") ~ "Vaccination",
                               pathogen %in% c( "Staphylococcus aureus", "Escherichia coli") ~ "Infection",
                               pathogen %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury")) %>% 
  arrange(timepoint_1) %>%
  #Annotate Inter and intra-species comparisons
  mutate(comparison_set = case_when(organism_1 != organism_2 ~ "Inter-species",
                                    organism_1 == organism_2 ~ "Intra-species",
                                    T ~ NA)) 

#### Create Biological null distribution
biological_null_degs = readRDS(file = here("tables", "Differential gene expression", "duchenne_mouse_dge_limma_degs.rds")) %>% 
  select(-comparison) %>% 
  rename(mean_log2fc = mean_l2fc, 
         p_adj = adj_p_val, 
         tstatistic = t, 
         bstatistic = b) %>% 
  rename_with(~ paste0(.x, "_2"), all_of(c( "pathogen", "organism", "treatment", "condition", "timepoint", "mean_log2fc", "ci_lower", "ci_upper", "ave_expr", "tstatistic", "p_value", "p_adj", "bstatistic", "se", "sd")))

# Create Biological null distribution =====


# 1. Get the target data using the arguments
target_df <- human_mouse_log2fc_avg_wide_all %>%
  filter(organism_1 == "Human", organism_2 == "Mouse") %>%
  mutate(diff_mean_log2fc = mean_log2fc_1 - mean_log2fc_2, model_type = "Mouse") %>% 
  rename(pathogen_1 = pathogen, treatment_1 = treatment) %>% 
  mutate(pathogen_2 = pathogen_1, treament_2 = treatment_1)

# 2. Get the biological null data and join
null_df <- biological_null_degs %>%
  filter(condition_2 == "DMD_Day28") %>%
  inner_join(target_df %>% select(!ends_with("_2")), by = "human_symbol", suffix = c("_2", "_1")) %>%
  mutate(diff_mean_log2fc = mean_log2fc_1 - mean_log2fc_2,
         model_type = "Control") %>% 
  # Compute expression divergence and volatility metrics
  mutate(
    # Cross-species effect size delta 
    mean_log2fc_diff = mean_log2fc_1 - mean_log2fc_2,
    
    # Coefficient of Variation - CV
    cv_1 = sd_1 / abs(mean_log2fc_1),
    cv_2 = sd_2 / abs(mean_log2fc_2),
    cv_mean  = (cv_1 + cv_2) / 2,
    
    # Statistical Weights for Downstream Weighted Correlation
    weight_sd_inverse = 1 / (sd_1^2 + sd_2^2),
    weight_se_inverse = 1 / (se_1^2 + se_2^2),
    
    se_combined = sqrt(se_1^2 + se_2^2),
    weight_se_inverse_combined = 1/(se_combined^2)
  ) 

# 3. Return combined rows
human_mouse_log2fc_avg_wide_interspecies_bionull = bind_rows(target_df,
                                                             null_df) %>% 
    #Rename 1 and 2 as human and mouse, respectively
     rename_with(
    ~ str_replace(.x, "_1$", "_human"), 
    ends_with("_1")
  ) %>% 
  rename_with(
    ~ str_replace(.x, "_2$", "_mouse"), 
    ends_with("_2")
  ) %>% 
  mutate(comparison = "Human vs. Mouse",
         timepoint_comparison = paste0(timepoint_human,  sep = "-", timepoint_mouse),
         timepoint_comparison = str_replace_all(timepoint_comparison, "\n", " "), 
         pathogen_comparison = paste0(pathogen_human, "-", pathogen_mouse)) %>% 
  
  #Standardize factors
     mutate(
       #Human
         timepoint_human = str_replace(timepoint_human,"Early hours", "Early\nhours"),
         timepoint_human = str_replace(timepoint_human,"Day ", "Day\n"),
         timepoint_human = fct_relevel(timepoint_human, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day\n1", "Day\n3", "Day\n7")),
         pathogen_human = case_when(pathogen_human == "Hepatitis B" ~ "HepB",
                              pathogen_human == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen_human == "Escherichia coli" ~ "E. coli",
                              T ~pathogen_human),
          pathogen_human = fct_relevel(pathogen_human, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment_human = case_when(pathogen_human %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen_human %in% c( "S. aureus", "E. coli") ~ "Infection",
                               pathogen_human %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen_human),
         treatment_human = fct_relevel(treatment_human, "Vaccination", "Infection", "Injury"),
         
      #Mouse
         timepoint_mouse = str_replace(timepoint_mouse,"Early hours", "Early\nhours"),
         timepoint_mouse = str_replace(timepoint_mouse,"Day ", "Day\n"),
         timepoint_mouse = fct_relevel(timepoint_mouse, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day\n1", "Day\n3", "Day\n7")),
         pathogen_mouse= case_when(pathogen_mouse == "Hepatitis B" ~ "HepB",
                              pathogen_mouse == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen_mouse == "Escherichia coli" ~ "E. coli",
                              T ~pathogen_mouse),
          pathogen_mouse = fct_relevel(pathogen_mouse, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment_mouse = case_when(pathogen_mouse %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen_mouse %in% c( "S. aureus", "E. coli") ~ "Infection",
                               pathogen_mouse %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen_mouse),
         treatment_mouse = fct_relevel(treatment_mouse, "Vaccination", "Infection", "Injury")) %>% 
  #Categorize in difference in log2fc and sign difference
    
   mutate(  
    sign_human = case_when(
      mean_log2fc_human > 0 ~ "Pos.",
      mean_log2fc_human < 0 ~ "Neg.",
      TRUE                  ~ "Zero"
    ),
    sign_mouse = case_when(
      mean_log2fc_mouse > 0 ~ "Pos.",
      mean_log2fc_mouse < 0 ~ "Neg.",
      TRUE                  ~ "Zero"
    ),
    diff_sign_binary =  if_else(sign_human == sign_mouse, "Same", "Different"),
    diff_sign        = stringr::str_c(sign_human, "_", sign_mouse))

#Check
human_mouse_log2fc_avg_wide_interspecies_bionull %>% 
  dplyr::count(pathogen_human, pathogen_mouse, timepoint_human, timepoint_mouse, comparison_set)

human_mouse_log2fc_avg_wide_interspecies_bionull %>% 
  select(pathogen_human, pathogen_mouse, timepoint_comparison, model_type, human_symbol, mean_log2fc_diff) %>% 
  distinct() %>% 
  filter(human_symbol == "IL7R") %>% 
  arrange(pathogen_human, pathogen_mouse)

#Get immune and non immune genes ==========
human_mouse_log2fc_avg_wide_degs_all_immune_nonimmune = 
  
  bind_rows(
    
  #All genes
  human_mouse_log2fc_avg_wide_interspecies_bionull %>%
  mutate(gene_set = "All genes"),
    
  #Same DEGs
  human_mouse_log2fc_avg_wide_interspecies_bionull %>%
  filter(p_adj_human <= 0.05 & p_adj_mouse <= 0.05) %>% 
  mutate(gene_set = "DEGs"), 
  
  #Classify as immune or non_immune 
  human_mouse_log2fc_avg_wide_interspecies_bionull %>% 
  mutate(gene_set = if_else(human_symbol %in% c(btm_annotation_genes  %>% 
                               pull(module_member_genes)),
                             "Immune",
                             "Not-immune")) %>% 
  filter(p_adj_human <= 0.05 & p_adj_mouse <= 0.05)
  
  ) 


human_mouse_log2fc_avg_wide_degs_all_immune_nonimmune %>% 
  dplyr::count(pathogen_human, pathogen_mouse, comparison_set, model_type)

# #Save
human_mouse_log2fc_avg_wide_degs_all_immune_nonimmune %>%
  saveRDS(here("tables", "Differential gene expression", "human_mouse_log2fc_avg_wide_degs_all_immune_nonimmune.rds"))

```

### Distributions

#### Log2FC differences

```{r}
#| fig-height: 3
#| fig-width: 12
#| message: false
#| warning: false
#Import
human_mouse_log2fc_avg_wide_degs_all_immune_nonimmune = 
  readRDS(here("tables", "Differential gene expression", "human_mouse_log2fc_avg_wide_degs_all_immune_nonimmune.rds"))

#### Difference in means vs. status
log2fc_wide_allconditions_control =  
  
  bind_rows(
  human_mouse_log2fc_avg_wide_degs_all_immune_nonimmune %>%
               filter(gene_set == "All genes") %>% 
               rename(pathogen = pathogen_human,
                      treatment = treatment_human, 
                      gene = human_symbol) %>% 
    filter(model_type == "Mouse") %>% 
    filter((pathogen == "E. coli" & timepoint_comparison == "24h-12h" |
            pathogen == "S. aureus" & timepoint_comparison == "24h-12h" |
            pathogen == "Burn" & timepoint_comparison == "Early hours-Day 7"|
            pathogen == "Trauma" & timepoint_comparison == "Day 7-Day 3"|
            pathogen == "Influenza" & timepoint_comparison == "Day 1-Day 7" |
            pathogen == "HepB" & timepoint_comparison == "Day 7-Day 7")),
  human_mouse_log2fc_avg_wide_degs_all_immune_nonimmune %>%
               filter(gene_set == "All genes") %>% 
               rename(pathogen = pathogen_human,
                      treatment = treatment_human, 
                      gene = human_symbol) %>% 
    filter(model_type == "Control")
  )
  

# Calculate statistical significance tests per condition
stat_results <- log2fc_wide_allconditions_control %>%
  # Filter significant human DEGs (use drop_na to ensure valid filtering)
  filter(!is.na(p_adj_human) & p_adj_human <= 0.05) %>%
  mutate(abs_diff = abs(mean_log2fc_diff)) %>%
  group_by(treatment, pathogen) %>%
  summarise(
    # Match exact factor levels present in your model_type column
    n_target      = sum(model_type == "Mouse" & !is.na(abs_diff)),
    n_control     = sum(model_type == "Control" & !is.na(abs_diff)),
    
    median_target = if (n_target > 0) median(abs_diff[model_type == "Mouse"], na.rm = TRUE) else NA_real_,
    median_null   = if (n_control > 0) median(abs_diff[model_type == "Control"], na.rm = TRUE) else NA_real_,
    
    # Run Wilcoxon test safely
    wilcox_p      = if (n_target >= 1 && n_control >= 1) {
      wilcox.test(
        abs_diff[model_type == "Mouse"],
        abs_diff[model_type == "Control"],
        alternative = "less"
      )$p.value
    } else {
      NA_real_
    },
    
    .groups = "drop"
  ) %>%
  mutate(
    fdr_wilcox = p.adjust(wilcox_p, method = "BH"),
    fdr_wilcox_sci = sprintf("%.2e", fdr_wilcox)
  )

# Inspect statistical summary table
print(stat_results)

#Plot
log2fc_distribution_control_condition_plot = log2fc_wide_allconditions_control %>%
  filter(p_adj_human <= 0.05) %>% 
  inner_join(stat_results,
             join_by(pathogen, treatment)) %>% 
  ggplot() +
  aes(
    x     = abs(mean_log2fc_diff),
    color = model_type,
    fill = model_type
  ) +
  facet_nested_wrap(
    ~ treatment + pathogen + paste0("padj = ", fdr_wilcox_sci),
    nrow   = 1,
  ) +
  geom_density(
    linewidth = 1,
    data = . %>% filter(model_type == "Mouse")
  ) +
  geom_density(
    linewidth = 1,
    data = . %>% filter(model_type == "Control")
  ) +
  geom_boxplot(aes(y = -0.1),
               width = 0.2, 
               outliers = F,
               color = "black") +
  scale_x_continuous(limits = c(0, 2), expand = c(0, 0)) +
  scale_y_continuous(expand = c(0.05, 0.05)) +
  scale_color_manual(
    values = c(
      "Control" = "black",
      "Mouse" = "#4DBBD5FF"
    )
  ) +
  scale_fill_manual(
    values = c(
      "Control" = NA,
      "Mouse" = "#4DBBD5FF"
    )
  ) +
  theme_vaxgo() +
  theme(legend.position = "top",
        panel.spacing = grid::unit(0.7, "cm"),
        plot.margin          = ggplot2::margin(0.1, 0.5, 0.1, 0.1, "cm"),
        axis.line.y = element_blank(),
        panel.grid.major.y = element_line(linewidth = 0.1, color = "black"))  +
  labs(x = "|ΔLog2FC|",
       y = "Density",
       fill = "Model",
       color = "Model")

#Print
log2fc_distribution_control_condition_plot 

#Save
log2fc_distribution_control_condition_plot %>%
  ggsave(filename = here("Figures", "all_human_mouse_dge_degs_log2fcdiff_distribution.png"),
         width = 12,
         height = 3)


```

```{r}
#| fig-height: 3
#| fig-width: 12
#Plot as boxplot + violin
log2fc_distribution_control_condition_boxplot = log2fc_wide_allconditions_control %>%
  filter(p_adj_human <= 0.05) %>% 
  inner_join(stat_results,
             join_by(pathogen, treatment)) %>% 
  ggplot() +
  aes(
    y     = abs(mean_log2fc_diff),
    # y     = mean_log2fc_diff,
    x = model_type,
    color = model_type,
    fill = model_type
  ) +
  # Create nested wrap facets across treatment and pathogen strata
  facet_nested_wrap(
    ~ treatment + pathogen + paste0("padj = ", fdr_wilcox_sci),
    nrow   = 1,
    
  ) +
  # Render density curves with linewidth formatting
  geom_violin(width = 1) +
  geom_boxplot(fill = NA,
               outliers = F, 
               color = "black", 
               width = 0.1) +
  scale_y_continuous(expand = c(0.01, 0,01),
                     limits = c(0, 2)
                     ) +
  # Set manual color palette for model types
  scale_fill_manual(
    values = c(
      "Control" = NA,
      "Mouse" = "#4DBBD5FF"
    )
  ) +
  scale_color_manual(
    values = c(
      "Control" = "black",
      "Mouse" = "#4DBBD5FF"
    )
  ) +
  theme_vaxgo() +
  theme(legend.position = "none") +
  labs(x = "Model",
       y = "|ΔLog2FC|")

log2fc_distribution_control_condition_boxplot
```

#### Log2FC diff vs. 1/SE

```{r}
#| fig-height: 3
#| fig-width: 4.5
#| message: false
#| warning: false
# Difference in means vs. status
log2fc_wide_allconditions = human_mouse_log2fc_avg_wide_degs_all_immune_nonimmune %>%
               filter(gene_set == "All genes") %>%
               rename(pathogen = pathogen_human,
                      treatment = treatment_human,
                      gene = human_symbol) %>%
    mutate(diff_sign_binary = if_else(sign(mean_log2fc_human) == sign(mean_log2fc_mouse), "Same", "Different"),
         diff_sign = paste0(sign(mean_log2fc_human), sign(mean_log2fc_mouse)),
         is_deg = case_when(p_adj_human <= 0.05 & p_adj_mouse <= 0.05 ~ "DEG, both",
                            p_adj_human <= 0.05 & p_adj_mouse > 0.05 ~ "DEG, human",
                            p_adj_human > 0.05 & p_adj_mouse <= 0.05 ~ "DEG, mouse",
                            p_adj_human > 0.05 & p_adj_mouse > 0.05 ~ "not DEG, both",
                            TRUE ~ NA))

# Categorize in difference in log2fc and sign difference
log2fc_wide_effect_error_plot_df = log2fc_wide_allconditions %>%
    filter((pathogen == "E. coli" & timepoint_comparison == "24h-12h" |
            pathogen == "S. aureus" & timepoint_comparison == "24h-12h" |
            pathogen == "Burn" & timepoint_comparison == "Early hours-Day 7" |
            pathogen == "Trauma" & timepoint_comparison == "Day 7-Day 3" |
            pathogen == "Influenza" & timepoint_comparison == "Day 1-Day 7" |
            pathogen == "HepB" & timepoint_comparison == "Day 7-Day 7"))

# Combine original data with Pooled summary group
log2fc_wide_effect_error_plot_df_pooled <- log2fc_wide_effect_error_plot_df %>%
  bind_rows(
    log2fc_wide_effect_error_plot_df %>%
      mutate(treatment = "Pooled", pathogen = "")
  ) %>%
  mutate(treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury", "Pooled"))

## Filter
df_base <- log2fc_wide_effect_error_plot_df_pooled %>% 
  select(treatment, pathogen, timepoint_comparison, is_deg, diff_sign_binary, mean_log2fc_diff, se_mouse, weight_se_inverse_combined) %>%
  distinct() %>% 
  filter(treatment == "Pooled")

# 2. Compute statistics including the overall "All genes" group
cor_stats <- bind_rows(
    df_base,
    df_base %>% mutate(is_deg = "All genes")
  ) %>% 
  distinct() %>% 
  mutate(x_val = 1 / se_mouse, y_val = abs(mean_log2fc_diff)) %>%
  group_by(treatment, pathogen, is_deg) %>% 
  cor_test(vars = "x_val", vars2 = "y_val", method = "spearman") %>%
  mutate(
    p_format = case_when(
      p < 2.2e-16 ~ "p < 2.2e-16",
      TRUE        ~ paste0("p = ", sprintf("%.2e", p))
    ),
    label_text = paste0("ρ=", round(cor, 2), ", ", p_format),
    # Reorder so "All genes" appears first at the top of the plot stack
    is_deg = fct_relevel(as.factor(is_deg), "All genes") 
  )



# 2. Plot the figure and add the correlation labels with geom_label
log2fc_wide_effect_error_plot <- df_base %>% 
  bind_rows(df_base %>%  mutate(is_deg = "All genes")) %>% 
  distinct() %>% 
  ggplot(aes(
    x = 1 / se_mouse,
    y = abs(mean_log2fc_diff),
    color = is_deg,
    fill = is_deg
  )) +

  facet_nested_wrap(~ treatment + pathogen, nrow = 1, scales = "free_x") +
  # Group-specific Linear Regressions
  geom_smooth(
    method = "lm",
    mapping = aes(color = is_deg, fill = is_deg),
    se = TRUE
  ) +
   geom_text(
    data = cor_stats,
    aes(
      x = 1,              
      y = 0.2,          
      label = label_text,
      color = is_deg
    ),
    hjust = 0,
    vjust = 1.2 + (as.numeric(as.factor(cor_stats$is_deg)) * 1.5), 
    # color = "black",
    size = 3,
    inherit.aes = FALSE     
  )+

  # Scales & Formatting
  scale_x_continuous(expand = c(0, 0), limits = c(0, NA)) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_manual(values = colors$degs_comparison_line) +
  scale_color_manual(values = colors$degs_comparison_line) +
  labs(
    x = "1/SE (mouse)",
    y = "|ΔLog2FC|",
    fill = ""
  ) +
  theme_vaxgo() +
  theme(
    legend.position = "right",
    panel.spacing = grid::unit(0.7, "cm")
  ) +
  guides(
    color = "none",
    fill = guide_legend(
      title.position = "top",
      # byrow = TRUE,
      # nrow = 1,
      override.aes = list(shape = 22, alpha = 1, size = 3, linetype = 0)
    )
  )

log2fc_wide_effect_error_plot 

#Save
log2fc_wide_effect_error_plot %>%
  ggsave(filename = here("Figures", "all_human_mouse_dge_degs_se_log2fc_scatter.png"),
         width = 4,
         height = 2.5)
  
```

Boxplots

```{r}
#| fig-height: 3
#| fig-width: 12
#Plot as boxplot + violin
# 1. Compute BH-adjusted p-values per facet/group
stat_test <- log2fc_wide_effect_error_plot_df_pooled %>%
  mutate(inverse_se_mouse = 1/se_mouse) %>% 
  group_by(treatment, pathogen) %>%
  rstatix::wilcox_test(
    inverse_se_mouse ~ is_deg, 
    ref.group = "DEG, both", 
    p.adjust.method = "BH"
  ) %>%
  rstatix::add_y_position(fun = "max", step.increase = 0.1)

# 2. Plot the figure using the test results
error_plot_df_boxplot = log2fc_wide_effect_error_plot_df_pooled %>% 
  mutate(inverse_se_mouse = 1/se_mouse) %>% 
  ggplot() +
  aes(
    y     = inverse_se_mouse,
    x     = is_deg,
    color = is_deg,
    fill  = is_deg
  ) +
  facet_nested_wrap(
    ~ treatment + pathogen,
    nrow = 1
  ) +
  geom_violin(width = 1, color = "black", linewidth = 0.5) +
  geom_boxplot(fill = NA, outliers = FALSE, width = 0.1, show.legend = F) +
  
  # Add p-values positioned correctly over each bar/group
  stat_pvalue_manual(
    stat_test,
    label           = "p.adj", 
    tip.length      = 0.01,
    hide.ns         = FALSE,
    size = 3
  ) +
  
  scale_fill_manual(values = colors$degs_comparison) +
  scale_color_manual(values = colors$degs_comparison_boxplot) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.2))) + 
  
  theme_vaxgo() +
  theme(legend.position = "top", axis.text.x = element_blank()) +
  labs(
    x = "Filter",
    y = "1/SE, Mouse"
  )

#Print
error_plot_df_boxplot

#Save
error_plot_df_boxplot %>%
  ggsave(filename = here("Figures", "all_human_mouse_dge_degs_se_condition.png"),
         width = 12,
         height = 2.5)


```

#### Ranks

```{r}
#| fig-height: 4
#| fig-width: 12
#### Ranks
log2fc_wide_allconditions %>% 
  filter(
    (pathogen == "E. coli" & timepoint_comparison == "24h-12h") |
    (pathogen == "S. aureus" & timepoint_comparison == "24h-12h") |
    (pathogen == "Burn" & timepoint_comparison == "Early hours-Day 7") |
    (pathogen == "Trauma" & timepoint_comparison == "Day 7-Day 3") |
    (pathogen == "Influenza" & timepoint_comparison == "Day 1-Day 7") |
    (pathogen == "HepB" & timepoint_comparison == "Day 7-Day 7")
  ) %>%
   group_by(treatment, pathogen, timepoint_comparison) %>% 
  mutate( rank_log2fc_mouse  = percent_rank(mean_log2fc_mouse * -log10(pmax(p_adj_mouse, 1e-300))) * 100,
          rank_log2fc_human  = percent_rank(mean_log2fc_human * -log10(pmax(p_adj_human, 1e-300))) * 100,
          abs_rank_log2fc_mouse  = percent_rank(abs(mean_log2fc_mouse) * -log10(pmax(p_adj_mouse, 1e-300))) * 100,
          abs_rank_log2fc_human  = percent_rank(abs(mean_log2fc_human) * -log10(pmax(p_adj_human, 1e-300))) * 100) %>% 
  ungroup() %>% 
  ggplot() +
  aes(x = rank_log2fc_mouse, 
         y = rank_log2fc_human,
      # color = abs_log2fc_diff
      )+
  geom_point(alpha = 0.01, stroke = 0) +
  geom_point(alpha = 1,
             data = . %>% filter(p_adj_human < 0.05 | p_adj_mouse < 0.05),
             color = "blue") +
  facet_grid(~treatment + pathogen) +
  theme(legend.position = "top") +
  scale_colour_gradient(
    low = "blue", 
    high = "white"
  ) +
  ggpubr::stat_cor(method = "spearman", fill = "white", color = "black", label.y = 115) +
  ggpubr::stat_cor(method = "spearman", fill = "white", color = "black", label.y = 105, 
                    data = . %>% filter(p_adj_human < 0.05| p_adj_mouse < 0.05)) +
  theme_vaxgo()
  
  
```

#### Log2FC vs Log2FC

```{r}
#| fig-height: 4
#| fig-width: 12
#### Log2FC vs Log2FC
log2fc_wide_allconditions %>% 
  filter(
    (pathogen == "E. coli" & timepoint_comparison == "24h-12h") |
    (pathogen == "S. aureus" & timepoint_comparison == "24h-12h") |
    (pathogen == "Burn" & timepoint_comparison == "Early hours-Day 7") |
    (pathogen == "Trauma" & timepoint_comparison == "Day 7-Day 3") 
  ) %>%
   group_by(treatment, pathogen, timepoint_comparison) %>% 
  mutate( rank_log2fc_mouse  = percent_rank(mean_log2fc_mouse * -log10(pmax(p_adj_mouse, 1e-300))) * 100,
          rank_log2fc_human  = percent_rank(mean_log2fc_human * -log10(pmax(p_adj_human, 1e-300))) * 100,
          abs_rank_log2fc_mouse  = percent_rank(abs(mean_log2fc_mouse) * -log10(pmax(p_adj_mouse, 1e-300))) * 100,
          abs_rank_log2fc_human  = percent_rank(abs(mean_log2fc_human) * -log10(pmax(p_adj_human, 1e-300))) * 100) %>% 
  ungroup() %>%
  ggplot() +
  aes(x = mean_log2fc_mouse, 
         y = mean_log2fc_human,
      # color = abs_log2fc_diff
      )+
  geom_point(alpha = 0.5,  color = "black") +
  geom_point(alpha = 0.8,
             data = . %>% filter(p_adj_human < 0.05 , p_adj_mouse < 0.05),
             color = "blue") +
  facet_wrap(~treatment + pathogen, nrow = 1) +
  theme(legend.position = "top") +
  ggpubr::stat_cor(method = "pearson", fill = "white", color = "black", 
                   label.y = 6
                   ) +
  ggpubr::stat_cor(method = "spearman", fill = "white", color = "black", 
                   label.y = 5.5
                   ) +
  theme_vaxgo() +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0)+
  geom_abline()
  
  
```

### Volcano plot

```{r}
#| fig-height: 16
#| fig-width: 16
#Import
all_degs_human_mouse = readRDS(file = here("tables", "Differential gene expression", "all_human_mouse_dge_limma_degs.rds"))

#Volcano plot -------
volcano_data <- all_degs_human_mouse %>% 
   filter((pathogen == "E. coli" & timepoint_comparison == "24h-12h" |
            pathogen == "S. aureus" & timepoint_comparison == "24h-12h" |
            pathogen == "Burn" & timepoint_comparison == "Early hours-Day 7"|
            pathogen == "Trauma" & timepoint_comparison == "Day 7-Day 3"|
            pathogen == "Influenza" & timepoint_comparison == "Day 1-Day 7" |
            pathogen == "HepB" & timepoint_comparison == "Day 7-Day 7")) %>% 
  mutate(logp = -log10(adj_p_val),
         sig = case_when(mean_l2fc > 0 & logp > 1.3 ~ "Up",
                         mean_l2fc < 0 & logp > 1.3 ~ "Down",
                         TRUE ~ "NS"))

volcano_plot = volcano_data %>% 
  mutate(pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "Staphylococcus aureus", "Escherichia coli", "Burn", "Trauma")) %>% 
ggplot() +
  aes(x = mean_l2fc, y = logp, fill = sig) +
  facet_nested_wrap(~pathogen+timepoint+organism, 
               scales = "free", 
               axes = "all",
               nest_line = element_line(linetype = 1, color = "black")) +
  geom_point(aes(fill = sig,
                 color = sig),
             pch = 21, 
             size = 1, 
             alpha = 1) +
  geom_vline(xintercept = 0, color = "black", linetype = 1, size = 0.5) +
  geom_hline(yintercept = 1.3, color = "black", linetype = 2, size = 0.5) +
  scale_y_continuous(n.breaks = 5,
                     # limits = c(0, 70),
                     expand = c(0.01, 0.01)) +
  guides(color = F) +
  scale_color_manual(
    values = c(
      "Up" = "#165289", 
      "Down" = "#45efc6",
      "NS" = "gray80")) +
  scale_fill_manual(
    values = c(
      "Up" = "#165289", 
      "Down" = "#45efc6",
      "NS" = "gray80")) +
  labs(x = "Log2-Fold Change",
       y = "-log(FDR)",
       fill = "Direction") +
  theme_vaxgo() +
  theme(legend.position = "top",
        legend.direction = "horizontal",
        legend.title = element_blank(),
        legend.key = element_blank(),
        strip.text = element_text(size = 10),
        axis.text = element_text(size = 10),
        axis.text.x = element_text(angle = 0))  +
  guides(fill = guide_legend(override.aes = list(size = 4)),
         color = "none")

#Save
volcano_plot %>%
  ggsave(filename = here("Figures", "all_human_mouse_qc_volcano_plot_raw.png"),
         width = 16,
         height = 16)

#Print
volcano_plot
```

### Overlapping DEGs

```{r}
#Import datasets
all_degs_human_mouse = readRDS(here("tables", "Differential gene expression", "all_human_mouse_dge_limma_degs_matched.rds"))

all_degs_human_mouse_unmatched = readRDS(here("tables", "Differential gene expression", "all_human_mouse_dge_limma_degs.rds"))

btm_annotation_genes = read_csv(file = here("tables", "Gene sets", "btm_annotation_genes.csv"))
```

Unmatched timepoints

```{r}
# 1. Data Preparation (single cleanup step)

degs_summary <- all_degs_human_mouse_unmatched %>%
  filter(adj_p_val <= 0.05) %>%
    mutate(pathogen = recode(pathogen, 
                           "Staphylococcus aureus" = "S. aureus",
                           "Escherichia coli" = "E. coli",
                           "Hepatitits B" = "HepB")) %>%
  select(human_symbol, pathogen, timepoint, organism) %>%
  distinct()

all_degs_human_mouse_unmatched_ndegs = degs_summary %>%
  group_by(pathogen, organism,  timepoint) %>% 
  summarise(n_genes = n())

# 2. Identify overlap 
# Create a table that indicates if a gene is a DEG in each organism
degs_human_mouse_overlap_genes_summary = degs_summary %>%
  filter(organism == "Human") %>%
  full_join(
    degs_summary %>% 
      filter(organism == "Mouse"),
    by = c("human_symbol", "pathogen"),
    suffix = c("_human", "_mouse")
  ) %>%
  mutate(
    status = case_when(
      !is.na(organism_human) & !is.na(organism_mouse) ~ "Shared",
      !is.na(organism_human) & is.na(organism_mouse)  ~ "Human\nonly",
      is.na(organism_human) & !is.na(organism_mouse)  ~ "Mouse\nonly"
    )
  ) %>% 
  group_by(pathogen, timepoint_human, timepoint_mouse, status) %>%
  summarise(n_genes = n(), .groups = "drop") %>%
    mutate(timepoint_mouse = if_else(is.na(timepoint_mouse), "Not\nShared", timepoint_mouse),
         timepoint_human = if_else(is.na(timepoint_human), "Not\nShared", timepoint_human))
  
#Calculate background number of genes
background_n <- all_degs_human_mouse_unmatched %>%
  select(human_symbol, organism, pathogen) %>%
  distinct() %>%
  group_by(human_symbol) %>%
  filter(n() == 2) %>% 
  nrow()


#Unite human
  
# 1. Build the totals table in a single pass
total_genes <- all_degs_human_mouse_unmatched_ndegs %>%
  mutate(organism = str_to_lower(organism)) %>%
  pivot_wider(
    names_from = organism, 
    values_from = n_genes, 
    names_prefix = "n_genes_total_"
  ) %>%
  # Fill the NAs so the totals are available
  group_by(pathogen) %>%
  fill(n_genes_total_human, n_genes_total_mouse, .direction = "downup") %>%
  ungroup()

# 2. Build the overlap table
shared_genes <- degs_human_mouse_overlap_genes_summary %>%
  filter(status == "Shared") %>%
  rename(n_genes_shared = n_genes) %>% 
  select(-status) %>% 
  distinct()

# 3. Join unificado
all_degs_human_mouse_unmatched_shared <- shared_genes %>%
  left_join(total_genes %>% 
              select(pathogen, timepoint, n_genes_total_human), by = c("pathogen", "timepoint_human" = "timepoint")) %>%
  left_join(total_genes %>% select(pathogen, timepoint, n_genes_total_mouse), by = c("pathogen", "timepoint_mouse" = "timepoint")) %>%
  # Final cleanup of the redundant columns produced by the join
  select(pathogen, timepoint_human, timepoint_mouse, 
         n_genes_shared, 
         n_genes_total_human, 
         n_genes_total_mouse) %>% 
  mutate(pct_shared_mouse = ((n_genes_shared/n_genes_total_mouse) *100) %>% round(0),
         pct_shared_human = ((n_genes_shared/n_genes_total_human) *100) %>% round(0)) %>% 
  #Factor cleaning
  mutate(pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", 
                                  "S. aureus", "E. coli", 
                                  "Burn", "Trauma"),
         treatment = case_when(pathogen %in% c("Hepatitis B", "Influenza") ~ "Vaccination",
                               pathogen %in% c( "S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury"), 
       
       #Break timepoint spaces
         timepoint_comparison = paste0(timepoint_mouse,  sep = "-", timepoint_human),
         timepoint_mouse = as.character(timepoint_mouse),
         timepoint_human = as.character(timepoint_human),
         timepoint_human = str_replace(timepoint_human,"Early hours", "Early\nhours"),
         timepoint_human = str_replace(timepoint_human,"Day ", "Day\n"),
         timepoint_human = fct_relevel(timepoint_human, c("Human\nonly", "Early\nhours", 
                                                          "2h", "4h", "6h", "12h", "24h",
                                                          "Day\n1", "Day\n3", "Day\n7")),
#Mouse
         timepoint_mouse = str_replace(timepoint_mouse,"Early hours", "Early\nhours"),
         timepoint_mouse = str_replace(timepoint_mouse,"Day ", "Day\n"),
         timepoint_mouse = fct_relevel(timepoint_mouse, c("Mouse\nonly", "Early\nhours", 
                                                          "2h", "4h", "6h", "12h", "24h",
                                                          "Day\n1", "Day\n3", "Day\n7"))
         ) %>% 
  arrange(timepoint_mouse)

```

Heatmap

```{r}
#| fig-height: 2.5
#| fig-width: 10
all_degs_human_mouse_unmatched_shared_heatmap = all_degs_human_mouse_unmatched_shared %>% 
  ggplot() +
  aes(x = timepoint_mouse,
      y = timepoint_human) +
  facet_nested_wrap(~treatment+pathogen,
                    scales = "free", 
                    nrow = 1, 
                    axes = "all",
                    remove_labels = "none",
                    strip = strip_nested(size = "variable")) +
  labs(x = "Mouse",
       y = "Human") +
  geom_point(aes(fill = pct_shared_human,
                 size = n_genes_shared),
            color = "black",
            pch = 21) +
  scale_size_continuous(range = c(2, 10), name = "Total") +
  scale_fill_gradientn(
        colors = c("white", "#4DBBD5FF","#4361ee"),
        values = scales::rescale(seq(0, 100, length.out = 10)),
        # limits = c(-1, 1),
        name = "%DEGs shared with Human",
        guide = guide_colorbar(
          barwidth = 15,
          barheight = 0.5,
          nbin = 10,
          raster = FALSE
        )
      ) +
  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =8),
        axis.text.y = element_text(size = 8),
        strip.text.x = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(0,0,-10,0),  
        legend.title = element_text(size = 10),
        panel.spacing.y = unit(0, "lines"),
        panel.spacing.x = unit(0.5, "lines"),
        plot.margin = margin(0, 0, 0, 0, unit = "cm"),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) +
  geom_text(aes(label = paste0(pct_shared_human, "%")),
            size = 2.5,
            vjust = 3, color = "black",
            data = . %>% filter(treatment == "Vaccination")) +
  geom_text(aes(label = paste0(pct_shared_human, "%")),
            size = 2.5, color = "black", fontface = "bold",
            data = . %>% filter(treatment != "Vaccination",
                                pct_shared_human < 20)) +
  geom_text(aes(label = paste0(pct_shared_human, "%")),
            size = 2.5, color = "white", fontface = "bold",
            data = . %>% filter(treatment != "Vaccination",
                                pct_shared_human >= 20))
#Display
all_degs_human_mouse_unmatched_shared_heatmap

#Save
all_degs_human_mouse_unmatched_shared_heatmap %>% 
  ggsave(filename = here("Figures", "all_human_mouse_dge_degs_heatmap_horizontal.png"),
         width = 10, height = 2.5)

```

### Lineplot

```{r}
#| fig-height: 2
#| fig-width: 10
all_degs_human_mouse_unmatched_ndegs = all_degs_human_mouse_unmatched %>%
  filter(adj_p_val <= 0.05) %>%
    mutate(pathogen = recode(pathogen, 
                           "Staphylococcus aureus" = "S. aureus",
                           "Escherichia coli" = "E. coli",
                           "Hepatitits B" = "HepB")) %>%
    group_by(pathogen, organism,  timepoint) %>% 
   summarise(
    n_genes           = sum(!is.na(mean_l2fc)),
    mean_log2fc_genes = mean(mean_l2fc, na.rm = TRUE),
    mean_se           = mean(se,   na.rm = TRUE),   
    median_se         = median(se, na.rm = TRUE), 
    all_inverse_se    = 1 / median_se,
    .groups = "drop"
  ) %>% 
   mutate(pathogen = fct_relevel(pathogen, "Influenza", "HepB", 
                                  "S. aureus", "E. coli", 
                                  "Burn", "Trauma"),
         treatment = case_when(pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen %in% c( "S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury"), 
         timepoint = as.character(timepoint),
         timepoint = str_replace(timepoint,"Early hours", "Early\nhours"),
         timepoint = str_replace(timepoint,"Day ", "Day\n"),
         timepoint = fct_relevel(timepoint, c("Early\nhours", 
                                                          "2h", "4h", "6h", "12h", "24h",
                                                          "Day\n1", "Day\n3", "Day\n7")))

all_degs_human_mouse_unmatched_ndegs 
```

```{r}
#| fig-height: 2
#| fig-width: 10
##### Total DEGs
all_degs_human_mouse_unmatched_ndegs_plot = 
  all_degs_human_mouse_unmatched_ndegs %>% 
  ggplot() +
  
  aes(x = timepoint,
      y = n_genes,
      fill = organism,
      color = organism,
      group = organism) + 
  
  facet_nested_wrap(~treatment+pathogen, 
                    axes = "all",
                    nrow = 1, 
                    scales = "free_x", 
                    # remove_labels = "all",
               nest_line = element_line(linetype = 1, color = "black")) +
  geom_line(width = 1) +
  
  geom_point(
             size = 2,
             pch = 21,
             color = "black",
             stroke = 0.5) + 
  geom_text_repel(aes(label = n_genes), 
                  size = 2.5, 
                  # direction = "y", 
                  segment.alpha = 1) +

  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_continuous(expand = c(0.05, 0.05),
                     labels = label_number(scale_cut = cut_short_scale())) +
  theme_vaxgo() +
  scale_fill_manual(values = colors$organism)    +
  scale_color_manual(values = colors$organism)    +
  labs(x = "Time",
       y = "Total",
       fill = "",
       color = "") +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
          axis.text.y = element_text(size = 8),
          axis.title.y = element_text(size = 10),
          axis.title.x = element_text(size = 10),
          strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
          strip.text = element_text(size = 10, margin = margin(1, 0, 1, 0)),
          panel.spacing = unit(1.1, "lines"),
          plot.margin = margin(0, 0.5, 0, 0, unit = "cm"),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
          legend.text = element_text(size = 10, hjust = 0),
          legend.title = element_text(size = 10),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)
    ) +
  guides(color = "none")

#Display
all_degs_human_mouse_unmatched_ndegs_plot 


#Inverse SE
all_degs_human_mouse_unmatched_inverseSE_plot = 
  all_degs_human_mouse_unmatched_ndegs %>% 
  ggplot() +
  aes(x = timepoint,
      y = all_inverse_se,
      fill = organism,
      color = organism,
      group = organism) + 
  
  facet_nested_wrap(~treatment+pathogen, 
                    axes = "all",
                    nrow = 1, 
                    scales = "free", 
                    # remove_labels = "all",
               nest_line = element_line(linetype = 1, color = "black")) +
  geom_line(width = 1) +
  
  geom_point(
             size = 2,
             pch = 21,
             color = "black",
             stroke = 0.5) + 
  geom_text_repel(aes(label = n_genes), 
                  size = 2.5, 
                  # direction = "y", 
                  segment.alpha = 1) +

  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_continuous(expand = c(0.05, 0.05),
                     labels = label_number(scale_cut = cut_short_scale())) +
  theme_vaxgo() +
  scale_fill_manual(values = colors$organism)    +
  scale_color_manual(values = colors$organism)    +
  labs(x = "Time",
       y = "1/medianSE",
       fill = "",
       color = "") +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
          axis.text.y = element_text(size = 8),
          axis.title.y = element_text(size = 10),
          axis.title.x = element_text(size = 10),
          strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
          strip.text = element_text(size = 10, margin = margin(1, 0, 1, 0)),
          panel.spacing = unit(1.1, "lines"),
          plot.margin = margin(0, 0.5, 0, 0, unit = "cm"),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
          legend.text = element_text(size = 10, hjust = 0),
          legend.title = element_text(size = 10),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)
    ) +
  guides(color = "none")


all_degs_human_mouse_unmatched_inverseSE_plot 

```

Merge plots

```{r}
#| fig-height: 4
#| fig-width: 10
#Merge plots
all_degs_human_mouse_nDEGs_InverseSE = (all_degs_human_mouse_unmatched_ndegs_plot/
all_degs_human_mouse_unmatched_inverseSE_plot) +
  plot_layout(axes = "collect",
              axis_titles = "collect",
              guides = "collect") +
  plot_annotation(theme = theme(legend.position = "top"))

all_degs_human_mouse_nDEGs_InverseSE %>% 
  ggsave(filename= here("Figures", "all_degs_human_mouse_nDEGs_InverseSE_lineplots.png"),
         width = 10, height = 4)

```

Matched timepoints

```{r}
# 1. Data Preparation (single cleanup step)
degs_summary <- all_degs_human_mouse %>%
  mutate(pathogen = recode(pathogen, 
                           "Staphylococcus aureus" = "S. aureus",
                           "Escherichia coli" = "E. coli",
                           "Hepatitits B" = "HepB")) %>%
  filter(adj_p_val <= 0.05) %>%
  # Keep only essential columns
  select(human_symbol, pathogen, timepoint, timepoint_comparison, organism) %>%
  distinct()

# 2. Identify overlap intelligently
# Create a table that indicates if a gene is a DEG in each organism
overlap_status <- degs_summary %>%
  separate(timepoint_comparison, into = c("timepoint_mouse", "timepoint_human"), 
           sep = "-", remove = FALSE) %>%
  group_by(pathogen, 
           timepoint_mouse,
           timepoint_human,
           human_symbol) %>%
  summarise(
    is_human = any(organism == "Human"),
    is_mouse = any(organism == "Mouse"),
    .groups = "drop", 
  ) %>%
  mutate(condition_new = case_when(
    is_human & is_mouse ~ "Shared",
    is_human            ~ "Human only",
    is_mouse            ~ "Mouse only"
  )) %>% 
  group_by(pathogen, condition_new,
           timepoint_mouse, timepoint_human) %>%
  summarise(n = n(), .groups = "drop") 


# 3. Final Summary
degs_human_mouse_overlap_genes_summary <- overlap_status %>%
  filter(!c(pathogen == "Staphylococcus aureus" & timepoint_mouse == "24h")) %>% 
  
  #Factor cleaning
  mutate(condition_new = fct_relevel(condition_new,  "Human only",  "Mouse only", "Shared"),
         pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", 
                                  "Staphylococcus aureus", "Escherichia coli", 
                                  "Burn", "Trauma"),
       timepoint_comparison = paste0(timepoint_mouse,  sep = "-", timepoint_human)) %>% 
  mutate(timepoint_mouse = as.character(timepoint_mouse),
         timepoint_mouse = fct_relevel(timepoint_mouse, c("Early", 
                                                          "Early hours", 
                                                          "2h", "4h", "6h", "12h", "24h", 
                                                          "Day 1", "Day 3", "Day 7")),
         timepoint_human = as.character(timepoint_human),
         timepoint_human = fct_relevel(timepoint_human, c("Early", 
                                                          "Early hours", 
                                                          "2h", "4h", "6h", "12h", "24h", 
                                                          "Day 1", "Day 3", "Day 7")),
         treatment = case_when(pathogen %in% c("Hepatitis B", "Influenza") ~ "Vaccination",
                               pathogen %in% c( "S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury")) %>% 
  arrange(timepoint_mouse)


# 4. Range calculation (kept as requested)
ranges_y <- degs_human_mouse_overlap_genes_summary %>%
  group_by(pathogen) %>%
  summarise(max = max(n))
```

### Lineplot of overlapped genes

Horizontal

```{r}
#| fig-height: 2
#| fig-width: 10
# Horizontal
degs_human_mouse_overlap_lineplot_horizontal = degs_human_mouse_overlap_genes_summary %>%
  distinct() %>% 
  ggplot() +
  
  aes(x = timepoint_mouse,
      y = n,
      fill = condition_new,
      color = condition_new,
      group = condition_new) + 
  
  facet_nested_wrap(~treatment+pathogen, 
                    axes = "all",
                    nrow = 1, 
                    scales = "free", 
                    # remove_labels = "all",
               nest_line = element_line(linetype = 1, color = "black")) +
  geom_line(width = 1) +
  
  geom_point(size = 2,
             pch = 21,
             color = "black",
             stroke = 0.5) + 
  geom_text_repel(aes(label = n), 
                  size = 2, 
                  # direction = "y", 
                  segment.alpha = 1) +

  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_continuous(expand = c(0.05, 0.05),
                     labels = label_number(scale_cut = cut_short_scale())) +
  theme_vaxgo() +
  scale_fill_manual(values = colors$comparison)    +
  scale_color_manual(values = colors$comparison)    +
  labs(x = "Time",
       y = "Total",
       fill = "",
       color = "") +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
          axis.text.y = element_text(size = 8),
          axis.title.y = element_text(size = 8),
          axis.title.x = element_text(size = 8),
          strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 8),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 8),
          strip.text = element_text(size = 8, margin = margin(1, 0, 1, 0)),
          panel.spacing = unit(1.1, "lines"),
          plot.margin = margin(0, 0.5, 0, 0, unit = "cm"),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
          legend.text = element_text(size = 8, hjust = 0),
          legend.title = element_text(size = 8),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)
    ) +
  guides(color = "none")



degs_human_mouse_overlap_lineplot_horizontal

# Save
degs_human_mouse_overlap_lineplot_horizontal %>%
  ggsave(filename = here("Figures", "all_human_mouse_dge_degs_overlap_lineplot_horizontal.png"),
         width = 10,
         height = 2)
```

## 2. Gene expression correlation analysis

### 2.1. Correlation: Genes, means, Overall

#### Pivot DEGs table

```{r}
#| fig-height: 20
#| fig-width: 20
#Import datasets ==============
#DEGs (matched)
all_degs_human_mouse =
  readRDS(here("tables", "Differential gene expression", "all_human_mouse_dge_limma_degs_matched.rds"))
#DEGs (unmatched)
all_degs_human_mouse_unmatched =
  readRDS(here("tables", "Differential gene expression", "all_human_mouse_dge_limma_degs.rds"))

#Check
all_degs_human_mouse %>% 
  dplyr::count(timepoint, organism, treatment, pathogen) 

#Functional modules
btm_annotation_genes = read_csv(file = here("tables", "Gene sets", "btm_annotation_genes.csv"))

human_mouse_log2fc_avg_wide_degs_all_immune_nonimmune =
  readRDS(here("tables", "Differential gene expression", "human_mouse_log2fc_avg_wide_degs_all_immune_nonimmune.rds"))

```

#### Compute correlation

```{r}
#| fig-height: 20
#| fig-width: 20
#####################
#Correlation (Weighted)
#####################

# Inter-species ==========
correlation_filtered_inter_weighted =      
  human_mouse_log2fc_avg_wide_degs_all_immune_nonimmune %>%
  filter(comparison_set == "Inter-species") %>% 
  group_by(comparison_set, treatment_human, treatment_mouse, pathogen_human, pathogen_mouse, 
           timepoint_human, timepoint_mouse, model_type,
           gene_set) %>%
  summarise({
    # Compute weighted correlation
    res <- wtd.cor(x = rank(mean_log2fc_human), 
                   y = rank(mean_log2fc_mouse), 
                   weight = weight_se_inverse)
    
    # Extract statistics
    data.frame(
      rho = res[1, "correlation"],
      se  = res[1, "std.err"],
      t   = res[1, "t.value"],
      p   = res[1, "p.value"],
      n_genes   = n()
    )
  }, .groups = 'drop')  %>% 
  mutate(
    #Significance
    p_adj = p.adjust(p, method = "BH"),
    p_adj_label = case_when(
      is.na(p_adj)      ~ "ns",
      p_adj <= 0.01     ~ "***",
      p_adj <= 0.05     ~ "**",
      p_adj <= 0.10     ~ "*",
      TRUE                ~ "ns"
    ),
    label = paste0("rho=", round(rho, 2), ", n=", n_genes, ", ", p_adj_label)
  ) %>% 
  
  #Standardize factors
     mutate(
       comparison = "Human vs. Mouse",
       timepoint_comparison = paste0(timepoint_human, sep = "-", timepoint_mouse),
       timepoint_comparison = str_replace_all(timepoint_comparison, "\n", " "),
       pathogen_comparison = paste0(pathogen_human, "-", pathogen_mouse),
       
       #Human
         timepoint_human = str_replace(timepoint_human,"Early hours", "Early\nhours"),
         timepoint_human = str_replace(timepoint_human,"Day ", "Day\n"),
         timepoint_human = fct_relevel(timepoint_human, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day\n1", "Day\n3", "Day\n7")),
         pathogen_human = case_when(pathogen_human == "Hepatitis B" ~ "HepB",
                              pathogen_human == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen_human == "Escherichia coli" ~ "E. coli",
                              T ~pathogen_human),
          pathogen_human = fct_relevel(pathogen_human, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment_human = case_when(pathogen_human %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen_human %in% c( "S. aureus", "E. coli") ~ "Infection",
                               pathogen_human %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen_human),
         treatment_human = fct_relevel(treatment_human, "Vaccination", "Infection", "Injury"),
         
      #Mouse
         timepoint_mouse = str_replace(timepoint_mouse,"Early hours", "Early\nhours"),
         timepoint_mouse = str_replace(timepoint_mouse,"Day ", "Day\n"),
         timepoint_mouse = fct_relevel(timepoint_mouse, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day\n1", "Day\n3", "Day\n7")),
         pathogen_mouse= case_when(pathogen_mouse == "Hepatitis B" ~ "HepB",
                              pathogen_mouse == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen_mouse == "Escherichia coli" ~ "E. coli",
                              T ~pathogen_mouse),
          pathogen_mouse = fct_relevel(pathogen_mouse, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment_mouse = case_when(pathogen_mouse %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen_mouse %in% c( "S. aureus", "E. coli") ~ "Infection",
                               pathogen_mouse %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen_mouse),
         treatment_mouse = fct_relevel(treatment_mouse, "Vaccination", "Infection", "Injury")) 

#Check
correlation_filtered_inter_weighted %>% 
  dplyr::count(pathogen_human, pathogen_mouse, timepoint_human, timepoint_mouse)

#Save
correlation_filtered_inter_weighted %>% 
  saveRDS(file = here("tables", "Differential gene expression", "all_human_mouse_correlation_filtered_inter_weighted.rds"))

#Summary
degs_immune_nonimmune_summary = human_mouse_log2fc_avg_wide_degs_all_immune_nonimmune %>%
  filter(comparison_set == "Inter-species") %>%  
  group_by(pathogen_human, pathogen_mouse, timepoint_comparison, gene_set) %>% 
  summarise(n_genes = n(), .groups = "drop")

#Filter comparisons with highest correlation ==============
correlation_filtered_inter_weighted_top = correlation_filtered_inter_weighted %>% 
  filter(gene_set == "DEGs") %>% 
  group_by(treatment_human, treatment_mouse, 
           pathogen_human, pathogen_mouse) %>% 
  slice_max(rho, n = 1) %>% 
  ungroup()

correlation_filtered_inter_weighted_filtered = correlation_filtered_inter_weighted %>% 
  inner_join(correlation_filtered_inter_weighted_top %>% 
               select(pathogen_human, pathogen_mouse, 
                      timepoint_human, timepoint_mouse),
             by = join_by(pathogen_human, pathogen_mouse, 
                          timepoint_human, timepoint_mouse))

#Save
correlation_filtered_inter_weighted_filtered %>% 
  saveRDS(file = here("tables", "Differential gene expression", "all_human_mouse_correlation_filtered_inter_weighted_filtered.rds"))


#Annotate facets  ==============
cor_data_facetted <- correlation_filtered_inter_weighted_filtered %>% 
  mutate(
    gene_set = factor(gene_set, levels = c("All genes", "DEGs", "Not-immune", "Immune")),
    facet_label = paste0(gene_set, ": ", label)) %>% 
  arrange(gene_set) %>% 
   group_by(pathogen_human, pathogen_mouse, timepoint_comparison) %>%
  summarise(
    facet_label = paste(facet_label, collapse = "\n"),
    .groups = "drop"
  )  %>% 
  mutate(facet_label = paste0(timepoint_comparison, "\n", facet_label)) %>% 
  arrange(timepoint_comparison) %>% 
  mutate(facet_label = fct_inorder(facet_label),
         pathogen_comparison = paste0(pathogen_human, " vs. ", pathogen_mouse)) %>% 
  arrange(pathogen_human) 

#Join with Expression data
human_mouse_log2fc_avg_wide_labels = human_mouse_log2fc_avg_wide_interspecies_bionull %>% 
  select(-pathogen_comparison) %>% 
  inner_join(cor_data_facetted , 
             by = join_by(pathogen_human, pathogen_mouse, timepoint_comparison)) %>% 
  mutate(pathogen_human = fct_relevel(pathogen_human, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma")) %>% 
  distinct() 

#Save
human_mouse_log2fc_avg_wide_labels %>% 
  saveRDS(file = here("tables", "Differential gene expression", "all_human_mouse_log2fc_avg_wide_labels.rds"))



```

#### Matrix plot

##### Interspecies

###### All genes + DEGs + Immune + Not-Immune

```{r}
#| fig-height: 10
#| fig-width: 11
#Import correlation
correlation_filtered_inter_weighted = 
  readRDS(file = here("tables", "Differential gene expression", "all_human_mouse_correlation_filtered_inter_weighted.rds"))
correlation_filtered_inter_weighted_filtered =
  readRDS(file = here("tables", "Differential gene expression", "all_human_mouse_correlation_filtered_inter_weighted_filtered.rds"))
human_mouse_log2fc_avg_wide_labels = 
  readRDS(file = here("tables", "Differential gene expression", "all_human_mouse_log2fc_avg_wide_labels.rds"))


# Plot Matrices
# All
correlation_heatmap_all = correlation_filtered_inter_weighted %>% 
  filter(pathogen_mouse != "DMD") %>% 
  ggplot() +
  aes(x = timepoint_mouse,
      y = timepoint_human) +
  #Facet
    facet_nested_wrap(~gene_set+treatment_mouse+pathogen_mouse,
                    scales = "free",
                    nrow = 4, 
                    axes = "all",
                    remove_labels = "none",
                    strip = strip_nested(size = "variable")) +
  labs(x = "Mouse",
       y = "Human") +
  geom_tile(aes(fill = rho)) +
  geom_text(aes(label = round(rho, 2)), size = 3, 
                  color = "black") +
  scale_x_discrete(expand = c(0,0)) +
  scale_y_discrete(expand = c(0,0)) +
  scale_fill_gradientn(
        colors = c("#4DBBD5FF", "white","#4361ee"),
        values = scales::rescale(seq(-1, 1, length.out = 10)),
        limits = c(-1, 1),
        name = "Spearman's rho",
        guide = guide_colorbar(
          barwidth = 15,
          barheight = 1,
          nbin = 10,
          raster = FALSE
        )
      ) +
  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.x = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(0,0,-10,0),
        panel.spacing.y = unit(0, "lines"),
        panel.spacing.x = unit(0.5, "lines"),
        plot.margin = margin(0, 0, 0, 0, unit = "cm"),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) 

correlation_heatmap_all 

correlation_heatmap_all %>%
  ggsave(filename = here("Figures", "all_human_mouse_dge_correlation_interspecies_matrixplot_weighted.png"),
         width = 11, height = 8)

```

###### All Genes + DEGs

```{r}
#| fig-height: 6
#| fig-width: 12
#Except immune
correlation_heatmap_all_degs = correlation_filtered_inter_weighted %>% 
  filter(pathogen_mouse != "DMD") %>% 
  filter(!(gene_set %in% c("Immune", "Not-immune"))) %>% 
  ggplot() +
  aes(x = timepoint_mouse,
      y = timepoint_human) +
  #Facet
  facet_nested_wrap(~gene_set+treatment_mouse+pathogen_mouse,
                    scales = "free",
                    ncol = 6, 
                    axes = "all",
                    remove_labels = "none",
                    strip = strip_nested(size = "variable")) +
  labs(x = "Mouse",
       y = "Human") +
  geom_tile(aes(fill = rho)) +
  geom_text(aes(label = round(rho, 2)), size = 3, 
                  color = "black") +
  scale_x_discrete(expand = c(0,0)) +
  scale_y_discrete(expand = c(0,0)) +
  scale_fill_gradientn(
        colors = c("#4DBBD5FF", "white","#4361ee"),
        values = scales::rescale(seq(-1, 1, length.out = 10)),
        limits = c(-1, 1),
        name = "Spearman's rho",
        guide = guide_colorbar(
          barwidth = 15,
          barheight = 1,
          nbin = 10,
          raster = FALSE
        )
      ) +
  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.x = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(0,0,-10,0),
        panel.spacing.y = unit(0, "lines"),
        panel.spacing.x = unit(0.5, "lines"),
        plot.margin = margin(0, 0, 0, 0, unit = "cm"),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) 

correlation_heatmap_all_degs 

 correlation_heatmap_all_degs %>%
  ggsave(filename = here("Figures", "all_human_mouse_dge_correlation_interspecies_all_degs_matrixplot.png"),
         width = 12, height = 6)
```

###### Only DEGs

```{r}
#| fig-height: 3
#| fig-width: 12
#Only DEGs
correlation_heatmap_degs = correlation_filtered_inter_weighted %>% 
  filter(gene_set == "DEGs") %>% 
  filter(pathogen_mouse != "DMD") %>% 
  ggplot() +
  aes(x = timepoint_mouse,
      y = timepoint_human) +
  #Facet
  facet_nested_wrap(~gene_set+treatment_mouse+pathogen_mouse,
                    scales = "free",
                    nrow = 1, 
                    axes = "all",
                    remove_labels = "none",
                    strip = strip_nested(size = "variable")) +
  labs(x = "Mouse",
       y = "Human") +
  geom_tile(aes(fill = rho),
            color = "white",
            size = 0.5) +
  geom_text(aes(label = round(rho, 2)), size = 3, 
                  color = "black") +
  scale_x_discrete(expand = c(0,0)) +
  scale_y_discrete(expand = c(0,0)) +
  scale_fill_gradientn(
        colors = c("#4DBBD5FF", "white","#4361ee"),
        values = scales::rescale(seq(-1, 1, length.out = 10)),
        limits = c(-1, 1),
        name = "Spearman's rho",
        guide = guide_colorbar(
          barwidth = 15,
          barheight = 0.5,
          nbin = 10,
          raster = FALSE
        )
      ) +
  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.x = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(0,0,-10,0),
        panel.spacing.y = unit(0, "lines"),
        panel.spacing.x = unit(0.5, "lines"),
        plot.margin = margin(0, 0, 0, 0, unit = "cm"),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) 

correlation_heatmap_degs 

correlation_heatmap_degs %>%
  ggsave(filename = here("Figures", "all_human_mouse_dge_correlation_interspecies_matrixplot_degs_weighted.png"),
         width = 12, height = 3)



```

###### Only DEGs

```{r}
#| fig-height: 3
#| fig-width: 12
#Only DEGs
correlation_dotplot_degs = correlation_filtered_inter_weighted %>% 
  filter(gene_set == "DEGs") %>% 
  filter(pathogen_mouse != "DMD") %>% 
  mutate(ci_upper = rho + 1.96 * se,
         ci_lower = rho - 1.96 * se,
         ci_width = ci_upper - ci_lower) %>% 
  ggplot() +
  aes(x = timepoint_mouse,
      y = timepoint_human) +
  #Facet
  facet_nested_wrap(~gene_set+treatment_mouse+pathogen_mouse,
                    scales = "free",
                    nrow = 1, 
                    axes = "all",
                    remove_labels = "none",
                    strip = strip_nested(size = "variable")) +
  labs(x = "Mouse",
       y = "Human") +
  geom_point(aes(fill = rho,
                 size = 1/se),
            color = "black",
            pch = 22) +
  scale_size_continuous(range = c(2, 12)) +
  # scale_x_discrete(expand = c(0,0)) +
  # scale_y_discrete(expand = c(0,0)) +
  scale_fill_gradientn(
        colors = c("#4DBBD5FF", "white","#4361ee"),
        values = scales::rescale(seq(-1, 1, length.out = 10)),
        limits = c(-1, 1),
        name = "Spearman's rho",
        guide = guide_colorbar(
          barwidth = 15,
          barheight = 0.5,
          nbin = 10,
          raster = FALSE
        )
      ) +
  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.x = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(0,0,-10,0),
        panel.spacing.y = unit(0, "lines"),
        panel.spacing.x = unit(0.5, "lines"),
        plot.margin = margin(0, 0, 0, 0, unit = "cm"),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1))  +
  # Headers
  #Vaccination
    geom_text(aes(label = round(rho, 2) %>% str_remove(., "0")), fontface = "bold", 
                  size = 3, 
                  vjust = -1.5,
                  color = "black",
              data = . %>% filter(treatment_human == "Vaccination")) +
  #Other
  geom_text(aes(label = round(rho, 2) %>% str_remove(., "0")), fontface = "bold", 
                  size = 3, 
                  color = "black",
            data = . %>% filter(treatment_human != "Vaccination")) 

#Print
correlation_dotplot_degs

#Save
correlation_dotplot_degs %>%
  ggsave(filename = here("Figures", "all_human_mouse_dge_correlation_interspecies_dotplot_degs_weighted.png"),
         width = 10, height = 3)



```

##### Intraspecies

All

```{r}
#| fig-height: 12
#| fig-width: 18
# # Intra-species ==========
# correlation_filtered_intra_weighted <- human_mouse_log2fc_avg_wide_degs_all_immune_nonimmune %>%
#   filter(comparison_set == "Intra-species") %>% 
#   group_by(
#     comparison_set, organism_1, treatment, pathogen,
#     timepoint_1, timepoint_2, gene_set
#   ) %>%
#   summarise({
#     # Capture the data for the current group
#     d <- cur_data()
#     
#     # Clean: filter out NAs and invalid weights
#     valid_data <- d %>% 
#       filter(!is.na(mean_log2fc_1), 
#              !is.na(mean_log2fc_2), 
#              weight_se_inverse > 0)
#     
#     # Check if we have enough observations (e.g., at least 3)
#     if (nrow(valid_data) >= 3) {
#       res <- wtd.cor(
#         x = rank(valid_data$mean_log2fc_1), 
#         y = rank(valid_data$mean_log2fc_2), 
#         weight = valid_data$weight_se_inverse
#       )
#       
#       data.frame(
#         rho = res[1, "correlation"],
#         se  = res[1, "std.err"],
#         t   = res[1, "t.value"],
#         p   = res[1, "p.value"]
#       )
#     } else {
#       # Return NAs if the group is too small or invalid
#       data.frame(rho = NA, se = NA, t = NA, p = NA)
#     }
#   }, .groups = 'drop') %>% 
#   mutate(comparison = paste0(organism_1, " vs. ", organism_1),
#          timepoint_comparison = paste0(timepoint_1, "-", timepoint_2)) %>%
# 
#   #Standardize factors
#   mutate(timepoint_label = if_else(pathogen %in% c("Staphylococcus aureus", "Escherichia coli"),
#                                    timepoint_2, timepoint_1),
#          timepoint_label = as.character(timepoint_label),
#          timepoint_label = str_replace(timepoint_label,"Early hours", "Early\nhours"),
#          timepoint_1 = str_replace(timepoint_1,"Early hours", "Early\nhours"),
#          timepoint_2 = str_replace(timepoint_2,"Early hours", "Early\nhours"),
#          timepoint_2 = str_replace(timepoint_2,"Day ", "Day\n"),
#          timepoint = fct_relevel(timepoint_label, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
#          timepoint_1 = fct_relevel(timepoint_1, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
#          timepoint_2 = fct_relevel(timepoint_2, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day\n1", "Day\n3", "Day\n7")),
#          pathogen = as.character(pathogen),
#          pathogen = case_when(pathogen == "Hepatitis B" ~ "HepB",
#                               pathogen == "Staphylococcus aureus" ~ "S. aureus",
#                               pathogen == "Escherichia coli" ~ "E. coli",
#                               T ~pathogen),
#           pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
#          treatment = case_when(pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
#                                pathogen %in% c( "S. aureus", "E. coli") ~ "Infection",
#                                pathogen %in% c( "Burn", "Trauma") ~ "Injury",
#                                T ~ pathogen),
#          treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury"))
```

```{r}
#| fig-height: 12
#| fig-width: 18
# #Matrix plot for intraspecies correlation =======
# #Human vs Human
# correlation_heatmap_intra_human_all = correlation_filtered_intra_weighted %>% 
#   # filter(!(pathogen %in% c("S. aureus", "E. coli") & organism_1 == "Human")) %>%
#   filter(organism_1 == "Human",
#          # gene_set != "Immune"
#          ) %>% 
#   ggplot() +
#   aes(x = timepoint_2,
#       y = timepoint_1) +
#   labs(x = "Timepoint",
#        y = "Timepoint") +
#   #Facet
#   facet_nested_wrap(~organism_1+gene_set+treatment+pathogen,
#                     scales = "free", trim_blank = F, ncol = 6,
#                     strip = strip_nested(size = "variable")) +
#   geom_tile(aes(fill = rho),
#             color = "white",
#             size = 0.5) +
#     geom_text(aes(label = round(rho, 2), 
#               color = ifelse(abs(rho) > 0.5, "white", "black")), 
#           size = 4, 
#           fontface = "bold") +
#   scale_color_identity() +
#   scale_x_discrete(expand = c(0,0)) +
#   scale_y_discrete(expand = c(0,0)) +
#   scale_fill_gradientn(
#         colors = c("#4DBBD5FF", "white","#4361ee"),
#         na.value = "white",
#         values = scales::rescale(seq(-0.999999, 0.999999, length.out = 10)),
#         limits = c(-0.999999, 0.99999),
#         name = "Spearman's rho",
#         guide = guide_colorbar(
#           barwidth = 15,
#           barheight = 0.5,
#           nbin = 10,
#           raster = FALSE
#         )
#       ) +
#   #Theme
#   theme_vaxgo() +
#   theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
#         axis.text.y = element_text(size = 10),
#         strip.text.x = element_text(size = 10),
#         strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
#         legend.position = "top",
#         legend.direction = "horizontal",
#         legend.margin=margin(0,0,0,0),
#         legend.box.margin=margin(0,0,-10,0),
#         panel.spacing.y = unit(0, "lines"),
#         panel.spacing.x = unit(0.2, "lines"),
#         plot.margin = margin(0, 0, 0, 0, unit = "cm"),
#         ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) 
# #Print
# correlation_heatmap_intra_human_all
# 
# #Save
# correlation_heatmap_intra_human_all %>%
#   ggsave(filename = here("Figures", "all_human_mouse_dge_correlation_intraspecies_matrixplot_human_weighted.png"),
#          width = 18, height = 10)
# 
# 
# #Mouse vs. Mouse
# correlation_heatmap_intra_mouse_all = correlation_filtered_intra_weighted %>% 
#   filter(!(pathogen %in% c("S. aureus", "E. coli") & organism_1 == "Human")) %>% 
#   filter(organism_1 == "Mouse",
#           # gene_set != "Immune"
#          ) %>% 
#   # View()
#   ggplot() +
#   aes(x = timepoint_2,
#       y = timepoint_1) +
#   labs(x = "Timepoint",
#        y = "Timepoint") +
#   #Facet
#   facet_nested_wrap(~organism_1+gene_set+treatment+pathogen,
#                     scales = "free",
#                     trim_blank = F,
#                     ncol = 6,
#                     strip = strip_nested(size = "variable")) +
#   geom_tile(aes(fill = rho),
#             color = "white",
#             size = 0.5) +
#   geom_text(aes(label = round(rho, 2), 
#               color = ifelse(abs(rho) > 0.5, "white", "black")), 
#           size = 4, 
#           fontface = "bold") +
#   scale_color_identity() +
#   scale_x_discrete(expand = c(0,0)) +
#   scale_y_discrete(expand = c(0,0)) +
#   scale_fill_gradientn(
#         colors = c("#4DBBD5FF", "white","#4361ee"),
#         na.value = "white",
#         values = scales::rescale(seq(-0.999999, 0.999999, length.out = 10)),
#         limits = c(-0.999999, 0.99999),
#         name = "Spearman's rho",
#         guide = guide_colorbar(
#           barwidth = 15,
#           barheight = 0.5,
#           nbin = 10,
#           raster = FALSE
#         )
#       ) +
#   #Theme
#   theme_vaxgo() +
#   theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
#         axis.text.y = element_text(size = 10),
#         strip.text.x = element_text(size = 10),
#         strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
#         legend.position = "top",
#         legend.direction = "horizontal",
#         legend.margin=margin(0,0,0,0),
#         legend.box.margin=margin(0,0,-10,0),
#         panel.spacing.y = unit(0, "lines"),
#         panel.spacing.x = unit(0.2, "lines"),
#         plot.margin = margin(0, 0, 0, 0, unit = "cm"),
#         ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) 
# #Print
# correlation_heatmap_intra_mouse_all
# 
# #Save
# correlation_heatmap_intra_mouse_all %>%
#   ggsave(filename = here("Figures", "all_human_mouse_dge_correlation_intraspecies_matrixplot_mouse_weighted.png"),
#          width = 18, height = 10)
# 
# 
# correlation_filtered_intra_weighted %>% 
#   filter(!(pathogen %in% c("S. aureus", "E. coli") & organism_1 == "Human")) %>% 
#   filter(organism_1 == "Mouse",
#           gene_set != "Immune") 
# 
# ######### Wrap plots
# design <- 
# "
# AAAABBBB
# AAAABBBB
# AAAA    
# "
# correlation_heatmap_intra_human_mouse_all = wrap_plots(A = correlation_heatmap_intra_mouse_all, 
#            B = correlation_heatmap_intra_human_all, 
#            design = design) +
#   plot_layout(guides = "collect") +
#   plot_annotation(theme = theme(legend.position = "top"))
#   
# correlation_heatmap_intra_human_mouse_all
# 
# #Save
# correlation_heatmap_intra_human_mouse_all %>%
#   ggsave(filename = here("Figures", "all_human_mouse_dge_correlation_intraspecies_matrixplot_all_weighted.png"),
#          width = 18, height = 10)


```

DEGs only

```{r}
#| fig-height: 6
#| fig-width: 12
#Matrix plot for intraspecies correlation =======
#Human vs Human
correlation_heatmap_intra_human = correlation_filtered_intra %>% 
  filter(!(pathogen %in% c("S. aureus", "E. coli") & organism1 == "Human")) %>% 
  filter(organism1 == "Human") %>% 
  filter(gene_set == "DEGs") %>% 
  # View()
  ggplot() +
  aes(x = timepoint2,
      y = timepoint1) +
  labs(x = "Timepoint",
       y = "Timepoint") +
  #Facet
  facet_nested_wrap(~organism1+treatment+pathogen,
                    scales = "free",
                    nrow = 1, trim_blank = F,
                    strip = strip_nested(size = "variable")) +
  geom_tile(aes(fill = rho),
            color = "white",
            size = 0.5) +
    geom_text(aes(label = round(rho, 2), 
              color = ifelse(abs(rho) > 0.2, "white", "black")), 
          size = 3, 
          fontface = "bold") +
  scale_color_identity() +
  scale_x_discrete(expand = c(0,0)) +
  scale_y_discrete(expand = c(0,0)) +
  scale_fill_gradientn(
        colors = c("#4DBBD5FF", "white","#4361ee"),
        values = scales::rescale(seq(-1, 1, length.out = 10)),
        limits = c(-1, 1),
        name = "Spearman's rho",
        guide = guide_colorbar(
          barwidth = 15,
          barheight = 0.5,
          nbin = 10,
          raster = FALSE
        )
      ) +
  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.x = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(0,0,-10,0),
        panel.spacing.y = unit(0, "lines"),
        panel.spacing.x = unit(0.2, "lines"),
        plot.margin = margin(0, 0, 0, 0, unit = "cm"),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) 

# correlation_heatmap_intra_human

#Mouse vs. Mouse
correlation_heatmap_intra_mouse = correlation_filtered_intra %>% 
  filter(!(pathogen %in% c("S. aureus", "E. coli") & organism1 == "Human")) %>% 
  filter(organism1 == "Mouse") %>% 
  filter(gene_set == "DEGs") %>% 
  # View()
  ggplot() +
  aes(x = timepoint2,
      y = timepoint1) +
  labs(x = "Timepoint",
       y = "Timepoint") +
  #Facet
  facet_nested_wrap(~organism1+treatment+pathogen,
                    scales = "free",
                    nrow = 1, trim_blank = F,
                    strip = strip_nested(size = "variable")) +
  geom_tile(aes(fill = rho),
            color = "white",
            size = 0.5) +
    geom_text(aes(label = round(rho, 2), 
              color = ifelse(abs(rho) > 0.2, "white", "black")), 
          size = 3, 
          fontface = "bold") +
  scale_color_identity() +
  scale_x_discrete(expand = c(0,0)) +
  scale_y_discrete(expand = c(0,0)) +
  scale_fill_gradientn(
        colors = c("#4DBBD5FF", "white","#4361ee"),
        values = scales::rescale(seq(-1, 1, length.out = 10)),
        limits = c(-1, 1),
        name = "Spearman's rho",
        guide = guide_colorbar(
          barwidth = 15,
          barheight = 0.5,
          nbin = 10,
          raster = FALSE
        )
      ) +
  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.x = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(0,0,-10,0),
        panel.spacing.y = unit(0, "lines"),
        panel.spacing.x = unit(0.2, "lines"),
        plot.margin = margin(0, 0, 0, 0, unit = "cm"),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) 

# correlation_heatmap_intra_mouse

######### Wrap plots
design <- "
  AAAAAA
  BBBB##
"
correlation_heatmap_intra_human_mouse = wrap_plots(A = correlation_heatmap_intra_mouse, 
           B = correlation_heatmap_intra_human, 
           design = design) +
  plot_layout(guides = "collect") +
  plot_annotation(theme = theme(legend.position = "top"))
  
correlation_heatmap_intra_human_mouse

#Save
correlation_heatmap_intra_human_mouse %>%
  ggsave(filename = here("Figures", "all_human_mouse_dge_correlation_intraspecies_matrixplot_degs.png"),
         width = 12, height = 6)


```

#### Scatter plots

##### Log2FC vs Log2fc

```{r}
#| fig-height: 12
#| fig-width: 12
#Unite gene data and correlation/facets
human_mouse_log2fc_avg_wide_labels_filters = 
  bind_rows(
    #All genes
    human_mouse_log2fc_avg_wide_labels %>% 
  mutate(color_group = "All genes"),

    #All DEGs
   human_mouse_log2fc_avg_wide_labels %>% 
  filter(p_adj_human <= 0.05 | p_adj_mouse <= 0.05) %>% 
  mutate(color_group = "All DEGs"),

   #Shared DEGs
   human_mouse_log2fc_avg_wide_labels %>% 
  filter(p_adj_human <= 0.05 & p_adj_mouse <= 0.05) %>% 
  mutate(color_group = "Shared DEGs"),

  #Immune DEGs
    human_mouse_log2fc_avg_wide_labels %>% 
  filter(human_symbol %in% c(btm_annotation_genes %>% 
                               clean_names() %>% 
                               filter(module_category == "immune") %>% 
                               pull(module_member_genes)),
         p_adj_human <= 0.05 & p_adj_mouse <= 0.05) %>% 
  mutate(color_group = "Immune")
  ) %>% 
  mutate(color_group = fct_relevel(color_group, "All genes", "All DEGs", "Shared DEGs", "Immune"))
```

```{r}
#| fig-height: 10
#| fig-width: 16
# Log2FC vs Log2fc
human_mouse_log2fc_log2fc_avg_wide_scatterplot = human_mouse_log2fc_avg_wide_labels_filters %>% 
  mutate(comparison_facet = if_else(str_detect(pathogen_comparison, "DMD"), "Control", "Experiment"),
         comparison_facet = fct_relevel(comparison_facet, "Experiment", "Control")) %>% 
  
  #Plot
  ggplot() +
  aes(x = mean_log2fc_mouse,
      y = mean_log2fc_human) +
  #Facet
  facet_nested_wrap(~comparison_facet+treatment_human+pathogen_human+facet_label,
                    # scales = "free",
                    ncol = 6, axes = "x",
                    strip = strip_nested(size = "variable")) +
  #Points 
  geom_point(aes(fill = color_group),
             alpha = 0.2,
             stroke = 0,
             pch = 21,
             size = 1.5,
             data = . %>% filter(color_group == "All genes")) +
  
  geom_point(aes(fill = color_group),
             alpha = 0.2,
             stroke = 0,
             pch = 21,
             size = 1.5,
             data = . %>% filter(color_group == "All DEGs")) +
  
  geom_point(aes(fill = color_group),
             alpha = 0.5,
             pch = 21,
             stroke = 0,
             size = 1.5,
             data = . %>% filter(!color_group %in% c("All genes", "All DEGs"))) +
    # Regression
  geom_smooth(
    method = "lm",
    mapping = aes(color = color_group, fill = color_group),
    se = TRUE
  ) +
  #Marginal plots
  # TOP SIDE 
  geom_violin(
    mapping = aes(
      color = color_group, fill = color_group,
      y = 8,
    ),
    outliers = FALSE,
    outlier.shape = NA,
    alpha = 1,
    width = 4,
    color = NA,
    show.legend = FALSE
  ) +
  geom_boxplot(
    mapping = aes(
      color = color_group, fill = color_group,
      y = 8,
    ),
    outliers = FALSE,
    outlier.shape = NA,
    alpha = 0, linewidth = 0.2,
    color = "black",
    width = 4,
    show.legend = FALSE
  ) +
  # Right SIDE 
  geom_violin(
    mapping = aes(
      color = color_group, fill = color_group,
      x = 8,
    ),
    outliers = FALSE,
    outlier.shape = NA,
    alpha = 1,
    width = 4,
    color = NA,
    show.legend = FALSE
  ) +
  geom_boxplot(
    mapping = aes(
      color = color_group, fill = color_group,
      x = 8,
    ),
    outliers = FALSE,
    outlier.shape = NA,
    alpha = 0, linewidth = 0.2,
    color = "black",
    width = 4,
    show.legend = FALSE
  ) +
  guides(fill = guide_legend(override.aes = list(size = 4))) +
  scale_fill_manual(values =  colors$shared) +
  scale_color_manual(values =  colors$shared) +
  #Add lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_hline(yintercept = 0, alpha = 1, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 1, linetype = 2) +
  #Theme
  theme_vaxgo() +
  theme(legend.position = "top",
        legend.direction = "horizontal",
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 12),
        strip.text.x           = element_text(size = 11, color = "black", hjust = 0, margin = margin(5, 0, 5, 0))) +
  labs(x = "Mouse L2FC",
       y = "Human L2FC",
       fill = "Class",
       color = "Class",
       title = "Correlation plot,\nHuman vs Mouse")


human_mouse_log2fc_log2fc_avg_wide_scatterplot
  

#Save
# human_mouse_log2fc_log2fc_avg_wide_scatterplot %>%
#   ggsave(filename = here("Figures", "all_human_mouse_dge_degs_scatterplot_withControl.png"),
#          width = 16, height = 10)

```
