---
title: "Comparing Human vs Mouse — Functional analysis"
author: Wasim Aluísio Prates-Syed
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Setup

```{r}
source(here("scripts_notebooks", "required.R"))
```

# Functional enrichment

Unified pipeline for all 6 conditions (Fluad, Engerix B, S. aureus, E.
coli,Burn, Trauma). All conditions are loaded from the pre-merged DGE
file, then each analysis (BTM means, BTM GSEA, Hallmarks means,
Hallmarks GSEA) runs once on the combined data and produces a single
output file.

## Import all conditions

Load the pre-merged DGE file that already contains all 6 conditions with
consistent column names and pathogen labels.

```{r}
#Import gene sets
btm_annotation_genes <- read_csv(here("tables", "Gene sets", "btm_annotation_genes.csv"))
msigdbr_grouped <- read_csv(here("tables", "Gene sets", "msigdb_hallmarks_grouped_genes.csv"))

#Import DEGs table
#Save
all_human_mouse_dge_limma_degs_matched_control = readRDS(file = here("tables", "Differential gene expression", "all_human_mouse_dge_limma_degs_matched_control.rds"))

#Filter out matched Control (DMD) and add only one timepoint
all_degs = all_human_mouse_dge_limma_degs_matched_control %>% 
  filter(organism != "DMD") %>%
  bind_rows(all_human_mouse_dge_limma_degs_matched_control %>%
              filter(organism == "DMD",
                     pathogen == "Burn",
                     timepoint == "Day 1") %>%
              mutate(timepoint = "Day 28",
                     pathogen = "DMD",
                     organism = "Mouse",
                     treatment = "Control",
                     condition = "DMD_Day 28")) %>%
  select(organism:human_symbol, mean_l2fc:sd, n) %>% 
  distinct()

# Extract unique condition labels for downstream filtering --------
annotation_conditions <- all_degs %>%
  select(condition, timepoint,  organism) %>%
  distinct()
```

## Blood Transcriptional Modules

### t-Test

Join DEGs with BTM annotation, collapse to module-level means per
organism and timepoint, test each module against zero, and save the
unified output.

```{r}
# Annotate each genes with its BTM membership by joining on gene symbol --------
dge_btm_process_genes <- all_degs %>%
  rename(symbol = human_symbol) %>%
  ungroup() %>%
  inner_join(btm_annotation_genes, by = "symbol") %>%
  filter(organism %in% c("Human", "Mouse"))

# Pivot the gene-level BTM table wide
#Create timepoint pairs
timepoint_map <- dge_btm_process_genes %>%
  filter(pathogen != "DMD") %>%
  select(pathogen, organism, timepoint) %>%
  distinct() %>%
  pivot_wider(
    names_from = organism,
    values_from = timepoint,
    values_fn = list
  )

#Pivot and bind control
dge_btm_process_genes_wide <- dge_btm_process_genes %>%
  filter(pathogen != "DMD") %>%
  select(symbol, mean_l2fc, adj_p_val, t, se, n_participants = n, pathogen, organism, timepoint, process, group) %>%
  distinct() %>%
  
  # Map and unnest cross-species timepoint combinations
  left_join(timepoint_map, by = "pathogen") %>%
  mutate(opposite_timepoint = if_else(organism == "Human", Mouse, Human)) %>%
  unnest(opposite_timepoint) %>%
  mutate(
    timepoint_human = if_else(organism == "Human", timepoint, opposite_timepoint),
    timepoint_mouse = if_else(organism == "Mouse", timepoint, opposite_timepoint),
    timepoint_comparison = stringr::str_c(timepoint_human, "-", timepoint_mouse)
  ) %>%
  select(-Human, -Mouse, -opposite_timepoint, -timepoint_human, -timepoint_mouse, -timepoint) %>%
  
  # Pivot the table wide
  pivot_wider(
    names_from = organism,
    values_from = c(mean_l2fc, adj_p_val, t, se, n_participants),
    values_fn = median
  ) %>%
  mutate(group = fct_relevel(group, immune_order)) %>%
  drop_na(mean_l2fc_Human, mean_l2fc_Mouse) %>% 
  inner_join(dge_btm_process_genes %>% 
                ungroup() %>%
              filter(pathogen == "DMD") %>%
              mutate(organism = "Control") %>% 
              select(symbol, mean_l2fc, adj_p_val, t, se, n_participants = n, organism, process, group) %>%
              distinct() %>%
              group_by(organism, process) %>%
              pivot_wider(names_from = organism,
                          values_from = c(mean_l2fc, adj_p_val, t, se, n_participants), values_fn = median) %>%
              ungroup() %>%
              mutate(group = fct_relevel(group, c(immune_order))),
             by = join_by(symbol, process, group)
  )

#Check
dge_btm_process_genes_wide

#Save
saveRDS(dge_btm_process_genes_wide, here("tables", "Functional analysis", "all_human_mouse_dge_btm_process_genes_wide.rds"))

# Collapse gene-level log2FC to module-level
dge_btm_mean_process <- dge_btm_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  summarise(mean_log2fc = mean(mean_l2fc, na.rm = T),
            median_log2fc = median(mean_l2fc, na.rm = T),
            sd_log2fc = sd(mean_l2fc, na.rm = T),
            var_log2fc = var(mean_l2fc, na.rm = T),
            n_genes = n(),
            n_participants = dplyr::first(n),
            group = dplyr::first(group),
            subgroup = dplyr::first(subgroup)) %>%
  distinct() %>%
  ungroup() %>%
  mutate(condition = paste0(organism, "_", pathogen, "_", timepoint))

# One-sample t-test per BTM vs zero with BH correction --------
pvals_vs_zero <- dge_btm_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>%
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance()

# Check
dge_btm_process_genes %>%
  dplyr::count(organism, pathogen, timepoint)

# Join module-level means with their t-test p-values --------
dge_btm_mean_process <- dge_btm_mean_process %>%
  left_join(
    pvals_vs_zero %>%
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  ) %>%
  clean_names()

#Pivot wide
dge_btm_mean_wide <- dge_btm_mean_process %>%
  select(organism, pathogen, timepoint, process, group,
         mean_log2fc, median_log2fc, sd_log2fc, var_log2fc, n_genes, p, p_adjust) %>%
  group_by(pathogen, timepoint, process) %>%
  pivot_wider(names_from = organism,
              values_from = c(mean_log2fc, median_log2fc, sd_log2fc, var_log2fc, n_genes, p, p_adjust),
              names_sep = "_") %>% 
  ungroup() 



dge_btm_mean_process %>% 
  dplyr::count(organism, pathogen, timepoint, n_participants) %>% 
  arrange(pathogen, timepoint)

#Save
saveRDS(dge_btm_mean_process, here("tables", "Functional analysis", "all_human_mouse_dge_btm_mean_process.rds"))
saveRDS(dge_btm_process_genes, here("tables", "Functional analysis", "all_human_mouse_dge_btm_process_genes.rds"))
```

### BTM GSEA

Rank genes with a composite metric (log2FC x -log(adj_p_val)) and run
GSEA against the BTM gene sets across all conditions simultaneously.

```{r message=FALSE, warning=FALSE}
# Prepare BTM term-to-gene mapping for GSEA
term2gene <- btm_annotation_genes %>%
  select(process, genes = symbol)

# Build ranked gene list with composite metric
df_genes2GSEA <- all_degs %>%
  distinct() %>%
  select(organism, pathogen, timepoint,
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val) %>%
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
  mutate(condition = paste0(organism, "_", pathogen, "_", timepoint),
         log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>%
  select(condition, genes, 
         #Rank metric
         rank = log2fold_change_logp)

# Run GSEA across all conditions
gsea_output <- autoGSEA(df = df_genes2GSEA,
                         TERM2GENE = term2gene,
                         geneset_name = "BTMs")

# Clean GSEA output and merge with module-level means --------
gsea_results <- gsea_output$gsea %>%
  as.data.frame() %>%
  clean_names() %>%
  rename(process = id) %>%
  filter(!is.na(process)) %>%
  rownames_to_column("del") %>% 
  select(-del, -description) %>% 
  full_join(dge_btm_mean_process %>% 
               rename(p_adjust_mean = p_adjust),
            by = join_by("condition", "process"))

#Check
gsea_results %>% dplyr::count(pathogen, organism, timepoint)


#Save
saveRDS(gsea_results, here("tables", "Functional analysis", "all_human_mouse_gsea_btm_results.rds"))
```


```{r message=FALSE, warning=FALSE}
#Compare humans to mice
gsea_results_human = gsea_results %>% 
  filter(organism == "Human",
         pathogen != "DMD") %>% 
  select(-organism) %>% 
  distinct() %>% 
  rename_with( .fn = ~ stringr::str_c(.x, "_Human"),
               cols = c(enrichment_score:condition, timepoint, mean_log2fc:p_adjust_mean)) %>% 
  rename(process = process_Human, gsea_enrichment = gsea_enrichment_Human,
         pathogen = pathogen_Human, group = group_Human, subgroup = subgroup_Human)

gsea_results_mouse = gsea_results %>% 
  filter(organism == "Mouse",
         pathogen != "DMD") %>% 
  select(-organism) %>% 
  distinct() %>% 
  rename_with( .fn = ~ stringr::str_c(.x, "_Mouse"),
               cols = c(enrichment_score:condition, timepoint, mean_log2fc:p_adjust_mean)) %>% 
  rename(process = process_Mouse, gsea_enrichment = gsea_enrichment_Mouse,
         pathogen = pathogen_Mouse, group = group_Mouse, subgroup = subgroup_Mouse)

gsea_results_control = gsea_results %>% 
  filter(organism == "Mouse",
         pathogen == "DMD") %>% 
  mutate(organism = "Control",
         pathogen = "Control") %>% 
  select(-organism, -timepoint, -pathogen,) %>% 
  distinct() %>% 
  rename_with( .fn = ~ stringr::str_c(.x, "_Control"),
               cols = c(enrichment_score:condition, timepoint, mean_log2fc:p_adjust_mean)) %>% 
  rename(process = process_Control, gsea_enrichment = gsea_enrichment_Control, group = group_Control, subgroup = subgroup_Control)

dge_btm_gsea_mean_wide = gsea_results_human %>% 
  #Add mouse
  full_join(gsea_results_mouse,
            by = join_by(process, gsea_enrichment, pathogen, group, subgroup)) %>% 
  #Add Control
  full_join(gsea_results_control,
            by = join_by(process, gsea_enrichment, group, subgroup)) %>% 
  drop_na(timepoint_Human, timepoint_Mouse) %>% 
  mutate(timepoint_comparison = paste0(timepoint_Human, "-", timepoint_Mouse)) %>% 
  
  
#Calculate statistics
  mutate(
    # Cross-species effect size delta 
    mean_log2fc_diff = mean_log2fc_Human - mean_log2fc_Mouse,
    
    # Joint FDR significance weighting (GSEA)
    weight_human_gsea_fdr = -log10(pmax(p_adjust_Human, 1e-10)),
    weight_mouse_gsea_fdr = -log10(pmax(p_adjust_Mouse, 1e-10)),
    weight_control_gsea_fdr = -log10(pmax(p_adjust_Control, 1e-10)),
    
    weight_joint_gsea_fdr = sqrt(weight_human_gsea_fdr * weight_mouse_gsea_fdr),
    weight_joint_gsea_fdr_ctrl = sqrt(weight_human_gsea_fdr * weight_control_gsea_fdr),
    
    # Statistical Weights for Weighted Correlation
    se_sq_Human = var_log2fc_Human / n_genes_Human,
    se_sq_Mouse = var_log2fc_Mouse / n_genes_Mouse,
    se_sq_Control = var_log2fc_Control / n_genes_Control,
    
    # Combined variance of the cross-species mean difference
    var_combined = se_sq_Human + se_sq_Mouse,
    var_combined_ctrl = se_sq_Human + se_sq_Control,
    # Correct inverse variance weight
    weight_var_inverse = 1 / var_combined,
    weight_var_inverse_ctrl = 1 / var_combined_ctrl,
  )

#Save
saveRDS(dge_btm_gsea_mean_wide, here("tables", "Functional analysis", "all_human_mouse_dge_btm_gsea_mean_wide.rds"))


#Pivot longer
gsea_btm_results_all_matched = dge_btm_gsea_mean_wide %>% 
  tidyr::pivot_longer(
    # Select all organism-suffixed columns
    cols = dplyr::ends_with(c("_Human", "_Mouse", "_Control")),
    
    # Split column names into the metric name (.value) and the organism level
    names_to = c(".value", "organism"),
    names_sep = "_(?=[^_]+$)" # Regex splits strictly at the final underscore
  )

#Save
saveRDS(gsea_btm_results_all_matched, here("tables", "Functional analysis", "all_human_mouse_gsea_btm_results_matched.rds"))



```


