---
title: "5_EvolutionaryAnalysis"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    embed-resources: true
    fig-cap-location: bottom
    theme: [cosmo, mouse2human-html.scss]
editor: 
  markdown: 
    wrap: 72
---
# Evolutionary analysis

# Setup

```{r}
#| message: false
#| warning: false
# Load core packages for the analysis session
library(here)

# Source project-wide functions, themes, and colour palettes
source(here("scripts_notebooks", "required.R"))
```

# Protein identity

## Annotate with sequence identity

Import human-to-mouse sequence identity data from Ensembl BioMart and
annotate BTM core enrichment genes with homology information. This
annotation enables correlation of sequence conservation with
cross-species transcriptional concordance.

```{r}
#| fig-height: 5
#| fig-width: 12
# Download sequence identity data from Ensembl BioMart
# http://www.ensembl.org/biomart/martview/cdbaeb7d08ea2dbc5bae9a1b0d6ad871?VIRTUALSCHEMANAME=default&ATTRIBUTES=hsapiens_gene_ensembl.default.homologs.ensembl_gene_id|hsapiens_gene_ensembl.default.homologs.ensembl_gene_id_version|hsapiens_gene_ensembl.default.homologs.ensembl_transcript_id|hsapiens_gene_ensembl.default.homologs.ensembl_transcript_id_version|hsapiens_gene_ensembl.default.homologs.mmusculus_homolog_perc_id_r1|hsapiens_gene_ensembl.default.homologs.mmusculus_homolog_perc_id|hsapiens_gene_ensembl.default.homologs.mmusculus_homolog_associated_gene_name|hsapiens_gene_ensembl.default.homologs.mmusculus_homolog_orthology_type|hsapiens_gene_ensembl.default.homologs.mmusculus_homolog_canonical_transcript_protein|hsapiens_gene_ensembl.default.homologs.mmusculus_homolog_ensembl_gene&FILTERS=&VISIBLEPANEL=attributepanel
# Import and annotate

# Import BTM core enrichment results
gsea_btm_results_core_joined =
  readRDS(file = here("tables", "Functional analysis", "all_human_mouse_gsea_btm_results_core_joined.rds")) %>%
  distinct()

#Import modules and LEGs
dge_btm_process_module_diff_and_legs = 
  readRDS(file = here("tables", "Functional analysis", "all_human_mouse_gsea_btm_legs.rds"))

# Import BTM gene-level correlation data
dge_btm_process_genes_diff_bygene_clean_filtered =  readRDS(file = here("tables", "Functional analysis", "dge_btm_process_genes_diff_bygene_clean_filtered.rds"))

# Import homolog alignment data
all_alignments = read_csv(here("tables", "Gene and Protein sequences", "human_to_mouse_alignment_mart_export.txt"))

# Import BTM gene annotations
btm_annotation_genes = read_csv(file = here("tables", "Gene sets", "btm_annotation_genes.csv"))

# Annotate genes with sequence identity and BTM membership
# Build alignment-annotated dataset for core enrichment gene

# Reference: all high-confidence orthologs in the database (unfiltered)
alignments_human2mouse_all = all_alignments %>%
  clean_names() %>%
  filter(mouse_orthology_confidence_0_low_1_high == 1) %>%
  drop_na(mouse_gene_name) %>%
  select(gene = mouse_gene_name, mouse_homology_type,
         identity_human2mouse = percent_id_query_gene_identical_to_target_mouse_gene,
         identity_mouse2human = percent_id_target_mouse_gene_identical_to_query_gene) %>%
  mutate(gene = str_to_upper(gene)) %>%
  # Resolve multiple identity scores by taking the median per gene
  group_by(gene) %>%
  summarise(identity_human2mouse = median(identity_human2mouse)) %>%
  ungroup()

alignments_human2mouse_all %>% 
  saveRDS(file = here("tables", "Gene and Protein sequences", "alignments_human2mouse_all.rds"))

# Filter to only BTM core enrichment genes and append process membership
alignments_human2mouse = alignments_human2mouse_all %>% 
  inner_join(dge_btm_process_module_diff_and_legs,
             by = join_by(gene == symbol))  %>% 
  mutate(status = if_else(status == "Shared", status, "Not-shared"))

#Group by process
protein_identity_process = alignments_human2mouse %>% 
  group_by(treatment, pathogen, timepoint_comparison, process, status) %>% 
  summarise(facet_label = dplyr::first(facet_label), group = dplyr::first(group),
            mean_identity = mean(identity_human2mouse, na.rm = T))

# Import pre-computed human gene information and join with alignment data
gene_info_human =
  alignments_human2mouse_all %>% 
  #Add difference in log2FC
  inner_join(dge_btm_process_genes_diff_bygene_clean_filtered,
             by = join_by(gene == hgnc_symbol)) 

gene_info_human %>% 
  dplyr::count(pathogen, status)


```

### Protein id% vs. Diff. in means

```{r}
#| fig-height: 3
#| fig-width: 8
#| message: false
#| warning: false
#### Difference in means vs. status
gene_info_human_clean_dge <- gene_info_human %>%
  # filter(treatment != "Vaccination") %>% 
  select(treatment, pathogen, status, timepoint_comparison, gene, identity_human2mouse, abs_log2fc_diff, inverse_se_mouse) %>% 
  distinct()

#Save
gene_info_human_clean_dge %>% 
  saveRDS(here("tables", "Gene and Protein sequences", "human_mouse_gene_info_dge_clean.rds"))
```

```{r}
#| fig-height: 3
#| fig-width: 8
#| message: false
#| warning: false
#Import
gene_info_human_clean_dge =
  readRDS(here("tables", "Gene and Protein sequences", "human_mouse_gene_info_dge_clean.rds"))

#Plot
alignments_human2mouse_legs_identity_log2fc_plot = gene_info_human_clean_dge %>% 

  filter(status != "Not LEG",
         treatment %in% c("Infection", "Injury")) %>% 
  bind_rows(gene_info_human_clean_dge %>% 

  filter(status != "Not LEG",
         treatment %in% c("Infection", "Injury")) %>% 
    mutate(treatment = "Pooled",
           pathogen = "")) %>% 

    ggplot() +
  aes(
    x = identity_human2mouse,
    y = abs_log2fc_diff,
    color = status,
    fill = status
  ) +
  
  facet_nested_wrap(~treatment+pathogen, nrow = 1) +
  
  # Regression
  geom_smooth(
    method = "lm",
    mapping = aes(color = status, fill = status),
    se = TRUE
  ) +
  #Marginal Boxplot
  # TOP  
  geom_violin(
    mapping = aes(
      x = identity_human2mouse,         
      y = 3,
    ),
    data = . %>% filter(status == "All genes"),
    outliers = FALSE,
    outlier.shape = NA,
    alpha = 0.5,
    width = 0.5,
    color = NA,
    show.legend = FALSE
  ) +
  geom_boxplot(
        data = . %>% filter(status == "All genes"),
    mapping = aes(
      x = identity_human2mouse,         
      y = 3,
    ),
    outliers = FALSE,
    outlier.shape = NA,
    alpha = 0, linewidth = 0.2,
    color = "black",
    width = 0.5,
    show.legend = FALSE
  ) +

  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = colors$comparison) +
  scale_x_continuous(expand = c(0,0)) +
  labs(
    x = "%Protein identity",
    y = "|ΔLog2FC|",
    fill = "LEG status"
  ) +
  theme_vaxgo() +
  theme(
    legend.position = "top",
  ) +
  guides(
    color = "none",
    fill = guide_legend(
      title.position = "left", 
      byrow = TRUE, 
      nrow = 1,
      override.aes = list(shape = 22, alpha = 1, size = 3, linetype = 0)
    )
  )

alignments_human2mouse_legs_identity_log2fc_plot 


# Save to file
alignments_human2mouse_legs_identity_log2fc_plot %>%
  ggsave(filename = here("Figures", "all_human_mouse_gsea_LEGs_alignment_btms_human2mouse_legs_identity_vs_meanlog2fc.png"),
         width = 8,
         height = 3)

```

```{r}
#| fig-height: 3
#| fig-width: 3
#| message: false
#| warning: false


#Plot
alignments_human2mouse_legs_identity_log2fc_pooled_plot = gene_info_human_clean_dge %>% 
    filter(status == "All genes",
         treatment %in% c("Infection", "Injury")) %>% 
    ggplot() +
  aes(
    x = identity_human2mouse,
    y = abs_log2fc_diff,
    color = status,
    fill = status
  ) +
  
  # Regression
  geom_smooth(
    method = "lm",
    mapping = aes(color = status, fill = status),
    se = TRUE,
  ) +
    #Marginal Boxplot
  # TOP  
  geom_violin(
    mapping = aes(
      x = identity_human2mouse,         
      y = 1,
    ),
    outliers = FALSE,
    outlier.shape = NA,
    alpha = 0.5,
    width = 0.1,
    color = NA,
    show.legend = T
  ) +
  geom_boxplot(
    mapping = aes(
      x = identity_human2mouse,         
      y = 1,
    ),
    outliers = FALSE,
    outlier.shape = NA,
    alpha = 0, 
    linewidth = 0.2,
    color = "black",
    width = 0.1,
    show.legend = FALSE
  ) +
    #Add correlation stats
  stat_cor( 
    method = "spearman",
    cor.coef.name = "rho", 
    parse = TRUE,      
    label.y = 0.10,
    label.x = 52,
    size = 3.5, 
    show.legend = F
  ) +
  # Line equation
  stat_regline_equation(
    aes(label = ..eq.label..),
    label.y = 0.20,
    label.x = 52,       
    size = 3.5
  ) +
  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = colors$comparison) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(
    x = "%Protein identity",
    y = "|ΔLog2FC|",
    fill = ""
  ) +
  theme_vaxgo() +
  theme(
    legend.position = "top",
    legend.location = "plot"
  ) +
  guides(
    color = "none",
    fill = guide_legend(
      title.position = "left", 
      byrow = TRUE, 
      nrow = 1,
      override.aes = list(shape = 22, alpha = 1, size = 3, linetype = 0)
    )
  ) 

alignments_human2mouse_legs_identity_log2fc_pooled_plot 

# Save 
alignments_human2mouse_legs_identity_log2fc_pooled_plot %>%
  ggsave(filename = here("Figures", "all_human_mouse_gsea_btms_legs_identity_vs_meanlog2fc_pooled.png"),
         width = 3,
         height = 3)

```

### Protein id% by Shared/NotShared: Boxplot

All

```{r}
#| fig-height: 3
#| fig-width: 8
# Protein identity by Shared/NotShared: Boxplot
protein_identity_genes_distribution = gene_info_human_clean_dge %>% 
  filter(status != "Not LEG",
         treatment %in% c("Infection", "Injury")) %>% 
  bind_rows(gene_info_human_clean_dge %>% 
  filter(status != "Not LEG",
         treatment %in% c("Infection", "Injury")) %>% 
    mutate(treatment = "Pooled",
           pathogen = "")) %>% 
  ggplot() +
  aes(
    x     = status,
    y     = identity_human2mouse,
    color = status,
    fill  = status
  ) +
  
  facet_nested_wrap(~treatment+pathogen, nrow = 1) +

  geom_violin(alpha = 0, ) +
  # Render boxplot layer without outliers
  geom_boxplot(
    width    = 0.3,
    alpha    = 0,
    outliers = FALSE,
    # color    = "black"
  ) +
  # Apply project custom palettes and labels
  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = colors$comparison) +
  
  scale_x_discrete(expand = expansion(mult = c(0.2, 0.2))) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.05)),
                     limits = c(50, 110)) +
  
  # Labels (Corrected X and Y axis mappings)
    labs(
    x     = "Status",
    y     = "% Identity",
    color = "",
    fill  = ""
  ) +
  
  # Theme and Legend Styling
  theme_vaxgo() +
  theme(
    legend.position = "top", legend.location = "plot",
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
    # STATISTICAL TEST LAYER
  stat_compare_means(
    comparisons = list(c("Shared", "Not-shared"),
                       c("Shared", "All genes"),
                       c("Not-shared", "All genes")),
    method = "wilcox.test",
    label = "p.adj", 
    vjust = 1.5,
    label.y.np = "top",
    size = 3, tip.length = 0,
    hide.ns = F
  ) 



protein_identity_genes_distribution


#Save
protein_identity_genes_distribution %>%
  ggsave(filename = here("Figures", "all_human_mouse_gsea_btms_legs_proteinsequenceidentity.png"),
         width = 8,
         height = 3)
```

Statistical analysis

```{r}
#| fig-width: 8
#| fig-height: 4
# # NON-PARAMETRIC TEST (Wilcoxon) + EFFECT SIZE (r)
# # Calculate Wilcoxon test + manual r effect size
# # 1. Run the Wilcoxon test
# stat_test <- protein_identity_process %>%
#   group_by(treatment, pathogen, timepoint_comparison) %>%
#   wilcox_test(mean_identity ~ status) %>%
#   add_significance()
# 
# # 2. Calculate the Effect Size (r)
# eff_size <- protein_identity_process %>%
#   group_by(treatment, pathogen, timepoint_comparison) %>%
#   wilcox_effsize(mean_identity ~ status)
# 
# # 3. Combine both results and create the custom label
# stat_test_group <- stat_test %>%
#   left_join(
#     eff_size %>% select(treatment, pathogen, timepoint_comparison, effsize, magnitude),
#     by = c("treatment", "pathogen", "timepoint_comparison")
#   ) %>%
#   mutate(
#     label_custom = paste0(p.signif, " (r = ", round(effsize, 2), ")")
#   )
# 
# # NON-PARAMETRIC TEST (Wilcoxon) + EFFECT SIZE (r)
# # Calculate Wilcoxon test + manual r effect size
# # 1. Run the Wilcoxon test
# stat_test <- protein_identity_process %>%
#   group_by(treatment, pathogen, timepoint_comparison) %>%
#   wilcox_test(mean_identity ~ status) %>%
#   add_significance()
# 
# # 2. Calculate the Effect Size (r)
# eff_size <- protein_identity_process %>%
#   group_by(treatment, pathogen, timepoint_comparison) %>%
#   wilcox_effsize(mean_identity ~ status)
# 
# # 3. Combine both results and create the custom label
# stat_test_group <- stat_test %>%
#   left_join(
#     eff_size %>% select(treatment, pathogen, timepoint_comparison, effsize, magnitude),
#     by = c("treatment", "pathogen", "timepoint_comparison")
#   ) %>%
#   mutate(
#     label_custom = paste0(p.signif, " (r = ", round(effsize, 2), ")")
#   )


```

Integrate sequence identity with cross-species gene correlation,
organized by functional hierarchy. This prepares data for comparing
sequence conservation with gene-level expression concordance.

```{r}
#| fig-height: 5
#| fig-width: 12
# Establish process order based on median sequence identity
process_order_median = alignments_human2mouse %>%
  select(process, group, gene, identity_human2mouse) %>%
  group_by(process) %>%
  summarise(median = median(identity_human2mouse)) %>%
  arrange(median)

group_order_median = alignments_human2mouse %>%
  select(process, group, gene, identity_human2mouse) %>%
  group_by(group) %>%
  summarise(median = median(identity_human2mouse)) %>%
  arrange(median)

# Prepare the detailed process-level dataframe with correlation stats
alignments_human2mouse_process = alignments_human2mouse %>%
  mutate(process = fct_relevel(process, process_order_median$process),
         group = fct_relevel(group, group_order_median$group) %>% fct_rev()
         ) %>%
  distinct()

```

Compare sequence identity distributions across high-level functional
groups (BTM groups). A violin-boxplot-pointrange composite visualizes
the density, quartiles, and bootstrapped mean of human-to-mouse sequence
identity per group.

```{r}
#| fig-height: 6
#| fig-width: 4
# Create the distribution plot (Violin + Boxplot + Pointrange)
alignments_human2mouse_pointrange_group_btm = alignments_human2mouse %>%
  mutate(group = fct_relevel(group, group_order_median$group)) %>%
  distinct() %>% 
  # Data aesthetics
  ggplot() +
  aes(x = identity_human2mouse,
      y = group,
      fill = group,
      color = group) +
  # Geoms
  geom_violin() +
  geom_boxplot(alpha = 0.1,
               width = 0.1,
               fill = NA,
               color = "white") +
  stat_summary(fun.data = "mean_cl_boot",
               geom = "pointrange",
               size = 1,
               fatten = 0.5,
               color = "black", show.legend = F) +
  # Scales
  scale_fill_manual(values = immune_colors) +
  scale_color_manual(values = immune_colors) +
  scale_y_discrete(expand = c(0.05, 0.05)) +
  scale_x_continuous(expand = c(0.01, 0.01)) +
  # Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 1, size = 10),
        axis.text.y = element_text(size = 10),
        strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
        panel.spacing = unit(0, "lines"),
        plot.margin = margin(0, 0.3, 0, 0, unit = "cm"),
        legend.position = "none",
        legend.direction = "horizontal",
        legend.margin = margin(0, 0, 0, 0),
        legend.box.margin = margin(0, 0, -10, 0),
        legend.text = element_text(size = 10, hjust = 0),
        legend.title = element_text(size = 10),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)
  ) +
  # Labels
  labs(x = "% Identity (Human-to-Mouse)",
       y = NULL,
       color = "Group",
       fill = "Group")

# Render plot
alignments_human2mouse_pointrange_group_btm

# Save to file
alignments_human2mouse_pointrange_group_btm %>%
  ggsave(filename = here("Figures", "all_human_mouse_gsea_LEGs_alignment_btms_human2mouse_violinplot_group.png"),
         width = 4,
         height = 6)
```






## Evolution of protein sequence

### Download sequences

```{r}
# ---- 1. Prerequisites & Installation ----
# Ensure all required packages are installed
# Gene expression
dge_btm_process_genes_diff_bygene_clean_filtered =  readRDS(file = here("tables", "Functional analysis", "dge_btm_process_genes_diff_bygene_clean_filtered.rds"))

gene_info_human_clean_dge = 
  readRDS(here("tables", "Gene and Protein sequences", "human_mouse_gene_info_dge_clean.rds"))

# ---- 3. Connect to Ensembl BioMart ----
message("Connecting to Ensembl BioMart...")
human_mart <- biomaRt::useEnsembl(
  biomart = "genes",
  dataset = "hsapiens_gene_ensembl",
  mirror  = "www" 
)

mouse_mart <- biomaRt::useEnsembl(
  biomart = "genes",
  dataset = "mmusculus_gene_ensembl",
  mirror  = "www" 
)


# listAttributes(human_mart) %>% view()

# ---- 4. Retrieve Human-Mouse 1:1 Orthologs Map ----
message("Fetching orthology mapping...")
ortholog_map <- getBM(
  attributes = c("ensembl_gene_id", 
                 "external_gene_name", 
                 "mmusculus_homolog_ensembl_gene", 
                 "mmusculus_homolog_associated_gene_name",
                 "mmusculus_homolog_orthology_type"),
  mart = human_mart
) %>% 
  as_tibble() %>%
  filter(mmusculus_homolog_orthology_type == "ortholog_one2one") %>%
  filter(mmusculus_homolog_ensembl_gene != "" & ensembl_gene_id != "")

# Filter genes
target_map <- ortholog_map %>% 
  filter(external_gene_name %in% c(dge_btm_process_genes_diff_bygene %>% pull(human_symbol) %>% unique()))
```

Functions

```{r}
#Functions =============
# Helper function to process gene chunks safely
fetch_cds_chunked <- function(df_part, part_label, chunk_size = 100) {
  message(stringr::str_c("=== Starting Part ", part_label, " ==="))
  
  df_chunked <- df_part %>%
    dplyr::mutate(chunk = (dplyr::row_number() - 1) %/% chunk_size + 1)
  
  total_chunks <- max(df_chunked$chunk)
  
  res_tb <- df_chunked %>%
    dplyr::group_split(chunk) %>%
    purrr::map_dfr(function(df_chunk) {
      chunk_id <- df_chunk$chunk[1]
      message(stringr::str_c("Part ", part_label, " | Fetching chunk ", chunk_id, " of ", total_chunks, "..."))
      
      # Set useCache = FALSE to prevent dbDisconnect / .lock2 SQLite errors
      biomaRt::getSequence(
        id       = df_chunk$ensembl_gene_id,
        type     = "ensembl_gene_id",
        seqType  = "coding",
        mart     = human_mart,
        useCache = FALSE
      ) %>%
        as_tibble()
    })
  
  return(res_tb)
}

# Helper function to process mouse CDS gene chunks safely
fetch_mouse_cds_chunked <- function(df_part, part_label, chunk_size = 100) {
  message(stringr::str_c("=== Starting Mouse Part ", part_label, " ==="))
  
  df_chunked <- df_part %>%
    dplyr::mutate(chunk = (dplyr::row_number() - 1) %/% chunk_size + 1)
  
  total_chunks <- max(df_chunked$chunk)
  
  res_tb <- df_chunked %>%
    dplyr::group_split(chunk) %>%
    purrr::map_dfr(function(df_chunk) {
      chunk_id <- df_chunk$chunk[1]
      message(stringr::str_c("Mouse Part ", part_label, " | Fetching chunk ", chunk_id, " of ", total_chunks, "..."))
      
      biomaRt::getSequence(
        id      = df_chunk$mmusculus_homolog_ensembl_gene,
        type    = "ensembl_gene_id",
        seqType = "coding",
        mart    = mouse_mart,
        useCache = FALSE
      ) %>%
        as_tibble()
    })
  
  return(res_tb)
}
```

Download sequences

```{r}
#Download Human sequences ====================
message("Downloading Mouse coding sequences in 3 batches")
# Split target map into 3 balanced partition objects
target_map_split <- target_map %>%
  dplyr::mutate(part_id = dplyr::ntile(dplyr::row_number(), 3)) %>%
  dplyr::group_split(part_id)

target_map_part1 <- target_map_split[[1]]
target_map_part2 <- target_map_split[[2]]
target_map_part3 <- target_map_split[[3]]

# Download 
human_cds_raw_part1 <- fetch_cds_chunked(target_map_part1, part_label = "1/3")
human_cds_raw_part2 <- fetch_cds_chunked(target_map_part2, part_label = "2/3")
human_cds_raw_part3 <- fetch_cds_chunked(target_map_part3, part_label = "3/3")

#Save temporarily. Connection may be lost during the process
human_cds_raw_part1 %>% saveRDS(file = here("tables", "human_cds_raw_part1.rds"))
human_cds_raw_part2 %>% saveRDS(file = here("tables", "human_cds_raw_part2.rds"))
human_cds_raw_part3 %>% saveRDS(file = here("tables", "human_cds_raw_part3.rds"))

human_cds_raw_part1  =  readRDS(file = here("tables", "human_cds_raw_part1.rds"))
human_cds_raw_part2  =  readRDS(file = here("tables", "human_cds_raw_part2.rds"))
human_cds_raw_part3 = readRDS(file = here("tables", "human_cds_raw_part3.rds"))

# Consolidate raw CDS partitions into single master table
human_cds_raw <- dplyr::bind_rows(
  human_cds_raw_part1,
  human_cds_raw_part2,
  human_cds_raw_part3
)

notify(
  title = "Human CDS downloaded!",
  msg   = c("All 3 parts processed and saved successfully.")
)

# Filter for Longest Canonical/Isoform CDS per Gene ----
# BioMart returns multiple rows per gene due to alternative splicing transcripts.
# We pick the longest sequence per stable gene ID to avoid truncation artifacts.
human_cds_clean <- human_cds_raw %>%
  filter(coding != "Sequence unavailable") %>%
  mutate(seq_len = nchar(coding)) %>%
  group_by(ensembl_gene_id) %>%
  slice_max(seq_len, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(ensembl_gene_id, human_cds = coding)

# Save
saveRDS(human_cds_clean, file = here::here("tables", "Gene and Protein sequences", "human_cds_clean.rds"))
```

```{r}
#Download Mouse ==========================
# Prepare valid mouse homolog target map and split into 3 balanced partitions
message("Downloading Mouse coding sequences in 3 batches...")

target_map_mouse_clean <- target_map %>%
  dplyr::filter(!is.na(mmusculus_homolog_ensembl_gene) & mmusculus_homolog_ensembl_gene != "") %>%
  dplyr::distinct(mmusculus_homolog_ensembl_gene, .keep_all = TRUE)

target_map_mouse_split <- target_map_mouse_clean %>%
  dplyr::mutate(part_id = dplyr::ntile(dplyr::row_number(), 3)) %>%
  dplyr::group_split(part_id)

target_map_mouse_part1 <- target_map_mouse_split[[1]]
target_map_mouse_part2 <- target_map_mouse_split[[2]]
target_map_mouse_part3 <- target_map_mouse_split[[3]]

# Download 
mouse_cds_raw_part1 <- fetch_mouse_cds_chunked(target_map_mouse_part1, part_label = "1/3", chunk_size = 100)
mouse_cds_raw_part2 <- fetch_mouse_cds_chunked(target_map_mouse_part2, part_label = "2/3", chunk_size = 100)
mouse_cds_raw_part3 <- fetch_mouse_cds_chunked(target_map_mouse_part3, part_label = "3/3", chunk_size = 100)

#Save temporarily. Connection may be lost during the process
mouse_cds_raw_part1 %>% saveRDS(file = here("tables", "mouse_cds_raw_part1.rds"))
mouse_cds_raw_part2 %>% saveRDS(file = here("tables", "mouse_cds_raw_part2.rds"))
mouse_cds_raw_part3 %>% saveRDS(file = here("tables", "mouse_cds_raw_part3.rds"))

mouse_cds_raw_part1  =  readRDS(file = here("tables", "mouse_cds_raw_part1.rds"))
mouse_cds_raw_part2  =  readRDS(file = here("tables", "mouse_cds_raw_part2.rds"))
mouse_cds_raw_part3 = readRDS(file = here("tables", "mouse_cds_raw_part3.rds"))

# Consolidate raw Mouse CDS partitions into single master table
mouse_cds_raw <- dplyr::bind_rows(
  mouse_cds_raw_part1,
  mouse_cds_raw_part2,
  mouse_cds_raw_part3
)

# Filter for Longest Canonical/Isoform CDS per Gene ----
# BioMart returns multiple rows per gene due to alternative splicing transcripts.
# We pick the longest sequence per stable gene ID to avoid truncation artifacts.

mouse_cds_clean <- mouse_cds_raw %>%
  filter(coding != "Sequence unavailable") %>%
  mutate(seq_len = nchar(coding)) %>%
  group_by(ensembl_gene_id) %>%
  slice_max(seq_len, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(mmusculus_homolog_ensembl_gene = ensembl_gene_id, mouse_cds = coding)

# Save
saveRDS(mouse_cds_clean, file = here::here("tables", "Gene and Protein sequences", "mouse_cds_clean.rds"))

# Send completion notification
notify(
  title = "Mouse CDS downloaded!",
  msg   = c("All 3 mouse parts processed and saved successfully.")
)
```

```{r}
#Import
mouse_cds_clean = readRDS(file = here::here("tables", "Gene and Protein sequences", "mouse_cds_clean.rds"))
human_cds_clean = readRDS(file = here::here("tables", "Gene and Protein sequences", "human_cds_clean.rds"))

# -Calculate distance ----
mouse_human_cds_clean <- ortholog_map %>%
  inner_join(human_cds_clean, by = "ensembl_gene_id") %>%
  inner_join(mouse_cds_clean, by = "mmusculus_homolog_ensembl_gene")

mouse_human_cds_distance_matrix <- mouse_human_cds_clean %>%
  dplyr::mutate(
    dna_metrics = purrr::map2(
      human_cds, 
      mouse_cds, 
      ~ calculate_pairwise_dna_distance(.x, .y)
    )
  ) %>%
  tidyr::unnest(dna_metrics)

human_mouse_cds_distance <- mouse_human_cds_distance_matrix %>%
  dplyr::select(
    hgnc_symbol    = external_gene_name,
    mgi_symbol = mmusculus_homolog_associated_gene_name,
    human_ensembl  = ensembl_gene_id,
    mouse_ensembl  = mmusculus_homolog_ensembl_gene,
    dist_jc69,
    dist_k80 
  )

print(head(human_mouse_cds_distance))

#Save
saveRDS(human_mouse_cds_distance, file = here::here("tables", "Gene and Protein sequences", "human_mouse_cds_distance.rds"))
notify(title = "Alignment done", msg = "Check it out!")

```

### Is LEG status explaned by the evolutionary distance in the CDS?

```{r}
#| fig-height: 3
#| fig-width: 3
#Import
# Gene expression
dge_btm_process_genes_diff_bygene_clean_filtered =  readRDS(file = here("tables", "Functional analysis", "dge_btm_process_genes_diff_bygene_clean_filtered.rds"))

human_mouse_cds_distance = 
  readRDS(file = here::here("tables", "Gene and Protein sequences", "human_mouse_cds_distance.rds"))

#Merge
human_mouse_cds_distance_dge = 
  #LEGs
  human_mouse_cds_distance %>% 
  inner_join(dge_btm_process_genes_diff_bygene_clean_filtered,
             by = "hgnc_symbol")

human_mouse_cds_distance_dge %>% 
  dplyr::count(pathogen, status)

```

Is LEG status explaned by the evolutionary distance in the CDS?

```{r}
#| fig-height: 3
#| fig-width: 8
#Is LEG status explaned by the evolutionary distance in the CDS?
distance_legs_plot =  human_mouse_cds_distance_dge %>% 
  filter(status != "Not LEG",
         treatment %in% c("Infection", "Injury")) %>% 
  bind_rows(human_mouse_cds_distance_dge %>% 
  filter(status != "Not LEG",
         treatment %in% c("Infection", "Injury")) %>% 
    mutate(treatment = "Pooled",
           pathogen = "")) %>% 
  
  ggplot() +
  aes(
    x     = status,
    y     = dist_k80,
    color = status,
    fill  = status
  ) +
  
  facet_nested_wrap(~treatment+pathogen, nrow = 1) +

  geom_violin(alpha = 0, ) +
  # Render boxplot layer without outliers
  geom_boxplot(
    width    = 0.3,
    alpha    = 0,
    outliers = FALSE,
    # color    = "black"
  ) +
  # Apply project custom palettes and labels
  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = colors$comparison) +
  labs(
    x     = "Status",
    y     = "Kimura distance",
    color = "",
    fill  = ""
  ) +
  # Set Y-axis view boundaries
  coord_cartesian(ylim = c(0, NA)) +
  theme_vaxgo() +
  theme(
    legend.position = "top", legend.location = "plot",
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
    # STATISTICAL TEST LAYER
  stat_compare_means(
    comparisons = list(c("Shared", "Not-shared"),
                       c("Shared", "All genes"),
                       c("Not-shared", "All genes")),
    method = "wilcox.test",
    label = "p.adj", 
    vjust = 1.5,
    label.y.np = "top",
    size = 3, tip.length = 0,
    hide.ns = F
  ) 

#Save
distance_legs_plot

#Save
distance_legs_plot %>%
  ggsave(filename = here("Figures", "all_human_mouse_gsea_btms_legs_kimuradistance.png"),
         width = 8,
         height = 3)

```

Is the absolute difference in Log2FC explained by sequence distance?

```{r}
#| fig-height: 3
#| fig-width: 3
difflog2fc_distance_scatter_plot <- human_mouse_cds_distance_dge %>%
  filter(status == "All genes") %>% 
  ggplot2::ggplot() +
  ggplot2::aes(
    x     = dist_k80,
    y     = abs_log2fc_diff,
    color = status,
    fill  = status
  ) +
  geom_smooth(method = "lm", se = TRUE,
              show.legend = F) +
    #Marginal Boxplot
  # TOP  
  geom_violin(
    mapping = aes(
      x = dist_k80,         
      y = 1,
    ),
    outliers = FALSE,
    outlier.shape = NA,
    alpha = 0.5,
    width = 0.1,
    color = NA,
    show.legend = T
  ) +
  geom_boxplot(
    mapping = aes(
      x = dist_k80,         
      y = 1,
    ),
    outliers = FALSE,
    outlier.shape = NA,
    alpha = 0, linewidth = 0.2,
    color = "black",
    width = 0.1,
    show.legend = FALSE
  ) +
  #Add correlation stats
  stat_cor( 
    method = "spearman",
    cor.coef.name = "rho", 
    parse = TRUE,      
    label.y = 0.10,
    label.x = 0.05,
    size = 3.5, 
    show.legend = F
  ) +
  # Line equation
  stat_regline_equation(
    aes(label = ..eq.label..),
    label.y = 0.20,
    label.x = 0.05,       
    size = 3.5
  ) +
  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = colors$comparison) +
  scale_x_continuous(expand = c(0.01, 0.01)) +
  scale_y_continuous(limits = c(0, 1.2)) +
  labs(
    x     = "Kimura Distance",
    y     = "|ΔLog2FC|",
    fill  = "",
    color = ""
  ) +
  theme_vaxgo() +
  theme(legend.position = "top",
        legend.location = "plot") +
  guides(fill = guide_legend(override.aes = list(alpha = 1)))

difflog2fc_distance_scatter_plot  +
  scale_y_continuous(limits = c(0, 1.1)) 
  


#Save
difflog2fc_distance_scatter_plot %>%
  ggsave(filename = here("Figures", "all_human_mouse_gsea_btms_legs_kimuradistance_divergence.png"),
         width = 3,
         height = 3)

```

```{r}
#| fig-height: 3
#| fig-width: 8
difflog2fc_distance_scatter_condition_plot <- human_mouse_cds_distance_dge %>% 
  filter(status != "Not LEG",
         treatment %in% c("Infection", "Injury")) %>% 
  bind_rows(human_mouse_cds_distance_dge %>% 
  filter(status != "Not LEG",
         treatment %in% c("Infection", "Injury")) %>% 
    mutate(treatment = "Pooled",
           pathogen = "")) %>% 
  ggplot() +
  aes(
    x     = dist_k80,
    y     = abs_log2fc_diff,
    color = status,
    fill  = status
  ) +
  facet_nested_wrap(~treatment+pathogen, nrow = 1) +
  geom_smooth(method = "lm", se = TRUE) +
    #Marginal Boxplot
  # TOP  
  geom_violin(
    mapping = aes(
      x = dist_k80,         
      y = 3,
    ),
    outliers = FALSE,
    outlier.shape = NA,
    alpha = 0.5,
    width = 0.5,
    color = NA,
    show.legend = FALSE
  ) +
  geom_boxplot(
    mapping = aes(
      x = dist_k80,         
      y = 3,
    ),
    outliers = FALSE,
    outlier.shape = NA,
    alpha = 0, linewidth = 0.2,
    color = "black",
    width = 0.5,
    show.legend = FALSE
  ) +
  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = colors$comparison) +
  scale_x_continuous(expand = c(0.01, 0.01)) +
  scale_y_continuous(expand = c(0.01, 0.01)) +
  labs(
    x     = "Kimura Distance",
    y     = "|ΔLog2FC|",
    fill  = "",
    color = ""
  ) +
  theme_vaxgo() +
  theme(legend.position = "top",
        legend.location = "plot") +
  guides(fill = guide_legend(override.aes = list(alpha = 1)))

difflog2fc_distance_scatter_condition_plot 

#Save
difflog2fc_distance_scatter_condition_plot %>%
  ggsave(filename = here("Figures", "all_human_mouse_gsea_btms_legs_kimuradistance_divergence_byCondition.png"),
         width = 8,
         height = 3)

```
