
#Glassplot

```{r}
library(ggplot2)
library(ggfx)
library(dplyr)
library(MASS)
library(here) # Para salvar

# ---------------------------------------------------------
# 1. GERAR DADOS
# ---------------------------------------------------------
set.seed(101)
n_genes <- 1000

# Correlação alvo (r = 0.9)
target_r <- 0.9
sigma_matrix <- matrix(c(1, target_r, target_r, 1), nrow = 2)

dados_brutos <- mvrnorm(n = n_genes, mu = c(0, 0), Sigma = sigma_matrix)

df_log2fc <- data.frame(
  human_logfc = dados_brutos[, 1] * 2.5, 
  mouse_logfc = dados_brutos[, 2] * 2.5 
) %>%
  mutate(
    # --- LÓGICA DE RANKING (Para as CAMADAS do Blur) ---
    distancia = abs(human_logfc) + abs(mouse_logfc),
    fake_pval = 1 / (distancia + runif(n(), 0, 2)), 
    pval_rank = rank(fake_pval), # 1 a 1000
    
    # --- LÓGICA DE COR (Para o FILL/COLOR) ---
    # Aqui criamos grupos que NÃO dependem do ranking.
    # Simula vias biológicas diferentes misturadas nos dados.
    pathway_group = sample(c("Inflammation", "Cell Cycle"), 
                           n(), replace = TRUE)
  )

# Calcular R² para o texto
r_squared <- round(cor(df_log2fc$human_logfc, df_log2fc$mouse_logfc)^2, 2)
label_text <- paste0("R² ≈ ", r_squared)

# ---------------------------------------------------------
# 2. DEFINIÇÃO DE CORES
# ---------------------------------------------------------
# Cores para os grupos (e não para o rank)
my_colors <- c(
  "Inflammation" = "#607d8b", # Vermelho
  "Cell Cycle"   = "#455a64"  # Azul
)

# ---------------------------------------------------------
# 3. PLOT COM GGFX
# ---------------------------------------------------------
glass_plot <- ggplot(df_log2fc, aes(x = human_logfc, y = mouse_logfc)) +
  
  # Guias de fundo
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey80") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey80") +
  
  # --- CAMADA 1: BACKGROUND (Todos os 1000 pontos) ---
  # Blur forte. Usamos shape 16 (sem borda) para um desfoque mais suave.
  with_blur(
    geom_point(
      data = . %>% filter(pval_rank > 500),
      aes(color = pathway_group), # Colore pelo grupo
      size = 3,
      alpha = 0.5,
      shape = 16
    ),
    sigma = 5 # Blur alto
  ) +
  
  # --- CAMADA 2: MIDGROUND (Top 500 Rank) ---
  # Blur médio. Filtrado pelo Rank, mas colorido pelo Grupo.
  with_blur(
    geom_point(
      data = . %>% filter(pval_rank <= 500),
      aes(color = pathway_group),
      size = 3,
      alpha = 0.8,
      shape = 16
    ),
    sigma = 2 # Blur médio
  ) +
  
  # --- CAMADA 3: FOREGROUND (Top 100 Rank - Foco) ---
  # Nítido. Usamos shape 21 para ter borda preta e preenchimento (fill) colorido.
  with_shadow(
    geom_point(
    data = . %>% filter(pval_rank <= 300),
    aes(fill = pathway_group,
        color = pathway_group), # Note que aqui usamos FILL, não color
    size = 3,
    # color = "black",
    shape = 21,                # Círculo preenchido
    stroke = 0.2,                # Espessura da borda preta
    alpha = 1
  ),
  x_offset = 2,
  y_offset = 2,
  sigma = 3
  ) +
  
  geom_smooth(method = "lm", se = FALSE, color = "#c59d45", size = 1.5) +
  
  # --- TEMA E ESCALAS ---
  scale_color_manual(values = my_colors) + # Para as camadas 1 e 2 (blur)
  scale_fill_manual(values = my_colors) +  # Para a camada 3 (foco)
  coord_cartesian(xlim = c(-7, 7), ylim = c(-7, 7)) +
  theme_classic() +
  labs(
    x = "Human Log2FC", 
    y = "Mouse Log2FC", 
    fill = "Pathway", 
    color = "Pathway"
  ) +
  theme(
    axis.line = element_line(color = "black", linewidth = 1),
    axis.text = element_text(size = 12, color = "black"),
    axis.title = element_text(size = 14, color = "black"),
    legend.position = "top"
  )

# Exibir o plot
print(glass_plot)

# Salvar
ggsave(filename = here("Figures", "glassplot_test.png"), plot = glass_plot, width = 6, height = 6)


```





### 2.2. Correlation: Genes, samples

```{r}
#Import
btm_annotation_genes = read_csv(file = here("tables", "Gene sets", "btm_annotation_genes.csv"))
all_human_mouse_metadata = readRDS(here("tables", "processed_data", "all_human_mouse_metadata.rds"))


#Filter conditions with too many human samples to reduce computation time
set.seed(420)
all_human_mouse_metadata_filtered =
    all_human_mouse_metadata %>% 
   filter(!(pathogen %in% c("Trauma", "Influenza", "Burn") & organism == "Human")) %>% 
  
   bind_rows(
     #Sample trauma patients
     all_human_mouse_metadata %>%
      filter(pathogen == "Trauma", treatment == "Trauma", organism == "Human") %>% 
      group_by(timepoint) %>% 
      slice_sample(n = 30, replace = F),
     #Sample Burn patients
     all_human_mouse_metadata %>%
      filter(pathogen == "Burn", treatment == "Burn", organism == "Human") %>% 
      group_by(timepoint) %>% 
      slice_sample(n = 30, replace = F),
     #Sample Fluad patients
     all_human_mouse_metadata %>% 
       filter(participant_id %in% c(all_human_mouse_metadata %>%
                                      filter(pathogen == "Influenza", organism == "Human") %>% 
                                      pull(participant_id) %>% 
                                      unique() %>% 
                                      sample(size = 30, replace = FALSE)))
  )

#Import samples and annotate with metadata
log2fc_samples = readRDS(here("tables", "all_human_mouse_samples_log2fc.rds")) 

log2fc_samples_annotated = log2fc_samples %>% 
  inner_join(all_human_mouse_metadata_filtered, 
             by = join_by("geo_sample"))   

#Check
log2fc_samples_annotated %>% 
  select(pathogen, organism, timepoint, participant_id) %>% 
  distinct() %>% 
  dplyr::count(pathogen, organism, timepoint) 


#DEGs
all_degs_human_mouse = 
  readRDS(file = here("tables", "Differential gene expression", "all_human_mouse_dge_limma_degs_matched.rds"))

all_degs_human_mouse %>% 
  dplyr::count(pathogen, organism, timepoint)


#Get DEGs from different organisms
mouse_degs = all_degs_human_mouse %>%
  group_by(pathogen, timepoint) %>%
  filter(adj_p_val <= 0.05, organism == "Mouse") %>%
  select(pathogen, human_symbol, timepoint) %>% 
  ungroup()

human_degs = all_degs_human_mouse %>%
  group_by(pathogen, timepoint) %>%
  filter(adj_p_val <= 0.05, organism == "Human") %>%
  select(pathogen, human_symbol, timepoint) %>% 
  ungroup()

#Get shared DEGs 
human_mouse = mouse_degs %>%
  full_join(human_degs %>%
               ungroup() %>%
               distinct(),
             by = join_by("human_symbol", "timepoint", "pathogen")) %>% 
  distinct()

human_mouse %>% 
  dplyr::count(pathogen, timepoint)

```

```{r}
# #Input
# filename = "all_human_mouse"
# human_mouse_log2fc_samples = log2fc_samples_annotated

#Input
filename = "all_human_mouse_DEGs"

human_mouse_log2fc_samples =
  log2fc_samples_annotated %>%
  inner_join(human_mouse %>%
               mutate(timepoint = as.character(timepoint)),
             by = join_by(timepoint, human_symbol, pathogen))
  

```



By pathogen, all vs all.
```{r}
library(tidyverse)
library(Hmisc)
library(here)
library(janitor)

# 1. Define Inputs and Gene Subsets -----------------------------------------
# We create a list to automate the loop over both sets
gene_sets_list <- list(
  "AllGenes" = human_mouse_log2fc_samples,
  "Immune"   = human_mouse_log2fc_samples %>% 
                 filter(human_symbol %in% (btm_annotation_genes %>% 
                                          filter(module_category == "immune") %>% 
                                          pull(symbol) %>% unique()))
)


# 2. Main Processing Loop ----------------------------------------------------
# map_dfr iterates through "AllGenes" and "Immune", stacking the results
final_correlations <- map_dfr(names(gene_sets_list), function(set_name) {
  
  message(paste("Processing gene set:", set_name))
  current_data <- gene_sets_list[[set_name]]
  
  # Data Preparation: Long format to Wide Matrix
  genes_samples_long <- current_data %>%
    select(pathogen, organism, human_symbol, participant_id, timepoint, log2fc) %>%
    distinct() %>%
    mutate(condition = paste0(pathogen, "_", participant_id, "_", timepoint))

  samples_annotation <- genes_samples_long %>%
    select(pathogen, organism, participant_id, timepoint, condition) %>%
    distinct() 

  genes_samples_wide <- genes_samples_long %>%
    select(human_symbol, condition, log2fc) %>% 
    distinct() %>%
    pivot_wider(names_from = "condition", values_from = "log2fc", values_fn = mean) %>%
    column_to_rownames("human_symbol")

  # 3. Pathogen-Specific Correlation Calculation -----------------------------
  pathogen_list <- unique(samples_annotation$pathogen)
  
  map_dfr(pathogen_list, function(path) {
    
    pathogen_samples <- samples_annotation %>% 
      filter(pathogen == path) %>% 
      pull(condition) 
    
    # Subset matrix and drop NAs
    matrix_subset <- genes_samples_wide %>%
      select(all_of(intersect(pathogen_samples, colnames(.)))) %>%
      # drop_na() %>%
      as.matrix()
    
    if(nrow(matrix_subset) <= 4 | ncol(matrix_subset) < 2) {
      warning(paste("Skipping pathogen:", path, "- only", nrow(matrix_subset), "shared genes found."))
      return(NULL)
    }
    
    # Calculate Spearman Correlation
    cor_obj <- rcorr(matrix_subset, type = "spearman")
    
    # Tidy extraction of Rho and P-values
    r_long <- as.data.frame(cor_obj$r) %>%
      rownames_to_column("condition1") %>%
      pivot_longer(-condition1, names_to = "condition2", values_to = "rho")
    
    p_long <- as.data.frame(cor_obj$P) %>%
      rownames_to_column("condition1") %>%
      pivot_longer(-condition1, names_to = "condition2", values_to = "p_val")
    
    # Join with Metadata and Filter Duplicates
    r_long %>%
      left_join(p_long, by = c("condition1", "condition2")) %>%
      filter(condition1 < condition2) %>%
      left_join(samples_annotation %>% select(condition, organism1 = organism, timepoint1 = timepoint), 
                by = c("condition1" = "condition")) %>%
      left_join(samples_annotation %>% select(condition, organism2 = organism, timepoint2 = timepoint), 
                by = c("condition2" = "condition")) %>%
      mutate(pathogen = path, gene_set = set_name)
  })
})

# 4. Statistical Summary per Comparison and Timepoint -----------------------
correlation_summary <- final_correlations %>% 
  # Keep Human as reference for cross-species comparisons
  filter(!(organism1 == "Mouse" & organism2 == "Human")) %>% 
  mutate(comparison = paste0(organism1, " vs. ", organism2),
         timepoint_comparison = paste0(timepoint1, "-", timepoint2)) %>% 
  group_by(gene_set, pathogen, comparison, timepoint1, timepoint2) %>% 
  summarise(
    mean_rho = mean(rho, na.rm = TRUE),
    sd_rho   = sd(rho, na.rm = TRUE),
    n        = n(),
    se       = sd_rho / sqrt(n),
    ci_low   = mean_rho - (qt(0.975, df = n - 1) * se),
    ci_high  = mean_rho + (qt(0.975, df = n - 1) * se),
    .groups = "drop"
  )

# 5. Final Filtering (Matched Timepoints and 24h Reference) -----------------
correlation_summary_filtered <- bind_rows(
  # Matched timepoints (e.g., 2h vs 2h)
  correlation_summary %>% 
    filter(!(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & comparison == "Human vs. Mouse"),
           timepoint1 == timepoint2),
  
  # 24h reference for acute infections
  correlation_summary %>% 
    filter(pathogen %in% c("Staphylococcus aureus", "Escherichia coli"),
           comparison == "Human vs. Mouse",
           timepoint1 == "24h")
) %>%
  mutate(timepoint_label = if_else(pathogen %in% c("Staphylococcus aureus", "Escherichia coli"), 
                                   timepoint2, timepoint1),
         timepoint_label = as.character(timepoint_label),
         timepoint_label = str_replace(timepoint_label,"Early hours", "Early\nhours"),
         timepoint = fct_relevel(timepoint_label, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
          pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "Staphylococcus aureus", "Escherichia coli", "Burn", "Trauma"),
         gene_set = str_replace(gene_set, "AllGenes", "All genes")) 

correlation_summary_filtered
# Save Output
saveRDS(correlation_summary_filtered, file = here("tables", paste0(filename, "_gene_sets_correlation_summary.rds")))
```


Horizontal
```{r fig.height=3, fig.width=16}
#Import
correlation_summary_filtered_DEGs = readRDS(file = here("tables", "Differential gene expression", "all_human_mouse_DEGs_gene_sets_correlation_summary.rds")) %>% 
  mutate(filter = "DEGs")
correlation_summary_filtered_all = readRDS(file = here("tables", "Functional analysis", "all_human_mouse_gene_sets_correlation_summary.rds")) %>% 
  mutate(filter = "No filter")

correlation_summaries =  
  bind_rows(correlation_summary_filtered_DEGs,
            correlation_summary_filtered_all) %>% 
  mutate(timepoint = str_replace(timepoint, "Day ", "Day\n"),
         timepoint = as.character(timepoint),
         timepoint = fct_relevel(timepoint, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day\n1", "Day\n3", "Day\n7")),
         filter = fct_relevel(filter, c("No filter", "DEGs")),
         pathogen = as.character(pathogen),
         pathogen = str_replace(pathogen, "Staphylococcus aureus", "S. aureus") %>% 
           str_replace("Escherichia coli", "E. coli"),
         pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "S. aureus", "E. coli", "Burn", "Trauma"))

#Plot
correlation_long_all_mouse_human_timepoints_summary_plot = correlation_summaries %>% 
  ggplot() +
  aes(x = timepoint,
      y = mean_rho,
      group = interaction(comparison, gene_set), 
      color = comparison,
      fill = comparison) +
  
  facet_nested_wrap(~filter + pathogen, scales = "free_x", nrow = 1) +
  geom_line(aes(linetype = gene_set,
                alpha = gene_set),
            linewidth = 0.5) +
    geom_errorbar(aes(ymin = ci_low, 
                    ymax = ci_high,
                    alpha = gene_set), 
                width = 0.2,
                size = 1,
                linetype = "solid") + 
   geom_point(aes(alpha = gene_set,
                  fill = comparison),
             size = 2, 
             pch = 21,
             color = "black",
             stroke = 0.5) + 

  scale_alpha_manual(values = c("Immune" = 1, "All genes" = 0.5)) +

  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_continuous(expand = c(0.01, 0.01),
                     limits = c(-0.1, 1)
                     ) +
  theme_vaxgo() +
  scale_color_manual(values = colors$comparison_versus) +
  scale_fill_manual(values = colors$comparison_versus) +
  labs(
    # title = gene_set_name,
       x = "Time",
       y = "Correlation",
       shape = "Type",
       linetype = "Gene category (Line)",
       alpha = "Gene category (Opacity)",
       color = "Comparison") +
  geom_hline(yintercept = 0, linetype = 2, color = "black") +
    theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
          axis.text.y = element_text(size = 8),
          strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
          panel.spacing = unit(1.1, "lines"),
          plot.margin = margin(0, 0.3, 0, 0, unit = "cm"),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
          legend.text = element_text(size = 10, hjust = 0),
          legend.title = element_text(size = 10),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)
    ) +
  guides(fill = "none",
         linetype = guide_legend(order = 1, linewidth = 5), 
        alpha = guide_legend(order = 2, linewidth = 3),
        color = guide_legend(override.aes = list(linewidth = 3)))

correlation_long_all_mouse_human_timepoints_summary_plot

correlation_long_all_mouse_human_timepoints_summary_plot %>%
  ggsave(filename = here("Figures", paste0("all_human_mouse_all_DEGs_correlation_timepoints_summary_plot_horizontal.png")),
         width = 16,
         height = 3)

```

Vertical
```{r fig.height=6, fig.width=10}
#Import
correlation_summary_filtered_DEGs = readRDS(file = here("tables", "Differential gene expression", "all_human_mouse_DEGs_gene_sets_correlation_summary.rds")) %>% 
  mutate(filter = "DEGs")
correlation_summary_filtered_all = readRDS(file = here("tables", "Functional analysis", "all_human_mouse_gene_sets_correlation_summary.rds")) %>% 
  mutate(filter = "No filter")

correlation_summaries =  
  bind_rows(correlation_summary_filtered_DEGs,
            correlation_summary_filtered_all) %>% 
  mutate(timepoint = str_replace(timepoint, "Day ", "Day\n"),
         timepoint = as.character(timepoint),
         timepoint = fct_relevel(timepoint, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day\n1", "Day\n3", "Day\n7")),
         filter = fct_relevel(filter, c("No filter", "DEGs")),
         pathogen = as.character(pathogen),
         pathogen = str_replace(pathogen, "Staphylococcus aureus", "S. aureus") %>% 
           str_replace("Escherichia coli", "E. coli"),
         pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "S. aureus", "E. coli", "Burn", "Trauma"))

#Plot
correlation_long_all_mouse_human_timepoints_summary_plot = correlation_summaries %>% 
  ggplot() +
  aes(x = timepoint,
      y = mean_rho,
      group = interaction(comparison, gene_set), 
      color = comparison,
      fill = comparison) +
  
  facet_nested_wrap(~filter + pathogen, scales = "free_x", nrow = 2) +
  geom_line(aes(linetype = gene_set,
                alpha = gene_set),
            linewidth = 0.5) +
    geom_errorbar(aes(ymin = ci_low, 
                    ymax = ci_high,
                    alpha = gene_set), 
                width = 0.2,
                size = 1,
                linetype = "solid") + 
   geom_point(aes(alpha = gene_set,
                  fill = comparison),
             size = 2, 
             pch = 21,
             color = "black",
             stroke = 0.5) + 

  scale_alpha_manual(values = c("Immune" = 1, "All genes" = 0.5)) +

  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_continuous(expand = c(0.01, 0.01),
                     limits = c(-0.1, 1)
                     ) +
  theme_vaxgo() +
  scale_color_manual(values = colors$comparison_versus) +
  scale_fill_manual(values = colors$comparison_versus) +
  labs(
    # title = gene_set_name,
       x = "Time",
       y = "Correlation",
       shape = "Type",
       linetype = "Gene category (Line)",
       alpha = "Gene category (Opacity)",
       color = "Comparison") +
  geom_hline(yintercept = 0, linetype = 2, color = "black") +
    theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
          axis.text.y = element_text(size = 8),
          strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
          panel.spacing = unit(1.1, "lines"),
          plot.margin = margin(0, 0.3, 0, 0, unit = "cm"),
          legend.position = "top",
          legend.direction = "horizontal",
          panel.spacing.y = unit(0, "lines"),
          legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
          legend.text = element_text(size = 10, hjust = 0),
          legend.spacing.y = unit(0, "cm"),
          legend.title = element_text(size = 10),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)
    ) +
  guides(fill = "none",
         linetype = guide_legend(order = 1, linewidth = 5, nrow = 2, byrow = TRUE), 
        alpha = guide_legend(order = 2, linewidth = 3, nrow = 2, byrow = TRUE,),
        color = guide_legend(override.aes = list(linewidth = 3), nrow = 3, byrow = TRUE,))

correlation_long_all_mouse_human_timepoints_summary_plot

correlation_long_all_mouse_human_timepoints_summary_plot %>%
  ggsave(filename = here("Figures", paste0("all_human_mouse_all_DEGs_correlation_timepoints_summary_plot_vertical.png")),
         width = 10,
         height = 6)

```







### 2.3. Correlation: RRHO, means

```{r}
# 1. Identify all unique pathogen and timepoint combinations
combinations <- human_mouse_log2fc_avg_wide %>%
  distinct(pathogen, timepoint_comparison)

# 2. Initialize a list to store results from each iteration
all_rrho_results <- list()

for(i in 1:nrow(combinations)) {
  
  curr_pathogen <- combinations$pathogen[i]
  curr_timepoint <- combinations$timepoint_comparison[i]
  
  # A. Filter and Prepare Differential Expression (DDE) Scores
  rrho_input_data <- human_mouse_log2fc_avg_wide %>% 
    filter(pathogen == curr_pathogen, timepoint_comparison == curr_timepoint) %>%
    filter(!is.na(p_adj_human), !is.na(p_adj_mouse), 
           !is.na(mean_log2fc_human), !is.na(mean_log2fc_mouse)) %>%
    mutate(
      # Handle p-values of 0 to avoid infinite logs
      p_adj_human = ifelse(p_adj_human == 0, .Machine$double.xmin, p_adj_human),
      p_adj_mouse = ifelse(p_adj_mouse == 0, .Machine$double.xmin, p_adj_mouse),
      # Calculate DDE score: -log10(p-value) * sign of Fold Change
      score_human = -log10(p_adj_human) * sign(mean_log2fc_human),
      score_mouse = -log10(p_adj_mouse) * sign(mean_log2fc_mouse)
    )

  # Skip if the gene list is too small (RRHO2 requires a minimum set)
  if(nrow(rrho_input_data) < 100) next 

  # B. Prepare Sorted Gene Lists for RRHO2
  # Sorting from highest to lowest DDE score is mandatory for RRHO analysis
  gene_list1 <- rrho_input_data %>% 
    arrange(desc(score_human)) %>%
    select(Genes = human_symbol, DDE = score_human) %>% 
    as.data.frame()
    
  gene_list2 <- rrho_input_data %>% 
    arrange(desc(score_mouse)) %>%
    select(Genes = human_symbol, DDE = score_mouse) %>% 
    as.data.frame()

  # Identify the Biological Split Point (where DDE score crosses zero)
  # This marks the boundary between Up-regulated and Down-regulated genes
  split_h_val <- sum(gene_list1$DDE > 0)
  split_m_val <- sum(gene_list2$DDE > 0)

  # C. Execute RRHO2
  # Using 'try' to skip iterations that might encounter mathematical errors
  rrho_obj <- try(RRHO2_initialize(gene_list2, gene_list1, 
                                   labels = c("Mouse", "Human"), 
                                   boundary = 0.01,
                                   log10.ind = TRUE, 
                                   method = "hyper"), silent = TRUE)
  
  if(inherits(rrho_obj, "try-error")) next

  # Transform the hypermat (matrix) into a long format dataframe for ggplot2
  temp_df <- as.data.frame(rrho_obj$hypermat) %>%
    mutate(Mouse_Rank = row_number()) %>%
    pivot_longer(cols = -Mouse_Rank, names_to = "Human_Rank", values_to = "LogP") %>%
    mutate(
      Human_Rank = as.numeric(gsub("V", "", Human_Rank)),
      pathogen = curr_pathogen,       
      timepoint = curr_timepoint,
      # Store the split points to draw exact quadrant boundaries later
      split_h = split_h_val,
      split_m = split_m_val
    )

  all_rrho_results[[i]] <- temp_df
}

# Combine all results into a single final dataframe
final_rrho_df <- bind_rows(all_rrho_results)

# Save the processed data for visualization
final_rrho_df %>% 
  saveRDS(file = here("tables", "all_human_mouse_rrho_df.rds"))
 

```

Extract RRHO2 results and gene lists by quadrant
```{r}
library(dplyr)
library(tidyr)
library(RRHO2) # Ensure the library is loaded

# 1. Identify all unique pathogen and timepoint combinations
combinations <- human_mouse_log2fc_avg_wide %>%
  distinct(pathogen, timepoint_comparison)

# 2. Initialize lists to store results
plot_data_list <- list()
gene_extraction_list <- list() # New list for the extracted genes
raw_objects_list <- list()

# --- MAIN LOOP ---
for(i in 1:nrow(combinations)) {
  
  curr_pathogen <- combinations$pathogen[i]
  curr_timepoint <- combinations$timepoint_comparison[i]
  # Create a unique name for each comparison
  iter_name <- paste(curr_pathogen, curr_timepoint, sep = "_")
  
  # --- A. Prepare Data ---
  rrho_input_data <- human_mouse_log2fc_avg_wide %>% 
    filter(pathogen == curr_pathogen, timepoint_comparison == curr_timepoint) %>%
    filter(!is.na(p_adj_human), !is.na(p_adj_mouse), 
           !is.na(mean_log2fc_human), !is.na(mean_log2fc_mouse)) %>%
    mutate(
      p_adj_human = ifelse(p_adj_human == 0, .Machine$double.xmin, p_adj_human),
      p_adj_mouse = ifelse(p_adj_mouse == 0, .Machine$double.xmin, p_adj_mouse),
      score_human = -log10(p_adj_human) * sign(mean_log2fc_human),
      score_mouse = -log10(p_adj_mouse) * sign(mean_log2fc_mouse)
    )

  if(nrow(rrho_input_data) < 100) next 

  gene_list_human <- rrho_input_data %>% 
    arrange(desc(score_human)) %>%
    select(Genes = human_symbol, DDE = score_human) %>% as.data.frame()
    
  gene_list_mouse <- rrho_input_data %>% 
    arrange(desc(score_mouse)) %>%
    select(Genes = human_symbol, DDE = score_mouse) %>% as.data.frame()

  split_human <- sum(gene_list_human$DDE > 0)
  split_mouse <- sum(gene_list_mouse$DDE > 0)

  # --- B. Execute RRHO2 ---
  rrho_obj <- try(RRHO2_initialize(gene_list_mouse, gene_list_human, 
                                   labels = c("Mouse", "Human"), 
                                   boundary = 0.01, log10.ind = TRUE, method = "hyper"), 
                  silent = TRUE)
  
  if(inherits(rrho_obj, "try-error")) next

  # Store the raw object
  raw_objects_list[[iter_name]] <- rrho_obj

  # --- C. Save Plot Data ---
  temp_df <- as.data.frame(rrho_obj$hypermat) %>%
    mutate(Mouse_Rank = row_number()) %>%
    pivot_longer(cols = -Mouse_Rank, names_to = "Human_Rank", values_to = "LogP") %>%
    mutate(
      Human_Rank = as.numeric(gsub("V", "", Human_Rank)),
      pathogen = curr_pathogen,        
      timepoint = curr_timepoint,
      split_h = split_human,
      split_m = split_mouse
    )
  plot_data_list[[iter_name]] <- temp_df
}

# --- 3. CONSOLIDATE INTO final_results ---
final_results <- list(
  plot_data = bind_rows(plot_data_list),
  objects = raw_objects_list
)


#Extract gene lists
# 1. Define the quadrants we want to extract
quadrants <- c("uu", "dd", "ud", "du")

# 2. Use map_dfr to iterate through each comparison and quadrant
all_extracted_genes <- imap_dfr(final_results$objects, function(obj, condition_name) {
  
  # For each object (comparison), iterate through the 4 quadrants
  map_dfr(quadrants, function(q) {
    
    # Construct the internal names used by RRHO2 (e.g., genelist_uu and gene_list1_uu)
    genelist_slot <- paste0("genelist_", q)
    gene_item_slot <- paste0("gene_list1_", q)
    
    # Extract the list of genes
    genes <- obj[[genelist_slot]][[gene_item_slot]]
    
    # Only create a dataframe if there are genes in that quadrant
    if (!is.null(genes) && length(genes) > 0) {
      data.frame(
        genes = as.character(genes),
        condition = condition_name,
        comparison = genelist_slot,
        stringsAsFactors = FALSE
      )
    } else {
      NULL
    }
  })
}) %>% 
  separate(condition, into = c("pathogen", "timepoint_comparison"), sep = "_") %>% 
  separate(comparison, into = c("del", "direction"), sep = "_") %>%
  select(-del) %>% 
  mutate(direction = case_when(direction == "uu" ~ "Up-Up",
                               direction == "ud" ~ "Up-Down",
                               direction == "dd" ~ "Down-Down",
                               direction == "du" ~ "Down-Up",
                               T~direction))


final_results_final = list(
  plotdata = final_results$plot_data, 
  rrho2_obj = final_results$objects, 
     genes = all_extracted_genes)


final_results_final %>% saveRDS(file = here("tables", "all_human_mouse_rrho_df.rds"))

```


#### Plot RRHO Heatmap
```{r fig.height=8, fig.width=12}
library(tidyverse)
library(ggh4x)

final_rrho_list = readRDS(file = here("tables", "all_human_mouse_rrho_df.rds"))

final_rrho_df_plot = final_results_final$plotdata

final_rrho_df_plot <- final_rrho_df %>%
  mutate(pathogen = as.character(pathogen),
    pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "Staphylococcus aureus", "Escherichia coli", "Burn", "Trauma")) %>%
    #Reorder timepoints
  separate(timepoint, into = c("timepoint_mouse", "timepoint_human"), remove = F) %>% 
  mutate(timepoint_mouse = fct_relevel(timepoint_mouse, c("2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7"))) %>% 
  arrange(timepoint_mouse) %>%
  mutate(timepoint = fct_inorder(timepoint)) %>% 
  mutate(
    LogP_Final = ifelse(LogP > 50, 50, LogP)
  )
    

# --- Step 2: Generate the Visualization ---
rrho_plot <- ggplot() +
    geom_raster(data = final_rrho_df_plot, 
              aes(x = Human_Rank, y = Mouse_Rank, fill = LogP_Final),
              interpolate = T) +
  
  # C. Faceting: Using ggh4x for nested labels
  facet_nested_wrap(~pathogen + timepoint,
                    scales = "free", 
                    nrow = 4) +
  
  # D. Scale Adjustments
  scale_x_reverse(expand = c(0,0)) + 
  scale_y_reverse(expand = c(0,0)) +

 scale_fill_gradientn(
    colors = c("white", "#4361ee", "#46eec6"),
    values = c(0, 0.2, 1, 1.3), 
    limits = c(0, 50),
    name = "-log10(p-value)",
    breaks = c(0, 25, 50),
    labels = c("0", "25", "≥50"),
    na.value = "black"
  ) +
  # F. Labels and Styling
  labs(
    title = "Comparative RRHO2 Analysis",
    x = "Human Gene Rank (Up-to-Down)",
    y = "Mouse Gene Rank (Up-to-Down)"
  ) +
  theme_vaxgo() + 
  theme(aspect.ratio = 1,
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.key.width = unit(0.5, "cm"),
        legend.key.height = unit(0.2, "cm"), 
        legend.text = element_text(size = 10, 
                                   # hjust = 1, vjust = 1, 
                                   angle = 0),
        legend.title = element_text(size = 10),
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(0,0,-10,0),
        plot.margin = margin(0, 0, 0, 0, unit = "cm"),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1))

# Display and Save
print(rrho_plot)

ggsave(rrho_plot, 
       filename = here("Figures", "all_human_mouse_rrho_heatmap.png"),
       width = 12, height = 8)
```



#### Plot raw ranks
```{r fig.height=8, fig.width=12}
# Prepare data
rank_data <- human_mouse_log2fc_avg_wide %>%
  filter(!is.na(mean_log2fc_human) | !is.na(mean_log2fc_mouse),
         !is.na(p_adj_human), !is.na(p_adj_mouse)) %>%
  group_by(pathogen, timepoint_comparison) %>%
  mutate(
    score_human = -log10(p_adj_human) * sign(mean_log2fc_human),
    score_mouse = -log10(p_adj_mouse) * sign(mean_log2fc_mouse),
    Human_Rank_Val = rank(score_human),
    Mouse_Rank_Val = rank(score_mouse),
  ) %>%
  ungroup() %>% 
  mutate(pathogen = as.character(pathogen),
    pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "Staphylococcus aureus", "Escherichia coli", "Burn", "Trauma"))

# Calcular os limites por grupo
thresholds <- rank_data %>%
  group_by(pathogen, timepoint_comparison) %>%
  summarise(
    # Encontra o maior rank entre os genes com log2fc negativo
    v_line = max(Human_Rank_Val[mean_log2fc_human < 0], na.rm = TRUE),
    h_line = max(Mouse_Rank_Val[mean_log2fc_mouse < 0], na.rm = TRUE),
    .groups = "drop"
  )

# 2. Plotar como Heatmap de Densidade (Raster)
rank_density_heatmap = rank_data %>% 
  ggplot() + 
  aes(x = Human_Rank_Val, y = Mouse_Rank_Val) +
  facet_nested_wrap(~pathogen + timepoint_comparison,
                    scales = "free", 
                    nrow = 4,
                    strip = strip_nested(size = "variable"), 
                    axes = "all") +
 stat_density_2d(aes(fill = after_stat(density / sum(density))), 
                  geom = "raster", 
                  contour = F, 
                  n = 50) +    
  scale_fill_gradientn(
  colors = c("white", "#4361ee", "#4DBBD5FF", "#46eec6"),
  na.value = "white", 
  labels = scales::label_percent(), 
  name = "Density (%)",
  guide = guide_colorbar(
    barwidth = 1, 
    barheight = 8, 
    title.position = "top", 
    title.hjust = 0.5,
    ticks.colour = "black",
    frame.colour = "black"
  )
) +
  scale_x_reverse(expand = c(0,0)) +
  scale_y_reverse(expand = c(0,0)) +
  # Divide by threshold
  geom_vline(data = thresholds, aes(xintercept = v_line), 
             linetype = "solid", color = "white", alpha = 1, linewidth = 2) +
  geom_hline(data = thresholds, aes(yintercept = h_line), 
             linetype = "solid", color = "white", alpha = 1, linewidth = 2) +
  labs(
    title = "Rank-Rank Density Heatmap",
    subtitle = "Normalized density per condition",
    x = "Human Gene Rank",
    y = "Mouse Gene Rank"
  ) +
  theme_vaxgo() + 
  theme(aspect.ratio = 1,
        axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
        legend.position = "right",
        legend.direction = "vertical",
        legend.key.width = unit(1, "cm"),
        legend.text = element_text(size = 8, 
                                   # hjust = 1, vjust = 1, 
                                   angle = 0),
        legend.title = element_text(size = 8),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1))


rank_density_heatmap


rank_density_heatmap %>% 
  ggsave(filename = here("Figures", "all_human_mouse_allgenes_rankrank_heatmap.png"),
         width = 12,
         height = 8)
```







#### Gene list analysis


```{r fig.height=10, fig.width=10}
final_rrho_list = readRDS(file = here("tables", "all_human_mouse_rrho_df.rds"))

gene_list_df = final_rrho_list$genes %>% 
  dplyr::count(pathogen, timepoint_comparison, direction) %>% 
  separate(timepoint_comparison, into = c("timepoint_mouse", "timepoint_human"), remove = F, sep = "-") %>% 
  mutate(timepoint_mouse = fct_relevel(timepoint_mouse, c("Early", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         timepoint_human = fct_relevel(timepoint_mouse, c("Early", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7"))) %>% 
  arrange(timepoint_mouse) %>%
  mutate(timepoint_comparison = fct_inorder(timepoint_comparison),
         direction = fct_relevel(direction, "Up-Up", "Up-Down", "Down-Down", "Down-Up")) %>% 
  #Calculate percentage
  inner_join(
    human_mouse_log2fc_avg_wide  %>% 
      select(pathogen, timepoint_comparison) %>%
      dplyr::count(pathogen, timepoint_comparison) %>%
      rename(n_total = n),
    by = join_by(pathogen, timepoint_comparison)
  ) %>% 
  mutate(pct = ((n/n_total)*100) %>% round(0)) %>% 
  mutate(pathogen = as.character(pathogen),
         pathogen = case_when(pathogen == "Staphylococcus aureus" ~ "S. aureus",
                             pathogen == "Escherichia coli" ~ "E. coli", 
                             T~pathogen),
    pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "S. aureus", "E. coli", "Burn", "Trauma"))


```

##### Line plot
```{r fig.height=3, fig.width=12}
gene_list_lineplot = gene_list_df %>% 
  ggplot() +
  aes(x = timepoint_mouse, y = pct) +
  labs(x = "Timepoint Mouse",
         y = "Shared genes (%)",
         fill = "Direction",
         color = "Direction",
        linetype = "Direction") +
  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_continuous(expand = c(0.01, 0.01),
                     limits = c(0, 100)
                     ) +
  facet_grid(~pathogen, space = "free_x", scales = "free_x", axes = "all", axis.labels = "margins") +
  geom_line(aes(color = direction, 
                group = direction,
                linetype = direction),
            size = 1) +
  geom_point(aes(fill = direction),
             alpha = 1,
             color = "black",
             stroke = 0.5,
             pch = 21,
             size = 3) +
  geom_text(aes(label = n), color = "black", size = 2, vjust = -0.8) +
  
  scale_color_manual(values = colors$direction) +
  scale_fill_manual(values = colors$direction) +
  scale_linetype_manual(values = c("Up-Up" = 1, 
                                "Up-Down" = 2, 
                                "Down-Down" = 1, 
                                "Down-Up" = 2)) +
  
    theme_vaxgo() + 
  theme(# axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.key.width = unit(1, "cm"),
        legend.key.height = unit(1, "cm"), 
        legend.text = element_text(size = 10, 
                                   # hjust = 1, vjust = 1, 
                                   angle = 0),
        legend.title = element_text(size = 10),
        legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1))  +

  guides(fill = "none",
         linetype = guide_legend(override.aes = list(linewidth = 1)),
         color = guide_legend(override.aes = list(size = 0)))

gene_list_lineplot %>% 
  ggsave(filename = here("Figures", "all_human_mouse_rrho_lineplot.png"),
         width = 12,
         height = 3)

gene_list_lineplot

  
```




##### Heatmap
```{r fig.height=10, fig.width=10}
gene_list_df_heatmap = final_rrho_list$genes %>% 
  separate(timepoint_comparison, into = c("timepoint_mouse", "timepoint_human"), remove = F, sep = "-") %>% 
  mutate(timepoint_mouse = fct_relevel(timepoint_mouse, c("Early", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         timepoint_human = fct_relevel(timepoint_mouse, c("Early", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7"))) %>% 
  arrange(timepoint_mouse) %>%
  mutate(timepoint_comparison = fct_inorder(timepoint_comparison),
         direction = fct_relevel(direction, "Up-Up", "Up-Down", "Down-Down", "Down-Up")) %>% 
  mutate(pathogen = case_when(pathogen == "Staphylococcus aureus" ~ "S. aureus",
                             pathogen == "Escherichia coli" ~ "E. coli", 
                             pathogen == "Hepatitis B" ~  "HepB",
                             T~pathogen),
    pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma")) %>% 
  mutate(timepoint_direction = paste0(direction, " (", timepoint_mouse,")")) %>% 
  arrange(timepoint_mouse, direction ) %>% 
  mutate(timepoint_direction = fct_inorder(timepoint_direction))


# Clustering
df_to_cluster <- gene_list_df_heatmap %>%
  mutate(val = case_when(
    direction == "Up-Up"     ~  2,
    direction == "Up-Down"   ~  1,
    direction == "Down-Up"   ~ -1,
    direction == "Down-Down" ~ -2,
    TRUE ~ 0
  )) %>%
  select(genes, timepoint_direction, pathogen, val) %>%
  mutate(cond = paste(pathogen, timepoint_direction, sep = "_")) %>%
  select(genes, cond, val) %>%
  distinct() %>%
  pivot_wider(names_from = cond, values_from = val, values_fill = 0) %>%
  column_to_rownames("genes")

gene_dist <- dist(as.matrix(df_to_cluster)) 
gene_hclust <- hclust(gene_dist, method = "ward.D2")

# 3. Extrair a ordem dos genes clusterizados
gene_order <- rownames(df_to_cluster)[gene_hclust$order]

# 4. Aplicar a ordem de volta ao dataframe original do ggplot
gene_list_df_heatmap_clustered <- gene_list_df_heatmap %>%
  mutate(genes = factor(genes, levels = gene_order))

```


```{r fig.height=7, fig.width=12}
#Heatmap
gene_list_df_heatmap_plot = gene_list_df_heatmap %>% 
  mutate(direction_short = as.character(direction),
         direction_short = case_when(direction_short == "Up-Up" ~ "UU",
                               direction_short == "Up-Down" ~ "UD",
                               direction_short == "Down-Down" ~ "DD",
                               direction_short == "Down-Up" ~ "DU",
                               T~direction_short),
         direction_short = fct_relevel(direction_short, "UU", "UD", "DD", "DU"),
        timepoint_direction = paste0(timepoint_mouse, " (", direction,")")) %>% 
  arrange(timepoint_mouse, direction ) %>% 
  mutate(timepoint_mouse = fct_rev(timepoint_mouse)) %>% 
  # filter(pathogen == "Influenza") %>% 
  ggplot() +
  aes(x = genes, 
      y = timepoint_mouse,
       label = genes,
      fill = direction) +
  labs(x = "Genes",
         y = "Timepoint",
         fill = "Genes (%)") +
  facet_grid(~pathogen~direction, 
             scales = "free", 
             axes = "all", 
             space = "free",
             axis.labels = "margins",
             switch = "y") +
  geom_tile() +
  scale_fill_manual(values = colors$direction) +
  theme_vaxgo() + 
  theme(axis.text.y = element_text(hjust = 1, size =8),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        strip.text.y = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 8),
        panel.spacing = unit(0, "lines"),
        plot.margin = margin(0, 0, 0, 0, unit = "cm"),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.key.width = unit(1, "cm"),
        legend.text = element_text(size = 8, 
                                   # hjust = 1, vjust = 1, 
                                   angle = 0),
        legend.title = element_text(size = 8)) 


gene_list_df_heatmap_plot

gene_list_df_heatmap_plot %>%
  ggsave(filename = here("Figures", "all_human_mouse_rrho_heatmap_genes_bydirection.png"),
         width = 12,
         height = 7)

```



##### ORA: Hallmarks
```{r}

msigdbr_grouped = read_csv(file = here("tables", "Gene sets", "msigdb_hallmarks_grouped_genes.csv"))

#Discretize by quartile
ora_df_input = gene_list_df_heatmap %>% 
  mutate(condition = paste0(pathogen, "_", timepoint_comparison, "_", direction)) %>% 
  select(genes, condition)

term2gene_df_input = msigdbr_grouped %>% 
  select(process, genes = symbol)

#Run ORA Hallmarks
ora_all_diff_hallmark <- enricher_condition(
  df = ora_df_input,
  term2gene_df = term2gene_df_input) %>% 
  mutate(log_p_adjust = -log(p_adjust),
         process = str_replace_all(description, "_", " ")) %>% 
  separate(condition, into = c("pathogen", "timepoint_comparison", "direction"), sep = "_")  %>% 
  separate(timepoint_comparison, sep = "-", into = c("timepoint_mouse", "timepoint_human")) %>% 
  inner_join(msigdbr_grouped %>% 
               select(process, group) %>% 
               distinct(),
             by = join_by(id == "process")) %>% 
  mutate(timepoint_mouse = fct_relevel(timepoint_mouse, c("Early", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         timepoint_human = fct_relevel(timepoint_mouse, c("Early", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         timepoint_direction = paste0(timepoint_mouse, " (", direction, ")")) %>% 
  arrange(fct_rev(direction), timepoint_mouse) %>% 
  mutate(timepoint_direction = fct_inorder(timepoint_direction)) 


ora_all_diff_hallmark2$timepoint_direction %>% levels()

#Save
ora_all_diff_hallmark %>% 
  saveRDS(file = here("tables", "all_human_mouse_rrho_ora_hallmark.rds"))

#Print
ora_all_diff_hallmark
  
```


```{r}
# Clustering
df_to_cluster <- ora_all_diff_hallmark %>%
  mutate(val = case_when(
    direction == "Up-Up"     ~  2,
    direction == "Up-Down"   ~  1,
    direction == "Down-Up"   ~ -1,
    direction == "Down-Down" ~ -2,
    TRUE ~ 0
  )) %>%
  select(description, timepoint_direction, pathogen, val) %>%
  mutate(cond = paste(pathogen, timepoint_direction, sep = "_")) %>%
  select(description, cond, val) %>%
  distinct() %>%
  pivot_wider(names_from = cond, values_from = val, values_fill = 0) %>%
  column_to_rownames("description")

gene_dist <- dist(as.matrix(df_to_cluster)) 
gene_hclust <- hclust(gene_dist, method = "ward.D2")

# 3. Extrair a ordem dos description clusterizados
gene_order <- rownames(df_to_cluster)[gene_hclust$order]

# 4. Aplicar a ordem de volta ao dataframe original do ggplot
ora_all_diff_hallmark_clustered <- ora_all_diff_hallmark %>%
  mutate(description = factor(description, levels = gene_order))
```




```{r fig.height=8, fig.width=20}
ora_all_diff_hallmark_heatmap = ora_all_diff_hallmark %>% 
  mutate(direction = fct_relevel(direction, "Up-Up", "Down-Down", "Up-Down", "Down-Up")) %>% 
  filter(log_p_adjust > 1) %>% 
  mutate(log_p_adjust = if_else(log_p_adjust >10, 10, log_p_adjust)) %>% 
  ggplot() +
  aes(x = timepoint_mouse ,
      y = description) +
  # facet_grid(~direction+pathogen, space = "free_x", scales = "free_x") +
    facet_nested_wrap(~direction+pathogen, scales = "free_x", nrow = 1) +
  geom_tile(aes(fill = log_p_adjust)) +
  theme_vaxgo() + 
  labs(x = "",
       y = "") +
  theme(axis.text.y = element_text(hjust = 1, size =8),
        axis.text.x= element_text(size =8, angle = 90, hjust = 1, vjust = 0.5),
        # axis.ticks.x = element_blank(),
        strip.text.x = element_text(hjust = 0.5,size = 10, color = "white", face = "bold"),
        strip.text.y = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 8),
        panel.spacing = unit(0.1, "lines"),
        strip.background.x = element_rect(fill = "black", color = "white"),
        panel.grid.major.x = element_line(linetype = 1, color = "gray90"), 
        panel.grid.major.y = element_line(linetype = 1, color = "gray90"), 
        plot.margin = margin(0, 0.3, 0, 0.3, unit = "cm"),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.key.width = unit(1, "cm"),
        axis.line.y.right = element_line(linetype = 1, size = 0.5, color = "black"),
        legend.text = element_text(size = 10, 
                                   # hjust = 1, vjust = 1, 
                                   angle = 0),
        legend.title = element_text(size = 10),
        ggh4x.facet.nestline = element_line(colour = "white", linetype = 1))  +
   scale_fill_gradient(
    low = "#4361ee", 
    high = "#46eec6",
    limits = c(0, 10),
    breaks = c(0, 1, 1.3, 2, 10), 
    labels = c("0", "1", "1.3", "2", "≥10"), 
    na.value = "white"
  )

ora_all_diff_hallmark_heatmap

ora_all_diff_hallmark_heatmap %>%
   ggsave(filename = here("Figures", "all_human_mouse_rrho_heatmap_process_hallmarks.png"),
         width = 20,
         height = 8)

```






## Gene biotype vs. Protein Identity%

Examine whether gene biotype categories (miRNA, protein-coding,
pseudogenes, lncRNA, snoRNA) differ in their sequence identity to mouse
orthologs. This tests whether conservation level relates to functional
gene class.

```{r fig.height=4, fig.width=8}
# Test whether gene biotype is associated with %Identity
gene_info_human %>% 
  filter(!is.na(status)) %>% 
  select(hgnc_symbol, gene_biotype, treatment, pathogen, status, identity_human2mouse) %>% 
  filter(gene_biotype %in% c("protein_coding")) %>% 
  mutate(gene_biotype = "Protein coding") %>% 
  # Violin with overlaid boxplot for each biotype
  ggplot() +
  aes(x = status,
      y = identity_human2mouse,
      fill = status) +
  facet_nested_wrap(~gene_biotype+treatment+pathogen,
                    nrow = 1) +
  geom_violin(alpha = 0.2,
              color = NA, 
              width = 1) +
  geom_boxplot(width = 0.3,
               fill = NA) +
  scale_fill_manual(values = colors$comparison) +
  theme_vaxgo() +
  theme(panel.spacing= grid::unit(0.2, "cm"),
        legend.position = "top") +
  labs(x = "%Identity",
       y = "",
       fill = "Status",
       title = "Biotype and %Identity of mouse orthologs")


```

## Cellular component ontology vs Protein Sequence Identity%

Evaluate sequence identity of Leading Edge Genes (LEGs) and other
categories across major subcellular compartments.

```{r fig.height=3, fig.width=12}
# How LEGs are associated with identity by gene cellular component ontology
# Point-range plot of %Identity across gene categories, faceted by subcellular compartment

gene_info_human %>%
  # 1. Integrate GO annotations with global identity and status data
  inner_join(
    alignments_human2mouse %>% 
      select(-identity_human2mouse),
    by = join_by("hgnc_symbol" == gene)
  ) %>% 
  
  # 2. Filter for specific Cellular Component domains
  filter(
    go_domain == "cellular_component",
    go_name %in% c("nucleus", "cytoplasm", "membrane", "extracellular region")
  ) %>% 
  
  # 3. Clean and reformat status labels for plotting
  select(hgnc_symbol, go_name, treatment, pathogen,  identity_human2mouse, status) %>% 
  distinct() %>% 
  mutate(status = str_replace(status, " ", "\n")) %>%
  mutate(status = fct_relevel(
    status,
    "Human\nonly", "Mouse\nonly", "Shared", "All", 
    "Non-Immune\nBTMs", "Immune\nBTMs"
  )) %>% 
  
  # 4. Visualization: Compare identity across GO categories using facets
  ggplot() +
  aes(x = status,
      y = identity_human2mouse,
      fill = status,
      color = status) +
  facet_wrap(~str_to_title(go_name)+treatment+pathogen,
             nrow = 1,
             axes = "all") +
  
  # 5. Statistical Summary: Mean and 95% CI via bootstrapping
  stat_summary(fun.data = "mean_cl_boot",
               geom = "pointrange") +
  
  # 6. Styling, Scales, and Legends
  theme_vaxgo() +
  theme(legend.position = "top",
        axis.text.x = element_blank()) +
  scale_color_manual(values = colors$comparison) +
  scale_fill_manual(values = colors$comparison) +
  labs(color = "Status",
       y = "%Identity",
       x = "") +
  guides(color = guide_legend(nrow = 1, byrow = T),
         fill = "none")
```

```{r fig.height=3, fig.width=12}
# Distribution, boxplot and violin
gene_info_human %>%
  # 1. Join with global filtered identity data
  inner_join(
    alignments_human2mouse_all_global_df_filtered %>% 
      select(-identity_human2mouse),
    by = join_by("hgnc_symbol" == gene)
  ) %>% 
  
  # 2. Filter for major subcellular domains (Gene Ontology)
  filter(
    go_domain == "cellular_component",
    go_name %in% c("nucleus", "cytoplasm", "membrane", "extracellular region")
  ) %>% 
  
  # 3. Clean and reorder categories for consistent visualization
  select(hgnc_symbol, go_name, identity_human2mouse, status) %>% 
  distinct() %>% 
  arrange(identity_human2mouse) %>% 
  mutate(status = str_replace(status, " ", "\n")) %>% 
  mutate(status = fct_relevel(
    status,
    "Human\nonly", "Mouse\nonly", "Shared", "All", 
    "Non-Immune\nBTMs", "Immune\nBTMs"
  )) %>% 
  
  # 4. Visualization: Layered distribution per compartment
  ggplot() +
  aes(x = status,
      y = identity_human2mouse,
      fill = status,
      color = status) +
  facet_wrap(~str_to_title(go_name),
             scales = "free",
             nrow = 1,
             axes = "all") +
  
  # 5. Geoms: Violin for density + Boxplot for summary stats
  geom_violin(alpha = 0.2, color = NA) +
  geom_boxplot(width = 0.2) +
  
  # 6. Styling and Custom Color Scales
  theme_vaxgo() +
  theme(axis.text.x = element_blank(),
        legend.position = "top",
        axis.title.x = element_blank()) +
  scale_color_manual(values = c(colors$comparison_color, "All" = "black")) +
  scale_fill_manual(values = c(colors$comparison, "All" = "black")) +
  
  # 7. Legend and Labels
  guides(color = "none",
         fill = guide_legend(nrow = 1, byrow = T)) +
  labs(fill = "Status",
       y = "%Identity",
       x = "")

```

Visualize the global distribution of sequence identity while
highlighting specific "exemplar" genes of interest

```{r fig.height=8, fig.width=12}
# Global Statistical Summary: Mean Conservation (Bootstrap)
alignments_human2mouse_pointrange_all_global = alignments_human2mouse_all_global_df_filtered %>%
  # 1. Clean labels and establish the logical factor order for the comparison
  mutate(
    status = str_replace(status, " ", "\n"),
    status = fct_relevel(
      status,
      "Human\nonly", "Mouse\nonly", "Shared", "All", 
      "Non-Immune\nBTMs", "Immune\nBTMs"
    )
  ) %>% 
  mutate(analysis = "Mean Conservation, Bootstrap") %>%
  
  # 2. Setup the visualization: Aesthetics and Faceting
  ggplot() +
  aes(x = status,
      y = identity_human2mouse,
      fill = status,
      color = status,
      group = status) +
  facet_wrap(~analysis) +
  
  # 3. Layer 1: Pointrange with bootstrapped Mean and CI
  stat_summary(fun.data = "mean_cl_boot",
               geom = "pointrange",
               size = 1,
               fatten = 2,
               stroke = 0.5,
               pch = 21,
               mapping = aes(color = status),
               position = position_dodge2(0)) +
  
  # 4. Layer 2: Text labels for the precise mean values
  stat_summary(fun.data = "mean_cl_boot", 
               geom = "text", 
               aes(label = after_stat(round(y, 1))), 
               hjust = -0.2, 
               size = 3,
               color = "black") +
  
  # 5. Custom scales for fill and color
  scale_fill_manual(values = c(colors$comparison,
                               "BTMs" = "#4361ee",
                               "All" = "white")) +
  scale_color_manual(values = c(colors$comparison,
                                "BTMs" = "#4361ee",
                                "All" = "black")) +
  
  # 6. Adjust scales and coordinate system
  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_continuous(expand = c(0.01, 0.01)) +
  
  # 7. Final Theme and Legend Cleanup
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 10),
        axis.text.y = element_text(size = 10),
        strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
        panel.spacing = unit(1.1, "lines"),
        plot.margin = margin(0.3, 0.3, 0, 0, unit = "cm"),
        legend.position = "none",
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) +
  labs(x = "Status",
       y = "%Identity (Human-to-Mouse)",
       color = "",
       fill = "")
# Display
alignments_human2mouse_pointrange_all_global
```

Plot boxplot with violin plot and unite with point-range plot

```{r fig.height=8, fig.width=12}
# Global Distribution of Conservation with Highlighted Exemplar Genes
# Violin + boxplot with highlighted exemplar genes overlaid
alignments_human2mouse_histogram_all_global = alignments_human2mouse_all_global_df_filtered %>%
  # 1. Prepare categories and factor levels
  mutate(analysis = "Distribution of conservation",
         status = str_replace(status, " ", "\n"),
         status = fct_relevel(status, "Human\nonly", "Mouse\nonly", "Shared", 
                              "All", "Non-Immune\nBTMs", "Immune\nBTMs")) %>% 
  
  ggplot() +
  aes(x = status, y = identity_human2mouse, fill = status, color = status, group = status) +
  facet_wrap(~analysis) +
  
  # 2. Main Layers: Violin for density and transparent Boxplot
  geom_violin(color = "black", alpha = 0.7) +
  geom_boxplot(mapping = aes(color = status), width = 0.1, fill = NA, outliers = F) +
  
  # 3. First Color Scale: Category status
  scale_fill_manual(name = "Status", values = c(colors$comparison, "All" = "white")) +
  scale_color_manual(name = "Status", 
                     values = c("Mouse\nonly" = "white", "Human\nonly" = "white", "Shared" = "white", 
                                "Immune\nBTMs" = "white", "Non-Immune\nBTMs" = "white", "All" = "black")) +
  
  # 4. Secondary Color Scale: Custom filters for highlighted genes
  new_scale_color() + 
  scale_color_manual(
     name = "Filter",
     values = c("High %Identity" = "#4361ee", "Positive correlation,\nHigh %Identity" = "#4DBBD5FF",
                "Negative correlation,\nHigh %Identity" = "#7209b7", "Low %Identity" = "red",
                "Positive correlation,\nLow %Identity" = "#CA6702", "Negative correlation,\nLow %Identity" = "#AE2012")) +
  
  # 5. Adding highlighted genes as points
  geom_point(data = alignments_human2mouse_genes_filtered %>% 
               group_by(status, filter) %>% 
               slice_head(n = 5), aes(color = filter), pch = 21, fill = "white") +
  
  # 6. Statistical annotation: Median identity labels
  stat_summary(fun = median, geom = "label", aes(label = after_stat(round(y, 1))),
               size = 3, hjust = -0.2, color = "black", fill = "white") +
  
  # 7. Gene Labeling: Repel prevents overlapping names
  geom_label_repel(data = alignments_human2mouse_genes_filtered %>% 
               group_by(status, filter) %>% 
               slice_head(n = 5),
                   aes(label = gene, color = filter), 
                   max.overlaps = 10000, size = 2, direction = "y", nudge_x = -0.4,
                   hjust = 0, fill = "white", box.padding = unit(0, "cm"), label.padding = unit(0.1, "cm")) +
  
  # 8. Scales, Theme, and Layout
  scale_x_discrete(expand = c(0.1, 0.1)) +
  scale_y_continuous(expand = c(0.01, 0.01), limits = c(50, 105)) +
  theme_vaxgo() +
  theme(axis.text = element_text(size = 10),
        legend.position = "top", legend.direction = "vertical",
        legend.margin = margin(0,0,0,0), legend.box.margin = margin(0,0,-10,0),
        plot.margin = margin(0.3, 0.3, 0, 0, unit = "cm")) +
  labs(x = "Status", y = "%Identity (Human-to-Mouse)", color = "", fill = "")

# Display
alignments_human2mouse_histogram_all_global

# Unite summary and distribution plots using patchwork
alignments_human2mouse_pointrange_histogram = (alignments_human2mouse_pointrange_all_global /
alignments_human2mouse_histogram_all_global) +
  plot_layout(heights = c(0.25, 0.75),
              guides = "collect",
              axis_titles = "collect"
              ) +
  plot_annotation(theme = theme(legend.position = "right",
                                legend.justification = "top"))
alignments_human2mouse_pointrange_histogram

# Save composite figure
alignments_human2mouse_pointrange_histogram %>%
ggsave(filename = here("Figures", "all_human_mouse_gsea_LEGs_alignment_btms_human2mouse_violinplot_all_global.png"),
         width = 8,
         height = 8)
```

Line plot of %Identity across all genes, faceted by treatment type and
pathogen. This shows how sequence conservation evolves over time for
each gene category.

```{r fig.height=3, fig.width=10}
# Prepare data: clean timepoints, assign treatment groups, filter outliers
alignments_human2mouse_pointrange_all_df = alignments_human2mouse %>% 
  mutate(timepoint = str_replace(timepoint," ", "\n"),
         timepoint = fct_relevel(timepoint, c("Ref", "Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day\n1", "Day\n3", "Day\n7")),
         treatment = case_when(pathogen %in% c("Hepatitis B", "Influenza") ~ "Vaccination",
                               pathogen %in% c( "Saureus", "Ecoli") ~ "Infection",
                               pathogen %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury")) %>% 
  filter(!(pathogen == "Saureus" & timepoint == "24h")) %>% 
  select(!c(process, group, set_size_human:mouse_core, group)) %>% 
  distinct()

# Compute reference identity means for Immune BTM, Non-Immune BTM, and All
mean_ref_identity = alignments_human2mouse_all_global_df_filtered %>% 
  filter(status %in% c("Immune BTMs", "Non-Immune BTMs", "All")) %>% 
  group_by(status) %>% 
  reframe(mean_cl_boot(identity_human2mouse))

# Build line plot with pointrange for each gene category across timepoints
alignments_human2mouse_pointrange_all_plot = alignments_human2mouse_pointrange_all_df %>% 
  ggplot() +
  aes(x = timepoint,
      y = identity_human2mouse,
      fill = status,
      color = status,
      group = status) +
  facet_grid(~treatment+pathogen, 
             scales = "free_x",
             space = "free_x") +
  # Line connecting mean values
  stat_summary(fun = mean,
               geom = "line",
               size = 0.5) +
  # Pointrange with bootstrapped CI
  stat_summary(fun.data = "mean_cl_boot",
               geom = "pointrange",
               size = 1,
               fatten = 1,
               mapping = aes(color = status))+
  # Label with number of genes per timepoint
  geom_text_repel(
    data = alignments_human2mouse_pointrange_all_df %>%
      group_by(treatment, pathogen, timepoint, status) %>%
      summarise(n_genes = n(),
                identity_human2mouse = mean(identity_human2mouse),
                .groups = "drop"),
    aes(x = timepoint,
        y = identity_human2mouse,
        label = n_genes,
        color = status),
    inherit.aes = FALSE,
    size = 2, segment.size = 0,
    box.padding = 0.5
  ) +
  # Reference pointrange for overall category mean at baseline
  geom_pointrange(mapping = aes(x = "Ref", y = y, ymin = ymin, ymax = ymax),
                  size = 1,
                  fatten = 1,
                  data = mean_ref_identity) +
  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = c(colors$comparison, 
                     "All" = "black")
                     ) +
  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_continuous(expand = c(0.01, 0.01)) +
  theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 10),
          axis.text.y = element_text(size = 10),
          strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
          panel.spacing = unit(1.1, "lines"),
          plot.margin = margin(0, 0.3, 0, 0, unit = "cm"),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
          legend.text = element_text(size = 10, hjust = 0),
          legend.title = element_text(size = 10),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)
    ) +
  labs(x = "Time",
       y = "%Identity (Human-to-Mouse)",
       color = "",
       fill = "") +
  guides(fill = "none")

# Print
alignments_human2mouse_pointrange_all_plot

# Save
alignments_human2mouse_pointrange_all_plot %>%
ggsave(filename = here("Figures", "all_human_mouse_gsea_LEGs_alignment_btms_human2mouse_lineplot_all.png"),
         width = 10,
         height = 3)
```

Point-range lineplot of %Identity across time, faceted by BTM functional
group and treatment. This provides a high-resolution view of
conservation dynamics within each immunological and metabolic module.

```{r fig.height=12, fig.width=12}
# Lineplot faceted by BTM group and treatment across timepoints
alignments_human2mouse_lineplot_all = 
  # Append an "All" category alongside individual BTM groups
  bind_rows(alignments_human2mouse %>% 
              mutate(group = "All"),
            alignments_human2mouse
            ) %>% 
  # Clean timepoint labels and assign treatment and factor orders
  mutate(timepoint = str_replace(timepoint,"Early hours", "Early\nhours"),
         timepoint = fct_relevel(timepoint, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         group = fct_relevel(group, "All", immune_order),
         treatment = case_when(pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen %in% c( "Saureus", "Ecoli") ~ "Infection",
                               pathogen %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury")) %>% 
  
  ggplot() +
  aes(x = timepoint,
      y = identity_human2mouse,
      fill = status,
      color = status,
      group = status) +
  # One row per BTM group, one column per treatment:pathogen
  facet_grid(group ~ treatment+pathogen, 
             scales = "free",
             space = "free_x") +
  # Line connecting mean identity values
  stat_summary(fun = mean, 
               geom = "line", 
               size = 0.8) +
  # Pointrange with bootstrapped 95% CI
  stat_summary(fun.data = "mean_cl_boot",
               geom = "pointrange",
               mapping = aes(color = status),
               size = 0.2
               ) +
  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_continuous(expand = c(0.01, 0.01)) +
  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = colors$comparison) +
  theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
          axis.text.y = element_text(size = 8),
          strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
          panel.spacing = unit(1.1, "lines"),
          plot.margin = margin(0, 0.3, 0, 0, unit = "cm"),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
          legend.text = element_text(size = 10, hjust = 0),
          legend.title = element_text(size = 10),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)
    ) +
  labs(x = "Time",
       y = "%Identity (Human-to-Mouse)",
       color = "",
       fill = "")

alignments_human2mouse_lineplot_all

# alignments_human2mouse_lineplot_all %>% 
#   ggsave(filename = here("Figures", "all_human_mouse_gsea_LEGs_alignment_btms_human2mouse_lineplot_bygroup.png"),
#          width = 12,
#          height = 12)

```

## Align most conserved genes

Load the shared LEG data and prepare for gene annotation and alignment.
The top exemplar genes are extracted from the filtered dataset for
downstream biomaRt queries and protein sequence alignment.

```{r}
# Import GSEA core results and BTM annotation table
gsea_btm_results_core_joined =
  readRDS(file = here("tables", "Functional analysis", "all_human_mouse_gsea_btm_results_core_joined.rds"))
btm_annotation_genes = read_csv(file = here("tables", "Gene sets", "btm_annotation_genes.csv"))

# Extract top exemplar genes from the filtered dataset
shared_genes_df = alignments_human2mouse_genes_filtered %>% 
  select(filter, status, everything()) %>% 
  group_by(filter, status) %>% 
  slice_head(n  = 1) %>% 
  ungroup()

shared_genes = shared_genes_df %>% 
  pull(gene) %>% 
  unique()

# 1. Setup biomaRt for gene annotation
## Gene annotation
# human_mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
# mouse_mart <- useMart("ensembl", dataset = "mmusculus_gene_ensembl")

# 2. Retrieve Pfam domains, GO terms, and Identity
# We include homology identity to correlate directly
domains_human <- getBM(
  attributes = c(
    "hgnc_symbol",
    "interpro",
    "interpro_start",
    "interpro_end",
    "interpro_short_description",
    # "tmhmm_start",
    # "tmhmm_end",
    "phobius",
    "phobius_start",
    "phobius_end"
  ),
  filters = "hgnc_symbol",
  values = shared_genes,
  # values = "ABCE1",
  mart = human_mart
)

sequences_human = getBM(
  attributes = c(
    "hgnc_symbol",
    "peptide",
    "uniprotswissprot"
  ),
  filters = "hgnc_symbol",
  values = shared_genes,
  # values = "ABCE1",
  mart = human_mart
) %>% 
  #Filter out non-canonical/annotated sequences
  filter(uniprotswissprot != "")

identity_human <- getBM(
  attributes = c("external_gene_name", 
                 "mmusculus_homolog_perc_id_r1",
                 "mmusculus_homolog_goc_score",
                 "mmusculus_homolog_orthology_confidence"),
  filters = "hgnc_symbol",
  # values = "ABCE1",
  values = shared_genes,
  mart = human_mart
) %>% 
  rename(hgnc_symbol = external_gene_name,
         identity = mmusculus_homolog_perc_id_r1)

#REGULATION
regulation_enhancer_human = getBM(
  attributes = c(
    "external_gene_name",
    "transcription_start_site"
  ),
  filters = "external_gene_name",
  values = "MARCKS",
  mart = human_mart
)
```

```{r}


# Inspect available BioMart attributes for human — used to identify which annotation columns can be retrieved
listAttributes(human_mart) %>% View()
```

```{r}
# Join the tables
domains_identity_sequence_human = 
  sequences_human %>% 
  inner_join(domains_human, 
             by = "hgnc_symbol") %>% 
  left_join(identity_human, 
            domains_human,
            by = "hgnc_symbol") %>% 
  group_by(hgnc_symbol, interpro) %>% 
  mutate(interpro_length = interpro_end - interpro_start) %>% 
  slice_max(interpro_length, n = 1)


#Mouse
genes_mouse = getBM(
  attributes = c("external_gene_name", 
                 "hsapiens_homolog_associated_gene_name"),
  filters = "external_gene_name",
  values = str_to_title(shared_genes),
  mart = mouse_mart
) %>% 
  rename(mgi_symbol = external_gene_name,
         hgnc_symbol = hsapiens_homolog_associated_gene_name)

domains_mouse <- getBM(
  attributes = c(
    "external_gene_name",
    "interpro",
    "interpro_start",
    "interpro_end",
    "interpro_short_description",
    # "tmhmm_start",
    # "tmhmm_end",
    "phobius",
    "phobius_start",
    "phobius_end"
    # "signalp_start",
    # "signalp_end"
  ), 
  filters = "external_gene_name",
  values = genes_mouse$mgi_symbol,
  mart = mouse_mart
)

sequences_mouse = getBM(
  attributes = c(
    "external_gene_name",
    "peptide",
    "uniprotswissprot"
  ),
  filters = "external_gene_name",
  values = genes_mouse$mgi_symbol,
  # values = "Abce1",
  mart = mouse_mart
) %>% 
  #Filter out non-canonical/annotated sequences
  filter(uniprotswissprot != "")

# enhancer_mouse = getBM(
#   attributes = c(
#     "external_gene_name",
#     "peptide",
#     "uniprotswissprot"
#   ),
#   filters = "external_gene_name",
#   values = genes_mouse$mgi_symbol,
#   # values = "Abce1",
#   mart = mouse_mart
# ) %>% 
#   #Filter out non-canonical/annotated sequences
#   filter(uniprotswissprot != "")


#Merge
domains_identity_sequence_mouse = 
  sequences_mouse %>% 
  inner_join(domains_mouse, 
             by = "external_gene_name") %>% 
  rename(mgi_symbol = external_gene_name) %>% 
  left_join(genes_mouse, 
            by = "mgi_symbol") %>% 
  arrange(interpro_start) %>% 
  group_by(mgi_symbol, interpro) %>% 
  mutate(interpro_length = interpro_end - interpro_start) %>% 
  slice_max(interpro_length, n = 1)


#Merge Human and Mouse
genes_domain_mouse_human <- domains_identity_sequence_mouse %>% 
  inner_join(domains_identity_sequence_human, 
             by = join_by(hgnc_symbol, interpro), 
             suffix = c("_mouse", "_human")) %>% 
  drop_na(peptide_mouse, peptide_human) %>% 
  inner_join(shared_genes_df %>% 
               select(filter, status, hgnc_symbol = gene, n, cor_value, p_value, r_squared, cor_label) %>% 
               distinct(),
             by = "hgnc_symbol")

glimpse(genes_domain_mouse_human)

genes_domain_mouse_human %>% 
  saveRDS(file = here("tables", "Functional analysis", "all_human_mouse_gsea_leganalysis_biomart_annotation.rds"))


# listAttributes(mouse_mart) %>% View()
# listAttributes(human_mart) %>% View()

```

Is there an association between domains and correlation?

```{r}
# Quick inspection: select distinct domain-level metrics per gene to check whether domain composition correlates with expression conservation
genes_domain_mouse_human %>%
  select(hgnc_symbol,
         interpro,
         interpro_short_description_human,
         identity, interpro_length_human,
         filter:cor_label) %>%
  distinct() %>%
  drop_na(cor_value)
```

Alignment

```{r}
genes_domain_mouse_human =
  readRDS(file = here("tables", "Functional analysis", "all_human_mouse_gsea_leganalysis_biomart_annotation.rds")) %>% 
  distinct()

selected_gene = "CD2"

# #Align domains
# sequences_align_domains = genes_domain_mouse_human %>% 
# mutate(
#     # Cut sequence: peptide[start : end]
#     domain_seq_human = stringr::str_sub(peptide_human, interpro_start_human, interpro_end_human),
#     domain_seq_mouse = stringr::str_sub(peptide_mouse, interpro_start_mouse, interpro_end_mouse)
#   )
# 
# #Align full sequence
# sequences_align_gene <- genes_domain_mouse_human %>%
#   filter(status == "Shared") %>% 
#   ungroup() %>% 
#   filter(hgnc_symbol == selected_gene) %>% 
#   select(hgnc_symbol, peptide_human, peptide_mouse, status, cor_label) %>%
#   distinct()
# 
# #Alignment
# gene_data <- sequences_align_gene
# 
# # Create AAStringSet to run on msa
# input_seqs <- AAStringSet(c(gene_data$peptide_human, gene_data$peptide_mouse))
# names(input_seqs) <- c("Human", "Mouse")
# 
# # Run MSA
# aln <- msa(input_seqs, method = "Muscle", type = "protein")
# 
# # Extract sequences
# # Convert to text
# seqs_alinhadas <- as.character(unmasked(aln))

# Pairwise sequence alignment
library(pwalign)
#Compare two original sequences
alignment <- pairwiseAlignment(
  pattern = gene_data$peptide_human, 
  subject = gene_data$peptide_mouse, 
  substitutionMatrix = "BLOSUM62",
  type = "global", # Garante alinhamento de ponta a ponta
  gapOpening = 10,  # Penalidade padrão
  gapExtension = 0.5
)

# Get identity score
identity_global <- pid(alignment, type = "PID1")

#Convert to a df
alignment_results_pairwise <- tibble(
  hgnc_symbol       = gene_data$hgnc_symbol,
  seq_human_orig    = gene_data$peptide_human,
  seq_mouse_orig    = gene_data$peptide_mouse,
  seq_human_aligned = as.character(pattern(alignment)),
  seq_mouse_aligned = as.character(subject(alignment)),
  identity          = identity_global
) %>% 
  inner_join(genes_domain_mouse_human %>% 
  ungroup() %>% 
  select(mgi_symbol, hgnc_symbol, identity_biomart = identity, peptide_mouse, peptide_human, uniprotswissprot_human,
         filter, status, n, cor_value, p_value, r_squared, cor_label) %>% 
  distinct(),
  by = "hgnc_symbol")



```

## Protein structure

### Prepare protein data for CD2

Fetch UniProt annotations (topology, domains, PTMs) and compute pairwise
sequence differences between human and mouse CD2 from the alignment.
These will feed into a multi-track protein annotation plot and 3D
visualisation.

```{r fig.height=4, fig.width=8}
# Load tidyverse for data processing
library(tidyverse)

# Prepare sequence and annotation data for the selected gene
# Extract the row corresponding to selected_gene and keep the longest isoforms
# to ensure we have full-length sequences for comparison
gene_info <- alignment_results_pairwise %>%
  filter(hgnc_symbol == selected_gene) %>%
  mutate(
    len_mouse = nchar(seq_mouse_orig),
    len_human = nchar(seq_human_orig)
  ) %>%
  slice_max(len_human, n = 1) %>%
  slice_max(len_mouse, n = 1)

# Create a "Full Protein" annotation track covering the entire sequence
# This provides the background rectangle spanning the whole protein length
full_protein <- gene_info %>%
  select(hgnc_symbol, len_mouse, len_human, status, cor_label,
         seq_human_orig, seq_mouse_orig, filter, identity, identity_biomart,
         seq_human_aligned, seq_mouse_aligned) %>%
  distinct() %>%
  pivot_longer(cols = starts_with("len_"), names_to = "species", values_to = "end") %>%
  group_by(hgnc_symbol, species) %>%
  slice_max(end, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    species = str_remove(species, "len_"),
    start = 1,
    interpro_short_description_human = "Full Protein",
    interpro = "Full",
    species_end = species
  ) %>%
  distinct()

# Retrieve UniProt feature annotations for CD2 (human and mouse orthologs)
# This fetches transmembrane domains, topology, and other protein features
prot_data <- drawProteins::get_features("P06729 P08920") %>%
  drawProteins::feature_to_dataframe() %>%
  mutate(description = case_when(type == "TRANSMEM" ~ "TM",
                                  type == "TOPO_DOM" & description == "Extracellular" ~ "EC",
                                  type == "TOPO_DOM" & description == "Cytoplasmic" ~ "CP",
                                  description == "NONE" ~ type,
                                  description == "" ~ type,
                                  TRUE ~ description)) %>%
  separate(entryName, sep = "_", into = c("gene", "organism"), remove = FALSE) %>%
  mutate(organism = str_to_title(organism))

# Identify sequence differences between human and mouse from the pairwise alignment
vec_h <- strsplit(unique(full_protein$seq_human_aligned), "")[[1]]
vec_m <- strsplit(unique(full_protein$seq_mouse_aligned), "")[[1]]

# Find positions where the aligned residues differ
diff_idx <- which(vec_h != vec_m)

# Map alignment positions back to real (ungapped) residue positions per species
mutations_base <- tibble(
  align_pos = diff_idx,
  res_h = vec_h[diff_idx],
  res_m = vec_m[diff_idx]
) %>%
  mutate(
    pos_real_human = map_int(align_pos, ~ sum(vec_h[1:.x] != "-")),
    pos_real_mouse = map_int(align_pos, ~ sum(vec_m[1:.x] != "-"))
  )

# Separate by species to track mutations in each organism's coordinate system
mutations_human <- mutations_base %>%
  filter(res_h != "-") %>%
  mutate(species = "human", x_pos = pos_real_human)

mutations_mouse <- mutations_base %>%
  filter(res_m != "-") %>%
  mutate(species = "mouse", x_pos = pos_real_mouse)

# Combine mutations into a single table formatted for ggplot annotation
mutations_plot <- bind_rows(mutations_human, mutations_mouse) %>%
  rename(organism = species) %>%
  mutate(gene = unique(gene_info$hgnc_symbol),
         organism = str_to_title(organism),
         type = "Alignment",
         description = "Mutations")

```

### Visualize structures

draw proteins

```{r fig.height=4, fig.width= 10, paged.print=TRUE}
# Get a human-readable protein name for the y-axis label
prot_name <- unique(prot_data$entryName)

# Split UniProt features into thematic tracks for multi-panel visualisation
# Track 1: Topology — transmembrane and subcellular localisation
df_topology <- prot_data %>%
  filter(type %in% c("TOPO_DOM", "TRANSMEM", "SIGNAL")) %>%
  mutate(y_pos = 1)

# Track 2: Domains — functional regions (Ig-like, receptors, etc.)
df_features <- prot_data %>%
  filter(type %in% c("DOMAIN", "REGION", "MOTIF", "RECEPT_DOM")) %>%
  mutate(y_pos = 1.5)

# Track 3: Modifications — PTMs and natural variants
df_modifications <- prot_data %>%
  filter(type %in% c("MOD_RES", "PHOSPHO", "VARIANT")) %>%
  mutate(y_pos = 1.61)

# Alignments-derived mutation data frame
df_mutations <- mutations_plot

# Create a summary dataframe to define the background line length for each facet
# This ensures the grey backbone scales to the correct length per organism
background_lines <- prot_data %>%
  group_by(organism, gene) %>%
  summarise(max_end = max(end), .groups = 'drop') %>%
  tidyr::crossing(y_val = c(1, 1.5))

# Initialise the multi-layer protein annotation ggplot
drawprotein_uniprot_ggplot <- ggplot() +

  # Track background: grey line spanning the full protein length per facet
  geom_segment(data = background_lines,
               aes(x = 1, xend = max_end, y = y_val, yend = y_val),
               color = "grey80", size = 2) +

  # Track 1: Topology rectangles — extracellular, cytoplasmic, transmembrane
  geom_rect(data = df_topology,
            aes(xmin = begin, xmax = end,
                ymin = y_pos - 0.1, ymax = y_pos + 0.1,
                fill = description),
            color = "black", size = 0.2) +

  # Facet by organism and gene for side-by-side species comparison
  facet_wrap(~ paste(organism, gene), ncol = 1) +

  # Custom fill scale for topology track
  scale_fill_manual(values = c("EC" = "#4cc9f0",
                                "CP" = "#3a86ff",
                                "TM" = "#90A4AEFF",
                                "SIGNAL" = "#f72585"),
                    name = "Topology") +

  # Reset fill scale for the next aesthetic layer
  ggnewscale::new_scale_fill() +

  # Track 2: Functional domains — semi-transparent rectangles with independent fill
  geom_rect(data = df_features,
            aes(xmin = begin, xmax = end,
                ymin = y_pos - 0.1, ymax = y_pos + 0.1,
                fill = description),
            color = "black", size = 0.2, alpha = 0.5) +
  scale_fill_viridis_d(option = "mako", name = "Domains") +

  # Track 3: Modification points — diamonds for phosphorylation / variants
  geom_point(data = df_modifications,
             aes(x = begin, y = y_pos + 0.02, color = type),
             size = 3, shape = 18) +
  scale_color_manual(values = c("PHOSPHO" = "yellow", "VARIANT" = "red"),
                     name = "PTMs") +

  # Reset color for mutation layer
  new_scale_color() +

  # Mutation layer: violin + points to show alignment divergence density
  geom_violin(data = mutations_plot,
              aes(x = x_pos, y = 1.5, color = "Mutations"),
              fill = NA, size = 0.4, width = 0.2, scale = "count", adjust = 0.05) +
  geom_point(data = mutations_plot,
             aes(x = x_pos, y = 1.5, color = "Mutations",
                 label = paste(align_pos, res_h, res_m)),
             shape = 124, size = 1, alpha = 0.6) +
  scale_color_manual(values = c("Mutations" = "red"), name = "Alignment") +

  # Axis formatting: map y_pos to track labels
  scale_y_continuous(breaks = c(unique(df_topology$y_pos), unique(df_features$y_pos)),
                     labels = c("Topology", "Domains")) +
  scale_x_continuous(expand = c(0.01, 0.01)) +

  # Annotate domain start and end coordinates on the plot
  geom_text(aes(label = begin, y = y_pos, x = begin),
            data = df_features,
            nudge_y = 0.15, hjust = 0.5, size = 3) +
  geom_text(aes(label = end, y = y_pos, x = end),
            data = df_features,
            nudge_y = -0.15, hjust = 0.5, size = 3) +
  geom_text(aes(label = begin, y = y_pos, x = begin),
            data = df_topology,
            nudge_y = 0.15, hjust = 0.5, size = 3) +
  geom_text(aes(label = end, y = y_pos, x = end),
            data = df_topology,
            nudge_y = -0.15, hjust = 0.5, size = 3) +

  # Custom lab theme
  theme_vaxgo() +
  theme(axis.text.y = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0),
        panel.grid.minor = element_blank(),
        legend.position = "right",
        legend.justification = "top",
        legend.spacing.y = unit(0, "cm"),
        plot.margin = margin(1, 1, 2, 1)) +

  # Labels: include identity and correlation metrics in the title
  labs(x = "Residue position",
       y = "Features",
       title = paste0("Gene: ", unique(prot_data$gene),
                      ", UNIPROT: ", unique(prot_data$accession),
                      ". Status: ", unique(gene_info$status),
                      ". Filter: ", str_replace(unique(gene_info$filter), "\n", " "),
                      ", Correlation: ", unique(gene_info$cor_label),
                      "\nIdentity (MUSCLE, BLOSUM62): ", round(unique(gene_info$identity), 1),
                      ". Identity (BioMart): ", round(unique(gene_info$identity_biomart), 1)))

# Display the plot
drawprotein_uniprot_ggplot

```

VIsualize 3d

```{r}
# Define hex colour codes for topology tracks — each type gets a distinct colour for the 3D viewer
topo_colors <- list(
  "EC" = "#4cc9f0",
  "CP" = "#3a86ff",
  "TM" = "black",
  "SIGNAL" = "#FF1493"
)

# Prepare topology data for NGLVieweR: assign colours and create residue-range selections
df_topology_3d <- df_topology %>%
  mutate(
    color_3d = as.character(topo_colors[description]),
    selection = paste0(begin, "-", end)
  )

# Prepare domain feature data for NGLVieweR: map mako colours and create selections
df_features_3d <- df_features %>%
  mutate(
    selection = paste0(begin, "-", end),
    color_hex = as.character(mako(n())[as.numeric(as.factor(description))])
  )

# Initialise NGLVieweR with the CD2 human AlphaFold structure
view_initial <- NGLVieweR(here::here("Structures", "CD2_HumanAF-P06729-F1-model_v6.pdb")) %>%
  stageParameters(backgroundColor = "white") %>%
  addRepresentation("ribbon",
                    param = list(color = "#D3D3D3", opacity = 0.2))

# Apply a small rotation for a better initial view angle
view_initial %>%
  setRotation(10, 0, 0)

view_initial
```

Topology

```{r}
# Add topology cartoon representations on top of the initial structure
# Each colour-coded segment (EC, CP, TM, SIGNAL) maps to a UniProt-annotated region
view_topology <- view_initial
for (i in seq_len(nrow(df_topology_3d))) {
  current_color <- df_topology_3d$color_3d[i]
  if (!is.na(current_color)) {
    view_topology <- view_topology %>%
      addRepresentation("cartoon", param = list(
        sele = df_topology_3d$selection[i],
        color = current_color,
        opacity = 0.5
      ))
  }
}
view_topology
```

domains

```{r}
library(NGLVieweR)
library(dplyr)

# Define distinct colours for each domain feature to ensure clear visual separation in the 3D structure
domain_colors <- colors_all

# Prepare the domain feature table with colour assignment and residue range selection
df_features_3d <- df_features %>%
  mutate(
    color_3d = domain_colors[1:n()],
    selection = paste0(begin, "-", end)
  )

# Initialise NGLVieweR with CD2 human AlphaFold structure
# Set orthographic camera, high sample level for antialiasing, and a transparent ribbon backbone
view_domain <- NGLVieweR(here::here("tables", "Structures", "CD2_HumanAF-P06729-F1-model_v6.pdb")) %>%
  stageParameters(
    backgroundColor = "white",
    cameraType = "orthographic",
    clipDist = 0,
    sampleLevel = 4,
    preserveDrawingBuffer = TRUE
  ) %>%
  addRepresentation("cartoon", param = list(color = "#D3D3D3", opacity = 0.2))

# Overlay each domain with its assigned colour as a cartoon representation
for (i in seq_len(nrow(df_features_3d))) {
  current_color <- as.character(df_features_3d$color_3d[i])
  current_sele  <- as.character(df_features_3d$selection[i])
  if (!is.na(current_color)) {
    view_domain <- view_domain %>%
      addRepresentation("cartoon", param = list(
        sele = current_sele,
        color = current_color,
        opacity = 1
      ))
  }
}

# Render and rotate for optimal viewing angle
view_domain
view_domain <- view_domain %>%
  NGLVieweR::setRotation(c(0, 1.57, 0)) %>%
  NGLVieweR::setSelection("*")
view_domain

# Annotate chains in the PDB object: assign a single-letter code per domain for downstream visualisation
for (i in seq_len(nrow(df_features_3d))) {
  desc_char <- toupper(substr(as.character(df_features_3d$description[i]), 1, 1))
  res_range <- as.numeric(unlist(strsplit(as.character(df_features_3d$selection[i]), "-")))
  sele <- atom.select(pdb, resno = res_range[1]:res_range[2])
  if (length(sele$atom) > 0) {
    pdb$atom$chain[sele$atom] <- desc_char
  }
}

# Save the chain-annotated PDB for later use
write.pdb(pdb, file = here("tables", "Structures", "cd2_com_chains.pdb"))
```

```{r}
# Render mutations from the pairwise alignment on the 3D structure
# Each divergence position between human and mouse CD2 is shown as a solid red cartoon segment
if (nrow(mutations_plot) > 0) {
  mut_residues <- unique(as.character(mutations_plot$x_pos))
  mut_sele <- paste(mut_residues, collapse = " OR ")
  view_mutations <- view_topology %>%
    addRepresentation("cartoon", param = list(
      sele = mut_sele,
      color = "#FF0000",
      opacity = 1
    ))
}
view_mutations
```

Display alignment with ggmsa

```{r fig.height=3, fig.width=16}
# Set the gene for visualisation
selected_gene <- "CD2"

# Extract the most complete row for CD2 from the alignment results
data_row <- plot_domains %>%
  select(hgnc_symbol, identity, identity_biomart,
          seq_human_orig, seq_mouse_orig,
          seq_mouse_aligned, seq_human_aligned) %>%
  distinct()

# Create an AAStringSet object from the human and mouse aligned sequences
seqs_to_plot <- AAStringSet(c(data_row$seq_human_aligned, data_row$seq_mouse_aligned))
names(seqs_to_plot) <- c("Human", "Mouse")

# Render the multiple sequence alignment with ggmsa
# Colour by amino acid chemistry, use Human as reference, and show consensus tracks
ggmsa(seqs_to_plot,
      ref = "Human",
      start = 1,
      end = 352,
      color = "Chemistry_AA",
      font = "DroidSansMono",
      char_width = 0.8,
      border = "white",
      seq_name = TRUE,
      consensus_views = TRUE,
      disagreement = FALSE,
      use_dot = FALSE,
      by_conservation = FALSE,
      ignore_gaps = FALSE) +
  facet_msa(field = 100) +
  theme(panel.spacing.y = unit(0, "lines"),
        legend.position = "right",
        axis.text.y = element_text(color = "black", size = 10),
        axis.text.x = element_text(color = "black", size = 10)) +
  labs(title = paste0(data_row$hgnc_symbol, ", Identity: ", round(data_row$identity, 1), "%")) 


```
















# Align sequences

```{r}


#Import homologous CREs from ENCODE 
homologous_cres = read_tsv(here("Genomic", "hg38-mm10-Homologous.tsv"),
  col_names = c("human_enhancer", "mouse_enhancer")
) 

#Import human CREs
cres_human <- read_tsv(
  here("Genomic", "GRCh38-cCREs.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>% 
   rename_with(~paste0(., "_human"))
              
              
#Import mouse CREs
cres_mouse <- read_tsv(
  here("Genomic", "mm10-cCREs.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>% rename_with(~paste0(., "_mouse"))


#Unite info
cres_annotated = homologous_cres %>% 
  full_join(cres_human, by = join_by(human_enhancer == id_ccre_human)) %>% 
  full_join(cres_mouse, by = join_by(mouse_enhancer == id_ccre_mouse)) 
  
```

```{r}
# Instalar ChIPseeker e TxDb se necessário
if (!require("ChIPseeker", quietly = TRUE)) {
  BiocManager::install("ChIPseeker")
}

library(ChIPseeker)
library(rtracklayer)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(org.Hs.eg.db)

# Look up data types and keytypes
columns(org.Hs.eg.db)
keytypes(org.Hs.eg.db)

# Convert to a GRanges object
cres_human_gr <- GRanges(
  seqnames = cres_human$chrom_human,
  ranges = IRanges(start = cres_human$start_human, end = cres_human$end_human)
)

# Get levels and seqinfo
seqlevels(cres_human_gr) <- seqlevels(BSgenome.Hsapiens.UCSC.hg38)
seqinfo(cres_human_gr) <- seqinfo(BSgenome.Hsapiens.UCSC.hg38)

# Annotate peak
cres_anno <- annotatePeak(
  cres_human_gr,
  tssRegion = c(-3000, 3000),
  TxDb = TxDb.Hsapiens.UCSC.hg38.knownGene,
  annoDb = "org.Hs.eg.db"
)

# Get sequences
sequences_human <- getSeq(BSgenome.Hsapiens.UCSC.hg38, cres_human_gr)

# Convert to dataframe
cres_anno_df <- as.data.frame(cres_anno)

cres_human_seq <- cres_human %>%
  mutate(
    sequence_human = as.character(sequences_human),
    gene_name = cres_anno_df$SYMBOL,
    gene_id = cres_anno_df$GENEID,
    annotation = cres_anno_df$annotation,
    distance_to_tss = cres_anno_df$distanceToTSS
  )
```

Annotate genes
```{r}
library(AnnotationHub)

# 1. Obter as coordenadas dos genes humanos (Ex: hg38)
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene
genes_gr <- genes(txdb)

#Get sequence in a range
promoters_custom <- promoters(genes_gr, upstream = 2500, downstream = 300)

# Inicializar o Hub
ah <- AnnotationHub()

# ---- A. ELEMENTOS TRANSPONÍVEIS (RepeatMasker) ----
rmsk_query <- query(ah, c("RepeatMasker", "Homo sapiens", "hg38"))
# Inspecione rmsk_query para escolher o ID correto (ex: AH61234 ou similar)
# Supondo que o ID retornado seja AH99341:
rmsk_gr <- ah[["AH99341"]] 

# ---- B. CONSERVAÇÃO FILOGENÉTICA (PhastCons) ----
# Buscar scores de PhastCons (ex: conservação entre 100 vertebrados)
phast_query <- query(ah, c("phastCons", "Homo sapiens", "hg38", "BigWig"))
# BigWigFiles retornam uma referência ao arquivo para consulta rápida por coordenada
phast_bw <- ah[["AH75265"]] # Substitua pelo ID real da sua busca


# Encontrar intersecções entre as janelas dos genes e os elementos repetitivos
overlaps_rmsk <- findOverlaps(promoters_custom, rmsk_gr)

# Criar um data.frame com o mapeamento
gene_te_mapping <- data.frame(
  gene_id = names(promoters_custom)[queryHits(overlaps_rmsk)],
  te_name = rmsk_gr$repName[subjectHits(overlaps_rmsk)],
  te_family = rmsk_gr$repClass[subjectHits(overlaps_rmsk)] # Ex: SINE/Alu
)

# Exemplo para calcular a densidade de elementos Alu por gene:
library(dplyr)
te_density_per_gene <- gene_te_mapping %>%
  group_by(gene_id) %>%
  summarise(
    total_TEs = n(),
    total_Alu = sum(te_family == "SINE/Alu")
  )


# Extrair os scores de conservação apenas para as regiões das nossas janelas promotoras
phast_scores <- import.bw(phast_bw, selection = BigWigSelection(promoters_custom))

# Calcular o score médio de conservação para a janela de cada gene
# Usamos aggregate para calcular a média do score sobrepondo as janelas
score_per_promoter <- aggregate(phast_scores, promoters_custom, score = mean)

# Adicionar o score ao objeto original
promoters_custom$mean_phastCons <- score_per_promoter$score


# Intersecção espacial
overlaps_cres <- findOverlaps(promoters_custom, cres_human_seq)


```









